### 2018-06-11 doc/intro.tex
eight models fit. doing model comparison/inspection. needs lots of RAM

\documentclass[12pt,letterpaper]{article}

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{docmute}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}

\frenchspacing

\begin{document}

\section{Introduction}

Species extinction risk is strongly linked to geographic range size.

A species' geographic range size changes over time.

How does extinction risk change over a species duration? What effect does geographic range have on this?

We want to know if including information about past geographic range size improves our inference of species extinction risk.

A lot of work on species ``rise and fall'' dynamics has focused on the change in species geographic range over timeand what the shape of the average trajectory is CITATION. Issues that aren't addressed, like non-random drop out. Species can go extinct at any point in their ``lifetime'', so we might not observe a ``pure'' change to geographic range. We have to take this into account. The models that have been used do not do that which can totally mess-up analysis of geographic ranges over time. They need a joint-model, but that hasn't penetrated the field yet.

Additionally, lack of mechanism to explain a pure ``rise-fall''. Importantly, ``rise and fall'' is expected from a Markov process that has only a single absorbing boundary; at some point it will go back to 0.

To demonstrate that species geographic range size is not a complete Markov process, it would be necessary to show that extinction risk depends not just on the current geographic range size, but the history of its geographic range size: has range size declined or increased between earlier and now?

We instead are focusing on how geographic range effects extinction risk of a species. We are not modeling geographic range through time, but instead modeling how species extinction is a function of geographic range over time.

\end{document}

### 2018-06-11 doc/methods.tex
eight models fit. doing model comparison/inspection. needs lots of RAM

\documentclass[12pt,letterpaper]{article}

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{docmute}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}

\frenchspacing

\begin{document}

\section{Data Specifications}

Microfossil occurrence data was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax}. Occurrence information was downloaded for calcareous nannofossils, diatoms, dinoflaggellants, foraminifera, and radiolarians from the entire globe and having occurred after 65 Mya.

Data was binned into 1 My bins based on the estimated age of the fossil occurrence. The estimated ages of each occurrence is a product of core-specific age-models and are overly precise. The hope is that binning the data overcomes the heterogenity in age-models between the cores. 
% binning, [t - 1, t)

The survival or extinction of a taxon is determined for each taxons' occurrences. For every occurrence except the last, the taxon has survived which is indicated by a 0. The last occurrence of the taxon is considered the bin in which the taxon has gone extinct. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous CITATIONS but is common with marine microfossil data CITATIONS.

A taxon's geographic range during a 1 My bin was calculated as the maximum great circle distance on an ellipsoid between any two occurrences, also called a geodesic. Geographic range was measured in kilometers.

Geographic range was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1.

\section{Model Specifications}

In survival analysis, the hazard funciton describes the instanenous rate of extinction at time \(t\) given that the species has lived up to that point. For discrete-time survival data, this is expressed as \(\lambda(t | X) = P(T = t | T \geq t, X)\) for \(t = 1, \cdots N\) where \(t\) is the status of a species at a time point, \(T\) is the time of extinction of the species, and \(i = 1, 2, \cdots, N\) with \(N\) being the total number of observations.

In survival analysis the hazard function describes the in stannous rate of extinction for a species given matrix of covariates X, \(\lambda(t | X)\). We parameterized the hazard function as a logistic regression/generalized linear model defined \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is an inverse logit link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\).

\(\Theta\) is then parameterized as a hierarchical model with multiple non-nested varying intercepts CITATION. We considered two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\). We also considered the interaction between geographic range during intervals \(t\) and \(t - 1\). At the first observation for a taxon, the diff of geographic range is 0.

Our model had up to three non-nested varying intercepts: age of observation \(a\), bin of observation \(b\), and taxonomic group of the observation \(c\). Age of observation is a critical varying intercept as this effect, when considered along with the intercept \(\beta_{0}\), captures changes to baseline hazard with species age. Because these factors are modeled as varying intercepts, we are assuming that the effects of geographic range (\(\beta_{1} \dots \beta_{3}\))

The logistic regression can thus be expressed as 
\begin{equation}
  \begin{aligned}
    t_{i} = logit^{-1}(\beta_{0} + \beta_{1} r_{i} + \beta_{2} d_{i} + \beta_{3} r_{i} d_{i} + a_{j[i]} + b_{k[i]} + c_{f[i]}) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation e.g. \(j[i]\) is the age of the \(i\)th observation. \(a_{j[i]}\) is taxon age in bins at time of observation where \(j = 1, 2, \cdots, J\) and \(J\) is the maximum observed age of any taxon. \(b_{k}\) is the time of observation where \(k = 1, 2, \cdots, K\) and \(K\) is the number of time bins observed. Finally, \(c_{f}\) is the taxonomic group of an observation where \(f = 1, 2, \cdots, F)\) and \(F\) is the total number of taxonomic groups analyzed.

\section{Model Parameter Estimation}

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language CITATION. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.99; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.

Posterior convergence was determined using the following general and HMC specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)). For futher explanation of these diagnostic criteria, see the Stan Manual cITATION.

\section{Model Selection and Adequacy}

We considered four variants of this model: 1) only age-based varying intercepts, 2) age and time of observation varying intercepts, 3) age and taxonomic group varying intercepts, and 4) all three considered varying intercepts. These models were compared using three measures of model performance: WAIC and LOOIC which are estimates of a model's out-of-sample performance, and ROC which is a measure of a model's predictive performance.

To determine how many non-nested varying-intercept terms were possible to include without overly biasing out models' parameter estimates, the four variant models described above were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior. CITATION. WAIC and LOOIC are interpreted similarly to AIC, with lower values indicating greater expected out-of-sample performance. Additionally, the Bayesian nature of WAIC and LOOIC mean that they are calculated with standard errors allowing for straight forward comparisons. We selected the model with lowest WAIC and LOOIC and all models within one standard error of its WAIC/LOOIC estimate.

Given these selected models, we did a further comparison between models including the change in geographic range from \(t - 1\) to \(t\), and its interaction with geographic range at \(t\), in the model to determine how much our predictions of extinction risk are improved by the inclusion of ``the past.''

The model adequacy was measured using the area under the reciever operating characteristic curve (AUC). This measure is commonly used in classification problems like this one as it has the desireable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION. This value ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance. AUC was calculated for the dataset as a whole, and for each of time bins.

\end{document}

### 2018-06-11 R/fit_model.r

@@ -34,23 +34,25 @@ counti <- prepare_analysis(counti)
 
 
 # fit the model
-form <- formula(event ~ maxgcd * lag1_maxgcd +
+form <- formula(event ~ maxgcd * diff_maxgcd +
                 (1 | fact_relage))
 forms <- list(form,
              update(form, ~ . + (1 | fact_mybin)),
              update(form, ~ . + (1 | fossil.group)),
              update(form, ~ . + (1 | fact_mybin) + (1 | fossil.group)))
+forms2 <- map(forms, ~ update(.x, ~ . - diff_maxgcd - maxgcd:diff_maxgcd))
+forms <- append(forms, forms2)
 
 disc_fit <- map(forms, ~ stan_glmer(.x, family = 'binomial', data = counti,
-                                    adapt_delta = 0.95))
+                                    adapt_delta = 0.999))
 write_rds(disc_fit, path = '../data/disc_fit.rds')
 
 
 
 # tree-based methods
 # survival tree with time-varying covariates
-counti <- as.data.frame(counti)  # this is fucking annoying but necessary
-form <- formula(Surv(time1, time2, event) ~ 
-                maxgcd + lag1_maxgcd + relage + mybin + fossil.group)
-tree_fit <- LTRCART(formula = form, data = counti)
-write_rds(tree_fit, path = '../data/tree_fit.rds')
+#counti <- as.data.frame(counti)  # this is fucking annoying but necessary
+#form <- formula(Surv(time1, time2, event) ~ 
+#                maxgcd + diff_maxgcd + relage + mybin + fossil.group)
+#tree_fit <- LTRCART(formula = form, data = counti)
+#write_rds(tree_fit, path = '../data/tree_fit.rds')

### 2018-06-11 R/inspect_model.r

@@ -40,31 +40,51 @@ inspect <- counti_trans %>%            # nice to see the super relevant columns
 disc_fit <- read_rds('../data/disc_fit.rds')
 tree_fit <- read_rds('../data/tree_fit.rds')
 
+# estimate of out-of-sample performance
+compare_loo <- map(disc_fit, loo)
+compare_loo_tab <- loo::compare(x = compare_loo)
+compare_waic <- map(disc_fit, waic)
+compare_waic_tab <- loo::compare(x = compare_waic)
+
+# posterior probability of observation surviving
+pp_est <- map(disc_fit, posterior_predict)
+#fitted_samples(disc_fit[[1]])
+pp_prob <- map(disc_fit, ~ posterior_linpred(.x, transform = TRUE))
+#predicted_samples(disc_fit[[1]])
+
+# adequacy of fit ROC
+pp_roc <- map(pp_est, ~ apply(.x, 1, function(y) auc(roc(counti_trans$event, y))))
+
+# for each time bin
+#pp_est
+#counti_trans$fact_mybin
 
 
-# glmer model
+# pic a model
+#disc_best <- disc_fit[[2]]
+disc_best <- disc_fit[[4]]
 # posterior intervals
-interval_est <- disc_fit %>%
+interval_est <- disc_best %>%
   spread_samples(`(Intercept)`, 
                  maxgcd, 
-                 lag1_maxgcd,
-                 `maxgcd:lag1_maxgcd`,
+                 diff_maxgcd,
+                 `maxgcd:diff_maxgcd`,
                  `Sigma[fact_mybin:(Intercept),(Intercept)]`,
                  `Sigma[fact_relage:(Intercept),(Intercept)]`,
                  `Sigma[fossil.group:(Intercept),(Intercept)]`) %>%
   median_qi(.prob = c(0.9, 0.5))
 
-interval_eye <- disc_fit %>%
+interval_eye <- disc_best %>%
   gather_samples(`(Intercept)`, 
                  maxgcd, 
-                 lag1_maxgcd) %>%
+                 diff_maxgcd) %>%
   ggplot(aes(y = term, x = estimate)) +
   geom_halfeyeh(.prob = c(0.9, 0.5))
 
-interval_range <- disc_fit %>%
+interval_range <- disc_best %>%
   gather_samples(`(Intercept)`, 
                  maxgcd, 
[…]

### 2018-06-13 doc/methods.tex
temperature data. just plotting the data for funzies

@@
 Microfossil occurrence data was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax}.
-Occurrence information was downloaded for calcareous nannofossils, diatoms, dinoflaggellants, foraminifera, and radiolarians from the entire globe and having occurred after 65 Mya.
+Occurrence information was downloaded for calcareous nannofossils, diatoms, dinoflaggellants, foraminifera, and radiolarians from the entire globe and having occurred between 120 and 0 Mya.
+This selection of species was then culled to just those species which have their first occurrence at most 63 My; this avoids those taxa which survived the K/Pg boundary, those taxa arrising just after the K/Pg boundary, and lines up the occurrences with the temperature time-series (discussed below).
 
@@
 
+Temperature data is based on estimates from Cramer et al CITATION.
+These estimates are based on Magnesian/Calcium isotope ratios and are a more accurate estimate of global temperature than the standard Oxygen isotope-based estimates because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
+metioric water).
+Cramer et al.
+provide temperature estimates for every 0.1 My from 0 to 63 Mya; we binned these estimates every 1 My in accordance with the way we binned all fossil occurrences.
+
 \section{Model Specifications}
@@
 \(\Theta\) is then parameterized as a hierarchical model with multiple non-nested varying intercepts CITATION.
-We considered two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\).
+First, we considered two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\).
 We also considered the interaction between geographic range during intervals \(t\) and \(t - 1\).
@@
 
+Second, we considered two aspects of global temperature: mean temperature during interval \(t\), and the mean temperature during interval \(t - 1\).
+
 Our model had up to three non-nested varying intercepts: age of observation \(a\), bin of observation \(b\), and taxonomic group of the observation \(c\).
 Age of observation is a critical varying intercept as this effect, when considered along with the intercept \(\beta_{0}\), captures changes to baseline hazard with species age.
-Because these factors are modeled as varying intercepts, we are assuming that the effects of geographic range (\(\beta_{1} \dots \beta_{3}\))
+Because these factors are modeled as varying intercepts, we are assuming that the effects of geographic range (\(\beta_{1} \dots \beta_{3}\)) and temperature (\(\beta_{4} \dots \beta_{5}\)) are constant over time.
 

### 2018-06-13 R/compare_model.r

@@ -1,0 +1,69 @@
+# data manipulation
+library(tidyverse)
+#devtools::install_github("mjskay/tidybayes")
+library(tidybayes)
+
+# parallel processing
+library(parallel)
+library(future)
+
+# survival
+library(survival)
+library(LTRCtrees)
+library(rpart.plot)
+library(partykit)
+
+# bayes
+library(arm)
+library(rstanarm)
+library(bayesplot)
+source('../R/stan_utility.R')
+
+# misc
+library(splines)
+library(pROC)
+source('../R/process_foo.r')
+
+# important constants
+options(mc.cores = parallel::detectCores())
+future::plan(strategy = multicore)
+
+# get data in
+longi <- read_rds('../data/longitude.rds')
+survi <- read_rds('../data/survival.rds')
+counti <- read_rds('../data/counting.rds')
+
+# form of data that was analyzed
+counti_trans <- prepare_analysis(counti)
+
+# read in model fits
+disc_fit <- read_rds('../data/disc_fit.rds')
+# estimate of out-of-sample performance
+# these should be parallelized
+cl <- map(disc_fit, ~ future::future(loo(.x)))
+cl <- map(cl, ~ future::value(.x))
+compare_loo_tab <- loo::compare(x = cl)
+wl <- map(disc_fit, ~ future::future(waic(.x)))
+wl <- map(wl, ~ future::value(.x))
+compare_waic_tab <- loo::compare(x = wl)
+
+
+# posterior probability of observation surviving
+pe <- map(disc_fit, ~ future::future(posterior_predict(.x)))
+pp_est <- map(pe, ~ future::value(.x))
+pp <- map(disc_fit, ~ future::future(posterior_linpred(.x, transform = TRUE)))
+pp_prob <- map(pe, ~ future::value(.x))
+
+# adequacy of fit ROC
+pp_roc <- map(pp_est, ~ apply(.x, 1, function(y) roc(counti_trans$event, y)))
+pp_auc <- map(pp_roc, function(y) map_dbl(y, ~ auc(.x)))
+
+
+# break up by point in time
+#tt <- map(pp_est, ~ split(.x, counti_trans$fact_mybin))
+#tt <- map(tt, ~ matrix(.x, nrow = 4000))
+#ee <- split(counti_trans$event, counti_trans$fact_mybin)
+#pp_auc_time <- mcMap(function(x, y) apply(x, 1, function(a) auc(roc(a, y))),
[…]

### 2018-06-13 R/data_process.r

@@ -10,21 +10,16 @@ library(XML)
 library(maptools)
 library(geosphere)
 
-# survival
-library(survival)
-library(flexsurv)
-library(survminer)
-
 # plotting stuff
-library(ggplot2)
 library(scales)
 
+# misc
 source('../R/process_foo.r')
 
 
 # constants
 bin_width <- 1
-age_max <- 65
+age_max <- 63
 bin_number <- age_max / bin_width
 
 # read in data file and clean column names
@@ -47,14 +42,18 @@ nano[nano$fossil.group == 'FALSE', 'fossil.group'] <- 'F'
 nano$longitude <- parse_double(nano$longitude)
 nano$latitude <- parse_double(nano$latitude)
 
-# basic properties
-nano <- nano %>%
+nano <- nano %>% 
+  filter(!is.na(plat), 
+         !is.na(plng)) %>%
+  group_by(taxon.id) %>%
+  filter(max(age) < age_max) %>%
+  ungroup() %>%
   # some don't have paleolat or long
-  filter(!is.na(plat), !is.na(plng)) %>%
+  # also need to limit to line up with mgca
   # then
   # assign million year bins (decreasing; youngest lowest)
   dplyr::arrange(desc(age)) %>%
-  dplyr::mutate(mybin = ntile(age, bin_number)) %>%
+  dplyr::mutate(mybin = break_my(age, by = 1)) %>%
   # then
   # make a species genus combo
   dplyr::mutate(fullname = str_c(genus, '_', species)) %>% 
@@ -108,9 +107,6 @@ nano <-
 
 # get all the important geographic range information
 # have to summarize the big matrix nano
-
-count.groups <- with(nano, table(fossil.group))
-
 sprange <- nano %>%
   group_by(fullname, mybin) %>%
   dplyr::summarise(nocc = n(),  # number of occurrences in bin
@@ -126,6 +122,23 @@ sprange <- nano %>%
          latext > 0)
 
 
+# want to add in the lear mgca data
+mgca <- read_tsv('../data/cramer/cramer_temp.txt')
+names(mgca) <- str_to_lower(names(mgca))
+names(mgca) <- str_remove_all(names(mgca), '[^[[:alnum:]]]')
+mgca <- mgca %>%
+  mutate(mybin = break_my(age)) %>%
+  group_by(mybin) %>%
+  summarize(temp = mean(temperature, na.rm = TRUE),
+            temp_sd = sd(temperature, na.rm = TRUE)) %>%
+  ungroup()
+
[…]

### 2018-06-13 R/fit_model.r

@@ -32,27 +32,16 @@ counti <- read_rds('../data/counting.rds')
 # transform the data for analysis
 counti <- prepare_analysis(counti)
 
-
 # fit the model
-form <- formula(event ~ maxgcd * diff_maxgcd +
-                (1 | fact_relage))
-forms <- list(form,
-             update(form, ~ . + (1 | fact_mybin)),
-             update(form, ~ . + (1 | fossil.group)),
-             update(form, ~ . + (1 | fact_mybin) + (1 | fossil.group)))
-forms2 <- map(forms, ~ update(.x, ~ . - diff_maxgcd - maxgcd:diff_maxgcd))
-forms <- append(forms, forms2)
+form <- formula(event ~ maxgcd * diff_maxgcd + 
+                temp + lag1_temp +
+                (1 | fact_relage) + 
+                (1 | fossil.group) + 
+                (1 | fact_mybin))
+form2 <- update(form, 
+                ~ . - diff_maxgcd - maxgcd:diff_maxgcd - lag1_temp)
+forms <- list(form, form2)
 
 disc_fit <- map(forms, ~ stan_glmer(.x, family = 'binomial', data = counti,
-                                    adapt_delta = 0.999))
+                                    adapt_delta = 0.999, thin = 2))
 write_rds(disc_fit, path = '../data/disc_fit.rds')
-
-
-
-# tree-based methods
-# survival tree with time-varying covariates
-#counti <- as.data.frame(counti)  # this is fucking annoying but necessary
-#form <- formula(Surv(time1, time2, event) ~ 
-#                maxgcd + diff_maxgcd + relage + mybin + fossil.group)
-#tree_fit <- LTRCART(formula = form, data = counti)
-#write_rds(tree_fit, path = '../data/tree_fit.rds')

### 2018-06-13 R/inspect_data.r

@@ -1,0 +1,94 @@
+# data manipulation
+library(tidyverse)
+
+# parallel processing
+library(parallel)
+
+# bayes
+library(arm)
+library(rstanarm)
+library(bayesplot)
+source('../R/stan_utility.R')
+
+# misc
+library(scales)
+library(ggridges)
+library(pROC)
+source('../R/process_foo.r')
+
+# important constants
+options(mc.cores = parallel::detectCores())
+
+# get data in
+longi <- read_rds('../data/longitude.rds')
+survi <- read_rds('../data/survival.rds')
+counti <- read_rds('../data/counting.rds')
+
+# form of data that was analyzed
+counti_trans <- prepare_analysis(counti)
+
+
+# occurrences through time, labeled if LAD
+octg <- counti %>%
+  ggplot(aes(x = mybin, fill = factor(event))) +
+  stat_bin() +
+  facet_grid(fossil.group ~ ., switch = 'y')
+
+# relative "abundance" of microfossil groups over time
+ocag <- counti %>%
+  ggplot(aes(x = mybin, fill = fossil.group)) +
+  geom_histogram(position = 'fill')
+
+# occurrences by relage
+ocrg <- counti %>% 
+  group_by(fullname) %>%
+  dplyr::summarize(maxage = max(relage),
+                   fossil.group = plurality(fossil.group),
+                   died = any(event == 1)) %>%
+  ggplot(aes(x = maxage, fill = died)) +
+  stat_bin() +
+  facet_grid(fossil.group ~ ., switch = 'y')
+
+# make a plot of a random selection of species
+srg <- counti %>% 
+  mutate(logmaxgcd = log1p(maxgcd)) %>%
+  group_by(fullname) %>%
+  sample_n_groups(size = 8) %>%
+  ungroup %>%
+  ggplot(aes(x = relage, y = logmaxgcd, group = fullname, colour = fullname)) +
+  geom_line() +
+  geom_point() +
+  theme(legend.position = 'bottom')
+
+# lots of little code here
+# for FAD/LAD accumulation curves
+ft <- counti %>%
+  group_by(fullname) %>%
+  summarize(fad = max(mybin),
+            fossil.group = plurality(fossil.group)) %>%
+  group_by(fossil.group, fad) %>%
+  summarize(n = n()) %>%
+  arrange(desc(fad)) %>%
+  mutate(nsum = cumsum(n),
[…]

### 2018-06-13 R/inspect_model.r

@@ -6,12 +6,6 @@ library(tidybayes)
 # parallel processing
 library(parallel)
 
-# survival
-library(survival)
-library(LTRCtrees)
-library(rpart.plot)
-library(partykit)
-
 # bayes
 library(arm)
 library(rstanarm)
@@ -19,7 +13,6 @@ library(bayesplot)
 source('../R/stan_utility.R')
 
 # misc
-library(splines)
 library(pROC)
 source('../R/process_foo.r')
 
@@ -33,33 +26,10 @@ counti <- read_rds('../data/counting.rds')
 
 # form of data that was analyzed
 counti_trans <- prepare_analysis(counti)
-inspect <- counti_trans %>%            # nice to see the super relevant columns
-  dplyr::select(event, relage, mybin, maxgcd, diff_maxgcd)
 
 # read in model fits
 disc_fit <- read_rds('../data/disc_fit.rds')
-
-# estimate of out-of-sample performance
-compare_loo <- map(disc_fit, loo)
-compare_loo_tab <- loo::compare(x = compare_loo)
-compare_waic <- map(disc_fit, waic)
-compare_waic_tab <- loo::compare(x = compare_waic)
-
-# limit
-disc_best <- disc_fit[[4]]
-# posterior probability of observation surviving
-pp_est <- posterior_predict(disc_best)
-pp_prob <- posterior_linpred(disc_best, transform = TRUE)
-# adequacy of fit ROC
-pp_auc <- apply(pp_est, 1, function(x) roc(counti_trans$event, x))
-
-
-# break up by point in time
-tt <- split(pp_est, counti_trans$fact_mybin)
-tt <- lapply(tt, function(x) matrix(x, nrow = 4000))
-ee <- split(counti_trans$event, counti_trans$fact_mybin)
-pp_auc_time <- map2(tt, ee, ~ apply(.x, 1, function(a) auc(roc(a, .y))))
-
+disc_best <- disc_fit[[1]]
 
 # posterior intervals
 interval_est <- disc_best %>%
@@ -67,6 +37,8 @@ interval_est <- disc_best %>%
                  maxgcd, 
                  diff_maxgcd,
                  `maxgcd:diff_maxgcd`,
+                 temp,
+                 lag1_temp,
                  `Sigma[fact_mybin:(Intercept),(Intercept)]`,
                  `Sigma[fact_relage:(Intercept),(Intercept)]`,
[…]

### 2018-06-13 src/run_analysis.sh

@@ -1,3 +1,4 @@
 Rscript ../R/data_process.r &
 Rscript ../R/fit_model.r &
+Rscript ../R/compare_model.r &
 Rscript ../R/inspect_model.r &

### 2018-06-13 R/fit_model.r
lots of plots looking at properties of the data

@@ -4,12 +4,6 @@ library(tidyverse)
 # parallel processing
 library(parallel)
 
-# survival
-library(survival)
-library(LTRCtrees)
-library(rpart.plot)
-library(partykit)
-
 # bayes
 library(arm)
 library(rstanarm)
@@ -17,8 +11,6 @@ library(bayesplot)
 source('../R/stan_utility.R')
 
 # misc
-library(splines)
-library(pROC)
 source('../R/process_foo.r')
 
 # important constants
@@ -43,5 +35,5 @@ form2 <- update(form,
 forms <- list(form, form2)
 
 disc_fit <- map(forms, ~ stan_glmer(.x, family = 'binomial', data = counti,
-                                    adapt_delta = 0.999, thin = 2))
+                                    adapt_delta = 0.999, thin = 4))
 write_rds(disc_fit, path = '../data/disc_fit.rds')

### 2018-06-13 R/inspect_data.r

@@ -92,3 +92,31 @@ ccg <- bind_rows(ft, lt, .id = 'type') %>%
   facet_grid(~ fossil.group) +
   labs(x = 'Time (My)', y = 'Cummulative count')
 
+
+# looking at the temperature data
+mgca <- read_tsv('../data/cramer/cramer_temp.txt')
+names(mgca) <- str_to_lower(names(mgca))
+names(mgca) <- str_remove_all(names(mgca), '[^[[:alnum:]]]')
+mgca <- mgca %>%
+  mutate(bin = break_my(age)) %>%
+  group_by(bin) %>%
+  mutate(temp_bin = mean(temperature, na.rm = TRUE)) %>%
+  ungroup
+
+# this is stupid but i'm bad at tidyr gather/spread
+a <- mgca %>% 
+  dplyr::select(age, temperature)
+b <- mgca %>% 
+  dplyr::select(bin, temp_bin) %>% 
+  mutate(bin = bin - 0.5) %>%
+  rename(age = bin,
+         temperature = temp_bin) 
+mg <- bind_rows(a, b, .id = 'scheme')
+
+# temperature over time based on Mg/Ca
+tpg <- mg %>%
+  mutate(scheme = plyr::mapvalues(scheme, 1:2, c('raw', 'bin'))) %>%
+  ggplot(aes(x = age, y = temperature, colour = scheme)) + 
+  geom_line() +
+  scale_colour_manual(values = c('goldenrod', 'skyblue')) +
+  labs(x = 'Time (My)', y = 'Temperature diff. from modern (C)')

### 2018-06-25 doc/methods.tex
lots of plots looking at properties of the data
rewrote model statement. need to expand

@@
 
-Second, we considered two aspects of global temperature: mean temperature during interval \(t\), and the mean temperature during interval \(t - 1\).
+Second, we considered two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e. mean temperature during interval \(t - 1\).)
 
-Our model had up to three non-nested varying intercepts: age of observation \(a\), bin of observation \(b\), and taxonomic group of the observation \(c\).
-Age of observation is a critical varying intercept as this effect, when considered along with the intercept \(\beta_{0}\), captures changes to baseline hazard with species age.
-Because these factors are modeled as varying intercepts, we are assuming that the effects of geographic range (\(\beta_{1} \dots \beta_{3}\)) and temperature (\(\beta_{4} \dots \beta_{5}\)) are constant over time.
+% lots of ways of varying
+% maxgcd * diff_maxgcd varies first by when then by fossil group.
+% non-nested varying intercept based on age, then by fossil group.
 
+% event ~ temp + lag1_temp + maxgcd * diff_maxgcd + 
+% (1 + maxgcd * diff_maxgcd | fact_mybin/fossil.group) +
+% (1 | fact_relage/fossil.group)
 The logistic regression can thus be expressed as 
+% scripts
+%   t = death? 0 if not last observation
+%   \beta = effect range vars (varies over time and by fossil group)
+%   \gamma = effect temp vars
+%   a = age effect
+% subscripts
+%   w = when in My
+%   f = fossil group
+%   d = age at observation
+% 
 \begin{equation}
   \begin{aligned}
-    t_{i} = logit^{-1}(\beta_{0} + \beta_{1} r_{i} + \beta_{2} d_{i} + \beta_{3} r_{i} d_{i} + a_{j[i]} + b_{k[i]} + c_{f[i]}) \\
+    t_{i} &= logit^{-1}(\beta_{w[i], f[i]} + \beta_{w[i], f[i]} range_{i} + \beta_{w[i], f[i]} rangediff_{i} + \beta_{w[i], f[i]} range_{i} rangediff_{i} + \gamma_{1} temp_{w[i]} + \gamma_{2} templag_{w[i]} + a_{d[i], f[i]}) \\
+    \beta_{w, f} &\sim MVN(\alpha_{f}, \Sigma) \\
+    \alpha_{f} &\sim MVN(0, \Sigma) \\
+    \gamma &\sim N(0, \sigma) \\
+    a_{w, f} &\sim MVN(\delta_{f}, \Sigma) \\
+    \delta_{f} &\sim MVN(0, \Sigma) \\
   \end{aligned}

### 2018-07-02 R/compare_model.r
no interaction model

@@ -41,6 +41,7 @@ counti_trans <- prepare_analysis(counti)
 
 # read in model fits
 model_key <- c('Past and vary', 
+               'Past and vary (no interaction)',
                'No past but vary', 
                'No past or vary')
 disc_fit <- read_rds('../data/disc_fit.rds')
@@ -68,8 +69,8 @@ wl <- map(disc_fit, ~ future::future(waic(.x)))
 wl <- map(wl, ~ future::value(.x))
 compare_waic_tab <- loo::compare(x = wl)
 # i should find a way to output these nicely
-compare_loo_tab
-compare_waic_tab
+#compare_loo_tab
+#compare_waic_tab
 
 
 # posterior probability of observation surviving

### 2018-07-02 R/fit_model.r

@@ -30,6 +30,10 @@ form <- formula(event ~ temp + lag1_temp + maxgcd * diff_maxgcd +
                 (1 + maxgcd * diff_maxgcd | fact_mybin/fossil_group) + 
                 (1 | fact_relage/fossil_group))
 
+form1 <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd + 
+                (1 + maxgcd + diff_maxgcd | fact_mybin/fossil_group) + 
+                (1 | fact_relage/fossil_group))
+
 form2 <- update(form, 
                 ~ . - diff_maxgcd - maxgcd:diff_maxgcd - 
                   lag1_temp - (1 + maxgcd * diff_maxgcd | fact_mybin/fossil_group) +
@@ -39,7 +43,7 @@ form3 <- update(form2,
                 ~ . - (1 + maxgcd | fact_mybin/fossil_group) + 
                   (1 | fact_mybin/fossil_group))
 
-forms <- list(form, form2, form3)
+forms <- list(form, form1, form2, form3)
 
 disc_fit <- map(forms, ~ stan_glmer(.x, family = 'binomial', data = counti,
                                     adapt_delta = 0.99, thin = 4))

### 2018-07-11 R/compare_model.r
5-fold cv implemented. calculate AUC for each fold-s prediction

@@ -8,12 +8,6 @@ library(tidybayes)
 library(parallel)
 library(future)
 
-# survival
-library(survival)
-library(LTRCtrees)
-library(rpart.plot)
-library(partykit)
-
 # bayes
 library(arm)
 library(rstanarm)
@@ -65,12 +59,21 @@ energy <- map2(np, ll, ~ mcmc_nuts_energy(.x, .y))
 cl <- map(disc_fit, ~ future::future(loo(.x)))
 cl <- map(cl, ~ future::value(.x))
 compare_loo_tab <- loo::compare(x = cl)
+rownames(compare_loo_tab) <- plyr::mapvalues(rownames(compare_loo_tab),
+                                             sort(rownames(compare_loo_tab)),
+                                             model_key)
+xtable::print.xtable(xtable::xtable(compare_loo_tab), 
+                     file = '../doc/loo_tab_draft.tex')
+
 wl <- map(disc_fit, ~ future::future(waic(.x)))
 wl <- map(wl, ~ future::value(.x))
 compare_waic_tab <- loo::compare(x = wl)
+rownames(compare_waic_tab) <- plyr::mapvalues(rownames(compare_waic_tab),
+                                              sort(rownames(compare_waic_tab)),
+                                              model_key)
+xtable::print.xtable(xtable::xtable(compare_waic_tab), 
+                     file = '../doc/waic_tab_draft.tex')
 # i should find a way to output these nicely
-#compare_loo_tab
-#compare_waic_tab
 
 
 # posterior probability of observation surviving
@@ -82,29 +85,21 @@ pp_prob <- map(pp, ~ future::value(.x))
 # adequacy of fit ROC
 pgc <- partial(get_cutpoints, target = counti_trans$event)
 list_cutpoint <- map(map(pp_prob, pgc), ~ map(.x, ~ .x[3]))
-
-pp_est_new <- list()
-for(jj in seq(length(pp_prob))) {
-  mm <- matrix(ncol = ncol(pp_prob[[jj]]), nrow = nrow(pp_prob[[jj]]))
-  for(ii in seq(nrow(mm))) {
-    mm[ii, ] <- pp_prob[[jj]][ii, ] > list_cutpoint[[jj]][[ii]]
-  }
-  pp_est_new[[jj]] <- mm * 1
-}
+pp_est_new <- cut_newpoint(pp_prob, list_cutpoint)
 # this resets the cutpoint for determining 0 vs 1 by setting it very low.
[…]

### 2018-07-11 R/data_process.r

@@ -48,19 +48,19 @@ nano <- nano %>%
 nano <- nano %>% 
   filter(!is.na(plat), 
          !is.na(plng)) %>%
-group_by(taxon_id) %>%
-filter(max(age) < age_max) %>%
-ungroup() %>%
-# some don't have paleolat or long
-# also need to limit to line up with mgca
-# then
-# assign million year bins (decreasing; youngest lowest)
-dplyr::arrange(desc(age)) %>%
-dplyr::mutate(mybin = break_my(age, by = 1)) %>%
-# then
-# make a species genus combo
-dplyr::mutate(fullname = str_c(genus, '_', species)) %>% 
-arrange(fullname)
+  group_by(taxon_id) %>%
+  filter(max(age) < age_max) %>%
+  ungroup() %>%
+  # some don't have paleolat or long
+  # also need to limit to line up with mgca
+  # then
+  # assign million year bins (decreasing; youngest lowest)
+  dplyr::arrange(desc(age)) %>%
+  dplyr::mutate(mybin = break_my(age, by = 1)) %>%
+  # then
+  # make a species genus combo
+  dplyr::mutate(fullname = str_c(genus, '_', species)) %>% 
+  arrange(fullname)
 
 # assign everything a geographic cell using paleocoordinates
 eq <- CRS("+proj=cea +lat_0=0 +lon_0=0 +lat_ts=30 +a=6371228.0 +units=m")

### 2018-07-11 R/fit_predict_model.r

@@ -1,0 +1,55 @@
+# data manipulation
+library(tidyverse)
+
+# parallel processing
+library(parallel)
+
+# bayes
+library(arm)
+library(rstanarm)
+library(bayesplot)
+source('../R/stan_utility.R')
+
+# misc
+library(pROC)
+library(ROCR)
+source('../R/process_foo.r')
+
+# important constants
+options(mc.cores = parallel::detectCores())
+
+# get data in
+longi <- read_rds('../data/longitude.rds')
+survi <- read_rds('../data/survival.rds')
+counti <- read_rds('../data/counting.rds')
+
+# transform the data for analysis
+counti_trans <- prepare_analysis(counti)
+
+# split the data into temporal folds
+# there are only 63 temporal units
+# assuming 5 folds, that's 12 time units per fold
+# train [1], test [2]
+# train [1,2], test [3]
+# train [1,2,3], test [4]
+# train [1,2,3,4], test [5]
+
+counti_trans <- counti_trans %>%
+  mutate(fold = break_my(mybin, number = 5))
+
+# 1 is young, 5 is old
+counti_fold <- rev(split(counti_trans, counti_trans$fold))  # each fold
+counti_accum <- accumulate(counti_fold, bind_rows) # combine some folds
+counti_accum <- counti_accum[-1]
+
+form1 <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd + 
+                 (1 + maxgcd + diff_maxgcd | fact_mybin/fossil_group) + 
+                 (1 | fact_relage/fossil_group))
+
+fit <- map(counti_accum, ~ stan_glmer(form1, 
+                                      family = 'binomial', 
+                                      data = .x,
+                                      adapt_delta = 0.999, 
+                                      thin = 4))
+
+write_rds(fit, path = '../data/training_fit.rds')

### 2018-07-11 R/inspect_data.r

@@ -27,6 +27,8 @@ counti <- read_rds('../data/counting.rds')
 # form of data that was analyzed
 counti_trans <- prepare_analysis(counti)
 
+counti_trans %>% dplyr::select(fullname, mybin, relage, maxgcd, diff_maxgcd)
+
 
 # occurrences through time, labeled if LAD
 octg <- counti %>%

### 2018-07-11 R/inspect_model.r

@@ -31,37 +31,38 @@ counti_trans <- prepare_analysis(counti)
 
 # read in model fits
 disc_fit <- read_rds('../data/disc_fit.rds')
-disc_best <- disc_fit[[1]]
+#disc_best <- disc_fit[[1]]
+disc_best <- disc_fit[[2]]
 
 # posterior intervals
 interval_est <- disc_best %>%
   spread_samples(`(Intercept)`, 
                  maxgcd, 
                  diff_maxgcd,
-                 `maxgcd:diff_maxgcd`,
+                 #`maxgcd:diff_maxgcd`,
                  temp,
                  lag1_temp,
                  `Sigma[fossil_group:fact_mybin:(Intercept),(Intercept)]`,
                  `Sigma[fossil_group:fact_mybin:maxgcd,(Intercept)]`,
                  `Sigma[fossil_group:fact_mybin:diff_maxgcd,(Intercept)]`,
-                 `Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,(Intercept)]`,
+                 #`Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,(Intercept)]`,
                  `Sigma[fossil_group:fact_mybin:maxgcd,maxgcd]`,
                  `Sigma[fossil_group:fact_mybin:diff_maxgcd,maxgcd]`,
-                 `Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,maxgcd]`,
+                 #`Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,maxgcd]`,
                  `Sigma[fossil_group:fact_mybin:diff_maxgcd,diff_maxgcd]`,
-                 `Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,diff_maxgcd]`,
-                 `Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,maxgcd:diff_maxgcd]`,
+                 #`Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,diff_maxgcd]`,
+                 #`Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,maxgcd:diff_maxgcd]`,
                  `Sigma[fossil_group:fact_relage:(Intercept),(Intercept)]`,
                  `Sigma[fact_mybin:(Intercept),(Intercept)]`,
[…]

### 2018-07-11 R/inspect_predict_model.r

@@ -1,0 +1,74 @@
+# data manipulation
+library(tidyverse)
+library(janitor)
+#devtools::install_github("mjskay/tidybayes")
+library(tidybayes)
+
+# parallel processing
+library(parallel)
+
+# bayes
+library(arm)
+library(rstanarm)
+library(bayesplot)
+source('../R/stan_utility.R')
+
+# misc
+library(pROC)
+library(ROCR)
+source('../R/process_foo.r')
+source('../R/plot_foo.r')
+
+# important constants
+options(mc.cores = parallel::detectCores())
+
+# get data in
+longi <- read_rds('../data/longitude.rds')
+survi <- read_rds('../data/survival.rds')
+counti <- read_rds('../data/counting.rds')
+
+# form of data that was analyzed
+counti_trans <- prepare_analysis(counti)
+
+# fold data prep
+counti_trans <- counti_trans %>%
+  mutate(fold = break_my(mybin, number = 5))
+
+# 1 is young, 5 is old
+counti_fold <- rev(split(counti_trans, counti_trans$fold))  # each fold
+counti_accum <- accumulate(counti_fold, bind_rows) # combine some folds
+counti_accum <- counti_accum[-1]
+
+
+# get in fits and posterior work
+fit <- read_rds('../data/training_fit.rds')
+
+# given trained models for the rolled-out folds, estimate the next fold
+# first: get the cutpoint from the training data set b/c class imbalance
+plin <- map2(fit, counti_accum, ~ posterior_linpred(object = .x, newdata = .y))
+cut_points <- map2(plin, counti_accum, ~ get_cutpoints(.x, .y$event)) %>%
+  map(., ~ .x[[1]][3])
+# second: using the new cutpoints, predict test data
+pred <- map2(fit, counti_fold[-1], 
+             ~ posterior_linpred(object = .x, newdata = .y)) %>%
+  cut_newpoint(., cut_points)
+# third: calculate ROC/AUC for the predictions of test data
+pred_roc <- map2(pred, counti_fold[-1], post_roc) %>%
+  map(., function(x) map_dbl(x, ~ auc(.x)))
+
+# now to make a plot about out-of-sample predictive accuracy
+oos_roc <- bind_cols(pred_roc) %>% 
+  gather() %>%
+  mutate(key = str_extract(key, '[0-9]')) %>%
+  ggplot(aes(x = value, fill = key)) + 
[…]

### 2018-07-11 R/plot_foo.r

@@ -24,7 +24,7 @@ plot_georange_compare <- function(data) {
 #' @return a ggplot object
 plot_taxon_covariate_time <- function(disc_best) {
   comp_const <- disc_best %>%
-    spread_samples(maxgcd, diff_maxgcd, `maxgcd:diff_maxgcd`)
+    spread_samples(maxgcd, diff_maxgcd)
 
   comp_var <- disc_best %>%
     spread_samples(b[i, f]) %>%
@@ -41,30 +41,28 @@ plot_taxon_covariate_time <- function(disc_best) {
   # first element is just temporal effect
   core <- full_join(comp_const, cv[[1]], by = c('.chain', '.iteration')) %>%
     mutate(eff_maxgcd = maxgcd.x + maxgcd.y,
-           eff_diff_maxgcd = diff_maxgcd.x + diff_maxgcd.y,
-           eff_interaction = `maxgcd:diff_maxgcd.x` + `maxgcd:diff_maxgcd.y`) %>%
+           eff_diff_maxgcd = diff_maxgcd.x + diff_maxgcd.y) %>%
   dplyr::select(.chain, .iteration, f, age, 
-                eff_maxgcd, eff_diff_maxgcd, eff_interaction) %>%
+                eff_maxgcd, eff_diff_maxgcd) %>%
   arrange(.chain, .iteration, age)
 
   # need to confirm correct line-up
   by_taxon <- map(cv[-1], ~ full_join(core, .x, by = c('.chain', '.iteration', 'age'))) %>%
     map(., ~ .x %>%
         mutate(taxon_eff_maxgcd = eff_maxgcd + maxgcd,
-               taxon_eff_diff_maxgcd = eff_diff_maxgcd + diff_maxgcd,
-               taxon_eff_interaction = eff_interaction + `maxgcd:diff_maxgcd`))
+               taxon_eff_diff_maxgcd = eff_diff_maxgcd + diff_maxgcd))
   by_taxon <- reduce(by_taxon, bind_rows)
   
   by_taxon <- by_taxon %>%
     gather(key, value, -.chain, -.iteration, -f.x, 
            -age, -f.y, -`(Intercept)`, -type,
-           -eff_maxgcd, -eff_diff_maxgcd, -eff_interaction,
-           -diff_maxgcd, -maxgcd, -`maxgcd:diff_maxgcd`) %>%
+           -eff_maxgcd, -eff_diff_maxgcd,
[…]

### 2018-07-11 R/process_foo.r

@@ -90,7 +90,6 @@ prepare_analysis <- function(x, fg = NULL) {
   tb
 }
 
-
 #' Break time data up into bins
 #' 
 #' Have fun with this. basic rules. greater than equal to base, less than top.
@@ -98,17 +97,17 @@ prepare_analysis <- function(x, fg = NULL) {
 #' @param x vector of ages
 #' @param by bin width
 #' @return vector of bin memberships
-break_my <- function(x, by = 1, number = NULL) {
+break_my <- function(x, by = NULL, number = NULL) {
   top <- ceiling(max(x))
   bot <- floor(min(x))
   if(!is.null(by)) {
     unt <- seq(from = bot, to = top, by = by)
-  } else {
-    unt <- seq(from = bot, to = top, length.out = number)
+  } else if(!is.null(number)) {
+    unt <- seq(from = bot, to = top, length.out = number + 1)
   }
   unt1 <- unt[-length(unt)]
   unt2 <- unt[-1]
-  uu <- map(map2(unt1, unt2, ~ x >= .x & x < .y), which)
+  uu <- map2(unt1, unt2, ~ which(between(x, left = .x, right = .y)))
 
   y <- x
   for(ii in seq(length(uu))) {
@@ -133,7 +132,7 @@ check_class <- function(x) class(x) != 'try-error'
 
 #' Get optimal cut from ROC results
 #' 
-#' Uses ROCR
+#' Uses ROCR. I got this from somewhere on the internet.
 #'
 #' @param perf performance object
 #' @param pred prediction object
@@ -160,3 +159,54 @@ get_cutpoints <- function(pp_prob, target) {
   perf <- map(pred, ~ performance(.x, measure = 'tpr', x.measure = 'fpr'))
   cutpoints <- map2(perf, pred, opt_cut)
 }
+
+
+#' Reassign class membership based on new cutpoints
+#'
+#' Give probabilities and new cutpoint level, reassigns classes.
+#'
+#' @param pp_prob list of posterior predictive draws, where each list element is the PPD for one model.
+#' @param list_cutpoint list of new cutpoints as estimated from get_cutpoints
+#' @return list of reassigned class predictions
+cut_newpoint <- function(pp_prob, list_cutpoint) {
+  pp_est_new <- list()
+  for(jj in seq(length(pp_prob))) {
[…]

### 2018-08-14 doc/intro.tex
gsa abstract

@@
 
+One of the great promises of paleobiology is that by studying the past we can better predict the future.
+This promise is particularly pertinent given suggestions that risk assessments for some modern species could be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
+Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)?
+(2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?
+
+To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
+Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature.
+Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.
+
+The effects of the historical covariates are extremely small and vary considerably over time.
+For example, the effect of change in geographic range size can either increase or decrease probability of extinction depending on when in the Cenozoic the observation takes place.
+The improvement in predictive power by including these historical covariates is modest at best and reflects the rarity of extinction events (i.e.
+class-imbalance) and the extremely stochastic nature of species survival.
+Correcting for class imbalance, we find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation.
+These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.
+This result implies that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
+The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.
+
+Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups.
+This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton.
+The greatest source of variance in survival probability is the timing of observation.
+Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age.
+
+Ultimately, we find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales.
+However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
+The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
+
+how predictable is extinction? how much does the past improve our predictions? how much does the past improve our predictions?
+
+we know that at almost all times, a species with geater than average geographic range is expected to have a lower extinction risk than a closely related species with a lower than average geographic range.
+what we don't know is if expanding or decreasing in geographic range over million-year time scales has any additional predictive information about a species' risk of going extinct.
+
 \end{document}

### 2018-08-14 doc/methods.tex
no interaction model
gsa abstract

@@
 The hope is that binning the data overcomes the heterogenity in age-models between the cores.
-% binning, [t - 1, t)
 
@@
 
-% lots of ways of varying
-% maxgcd * diff_maxgcd varies first by when then by fossil group.
-% non-nested varying intercept based on age, then by fossil group.
-
-% event ~ temp + lag1_temp + maxgcd * diff_maxgcd + 
-% (1 + maxgcd * diff_maxgcd | fact_mybin/fossil.group) +
-% (1 | fact_relage/fossil.group)
 The logistic regression can thus be expressed as 
-% scripts
-%   t = death? 0 if not last observation
-%   \beta = effect range vars (varies over time and by fossil group)
-%   \gamma = effect temp vars
-%   a = age effect
-% subscripts
-%   w = when in My
-%   f = fossil group
-%   d = age at observation
-% 
 \begin{equation}
   \begin{aligned}
-    t_{i} &= logit^{-1}(\beta_{w[i], f[i]} + \beta_{w[i], f[i]} range_{i} + \beta_{w[i], f[i]} rangediff_{i} + \beta_{w[i], f[i]} range_{i} rangediff_{i} + \gamma_{1} temp_{w[i]} + \gamma_{2} templag_{w[i]} + a_{d[i], f[i]}) \\
+    t_{i} &= logit^{-1}(\beta_{w[i], f[i]} + \beta_{w[i], f[i]} range_{i} + \beta_{w[i], f[i]} rangediff_{i} + \gamma_{1} temp_{w[i]} + \gamma_{2} templag_{w[i]} + a_{d[i], f[i]}) \\
     \beta_{w, f} &\sim MVN(\alpha_{f}, \Sigma) \\
@@
 Posterior convergence was determined using the following general and HMC specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)).
-For futher explanation of these diagnostic criteria, see the Stan Manual cITATION.
+For futher explanation of these diagnostic criteria, see the Stan Manual CITATION.
 
@@
 
-We considered four variants of this model: 1) only age-based varying intercepts, 2) age and time of observation varying intercepts, 3) age and taxonomic group varying intercepts, and 4) all three considered varying intercepts.
-These models were compared using three measures of model performance: WAIC and LOOIC which are estimates of a model's out-of-sample performance, and ROC which is a measure of a model's predictive performance.
+We considered four variants of this model: 1) historical covariates with varying intercepts and slopes, 2) no historical coviarates with varying intercepts and slopes, 3) historical covariates without varying intercepts and slopes, and 4)
 
@@
 
+\subsection{Prediction given class imbalance}
+
+When the classes are not approximately balanced, the base probability of 1 versus 0 is below 0.5.
+When the intercept term of the logistic regression is sufficiently negative (-2 or less), the only way to obtain a linear predictor value of 0.5 or greater is if the covariate effects are large or there the covariates are at extreme values.
+The 0.5 default cut-off point for binary prediction is much to high.
+A method for determining the new ``optimal'' cutpoint involves calculating the ROC curve for the data and finding the point along that curve which maximizes both true positive rate and the false positive rate; the combination with the greatest sum corresponds to the new optimal cutpoint CITATION.
+Class membership is then predicted given this new cutpoint.
+This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cutpoint.
+
+\subsection{Cross-validation}
+
+Cross-validation is a procedure for estimating the expected out-of-sample predictive error of a model CITATION.
+By dividing the data into sub-components (folds), fitting a model to some of the folds folds, and then predicting the response values for the held-out fold.
+For our dataset, expected out-of-sample AUC was estimated using five-fold cross-validation.
+For time-series data, the folds are approximately equal segments of time.
+The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold.
+With 63 time points, each of the five folds represents approximately 13 time points.
+Keep in mind, however, that each time point is represented by many (100-1000) individual observations.
+
+For each posterior draw of the cross-validation predictions, the optimal cutpoint was calculated as described above.
+Given these class assignments based on estimated optimal cutpoint, the AUC value of the out-of-sample estmates was calculated.
+AUC values are calculated for the entire fold and for each of the individual time points represented within that fold.
+
 \end{document}

### 2018-08-14 doc/results.tex
gsa abstract

--- 2018-06-13 doc/methods.tex
@@
 
-\section{Data Specifications}
+\section{Model choice and adequacy}
 
-Microfossil occurrence data was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax}.
-Occurrence information was downloaded for calcareous nannofossils, diatoms, dinoflaggellants, foraminifera, and radiolarians from the entire globe and having occurred between 120 and 0 Mya.
-This selection of species was then culled to just those species which have their first occurrence at most 63 My; this avoids those taxa which survived the K/Pg boundary, those taxa arrising just after the K/Pg boundary, and lines up the occurrences with the temperature time-series (discussed below).
+% loo table
 
-Data was binned into 1 My bins based on the estimated age of the fossil occurrence.
-The estimated ages of each occurrence is a product of core-specific age-models and are overly precise.
-The hope is that binning the data overcomes the heterogenity in age-models between the cores.
-% binning, [t - 1, t)
+% waic table
 
-The survival or extinction of a taxon is determined for each taxons' occurrences.
-For every occurrence except the last, the taxon has survived which is indicated by a 0.
-The last occurrence of the taxon is considered the bin in which the taxon has gone extinct.
-This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous CITATIONS but is common with marine microfossil data CITATIONS.
+% ROC model comparison
 
-A taxon's geographic range during a 1 My bin was calculated as the maximum great circle distance on an ellipsoid between any two occurrences, also called a geodesic.
-Geographic range was measured in kilometers.
+% ROC model comparison time series
 
-Geographic range was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1.
+\section{Cross-validation}
 
-Temperature data is based on estimates from Cramer et al CITATION.
-These estimates are based on Magnesian/Calcium isotope ratios and are a more accurate estimate of global temperature than the standard Oxygen isotope-based estimates because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
-metioric water).
-Cramer et al.
-provide temperature estimates for every 0.1 My from 0 to 63 Mya; we binned these estimates every 1 My in accordance with the way we binned all fossil occurrences.
+% ROC OOS estimate
 
-\section{Model Specifications}
+% ROC OOS estimate time series
 
-In survival analysis, the hazard funciton describes the instanenous rate of extinction at time \(t\) given that the species has lived up to that point.
-For discrete-time survival data, this is expressed as \(\lambda(t | X) = P(T = t | T \geq t, X)\) for \(t = 1, \cdots N\) where \(t\) is the status of a species at a time point, \(T\) is the time of extinction of the species, and \(i = 1, 2, \cdots, N\) with \(N\) being the total number of observations.
+\section{Parameter estimates}
 
-In survival analysis the hazard function describes the in stannous rate of extinction for a species given matrix of covariates X, \(\lambda(t | X)\).
-We parameterized the hazard function as a logistic regression/generalized linear model defined \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is an inverse logit link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\).
+% effects, average
 
-\(\Theta\) is then parameterized as a hierarchical model with multiple non-nested varying intercepts CITATION.
-First, we considered two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\).
-We also considered the interaction between geographic range during intervals \(t\) and \(t - 1\).
-At the first observation for a taxon, the diff of geographic range is 0.
+% effects, time series
 
-Second, we considered two aspects of global temperature: mean temperature during interval \(t\), and the mean temperature during interval \(t - 1\).
+% variance components
 
-Our model had up to three non-nested varying intercepts: age of observation \(a\), bin of observation \(b\), and taxonomic group of the observation \(c\).
-Age of observation is a critical varying intercept as this effect, when considered along with the intercept \(\beta_{0}\), captures changes to baseline hazard with species age.
-Because these factors are modeled as varying intercepts, we are assuming that the effects of geographic range (\(\beta_{1} \dots \beta_{3}\)) and temperature (\(\beta_{4} \dots \beta_{5}\)) are constant over time.
-
-The logistic regression can thus be expressed as 
-\begin{equation}
-  \begin{aligned}
-    t_{i} = logit^{-1}(\beta_{0} + \beta_{1} r_{i} + \beta_{2} d_{i} + \beta_{3} r_{i} d_{i} + a_{j[i]} + b_{k[i]} + c_{f[i]}) \\
-  \end{aligned}
-  \label{eq:core}
-\end{equation}
-with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation e.g.
-\(j[i]\) is the age of the \(i\)th observation.
-\(a_{j[i]}\) is taxon age in bins at time of observation where \(j = 1, 2, \cdots, J\) and \(J\) is the maximum observed age of any taxon.
-\(b_{k}\) is the time of observation where \(k = 1, 2, \cdots, K\) and \(K\) is the number of time bins observed.
-Finally, \(c_{f}\) is the taxonomic group of an observation where \(f = 1, 2, \cdots, F)\) and \(F\) is the total number of taxonomic groups analyzed.
-
-\section{Model Parameter Estimation}
-
-We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language CITATION.
-This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R.
-Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling.
-In order to prevent divergent transitions the adapt delta value was increased to 0.99; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.
-
-Posterior convergence was determined using the following general and HMC specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)).
-For futher explanation of these diagnostic criteria, see the Stan Manual cITATION.
-
-\section{Model Selection and Adequacy}
-
-We considered four variants of this model: 1) only age-based varying intercepts, 2) age and time of observation varying intercepts, 3) age and taxonomic group varying intercepts, and 4) all three considered varying intercepts.
-These models were compared using three measures of model performance: WAIC and LOOIC which are estimates of a model's out-of-sample performance, and ROC which is a measure of a model's predictive performance.
-
-To determine how many non-nested varying-intercept terms were possible to include without overly biasing out models' parameter estimates, the four variant models described above were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior.
-CITATION.
-WAIC and LOOIC are interpreted similarly to AIC, with lower values indicating greater expected out-of-sample performance.
-Additionally, the Bayesian nature of WAIC and LOOIC mean that they are calculated with standard errors allowing for straight forward comparisons.
-We selected the model with lowest WAIC and LOOIC and all models within one standard error of its WAIC/LOOIC estimate.
-
-Given these selected models, we did a further comparison between models including the change in geographic range from \(t - 1\) to \(t\), and its interaction with geographic range at \(t\), in the model to determine how much our predictions of extinction risk are improved by the inclusion of ``the past.''
-
-The model adequacy was measured using the area under the reciever operating characteristic curve (AUC).
-This measure is commonly used in classification problems like this one as it has the desireable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION.
-This value ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.
-AUC was calculated for the dataset as a whole, and for each of time bins.
+% risk estimate compared to change in geo-range
 

### 2018-08-14 doc/methods.tex
modified methods

@@
 
-Temperature data is based on estimates from Cramer et al CITATION.
-These estimates are based on Magnesian/Calcium isotope ratios and are a more accurate estimate of global temperature than the standard Oxygen isotope-based estimates because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
+Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from Cramer et al CITATION.
+These are considered a more accurate estimate of global temperature than the standard Oxygen isotope-based estimates because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
 metioric water).
-Cramer et al.
-provide temperature estimates for every 0.1 My from 0 to 63 Mya; we binned these estimates every 1 My in accordance with the way we binned all fossil occurrences.
+Our data source, Cramer et al., provides temperature estimates for every 0.1 My from 0 to 63 Mya; we binned these estimates every 1 My to match with how we binned all fossil occurrences.
+The mean of these values was used as the temperature estimate for that 1 My interval (Fig.
+\ref{fig:temp_curve}).
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=\0.5textheight,keepaspectratio=true]{figure/cramer_temp}
+  \caption{Comparison of initial temperature estimates from Cramer et al.
+(goldenrod) versus the binned values used in this analysis (blue).
+The initial values are for every 0.1 My while our bins are defined for every 1 My.}
+  \label{fig:temp_curve}
+\end{figure}
 
@@
 For discrete-time survival data, this is expressed as \(\lambda(t | X) = P(T = t | T \geq t, X)\) for \(t = 1, \cdots N\) where \(t\) is the status of a species at a time point, \(T\) is the time of extinction of the species, and \(i = 1, 2, \cdots, N\) with \(N\) being the total number of observations.
-
-In survival analysis the hazard function describes the in stannous rate of extinction for a species given matrix of covariates X, \(\lambda(t | X)\).
 We parameterized the hazard function as a logistic regression/generalized linear model defined \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is an inverse logit link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\).
@@
 At the first observation for a taxon, the diff of geographic range is 0.
-
-Second, we considered two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e. mean temperature during interval \(t - 1\).)
+Secondly, we considered two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e.
+mean temperature during interval \(t - 1\).)
 
@@
 
-We considered four variants of this model: 1) historical covariates with varying intercepts and slopes, 2) no historical coviarates with varying intercepts and slopes, 3) historical covariates without varying intercepts and slopes, and 4)
+We considered four variants of this model: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect).
 
@@
 With 63 time points, each of the five folds represents approximately 13 time points.
-Keep in mind, however, that each time point is represented by many (100-1000) individual observations.
+Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
 

### 2018-08-16 doc/methods.tex
specified priors. methods expanded

@@
 Geographic range was measured in kilometers.
-
 Geographic range was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1.
+This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly interpretable.
 
@@
 \ref{fig:temp_curve}).
+Temperature was transformed and standardized the in the same manner as geographic range.
 
@@
   \centering
-  \includegraphics[width=\textwidth,height=\0.5textheight,keepaspectratio=true]{figure/cramer_temp}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/cramer_temp}
   \caption{Comparison of initial temperature estimates from Cramer et al.
-(goldenrod) versus the binned values used in this analysis (blue).
+CITATION (goldenrod) versus the binned values used in this analysis (blue).
 The initial values are for every 0.1 My while our bins are defined for every 1 My.}
@@
 
-In survival analysis, the hazard funciton describes the instanenous rate of extinction at time \(t\) given that the species has lived up to that point.
-For discrete-time survival data, this is expressed as \(\lambda(t | X) = P(T = t | T \geq t, X)\) for \(t = 1, \cdots N\) where \(t\) is the status of a species at a time point, \(T\) is the time of extinction of the species, and \(i = 1, 2, \cdots, N\) with \(N\) being the total number of observations.
-We parameterized the hazard function as a logistic regression/generalized linear model defined \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is an inverse logit link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\).
+In survival analysis, the hazard funciton describes the instanenous rate of extinction of a species.
+Species occurrence are recorded for all \(k\) 1 My intervals.
+For the discrete time intervals \(T = 1, \cdots, k\), extinction is defined as \(T = t\).
+The discrete-time hazard function is then written
+\begin{equation}
+  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k \\
+  \label{eq:hazard}
+\end{equation}
+where \(X\) is a matrix of covariates. This statement is the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survied up until \(t\).
 
-\(\Theta\) is then parameterized as a hierarchical model with multiple non-nested varying intercepts CITATION.
+The corresponding survival function, describing the probability of a species going extinct after time \(t\) is expressed
+\begin{equation}
+  S(t | X) = P(T > t | X) = \prod{i = 1}{t}(1 - \lambda(i | X).
+  \label{eq:surv}
+\end{equation}
+
+Convinently, the hazard function (Eq.
+\ref{eq:hazard}) is easily reparameterized as logistic regression.
+This form of logistic regression model defined where \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is an logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\).
+\(\Theta\) is then parameterized as any regression model.
+
+In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes CITATION.
 First, we considered two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\).
@@
   \begin{aligned}
-    t_{i} &= logit^{-1}(\beta_{w[i], f[i]} + \beta_{w[i], f[i]} range_{i} + \beta_{w[i], f[i]} rangediff_{i} + \gamma_{1} temp_{w[i]} + \gamma_{2} templag_{w[i]} + a_{d[i], f[i]}) \\
+    t_{i} &\sim \text{Bernoulli}(\Theta) \\
+    \Theta &= logit^{-1}(\beta_{w[i], f[i]} + \beta_{w[i], f[i]} r_{i} + \beta_{w[i], f[i]} d_{i} + \gamma_{1} c_{w[i]} + \gamma_{2} l_{w[i]} + a_{d[i], f[i]}) \\
     \beta_{w, f} &\sim MVN(\alpha_{f}, \Sigma) \\
@@
 Finally, \(c_{f}\) is the taxonomic group of an observation where \(f = 1, 2, \cdots, F)\) and \(F\) is the total number of taxonomic groups analyzed.
+
+% priors
+
+The basic core of the model described above \ref{eq:core} can be written in R formula syntax as
+\begin{equation}
+  \begin{aligned}
+    \text{event} &\sim \text{range} + \text{range\_diff} + \text{temp} + \text{temp\_lag} + \\
+    &\quad (1 + \text{range} + \text{range\_diff} | \text{mybin} / \text{phylum}) + \\
+    &\quad (1 | \text{age} / \text{phylum}).
+  \end{aligned}
+\end{equation}
+This formula is then used with a call to \begin{texttt} stan\_glmer \end{texttt} as Bernoulli family model which, along with the relevant data and prior choices, yields parameter estimates.
 

### 2018-08-16 R/fit_model.r

@@ -22,9 +22,9 @@ survi <- read_rds('../data/survival.rds')
 counti <- read_rds('../data/counting.rds')
 
 # transform the data for analysis
-counti <- prepare_analysis(counti)
+counti_trans <- prepare_analysis(counti)
 
-# fit the model
+#pp fit the model
 
 # past and vary
 form <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd + 
@@ -49,6 +49,16 @@ form4 <- update(form3,
 
 forms <- list(form, form2, form3, form4)
 
-disc_fit <- map(forms, ~ stan_glmer(.x, family = 'binomial', data = counti,
-                                    adapt_delta = 0.99, thin = 4))
+# partial evaluation to fill in priors
+glmer_part <- 
+  partial(stan_glmer, 
+          prior = normal(0, 1, autoscale = FALSE),
+          prior_intercept = normal(0, 10, autoscale = FALSE),
+          prior_aux = cauchy(0, 5, autoscale = FALSE))
+disc_fit <- map(forms, ~ glmer_part(.x, 
+                                    family = 'binomial', 
+                                    data = counti_trans,
+                                    adapt_delta = 0.999, 
+                                    thin = 4))
+
 write_rds(disc_fit, path = '../data/disc_fit.rds')

### 2018-08-16 R/fit_predict_model.r

@@ -47,7 +47,12 @@ form1 <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd +
                  (1 + maxgcd + diff_maxgcd | fact_mybin/fossil_group) + 
                  (1 | fact_relage/fossil_group))
 
-fit <- map(counti_accum, ~ stan_glmer(form1, 
+glmer_part <- 
+  partial(stan_glmer, 
+          prior = normal(0, 1, autoscale = FALSE),
+          prior_intercept = normal(0, 10, autoscale = FALSE),
+          prior_aux = cauchy(0, 5, autoscale = FALSE))
+fit <- map(counti_accum, ~ glmer_part(form1, 
                                       family = 'binomial', 
                                       data = .x,
                                       adapt_delta = 0.9999, 

### 2018-08-16 R/inspect_model.r

@@ -34,24 +34,21 @@ disc_fit <- read_rds('../data/disc_fit.rds')
 disc_best <- disc_fit[[1]]
 
 
-# check prior shrinkage to get an idea of "information learned"
+# check how much information learned
 priors <- prior_summary(disc_best)
 prmn <- priors$prior$location
-prsc <- priors$prior$adjusted_scale
+prsc <- priors$prior$scale
 
-po <- disc_best %>% 
+posc <- disc_best %>% 
   spread_samples(maxgcd,
                  diff_maxgcd,
                  temp,
                  lag1_temp) %>% 
-  clean_names
-pomn <- po %>%
-  dplyr::summarize_at(vars(-chain, -iteration), funs(mean(.)))
-posc <- po %>% 
+  clean_names %>% 
   dplyr::summarize_at(vars(-chain, -iteration), funs(sd(.)))
 
-posterior_shrinkage <- 1 - (posc / prsc) ^ 2
 # 1 = lot of shrinkage, 0 = no shrinkage
+posterior_shrinkage <- 1 - (posc / prsc) ^ 2
 
 
 # posterior intervals

### 2018-08-16 R/process_foo.r

@@ -58,7 +58,7 @@ prepare_analysis <- function(x, fg = NULL) {
   tb <- tb %>%
     arrange(fullname, desc(mybin)) %>%
     mutate_at(.vars = vars(ncell, latext, maxgcd, area, temp),
-              .funs = ~ arm::rescale(log1p(.x))) %>%
+              .funs = ~ scale(log1p(.x))) %>%
     group_by(fullname) %>%
     mutate(diff_maxgcd = c(0, diff(maxgcd)),
            lag1_maxgcd = lag(maxgcd, default = 0),

### 2018-08-17 doc/methods.tex
grew methods more

@@
     t_{i} &\sim \text{Bernoulli}(\Theta) \\
-    \Theta &= logit^{-1}(\beta_{w[i], f[i]} + \beta_{w[i], f[i]} r_{i} + \beta_{w[i], f[i]} d_{i} + \gamma_{1} c_{w[i]} + \gamma_{2} l_{w[i]} + a_{d[i], f[i]}) \\
-    \beta_{w, f} &\sim MVN(\alpha_{f}, \Sigma) \\
-    \alpha_{f} &\sim MVN(0, \Sigma) \\
-    \gamma &\sim N(0, \sigma) \\
-    a_{w, f} &\sim MVN(\delta_{f}, \Sigma) \\
-    \delta_{f} &\sim MVN(0, \Sigma) \\
+    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + a_{d[i], p[i]}) \\
+    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
+    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
+    a_{w, p} &\sim MVN(\delta_{p}, \Sigma_{a}) \\
+    \delta_{p} &\sim MVN(0, \Sigma_{\delta}) \\
   \end{aligned}
@@
 \end{equation}
-with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation e.g.
-\(j[i]\) is the age of the \(i\)th observation.
-\(a_{j[i]}\) is taxon age in bins at time of observation where \(j = 1, 2, \cdots, J\) and \(J\) is the maximum observed age of any taxon.
-\(b_{k}\) is the time of observation where \(k = 1, 2, \cdots, K\) and \(K\) is the number of time bins observed.
-Finally, \(c_{f}\) is the taxonomic group of an observation where \(f = 1, 2, \cdots, F)\) and \(F\) is the total number of taxonomic groups analyzed.
+with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.
 
-% priors
+\(B\) is a matrix of regression coefficients indexed by both time of observation and phylum.
+
+\(\mu\) is a vector, the first element of which corresponds to the overall model intercept and the others being the overall average estimate of the regression coefficients.
+
+To complete the generative model, we need to assign final priors to the ``top level'' parameters
+\begin{equation}
+  \begin{aligned}
+    \mu_{c} &\sim 
+      \begin{cases}
+        N(0, 10) & \text{if } c = 1 \\
+        N(0, 1) & \text{if } c > 1 \\
+      \end{cases}
+    \delta &\sim N(0, 1) \\
+    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
+    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
+    \Sigma_{a} &= diag(\tau_{a}) \Omega_{a} diag(\tau_{a}) \\
+    \Sigma_{\delta} &= diag(\tau_{\delta}) \Omega_{\delta} diag(\tau_{\delta}) \\
+    \tau_{B} &\sim C^{+}(5) \\
+    \tau_{\alpha} &\sim C^{+}(5) \\
+    \tau_{a} &\sim C^{+}(5) \\
+    \tau_{\delta} &\sim C^{+}(5) \\
+    \Omega_{B} &\sim LKJ(1) \\
+    \Omega_{\alpha} &\sim LKJ(1) \\
+    \Omega_{a} &\sim LKJ(1) \\
+    \Omega_{\delta} &\sim LKJ(1). \\
+  \end{aligned}
+  \label{eq:priors}
+\end{equation}
+The decomposition of the covariance matrices (e.g \(\Sigma_{B}\)) allows priors the variance and correlation aspects of covariance to be defined separately.
+These priors are also slightly more interpretable than a, for example, inverse-Wishart distributed prior for the covariance matrices.
+This decomposition is recommended by CITATION.
 
@@
 This formula is then used with a call to \begin{texttt} stan\_glmer \end{texttt} as Bernoulli family model which, along with the relevant data and prior choices, yields parameter estimates.
+An example line of R code to fit this mode given a data.frame ``data'' could be written as:
+\begin{verbatim}
+  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
+                       (1 + range + range_diff | mybin/phylum) + 
+                       (1 | age/phylum), 
+             prior = normal(0, 1, autoscale = FALSE), 
+             prior_intercept = normal(0, 10, autoscale = FALSE), 
+             prior_aux = cauchy(0, 5, autoscale = FALSE), 
+             data = data, 
+             adapt_delta = 0.9999, 
+             thin = 4)
+\end{verbatim}
+
+See our code repository LINK for full code details.
+Our code uses ``tidyverse'' tools such as dplyr CITATION, purr CITATION, and tidybayes CITATION, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
 

### 2018-08-17 R/fit_model.r

@@ -58,7 +58,7 @@ glmer_part <-
 disc_fit <- map(forms, ~ glmer_part(.x, 
                                     family = 'binomial', 
                                     data = counti_trans,
-                                    adapt_delta = 0.999, 
+                                    adapt_delta = 0.9999, 
                                     thin = 4))
 
 write_rds(disc_fit, path = '../data/disc_fit.rds')

### 2018-08-20 doc/methods.tex
full methods ready to be read by other humans

@@
 
-Microfossil occurrence data was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax}.
-Occurrence information was downloaded for calcareous nannofossils, diatoms, dinoflaggellants, foraminifera, and radiolarians from the entire globe and having occurred between 120 and 0 Mya.
-This selection of species was then culled to just those species which have their first occurrence at most 63 My; this avoids those taxa which survived the K/Pg boundary, those taxa arrising just after the K/Pg boundary, and lines up the occurrences with the temperature time-series (discussed below).
+We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax}.
+We downloaded all occurrence information for calcareous nannofossils, diatoms, foraminifera, and radiolarians from the entire globe and having occurred between 120 and 0 Mya.
+This selection of occurrences was then culled to just those species which have their first occurrence at most 63 My; this avoids those taxa which survived the K/Pg boundary, those taxa arising just after the K/Pg boundary, and lines up the occurrences with the temperature time-series (discussed below).
 
-Data was binned into 1 My bins based on the estimated age of the fossil occurrence.
-The estimated ages of each occurrence is a product of core-specific age-models and are overly precise.
-The hope is that binning the data overcomes the heterogenity in age-models between the cores.
+All occurrences were binned into 1 My bins based on the estimated age of the fossil occurrence.
+Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources.
+We then determined for each species the million year bins when it had occurred.
+For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0.
+The last occurrence of that species is considered the bin in which the taxon has gone extinct.
+This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species CITATIONS.
+However, this practice is common with marine microfossil data CITATIONS.
 
-The survival or extinction of a taxon is determined for each taxons' occurrences.
-For every occurrence except the last, the taxon has survived which is indicated by a 0.
-The last occurrence of the taxon is considered the bin in which the taxon has gone extinct.
-This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous CITATIONS but is common with marine microfossil data CITATIONS.
-
-A taxon's geographic range during a 1 My bin was calculated as the maximum great circle distance on an ellipsoid between any two occurrences, also called a geodesic.
-Geographic range was measured in kilometers.
-Geographic range was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1.
-This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly interpretable.
+A taxon's geographic range was calculated for each of the 1 My bins in which it occurred.
+Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e.
+the globe) between any two occurrences; this measure is also called a geodesic.
+This distance was measured in kilometers.
+The geodesic distance was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1.
+This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly comparable in magnitude.
 
@@
 These are considered a more accurate estimate of global temperature than the standard Oxygen isotope-based estimates because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
-metioric water).
-Our data source, Cramer et al., provides temperature estimates for every 0.1 My from 0 to 63 Mya; we binned these estimates every 1 My to match with how we binned all fossil occurrences.
-The mean of these values was used as the temperature estimate for that 1 My interval (Fig.
+meteoric water).
+This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
+Our data source, Cramer et al., provides temperature estimates for every 0.1 My from 0 to 63 Mya.
+We binned these estimates into 1 My intervals as we did with the fossil occurrences.
+The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig.
 \ref{fig:temp_curve}).
-Temperature was transformed and standardized the in the same manner as geographic range.
+Temperature was then transformed and standardized the in the same manner as geographic range (above).
 
@@
 
-In survival analysis, the hazard funciton describes the instanenous rate of extinction of a species.
-Species occurrence are recorded for all \(k\) 1 My intervals.
+In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and relevant covariates.
+The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals.
 For the discrete time intervals \(T = 1, \cdots, k\), extinction is defined as \(T = t\).
-The discrete-time hazard function is then written
+The discrete time hazard function is defined as
 \begin{equation}
-  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k \\
+  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k. \\
   \label{eq:hazard}
 \end{equation}
-where \(X\) is a matrix of covariates. This statement is the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survied up until \(t\).
 
-The corresponding survival function, describing the probability of a species going extinct after time \(t\) is expressed
-\begin{equation}
-  S(t | X) = P(T > t | X) = \prod{i = 1}{t}(1 - \lambda(i | X).
-  \label{eq:surv}
-\end{equation}
+%The complement of the hazard function is called the survival function, which describes the probability of a species going extinct after time \(t\).
+Just as extinction is defined as when the current occurrence is the last occurrence, or \(T = t\), survival is defined as when the current occurrence is less than (in age) than the final occurrence, or \(T > t\).
+The discrete time survival function is expressed as
+%\begin{equation}
+%  S(t | X) = P(T > t | X) = \prod{i = 1}{t}(1 - \lambda(i | X).
+%  \label{eq:surv}
+%\end{equation}
 
-Convinently, the hazard function (Eq.
-\ref{eq:hazard}) is easily reparameterized as logistic regression.
-This form of logistic regression model defined where \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is an logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\).
-\(\Theta\) is then parameterized as any regression model.
+The hazard function (Eq.
+\ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\).
+\(h(\Theta)\) is then modeled as with any regression as it is defined for all real-values.
+In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes CITATION.
 
-In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes CITATION.
-First, we considered two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\).
-We also considered the interaction between geographic range during intervals \(t\) and \(t - 1\).
-At the first observation for a taxon, the diff of geographic range is 0.
-Secondly, we considered two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e.
-mean temperature during interval \(t - 1\).)
+Our covariates matrix \(X\) is a \(N \times D\) matrix where \(N\) is the total number of observations and \(D\) is the total number of covariates.
+The first column of \(X\) is entirely 1's as it corresponds to the intercept term in the regression model.
+The next two columns of \(X\) are two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\).
+The difference in geographic range was calculated from the transformed and standardized geographic range values; this means that change in geographic range is in units of changes in standard deviations.
+The final two columns are two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e.
+mean temperature during interval \(t - 1\).) As with change to geographic range, the lag of temperature is based on the transformed and standardized temperature estimates.
 
-The logistic regression can thus be expressed as 
+The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\).
+These regression coefficients are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\).
+These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept.
+
+This logistic regression model, minus the final selection of priors, is expressed as 
 \begin{equation}
@@
 
-\(B\) is a matrix of regression coefficients indexed by both time of observation and phylum.
-
-\(\mu\) is a vector, the first element of which corresponds to the overall model intercept and the others being the overall average estimate of the regression coefficients.
-
 To complete the generative model, we need to assign final priors to the ``top level'' parameters
@@
         N(0, 10) & \text{if } c = 1 \\
-        N(0, 1) & \text{if } c > 1 \\
+        N(0, 3) & \text{if } c > 1 \\
       \end{cases}
@@
 
-The basic core of the model described above \ref{eq:core} can be written in R formula syntax as
-\begin{equation}
-  \begin{aligned}
-    \text{event} &\sim \text{range} + \text{range\_diff} + \text{temp} + \text{temp\_lag} + \\
-    &\quad (1 + \text{range} + \text{range\_diff} | \text{mybin} / \text{phylum}) + \\
-    &\quad (1 | \text{age} / \text{phylum}).
-  \end{aligned}
-\end{equation}
-This formula is then used with a call to \begin{texttt} stan\_glmer \end{texttt} as Bernoulli family model which, along with the relevant data and prior choices, yields parameter estimates.
-An example line of R code to fit this mode given a data.frame ``data'' could be written as:
-\begin{verbatim}
-  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
-                       (1 + range + range_diff | mybin/phylum) + 
-                       (1 | age/phylum), 
-             prior = normal(0, 1, autoscale = FALSE), 
-             prior_intercept = normal(0, 10, autoscale = FALSE), 
-             prior_aux = cauchy(0, 5, autoscale = FALSE), 
-             data = data, 
-             adapt_delta = 0.9999, 
-             thin = 4)
-\end{verbatim}
-
-See our code repository LINK for full code details.
-Our code uses ``tidyverse'' tools such as dplyr CITATION, purr CITATION, and tidybayes CITATION, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
-
 \section{Model Parameter Estimation}
@@
 In order to prevent divergent transitions the adapt delta value was increased to 0.99; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.
+An example line of R code to fit this mode given a data.frame of all necessary data called ``data'' could be written as:
+\begin{verbatim}
+  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
+                       (1 + range + range_diff | mybin/phylum) + 
+                       (1 | age/phylum), 
+             prior = normal(0, 3, autoscale = FALSE), 
+             prior_intercept = normal(0, 10, autoscale = FALSE), 
+             prior_aux = cauchy(0, 5, autoscale = FALSE), 
+             data = data, 
+             adapt_delta = 0.99, 
+             thin = 4)
+\end{verbatim}
 
 Posterior convergence was determined using the following general and HMC specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)).
-For futher explanation of these diagnostic criteria, see the Stan Manual CITATION.
+For further explanation of these diagnostic criteria, see the Stan Manual CITATION.
 
@@
 
-To determine how many non-nested varying-intercept terms were possible to include without overly biasing out models' parameter estimates, the four variant models described above were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior.
-CITATION.
+To determine how many non-nested varying-intercept terms were possible to include without overly biasing out models' parameter estimates, the four variant models described above were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior CITATION.
 WAIC and LOOIC are interpreted similarly to AIC, with lower values indicating greater expected out-of-sample performance.
-Additionally, the Bayesian nature of WAIC and LOOIC mean that they are calculated with standard errors allowing for straight forward comparisons.
-We selected the model with lowest WAIC and LOOIC and all models within one standard error of its WAIC/LOOIC estimate.
+Additionally, the Bayesian nature of WAIC and LOOIC mean that they are calculated with standard errors allowing for a more nuanced understanding the actual amount of difference between the models.
+We selected the model with the lowest WAIC and LOOIC, which happened to be are most parameter rich model.
 
-Given these selected models, we did a further comparison between models including the change in geographic range from \(t - 1\) to \(t\), and its interaction with geographic range at \(t\), in the model to determine how much our predictions of extinction risk are improved by the inclusion of ``the past.''
-
-The model adequacy was measured using the area under the reciever operating characteristic curve (AUC).
-This measure is commonly used in classification problems like this one as it has the desireable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION.
+Model adequacy was measured using the area under the receiver operating characteristic curve (AUC).
+This measure is commonly used in classification problems like this one as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION.
 This value ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.
 AUC was calculated for the dataset as a whole, and for each of time bins.
+These values are measures of in-sample predictive accuracy.
 
@@
 
+One of the issues with our data set is extreme class imbalance; this means that there are many more survival observations than extinction observations.
 When the classes are not approximately balanced, the base probability of 1 versus 0 is below 0.5.
 When the intercept term of the logistic regression is sufficiently negative (-2 or less), the only way to obtain a linear predictor value of 0.5 or greater is if the covariate effects are large or there the covariates are at extreme values.
-The 0.5 default cut-off point for binary prediction is much to high.
-A method for determining the new ``optimal'' cutpoint involves calculating the ROC curve for the data and finding the point along that curve which maximizes both true positive rate and the false positive rate; the combination with the greatest sum corresponds to the new optimal cutpoint CITATION.
-Class membership is then predicted given this new cutpoint.
-This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cutpoint.
+The 0.5 default cut-off point for binary prediction is much to high when events are rare.
+A method for determining the new ``optimal'' cut-point involves calculating the receiver operating characterist (ROC) curve from the linear predictor estimates and then identifying the point on the ROC curve which maximizes both true positive rate and the false positive rate.
+The new cut-point is then back-calculated from the true positive and false positive rates.
+Class membership is then predicted given this new cut-point.
+This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cut-point.
 
@@
 
-Cross-validation is a procedure for estimating the expected out-of-sample predictive error of a model CITATION.
-By dividing the data into sub-components (folds), fitting a model to some of the folds folds, and then predicting the response values for the held-out fold.
+In addition to model selection and adequacy, we are also interested in understanding how well our model predicts species extinction given new,future data.
+We estimated out-of-sample predictive error using cross-validation.
 For our dataset, expected out-of-sample AUC was estimated using five-fold cross-validation.
-For time-series data, the folds are approximately equal segments of time.
+For time-series data, the folds (data partitions) are approximately equal segments of time.
 The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold.
@@
 
-For each posterior draw of the cross-validation predictions, the optimal cutpoint was calculated as described above.
-Given these class assignments based on estimated optimal cutpoint, the AUC value of the out-of-sample estmates was calculated.
+For each posterior draw of the cross-validation predictions, the optimal cut-point was calculated as described above.
+Given these class assignments based on estimated optimal cut-point, the AUC value of the out-of-sample estimates was calculated.
 AUC values are calculated for the entire fold and for each of the individual time points represented within that fold.
 
+See our code repository LINK for full code details.
+Our code uses ``tidyverse'' tools such as dplyr CITATION, purr CITATION, and tidybayes CITATION, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
+
 \end{document}

### 2018-08-20 R/fit_model.r

@@ -52,13 +52,13 @@ forms <- list(form, form2, form3, form4)
 # partial evaluation to fill in priors
 glmer_part <- 
   partial(stan_glmer, 
-          prior = normal(0, 1, autoscale = FALSE),
+          prior = normal(0, 3, autoscale = FALSE),
           prior_intercept = normal(0, 10, autoscale = FALSE),
           prior_aux = cauchy(0, 5, autoscale = FALSE))
 disc_fit <- map(forms, ~ glmer_part(.x, 
                                     family = 'binomial', 
                                     data = counti_trans,
-                                    adapt_delta = 0.9999, 
+                                    adapt_delta = 0.99, 
                                     thin = 4))
 
 write_rds(disc_fit, path = '../data/disc_fit.rds')

### 2018-08-27 doc/results.tex
regen figures with nondivergent output

@@
 
-\section{Model choice and adequacy}
+\section{Model selection and adequacy}
 
-% loo table
-
-% waic table
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{rrr}
+    \hline
+    Model & looic \\
+    \hline
+    Past and vary & 12811.58 & 12807.61 \\
+    No past but vary & 12834.53 & 12831.51 \\
+    Past but no vary & 12836.19 & 12833.13 \\
+    No past or vary & 12849.43 & 12847.08 \\
+    \hline
+  \end{tabular}
+  \label{tab:selection}
+\end{table}
 
 % ROC model comparison
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/roc_hist}
+  \caption{Comparison of AUC estimates between the four models. These estimates are based on in-sample values. Higher values indicate greater performance.}
+  \label{fig:roc_hist}
+\end{figure}
 
 % ROC model comparison time series
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/roc_ts}
+  \caption{Comparison of time-specific AUC values between the four models.
+These estimates correspond to the models ability to predict or fit that section of the data.
+Higher values indicate greater performance.}
+  \label{fig:roc_ts}
+\end{figure}
 
@@
 % ROC OOS estimate
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/fold_auc}
+  \caption{Approximate out-of-sample AUC values calculated from five-fold cross-validation of the time series.
+The AUC estimates from each fold are labeled.
+Folds are numbered from oldest to youngest.}
+  \label{fig:fold_auc}
+\end{figure}
 
 % ROC OOS estimate time series
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/fold_auc_time}
+  \caption{Approximate out-of-sample AUC values calculated for each of the My intervals using five-fold cross-validation of the time series.
+The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
+The AUC estimates from each fold are labeled and are numbered from oldest to youngest.}
+  \label{fig:fold_auc_time}
+\end{figure}
 
@@
 % effects, average
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/effect_est}
+  \caption{Estimates of the overall average covariate effects.
+These are averaged both over time and across phyla.
+The entire posterior is presented, with the median estimate labeled and 80\% credible intervals.}
+  \label{fig:effect_est}
+\end{figure}
 
 % effects, time series
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/effect_time_group}
+  \caption{Covariate effect estimates for the effect of geographic range and the change in geographic range for all time points and phyla.
+The red line corresponds to the median estimate at each time, the dark blue area corresponds to the 50\% credible interval, the medium blue the 80\% credible interval, and the light-blue the 95\% credible interval.}
+  \label{fig:<+label+>}
+\end{figure}<++>
 

### 2018-08-27 R/fit_model.r

@@ -58,7 +58,7 @@ glmer_part <-
 disc_fit <- map(forms, ~ glmer_part(.x, 
                                     family = 'binomial', 
                                     data = counti_trans,
-                                    adapt_delta = 0.99, 
+                                    adapt_delta = 0.9999, 
                                     thin = 4))
 
 write_rds(disc_fit, path = '../data/disc_fit.rds')

### 2018-08-27 R/fit_predict_model.r

@@ -49,13 +49,13 @@ form1 <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd +
 
 glmer_part <- 
   partial(stan_glmer, 
-          prior = normal(0, 1, autoscale = FALSE),
+          prior = normal(0, 3, autoscale = FALSE),
           prior_intercept = normal(0, 10, autoscale = FALSE),
           prior_aux = cauchy(0, 5, autoscale = FALSE))
 fit <- map(counti_accum, ~ glmer_part(form1, 
                                       family = 'binomial', 
                                       data = .x,
-                                      adapt_delta = 0.9999, 
+                                      adapt_delta = 0.99999, 
                                       thin = 4))
 
 write_rds(fit, path = '../data/training_fit.rds')

### 2018-08-27 doc/results.tex
adding details to results section

@@
 
+Our best model, as selected via LOOIC and WAIC, is the parameter rich model allowing parameters to vary through time and includes our historical covariates (Table \ref{tab:selection}).
+Our best model is an improvement over the next best model by approximately 13 LOOIC and 24 WAIC points.
+However, the actual differences between these models are minor as the wide the standard errors of LOOIC and WAIC reveal.
 \begin{table}[ht]
   \centering
-  \begin{tabular}{rrr}
+  \begin{tabular}{ r r r r r }
     \hline
-    Model & looic \\
+    Model & looic & looic se & waic & waic se \\
     \hline
-    Past and vary & 12811.58 & 12807.61 \\
-    No past but vary & 12834.53 & 12831.51 \\
-    Past but no vary & 12836.19 & 12833.13 \\
-    No past or vary & 12849.43 & 12847.08 \\
+    Past and vary & 12811.58 & 179.27 & 12807.61 & 170.21 \\
+    No past but vary & 12834.53 & 179.07 & 12831.51 & 179.02 \\
+    Past but no vary & 12836.19 & 179.20 & 12833.13 & 179.15 \\
+    No past or vary & 12849.43 & 179.46 & 12847.08 & 179.43 \\
     \hline
@@
 
+Comparison of the in-sample AUC estimates between the four models reveals similar results as the selection criteria (Fig.
+\ref{fig:roc_hist}).
+While the parameter rich ``past and vary'' model has the greatest mean AUC when compared to the other three models, the other three models are quite similar.
+We chose to base our conclusions on the parameter rich ``past and vary'' model but these comparisons of model selection and adequacy illustrate the very small gain in performance that is achieved by allowing parameter effects to vary through time and including the historical covariates, though it appears that each of these choices on their own yield approximately equivalent models.
 % ROC model comparison
@@
 
+Additional comparison of in-sample model performance at each of the time intervals reveals just how similar our four models are to each other (Fig.
+\ref{fig:roc_ts}).
+There are few if any obvious differences between the models in their performances.
 % ROC model comparison time series
@@
 
+Because of the parameter rich ``past and vary'' model displays some modest improvements over the other models we choose this model for all further analysis.
+
 \section{Cross-validation}
 
+The approximate out-of-sample predictive performance for our selected model, as measured by AUC, is approximately identical to our in-sample performance (Fig.
+\ref{fig:fold_auc}).
+The quality of performance, however, is not great, with average out-of-sample AUC estimated to be just above 0.7 which is far from perfect.
+This result means that while we expect our model to yield consistent results when provided with new data, we do not expect that our predictions will be very accurate.
 % ROC OOS estimate
@@
 
-% ROC OOS estimate time series
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/fold_auc_time}
-  \caption{Approximate out-of-sample AUC values calculated for each of the My intervals using five-fold cross-validation of the time series.
+%When our cross-validation estimates are presented for each time interval, 
+%% ROC OOS estimate time series
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/fold_auc_time}
+%  \caption{Approximate out-of-sample AUC values calculated for each of the My intervals using five-fold cross-validation of the time series.
 The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
 The AUC estimates from each fold are labeled and are numbered from oldest to youngest.}
-  \label{fig:fold_auc_time}
-\end{figure}
+%  \label{fig:fold_auc_time}
+%\end{figure}
 
 \section{Parameter estimates}
+
+The overall average probability of a species going extinct during any interval is very low (Fig.
+\ref{fig:effect_est}).
+This result means that even if covariate effects are very large on the log-odds scale, when compared on the probability scale will only produce a minor change in probability of extinction.
+This is one of the hardest parts of logistic regression for people to understand.
+Effect sizes are on the log-odds scale but prediction is on the probability scale.
+When the intercept term is very negative or positive (greater than 2), covariate effects are fighting over an increasingly small amount of probability.
+Logistic regressions are easiest to interpret when the intercept term is approximately 0; this is a fundamental difficulty of dealing with very uneven data sets.
+
+As expected, a species with greater than average geographic range is expected to have a greater probability of surviving than a species with average or less than average geographic range (Fig.
+\ref{fig:effect_est})
+
+We estimated an overall positive effect of the change in geographic range on extinction probability; this means that a species gaining in geographic range between that observation and their previous observed geographic range is associated with a greater extinction risk than a species which decreased in geographic range (Fig.
+\ref{fig:effect_est}).
+
+We estimated an overall positive effect of both global temperature and the lag of global temperature with species extinction probability (Fig.
+\ref{fig:effect_est}.
+This type of effect means that when global temperature is above the Cenozoic average, we would expect higher than average extinction risk.
+Similarly, if global temperature was above the Cenozoic average in the time interval before the current one, we would also expect higher than average extinction risk.
+This means that during a run of above Cenozoic average global temperatures (2+ intervals) we would expect an even greater extinction risk than if just the current interval is of above average temperature.
 
@@
 
+The overall effect averages described above do not reflect the between time-interval variance of the effects.
+While the effect of geographic range has a consistent sign for the entire Cenozoic, the effects of the other covariates do not have a consistent sign (Fig.
+\ref{fig:effect_time_group}).
+
 % effects, time series
@@
 The red line corresponds to the median estimate at each time, the dark blue area corresponds to the 50\% credible interval, the medium blue the 80\% credible interval, and the light-blue the 95\% credible interval.}
-  \label{fig:<+label+>}
-\end{figure}<++>
+  \label{fig:effect_time_group}
+\end{figure}
 
 % variance components
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/variance_components}
+  \caption{Contribution of multi-level components to unmodeled variance.
+Larger values indicate a greater contribution to overall variance.
+Variance components labeled as ``overall'' represent variance between grouping factors, while those labeled  ``within groups'' correspond to variance within grouping factors.
+If the overall component is greater than the within component, then the groupings do not structure the overall variance.
+If the within component is greater than the overall component, then the overall variance is structured by the groupings.}
+  \label{fig:variance_components}
+\end{figure}
 

### 2018-08-30 doc/results.tex
clean up methods and fill out figure captions

@@
 
-Our best model, as selected via LOOIC and WAIC, is the parameter rich model allowing parameters to vary through time and includes our historical covariates (Table \ref{tab:selection}).
+Our best model, as selected via LOOIC and WAIC, is the parameter rich ``past and vary'' model which includes our historical covariantes and allows all regression parameters to vary through time (Table \ref{tab:selection}).
 Our best model is an improvement over the next best model by approximately 13 LOOIC and 24 WAIC points.
-However, the actual differences between these models are minor as the wide the standard errors of LOOIC and WAIC reveal.
+However, the actual differences between these models are minor as revealed by the wide the standard errors.
+These results mean that while our ``past and vary'' model could be considered the best of the four, all four of the models are very close and only offer incremental improvement but not major differences.
 \begin{table}[ht]
   \centering
+  \caption{Model comparison using WAIC and LOOIC}
   \begin{tabular}{ r r r r r }
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/roc_hist}
-  \caption{Comparison of AUC estimates between the four models. These estimates are based on in-sample values. Higher values indicate greater performance.}
+  \caption{Posterior predictive AUC estimates for each of the four models being compared.
+These estimates are calculated from each of the models posterior predictive distribution compared to the empirical values.
+Models with a higher AUC values indicate better performance over models with lower AUC values.
+AUC is bounded between 0.5 and 1.}
   \label{fig:roc_hist}
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/roc_ts}
-  \caption{Comparison of time-specific AUC values between the four models.
-These estimates correspond to the models ability to predict or fit that section of the data.
-Higher values indicate greater performance.}
+  \caption{Comparison between the posterior predictive AUC estimates for each of the time intervals for each of the four models.
+These estimates are reflections of each model's ability to fit the various time intervals.
+The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals.
+In all cases, higher values AUC indicate greater performance versus lower AUC values.}
   \label{fig:roc_ts}
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/fold_auc}
-  \caption{Approximate out-of-sample AUC values calculated from five-fold cross-validation of the time series.
-The AUC estimates from each fold are labeled.
-Folds are numbered from oldest to youngest.}
+  \caption{Results from our five-fold cross-validation of the time-series.
+Each labeled distribution of AUC values correspond to expected out-of-sample performance as estimated from that fold.
+Each fold represents a section of data being predicted from a model fit to all data before the start of that fold.
+Given that there are only five folds, performance is measured from model predictions to 4 of the folds.
+Folds are labeled from oldest to youngest, with the oldest fold only being predicted by a single previous fold and the youngest fold being predicted by the four previous folds.}
   \label{fig:fold_auc}
@@
 \ref{fig:effect_est}).
+This sign of this result is consistent with virtually all previous analyses of the relation between geographic range and extinction.
 
@@
 This means that during a run of above Cenozoic average global temperatures (2+ intervals) we would expect an even greater extinction risk than if just the current interval is of above average temperature.
+These results imply that, on average, the species during Paleogene had a greater extinction probability that species during the Neogene.
 
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/effect_est}
-  \caption{Estimates of the overall average covariate effects.
-These are averaged both over time and across phyla.
-The entire posterior is presented, with the median estimate labeled and 80\% credible intervals.}
+  \caption{Posterior estimates for the top-level covariate effects.
+These effect estimates are effectively the weighted average of the estimates for each of the time intervals, for each of the phyla analyzed.
+Presented are posterior densities with the median estimate labeled along 80\% credible intervals at the bottom of the density.}
   \label{fig:effect_est}
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/effect_time_group}
-  \caption{Covariate effect estimates for the effect of geographic range and the change in geographic range for all time points and phyla.
-The red line corresponds to the median estimate at each time, the dark blue area corresponds to the 50\% credible interval, the medium blue the 80\% credible interval, and the light-blue the 95\% credible interval.}
+  \caption{Effects of the four covariates as estimated for all time intervals and for each phyla.
+For each time series, the red line corresponds to the median estimate for each interval, while the envelopes represent various credible intervals as labeled.}
   \label{fig:effect_time_group}

### 2018-09-04 R/fit_model.r
regen graphs

@@ -27,38 +27,53 @@ counti_trans <- prepare_analysis(counti)
 #pp fit the model
 
 # past and vary
-form <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd + 
-                (1 + maxgcd + diff_maxgcd | fact_mybin/fossil_group) + 
-                (1 | fact_relage/fossil_group))
+form <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd 
+                + (1 | fact_relage/fossil_group)
+                + (1 + temp + lag1_temp + maxgcd + diff_maxgcd | 
+                   fact_mybin/fossil_group))
 
 # past but no vary
-form2 <- update(form, 
-                ~ . - (1 + maxgcd + diff_maxgcd | fact_mybin/fossil_group)
-                + (1 + maxgcd | fact_mybin/fossil_group))
+form2 <- update(form, ~ . - (1 + temp + lag1_temp + maxgcd + diff_maxgcd | 
+                             fact_mybin/fossil_group) 
+                      + (1 | fact_mybin/fossil_group))
 
 # no past but vary
-form3 <- update(form, 
-                ~ . - diff_maxgcd - lag1_temp -
-                  (1 + maxgcd + diff_maxgcd | fact_mybin/fossil_group) +
-                  (1 + maxgcd | fact_mybin/fossil_group))
+form3 <- update(form, ~ . - diff_maxgcd - lag1_temp 
+                      - (1 + temp + lag1_temp + maxgcd + diff_maxgcd | 
+                         fact_mybin/fossil_group) 
+                      + (1 + temp + maxgcd | fact_mybin/fossil_group))
 
 # no past or vary
-form4 <- update(form3, 
-                ~ . - (1 + maxgcd | fact_mybin/fossil_group) + 
-                  (1 | fact_mybin/fossil_group))
+form4 <- update(form3, ~ . - (1 + temp + maxgcd | fact_mybin/fossil_group) 
+                       + (1 | fact_mybin/fossil_group))
 
 forms <- list(form, form2, form3, form4)
 
-# partial evaluation to fill in priors
-glmer_part <- 
-  partial(stan_glmer, 
-          prior = normal(0, 3, autoscale = FALSE),
-          prior_intercept = normal(0, 10, autoscale = FALSE),
-          prior_aux = cauchy(0, 5, autoscale = FALSE))
-disc_fit <- map(forms, ~ glmer_part(.x, 
-                                    family = 'binomial', 
[…]

### 2018-09-04 R/fit_predict_model.r

@@ -43,19 +43,20 @@ counti_fold <- rev(split(counti_trans, counti_trans$fold))
 counti_accum <- accumulate(counti_fold, bind_rows) 
 counti_accum <- counti_accum[-1]
 
-form1 <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd + 
-                 (1 + maxgcd + diff_maxgcd | fact_mybin/fossil_group) + 
-                 (1 | fact_relage/fossil_group))
+form <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd 
+                + (1 | fact_relage/fossil_group)
+                + (1 + temp + lag1_temp + maxgcd + diff_maxgcd | 
+                   fact_mybin/fossil_group))
 
 glmer_part <- 
   partial(stan_glmer, 
-          prior = normal(0, 3, autoscale = FALSE),
+          prior = normal(c(0, 0, -1, 0), rep(3, 4), autoscale = FALSE),
           prior_intercept = normal(0, 10, autoscale = FALSE),
           prior_aux = cauchy(0, 5, autoscale = FALSE))
-fit <- map(counti_accum, ~ glmer_part(form1, 
+fit <- map(counti_accum, ~ glmer_part(form, 
                                       family = 'binomial', 
                                       data = .x,
-                                      adapt_delta = 0.99999, 
+                                      adapt_delta = 0.999999, 
                                       thin = 4))
 
 write_rds(fit, path = '../data/training_fit.rds')

### 2018-09-04 doc/intro.tex

@@
 
-Species extinction risk is strongly linked to geographic range size.
-
-A species' geographic range size changes over time.
-
-How does extinction risk change over a species duration? What effect does geographic range have on this?
-
-We want to know if including information about past geographic range size improves our inference of species extinction risk.
-
-A lot of work on species ``rise and fall'' dynamics has focused on the change in species geographic range over timeand what the shape of the average trajectory is CITATION.
-Issues that aren't addressed, like non-random drop out.
-Species can go extinct at any point in their ``lifetime'', so we might not observe a ``pure'' change to geographic range.
-We have to take this into account.
-The models that have been used do not do that which can totally mess-up analysis of geographic ranges over time.
-They need a joint-model, but that hasn't penetrated the field yet.
-
-Additionally, lack of mechanism to explain a pure ``rise-fall''.
-Importantly, ``rise and fall'' is expected from a Markov process that has only a single absorbing boundary; at some point it will go back to 0.
-
-To demonstrate that species geographic range size is not a complete Markov process, it would be necessary to show that extinction risk depends not just on the current geographic range size, but the history of its geographic range size: has range size declined or increased between earlier and now?
-
-We instead are focusing on how geographic range effects extinction risk of a species.
-We are not modeling geographic range through time, but instead modeling how species extinction is a function of geographic range over time.
+How predictable is extinction? Forecasting species survival at million-year timescales.
 
 One of the great promises of paleobiology is that by studying the past we can better predict the future.
-This promise is particularly pertinent given suggestions that risk assessments for some modern species could be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
-Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)?
+This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
+Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)?
 (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?
@@
 
-how predictable is extinction? how much does the past improve our predictions? how much does the past improve our predictions?
+Species global geographic range changes over time due to both local and global processes.
 
-we know that at almost all times, a species with geater than average geographic range is expected to have a lower extinction risk than a closely related species with a lower than average geographic range.
-what we don't know is if expanding or decreasing in geographic range over million-year time scales has any additional predictive information about a species' risk of going extinct.
+An old assumption in paleobiology was that species quickly obtained their maximum global geographic range and then maintained that range till their decline eventual extinction.
+Observation of species distribution records does not support this assumption.
+Instead, changes to a species' global geographic range is currently theorized to follow one of two general patterns: hat or hump PERFECT TIME FOR A FIGURE.
+Both are symmetric patterns but have different shapes.
+The hat pattern is characterized by an increase in geographic range for the first half a species' duration followed by a brief peak in geographic range, and the a decrease in geographic range for the second half of that species' duration.
+The hump pattern is instead characterized by a brief period of increasing species geographic range, an extended period of that species' duration at its maximum geographic range, and finally a quick decline and extinction of the species.
+
+Measuring this pattern is extremely difficult for multiple reasons.
+Species are not observed continuously through time, species geographic ranges are based on what samples are avaliable, and species are harder to observe at the tails of their duration because they are rare (small geographic range, small population sizes).
+
+Additionally, neither of these patterns are necessarily different from a Markovian process with a lower bound.
+A random walk beginning at 0 will most likely be unimodal with a period of increase followed by a period of decrease and then eventually reaching 0.
+PERFECT PLACE FOR A SIMULATION.
 

### 2018-09-04 doc/methods.tex

@@
 These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept.
+\(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.
 
@@
 
-To complete the generative model, we need to assign final priors to the ``top level'' parameters
+To complete the generative model, we need to assign final priors to the ``top level'' parameters.
+In general we favored weakly informative priors which help regularize our estimates.
+In the case of a regression coefficient, this means a Normal distribution with mean 0 and a standard deviation of 3.
+For our scale parameters (e.g.
+standard deviations), we used half-Cauchy distributed priors with heavy tails but the majority of probability density near 0.
+
+Our top-level intercept was given a more diffuse prior than our regression coefficients, which reflects our greater degree of uncertainty about its value.
+Our top-level regression coefficient for the effect of geographic range was given an informative prior reflecting the overwhelming amount of evidence that species with a larger than average geographic range have a lower risk of extinction than species with an average or less than average geographic range.
+In the context of this analysis, this means that we are again using a weakly informative prior but instead of centering the density around -1 (i.e.
+larger than average geographic range decreases extinction risk).
+
+Instead of assigning a prior distribution for each of the covariance matrices in the model, we instead decomposed the covariance matrices (e.g.
+\(\Sigma_{B}\)) which allows us to assign independent priors for the scale and correlation aspects of covariance.
+The scale parameters were assigned half-Cauchy priors as described above in the context of all other scale parameters.
+The correlation matrices were assigned LKJ priors each with shape parameter set to 1.
+This choice of shape parameter produces a uniform distribution over possible correlation matrices.
+These priors are also slightly more interpretable than other common prior distributions for covariance matrices such as the inverse-Wishart distribution.
+This approach to assigning priors to a covariance matrix is recommended by CITATION.
+
+In total, these prior choices can be expressed as
 \begin{equation}
   \begin{aligned}
-    \mu_{c} &\sim 
+    \mu_{d} &\sim 
       \begin{cases}
-        N(0, 10) & \text{if } c = 1 \\
-        N(0, 3) & \text{if } c > 1 \\
+        N(0, 10) & \text{if } d = 1 \\
+        N(-1, 3) & \text{if } d = 2 \\
+        N(0, 3) & \text{if } d > 2 \\
       \end{cases}
@@
 \end{equation}
-The decomposition of the covariance matrices (e.g \(\Sigma_{B}\)) allows priors the variance and correlation aspects of covariance to be defined separately.
-These priors are also slightly more interpretable than a, for example, inverse-Wishart distributed prior for the covariance matrices.
-This decomposition is recommended by CITATION.
 
@@
 Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling.
-In order to prevent divergent transitions the adapt delta value was increased to 0.99; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.
+In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.
 An example line of R code to fit this mode given a data.frame of all necessary data called ``data'' could be written as:
@@
   stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
-                       (1 + range + range_diff | mybin/phylum) + 
+                       (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                        (1 | age/phylum), 
-             prior = normal(0, 3, autoscale = FALSE), 
+             prior = normal(c(-1, 0, 0, 0), rep(3, 4), autoscale = FALSE), 
              prior_intercept = normal(0, 10, autoscale = FALSE), 
@@
              data = data, 
-             adapt_delta = 0.99, 
+             adapt_delta = 0.9999, 
              thin = 4)
@@
 
-We considered four variants of this model: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect).
+We considered four variants of the model described above: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect).
+The second and fourth models modify the number of columns in \(X\) by removing two of the covariates.
+The third and fourth model simplify the first model to just varying intercepts without varying slopes.
 
-To determine how many non-nested varying-intercept terms were possible to include without overly biasing out models' parameter estimates, the four variant models described above were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior CITATION.
+The four variant models were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior CITATION.
 WAIC and LOOIC are interpreted similarly to AIC, with lower values indicating greater expected out-of-sample performance.
@@
 Model adequacy was measured using the area under the receiver operating characteristic curve (AUC).
-This measure is commonly used in classification problems like this one as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION.
+This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION.
 This value ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.
 AUC was calculated for the dataset as a whole, and for each of time bins.
-These values are measures of in-sample predictive accuracy.
+This measure was used in two contexts: measuring in-sample and out-of-sample predictive performance.
+``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
+``Out-of-sample'' is defined below.
 
-\subsection{Prediction given class imbalance}
+We are also interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data).
+To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation.
+For time-series data, the folds (data partitions) are approximately equal segments of time.
+The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold.
+With 63 time points, each of the five folds represents approximately 13 time points.
+Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
 
@@
 
-\subsection{Cross-validation}
-
-In addition to model selection and adequacy, we are also interested in understanding how well our model predicts species extinction given new,future data.
-We estimated out-of-sample predictive error using cross-validation.
-For our dataset, expected out-of-sample AUC was estimated using five-fold cross-validation.
-For time-series data, the folds (data partitions) are approximately equal segments of time.
-The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold.
-With 63 time points, each of the five folds represents approximately 13 time points.
-Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
-
-For each posterior draw of the cross-validation predictions, the optimal cut-point was calculated as described above.
-Given these class assignments based on estimated optimal cut-point, the AUC value of the out-of-sample estimates was calculated.
-AUC values are calculated for the entire fold and for each of the individual time points represented within that fold.
-
 See our code repository LINK for full code details.

### 2018-09-07 R/compare_model.r
regen graphs

@@ -53,6 +53,9 @@ steps <- map2(np, ll, ~ mcmc_nuts_stepsize(.x, .y))
 treed <- map2(np, ll, ~ mcmc_nuts_treedepth(.x, .y))
 energy <- map2(np, ll, ~ mcmc_nuts_energy(.x, .y))
 
+# bayes R2
+br2 <- map(disc_fit, bayes_R2)
+
 
 # estimate of out-of-sample performance
 # these should be parallelized; done using future

### 2018-09-07 R/fit_model.r

@@ -54,23 +54,23 @@ part_glmer <- partial(stan_glmer,
                       family = 'binomial',
                       data = counti_trans,
                       prior_intercept = normal(0, 10, autoscale = FALSE),
-                      prior_aux = cauchy(0, 5, autoscale = FALSE),
+                      prior_aux = cauchy(0, 3, autoscale = FALSE),
                       thin = 4, 
-                      adapt_delta = 0.99999999)
+                      adapt_delta = 0.999999)
 
 # some models have differen't priors
 list_part_glmer <- 
   list(partial(part_glmer, prior = normal(c(0, 0, -1, 0), 
-                                          rep(3, 4), 
+                                          rep(1, 4), 
                                           autoscale = FALSE)),
        partial(part_glmer, prior = normal(c(0, 0, -1, 0), 
-                                          rep(3, 4), 
+                                          rep(1, 4), 
                                           autoscale = FALSE)),
        partial(part_glmer, prior = normal(c(0, -1), 
-                                          rep(3, 2), 
+                                          rep(1, 2), 
                                           autoscale = FALSE)),
        partial(part_glmer, prior = normal(c(0, -1), 
-                                          rep(3, 2), 
+                                          rep(1, 2), 
                                           autoscale = FALSE)))
 
 # map the data to the models

### 2018-09-07 R/inspect_model.r

@@ -33,7 +33,6 @@ counti_trans <- prepare_analysis(counti)
 disc_fit <- read_rds('../data/disc_fit.rds')
 disc_best <- disc_fit[[1]]
 
-
 # check how much information learned
 priors <- prior_summary(disc_best)
 prmn <- priors$prior$location

### 2018-09-07 R/plot_foo.r

@@ -115,7 +115,7 @@ plot_risk_time <- function(data, model, nsp = 4) {
     ggplot(aes(x = relage, y = maxgcd)) + 
     geom_line() +
     facet_grid(~ fullname) + 
-    labs(x = 'Age (My)', y = 'P(T = t | T >= t, x)')
+    labs(x = 'Age (My)', y = 'Geographic range (standardized)')
 
   full_plot <- full_est %>%
     gather(key, value, -fullname, -iterations, -relage, -row) %>%

### 2018-09-10 doc/intro.tex
expanding results section to try and explain variance sources

@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/range_time}
+  \caption{Geographic range over time for a selection of eight random species included in this analysis.
+Geographic range is measured as the maximum geodesic distance between two occurrences of that species during a given time interval.
+The time axis corresponds to species age in millions of years (My).
+Because each species has a different duration, not all of these trajectories are of equivalent duration.}
+  \label{fig:range_real}
+\end{figure}
+
 Measuring this pattern is extremely difficult for multiple reasons.
-Species are not observed continuously through time, species geographic ranges are based on what samples are avaliable, and species are harder to observe at the tails of their duration because they are rare (small geographic range, small population sizes).
+Species are not observed continuously through time, species geographic ranges are based on what samples are available, and species are harder to observe at the tails of their duration because they are rare (small geographic range, small population sizes).
 
@@
 A random walk beginning at 0 will most likely be unimodal with a period of increase followed by a period of decrease and then eventually reaching 0.
-PERFECT PLACE FOR A SIMULATION.
+This can be demonstrated through a simple simulation PERFECT PLACE FOR A SIMULATION.
+
+Our simulations are of a random walk with a lower bound of 0.
+Each individual simulation begins with a value of 0, and for each time step that value is modified by a single realization from a standard Normal distribution.
+Each simulation had a maximum of 100 time steps beyond the initial point.
+For each simulation, the first time the value is less than or equal to 0 is considered the stopping point, with that last value set to 0.
+We retained all simulations which returned to 0, which would be the equivalent of a species going extinct.
+While we attempted 1000 simulations, not all of these fulfilled this criteria as they did not had a value greater than 0 at their final time step.
+Finally, the duration of all simulations was rescaled to be between 0 and 1 so that their trajectories are directly comparable.
+A binned average and moving-window average are calculated from the population of simulations.
+
+The binned average and moving-window averages are nearly identical to figure 1 from the original rise-and-fall paper by Foote et al.
+The results from these simulations show that for a random walk bounded at 0, the average trajectory is symmetrical and virtually identical to either the ``hat'' or ``hump'' models and was generated without any deterministic process; the pattern is unimodal and intermediate between ``strict'' hat or hump models.
+Thus, a hat or hump trajectory demonstrates nothing about process or determinism but could just result from entirely stochastic processes.
+
+ \begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/georange_sim_norm}
+  \caption{Simulation of a bounded random walk. The blue line is a binned average of all simulations while the red line is a moving-window average of those averages.}
+  \label{fig:range_sim}
+\end{figure}
+
+Instead of modeling geographic range through time, we instead model extinction probability as a function of change in geographic range.
+This formulation means that, if there if geographic range trajectory is deterministic and not entirely Markovian, that past geographic range, specifically the change in geographic range, and not just current geographic range predicts the probability of extinction.
+For example, is extinction probability affected by a species has been decreasing in geographic range over time or does it just matter if that species has below average geographic range regardless of how it got there?
 

### 2018-09-10 doc/results.tex
expanding results section to try and explain variance sources

@@
     \hline
-    Past and vary & 12811.58 & 179.27 & 12807.61 & 170.21 \\
-    No past but vary & 12834.53 & 179.07 & 12831.51 & 179.02 \\
-    Past but no vary & 12836.19 & 179.20 & 12833.13 & 179.15 \\
-    No past or vary & 12849.43 & 179.46 & 12847.08 & 179.43 \\
+    Past and vary & 12789.81 & 178.90 & 12785.51 & 178.84 \\
+    No past but vary & 12820.67 & 178.82 & 12817.49 & 178.77 \\ 
+    Past but no vary & 12847.18 & 179.40 & 12844.88 & 179.36 \\ 
+    No past or vary & 12851.43 & 179.48 & 12849.10 & 179.44 \\ 
     \hline
@@
 
+% effect of age at observation
+
+% average hazard
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_baseline}
+  \caption{<+caption text+>}
+  \label{fig:hazard_baseline}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_bygroup}
+  \caption{<+caption text+>}
+  \label{fig:hazard_bygroup}
+\end{figure}
+
 % variance components
+There are multiple levels and sources of variance in our model.
+Each of the grouping factors account for some amount of variance in the data.
+The first ``level'' of variance is the data level or the variance in the observed data (the model response).
+In the context of regular linear regression, this variance is equivalent to the residual variance.
+In the context of logistic regression, however, this value is not an inherent part of the model.
+
+There are then two non-nested sources of the second ``level'' of variance: intercept and slopes varying by time of observation, and a mean-zero age at observation varying intercept.
+Each of the elements of the vector of scale terms \(\tau_{B}\) describe the within-phyla variance for the intercept and the slope terms for the covariates.
+Similarly, the scale term \(\sigma_{a}\) is a scalar that describes the within-phyla variance of the effect of age at observation.
+
+The final ``level'' of variance the phyla-level: the between phyla variance for the intercepts and regression coefficients \(\tau_{\alpha}\), and the between phyla variance for the effect of age at observation \(\sigma_{\delta}\).
+
+The greatest source of variance accounted for by the grouping factors is the between phylum variance in the model intercept (Fig.
+\ref{fig:variance_components}); this is the first element of \(\tau_{B}\).
+All of the other scale terms have approximately equal estimates, which means that they contribute a similar amount to the structure underlying the data.
+For example, the within-phyla effect of geographic range on extinction probability is associated with approximately as much variance as there is between the phyla themselves.
+
+This result means that, given this model, the greatest contributor to changes in extinction risk over time is when a species is present.
+In comparison, the covariates we studied individually contribute only a limited amount to the variation in extinction.
+Similarly, while age at observation does contribute to some differences in extinction risk (Fig XXX age effect), this effect is most likely overwhelmed by the importance of when that observation takes place.
+
 \begin{figure}[ht]
@@
 % risk estimate compared to change in geo-range
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/relrisk_range}
+    \caption{A}
+    \label{fig:relrisk_range}
+  \end{subfigure}
+
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/relrisk_ext}
+    \caption{B}
+    \label{fig:relrisk_ext}
+  \end{subfigure}
+  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation.
+Geographic range is assumed observed without error.
+Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
+  \label{fig:relrisk_compare}
+\end{figure}
 

### 2018-09-10 R/sim_foo.r

@@ -8,7 +8,7 @@
 #'
 #' @param max_length scalar int for maximum number of discrete time steps.
 #' @return vector of of real values
-sim_range <- function(max_length = 10) {
+sim_walk <- function(max_length = 10) {
 
   # this is easily solved by a fold across a series 
   # that series begins at 0, has max_length changes
@@ -22,7 +22,8 @@ sim_range <- function(max_length = 10) {
   if(death == 0) {
     out <- run 
   } else if(death > 0) {
-    out <- run[1:(death - 1)]
+    out <- run[1:death]
+    out[death] <- 0
   }
 
   out

### 2018-09-10 R/sim_range.r

@@ -6,8 +6,7 @@ source('../R/sim_foo.r')
 
 nsim <- 1000
 nstep <- 100
-
-simul <- rerun(.n = nsim, sim_range(nstep)) %>%
+simul <- rerun(.n = nsim, sim_walk(nstep)) %>%
   imap(~ tibble(range = .x, 
                 time = seq(length(.x)), 
                 sim = .y)) %>%
@@ -24,7 +23,7 @@ simul <- simul %>%
   mutate(n = map(data, n_distinct)) %>%
   unnest(n = n) %>%
   filter(n > 1,
-         n < nstep) %>%
+         n < (nstep + 1)) %>%
   select(-n) %>%
   unnest() %>%
   group_by(sim) %>%
@@ -44,14 +43,20 @@ simul_avg <- simul %>%
 
 # basic plot of runs
 graph_basic <- simul %>%
-  ggplot(aes(x = time, y = range, group = sim)) + geom_line()
+  ggplot(aes(x = time, y = range, group = sim)) + 
+  geom_line()
+
+ggsave(filename = '../doc/figure/georange_sim_basic.png',
+       plot = graph_basic, width = 6, height = 6)
 
 # normalize by duration
 graph_norm <- simul %>%
   ggplot(aes(x = time_standard, y = range, group = sim)) +
-  geom_line() +
+  geom_line() + 
   geom_line(mapping = aes(x = time_standard, y = range_avg, group = NULL), 
             colour = 'blue', size = 1.2) +
   geom_line(mapping = aes(x = time_standard, y = range_roll, group = NULL),
             data = simul_avg, colour = 'red')
 
+ggsave(filename = '../doc/figure/georange_sim_norm.png',
+       plot = graph_norm, width = 6, height = 6)

### 2018-09-11 doc/methods.tex
expanding results section to try and explain variance sources
brms implementation included for funsies

@@
     \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
-    a_{w, p} &\sim MVN(\delta_{p}, \Sigma_{a}) \\
-    \delta_{p} &\sim MVN(0, \Sigma_{\delta}) \\
+    a_{w, p} &\sim \mathrm{N}(\delta_{p}, \sigma_{a}) \\
+    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
   \end{aligned}
@@
     \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
-    \Sigma_{a} &= diag(\tau_{a}) \Omega_{a} diag(\tau_{a}) \\
-    \Sigma_{\delta} &= diag(\tau_{\delta}) \Omega_{\delta} diag(\tau_{\delta}) \\
     \tau_{B} &\sim C^{+}(5) \\
     \tau_{\alpha} &\sim C^{+}(5) \\
-    \tau_{a} &\sim C^{+}(5) \\
-    \tau_{\delta} &\sim C^{+}(5) \\
     \Omega_{B} &\sim LKJ(1) \\
     \Omega_{\alpha} &\sim LKJ(1) \\
-    \Omega_{a} &\sim LKJ(1) \\
-    \Omega_{\delta} &\sim LKJ(1). \\
   \end{aligned}
@@
 In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.
-An example line of R code to fit this mode given a data.frame of all necessary data called ``data'' could be written as:
+
+An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as
 \begin{verbatim}
@@
                        (1 | age/phylum), 
+             data = data, 
+             family = 'binomial',
              prior = normal(c(-1, 0, 0, 0), rep(3, 4), autoscale = FALSE), 
              prior_intercept = normal(0, 10, autoscale = FALSE), 
-             prior_aux = cauchy(0, 5, autoscale = FALSE), 
-             data = data, 
-             adapt_delta = 0.9999, 
-             thin = 4)
+             prior_aux = cauchy(0, 3, autoscale = FALSE), 
+             chains = 4,
+             thin = 4,
+             adapt_delta = 0.9999).
 \end{verbatim}
 
-Posterior convergence was determined using the following general and HMC specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)).
+Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} as
+\begin{verbatim}
+priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
+            set_prior('normal(0, 1)', class = 'b'),
+            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
+            set_prior('cauchy(0, 3)', class = 'sd'),
+            set_prior('lkj(1)', class = 'cor'))
+brmfit <- brm(formula = bf(event ~ range + range_diff + temp + temp_lag +
+                           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
+                           (1 | age/phylum))
+              data = data, 
+              family = bernoulli(), 
+              prior = priors,
+              chains = 4, 
+              thin = 4,
+              control = list(adapt_delta = 0.999999).
+\end{verbatim}
+
+Posterior convergence was determined using the general and HMC-specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)).
 For further explanation of these diagnostic criteria, see the Stan Manual CITATION.
@@
 
+%\section{Variance Partioning and Intraclass Correlation Coeficient}
+%
+%There are multiple levels and sources of variance in this model.
+Each of the grouping factors account for some amount of variance in the data.
+Normalizing that variance by the total variance from all sources, we can see the percent contribution to variance.
+%
+%``level 1'' is the ``data level.'' in a regular linear regression problem, this would be the residual variance.
+that's not calculated in logistic regression, which poses a problem.
+how do we get this value?
+simulation?
+%
+%``level 2'' is the ``time level.'' the logistic regression coefficients \textbf{vary} through time, we are actually estimating that variance in \(\Sigma_{B}\) and \(\tau_{B}\).
+there is also the non-nested age-based varying-intercept.
+this source of variance is in \(\Sigma_{a}\) and \(\tau_{a}\).
+%
+%``level 3'' is the ``phyla level.'' the over-time averages of the regression coefficients further vary between phyla.
+that variance is part of \(\Sigma_{\alpha}\) and \(\tau_{\alpha}\).
+there is also the average age effect on extinction probability which is part of \(\Sigma_{\delta}\) and \(\tau_{\delta}\).
+%
+%that's a lot of variance. 
+%
+%we want to figure out what the total variance is, so we can determine the percent contribution for each level and segment of variance.
+the level 1 issue and the nonnestedness pose potential challenges.
+
 See our code repository LINK for full code details.

### 2018-09-13 doc/methods.tex
filling in more of results with supplement stuff

@@
     t_{i} &\sim \text{Bernoulli}(\Theta) \\
-    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + a_{d[i], p[i]}) \\
+    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{d[i], p[i]}) \\
     B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
     \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
-    a_{w, p} &\sim \mathrm{N}(\delta_{p}, \sigma_{a}) \\
+    % double check this\dots is A MVN dist?
+    A_{w, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
     \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
@@
     \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
+    \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
     \tau_{B} &\sim C^{+}(5) \\
     \tau_{\alpha} &\sim C^{+}(5) \\
+    \tau_{A} &\sim C^{+}(5) \\
     \Omega_{B} &\sim LKJ(1) \\
     \Omega_{\alpha} &\sim LKJ(1) \\
+    \Omega_{A} &\sim LKJ(1) \\
   \end{aligned}
@@
 When the intercept term of the logistic regression is sufficiently negative (-2 or less), the only way to obtain a linear predictor value of 0.5 or greater is if the covariate effects are large or there the covariates are at extreme values.
-The 0.5 default cut-off point for binary prediction is much to high when events are rare.
-A method for determining the new ``optimal'' cut-point involves calculating the receiver operating characterist (ROC) curve from the linear predictor estimates and then identifying the point on the ROC curve which maximizes both true positive rate and the false positive rate.
-The new cut-point is then back-calculated from the true positive and false positive rates.
-Class membership is then predicted given this new cut-point.
-This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cut-point.
+The 0.5 default cutpoint for binary prediction is much to high when events are rare.
+A method for determining the new ``optimal'' cutpoint involves calculating the receiver operating characteristic (ROC) curve from the linear predictor estimates and then identifying the point on the ROC curve which maximizes both true positive rate and the false positive rate.
+The new cutpoint is then back-calculated from the true positive and false positive rates.
+Class membership is then predicted given this new cutpoint.
+This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cutpoint.
 

### 2018-09-13 doc/results.tex
filling in more of results with supplement stuff

@@
 \end{table}
+
+As described above, our data is extremely unbalanced with respect to the number of extinction events relative to the total number of observations; this reality means that 0.5 is an inappropriate cutpoint for estimating if a species observation is their last (i.e.
+goes extinct).
+New, optimal cutpoint were calculated from the ROC curve as the level which maximized the true and false positive rates.
+The distributions of optimal cut-off point estimates for each of the four models are broadly similar, with all 50\% credible intervals overlapping.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/cut_plot}
+  \caption{Posterior estimates of the optimal cutpoints for each of the three models. The optimal cutpoint corresponds to the point which maximizes the true and false positive rates.}
+  \label{fig:cut_point}
+\end{figure}
 
@@
 % average hazard
+The hazard function describes how age of observation impacts probability of extinction.
+Hazard is estimated for each phylum for the range of ages observed within the phylum.
+The matrix \(A\) describes the effect of age on extinction probability for age at observation (rows) by phylum (columns).
+The phylum-level averages of the effect of age of observation on extinction \(\delta\) are themselves drawn from a distribution centered around 0.
+
+Figure \ref{fig:hazard_baselin} is a graph of the average hazard associated with age of observation.
+This graph illustrates how average extinction risk increases with age for the first few million years before leveling off.
+A species reaches highest age-related probability of extinction at 10 million years, after which the median becomes volatile until a duration over 30 million years.
+The flat-ness of the median estimate and the homogeneity of variance durations between approximately 30 million years to under 60 million years is due to the lack of species having survived over 30 million years, which inherently means that these estimates are based on less information and are shrunk towards the overall average while reflecting more and more of our prior uncertainty over time.
 
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_baseline}
-  \caption{<+caption text+>}
+  \caption{Average hazard estimates based on all all observed species ages in millions of years.
+Hazard in discrete-survival analysis is the probability of going extinct at a given age having survived up to that age.
+Overall hazard was calculated using the phylum average intercept and assuming all covariates equaled 0.
+The red line is the median hazard estimate along with 50\%, 80\%, and 95\% credible intervals.}
   \label{fig:hazard_baseline}
 \end{figure}
+
+The hazard associated with the individual phyla have broad similarities with the overall hazard estimates (Fig.
+\ref{fig:hazard_bygroup}).
+In all four cases, hazard increases over the first 10 million years of a species duration.
+However, the rate of increase in hazard varies between the phyla, with forams and radiolarians having the most pronounced increase with age.
+Radiolarians also have the lowest initial hazard of the phyla, followed by forams and calcareous nannoplankton.
+In contrast, diatoms have the highest initial hazard of the phyla while having the smallest change hazard with age.
+
+Optimal cutpoint between 0.1 and 0.11 percent. We estimate that the effect of age of observation on extinction can reach this threshold after XX My at ##\% posterior probability.
 
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_bygroup}
-  \caption{<+caption text+>}
+  \caption{Hazard estimates for each of the phyla analyzed here. The red line is the median hazard estimate along with 50\%, 80\%, and 95\% credible intervals.}
   \label{fig:hazard_bygroup}
@@
 Each of the elements of the vector of scale terms \(\tau_{B}\) describe the within-phyla variance for the intercept and the slope terms for the covariates.
-Similarly, the scale term \(\sigma_{a}\) is a scalar that describes the within-phyla variance of the effect of age at observation.
+Similarly, the scale term \(\sigma_{A}\) is a scalar that describes the within-phyla variance of the effect of age at observation.
 

### 2018-09-13 doc/results.tex
cut point caption

@@
     \hline
-    Past and vary & 12789.81 & 178.90 & 12785.51 & 178.84 \\
-    No past but vary & 12820.67 & 178.82 & 12817.49 & 178.77 \\ 
-    Past but no vary & 12847.18 & 179.40 & 12844.88 & 179.36 \\ 
-    No past or vary & 12851.43 & 179.48 & 12849.10 & 179.44 \\ 
+    Past and vary & 12784.04 & 178.71 & 12784.04 & 178.71 \\
+    No past but vary & 12819.01 & 178.80 & 12819.01 & 178.80 \\ 
+    Past but no vary & 12847.00 & 179.38 & 12847.00 & 179.38 \\ 
+    No past or vary & 12851.53 & 179.48 & 12851.53 & 179.48 \\ 
     \hline
@@
 New, optimal cutpoint were calculated from the ROC curve as the level which maximized the true and false positive rates.
+
 The distributions of optimal cut-off point estimates for each of the four models are broadly similar, with all 50\% credible intervals overlapping.
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/cut_plot}
-  \caption{Posterior estimates of the optimal cutpoints for each of the three models. The optimal cutpoint corresponds to the point which maximizes the true and false positive rates.}
+  \caption{Posterior estimates of the optimal cutpoints for each of the three models.
+The optimal cutpoint corresponds to the point which maximizes the true and false positive rates; this new cutpoint is used instead of 0.5 when estimating if an observation is the last observation of that species (i.e.
+goes extinct).}
   \label{fig:cut_point}

### 2018-09-21 R/fit_model.r
changed to lag1_maxgcd because it makes more sense to me. also the sign is logical.

@@ -24,7 +24,7 @@ counti <- read_rds('../data/counting.rds')
 # transform the data for analysis
 counti_trans <- prepare_analysis(counti)
 
-#priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
+#priors <- c(set_prior('normal(-2, 1)', class= 'Intercept'),
 #            set_prior('normal(0, 1)', class = 'b'),
 #            set_prior('normal(-1, 1)', class = 'b', coef = 'maxgcd'),
 #            set_prior('cauchy(0, 1)', class = 'sd'),
@@ -43,19 +43,19 @@ counti_trans <- prepare_analysis(counti)
 #              prior = priors)
 
 # past and vary
-form <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd 
+form <- formula(event ~ temp + lag1_temp + maxgcd + lag1_maxgcd 
                 + (1 | fact_relage/fossil_group)
-                + (1 + temp + lag1_temp + maxgcd + diff_maxgcd | 
+                + (1 + temp + lag1_temp + maxgcd + lag1_maxgcd | 
                    fact_mybin/fossil_group))
 
 # past but no vary
-form2 <- update(form, ~ . - (1 + temp + lag1_temp + maxgcd + diff_maxgcd | 
+form2 <- update(form, ~ . - (1 + temp + lag1_temp + maxgcd + lag1_maxgcd | 
                              fact_mybin/fossil_group) 
                       + (1 | fact_mybin/fossil_group))
 
 # no past but vary
-form3 <- update(form, ~ . - diff_maxgcd - lag1_temp 
-                      - (1 + temp + lag1_temp + maxgcd + diff_maxgcd | 
+form3 <- update(form, ~ . - lag1_maxgcd - lag1_temp 
+                      - (1 + temp + lag1_temp + maxgcd + lag1_maxgcd | 
                          fact_mybin/fossil_group) 
                       + (1 + temp + maxgcd | fact_mybin/fossil_group))
 
@@ -70,7 +70,7 @@ forms <- list(form, form2, form3, form4)
 part_glmer <- partial(stan_glmer, 
                       family = 'binomial',
                       data = counti_trans,
-                      prior_intercept = normal(0, 10, autoscale = FALSE),
+                      prior_intercept = normal(-2, 10, autoscale = FALSE),
                       prior_aux = cauchy(0, 3, autoscale = FALSE),
                       thin = 4, 
                       adapt_delta = 0.999999)

### 2018-09-21 R/inspect_model.r

@@ -39,12 +39,12 @@ prmn <- priors$prior$location
 prsc <- priors$prior$scale
 
 posc <- disc_best %>% 
-  spread_samples(maxgcd,
-                 diff_maxgcd,
-                 temp,
-                 lag1_temp) %>% 
-  clean_names %>% 
-  dplyr::summarize_at(vars(-chain, -iteration), funs(sd(.)))
+  spread_draws(maxgcd,
+               lag1_maxgcd,
+               temp,
+               lag1_temp) %>% 
+clean_names %>% 
+dplyr::summarize_at(vars(-chain, -iteration), funs(sd(.)))
 
 # 1 = lot of shrinkage, 0 = no shrinkage
 posterior_shrinkage <- 1 - (posc / prsc) ^ 2
@@ -52,72 +52,74 @@ posterior_shrinkage <- 1 - (posc / prsc) ^ 2
 
 # posterior intervals
 interval_est <- disc_best %>%
-  spread_samples(`(Intercept)`, 
-                 maxgcd, 
-                 diff_maxgcd,
-                 #`maxgcd:diff_maxgcd`,
-                 temp,
-                 lag1_temp,
-                 `Sigma[fossil_group:fact_mybin:(Intercept),(Intercept)]`,
-                 `Sigma[fossil_group:fact_mybin:maxgcd,(Intercept)]`,
-                 `Sigma[fossil_group:fact_mybin:diff_maxgcd,(Intercept)]`,
-                 #`Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,(Intercept)]`,
-                 `Sigma[fossil_group:fact_mybin:maxgcd,maxgcd]`,
-                 `Sigma[fossil_group:fact_mybin:diff_maxgcd,maxgcd]`,
-                 #`Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,maxgcd]`,
-                 `Sigma[fossil_group:fact_mybin:diff_maxgcd,diff_maxgcd]`,
-                 #`Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,diff_maxgcd]`,
-                 #`Sigma[fossil_group:fact_mybin:maxgcd:diff_maxgcd,maxgcd:diff_maxgcd]`,
-                 `Sigma[fossil_group:fact_relage:(Intercept),(Intercept)]`,
-                 `Sigma[fact_mybin:(Intercept),(Intercept)]`,
-                 `Sigma[fact_mybin:maxgcd,(Intercept)]`,
[…]

### 2018-10-08 doc/results.tex

@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/effect_time_group}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/eff_time_group}
   \caption{Effects of the four covariates as estimated for all time intervals and for each phyla.
@@
 
-Figure \ref{fig:hazard_baselin} is a graph of the average hazard associated with age of observation.
+Figure \ref{fig:hazard_baseline} is a graph of the average hazard associated with age of observation.
 This graph illustrates how average extinction risk increases with age for the first few million years before leveling off.
@@
 
-Optimal cutpoint between 0.1 and 0.11 percent. We estimate that the effect of age of observation on extinction can reach this threshold after XX My at ##\% posterior probability.
+Optimal cutpoint between 0.1 and 0.11 percent. We estimate that the effect of age of observation on extinction can reach this threshold after XX My at YY\% posterior probability.
 
@@
 
-This result means that, given this model, the greatest contributor to changes in extinction risk over time is when a species is present.
+This result means that, given this model, the greatest contributor to differences in extinction risk over time is when a species is present.
 In comparison, the covariates we studied individually contribute only a limited amount to the variation in extinction.
 Similarly, while age at observation does contribute to some differences in extinction risk (Fig XXX age effect), this effect is most likely overwhelmed by the importance of when that observation takes place.
+
+The variance in effects contributed by the phyla is approximately equal to the variance in those effects due to differences over time.
 
@@
 \end{figure}
+
+Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of
 

### 2018-10-08 doc/gsa2018_talk.tex
restructured dir. gsa2018 talk skeleton

\documentclass{beamer} 
\usepackage{amsmath,amsthm}
\usepackage{graphicx,microtype,parskip}
\usepackage{caption,subcaption,multirow}
\usepackage{attrib}

\frenchspacing

\usetheme{default}
\usecolortheme{whale}

\setbeamertemplate{navigation symbols}{}

\setbeamercolor{title}{fg=blue,bg=white}

\setbeamercolor{block title}{fg=white,bg=gray}
\setbeamercolor{block body}{fg=black,bg=lightgray}

\setbeamercolor{block title alerted}{fg=white,bg=darkgray}
\setbeamercolor{block body alerted}{fg=black,bg=lightgray}

\title{The changing functional composition of the North American species pool}
\subtitle{ }
\author{Peter D Smits, Seth Finnegan}
\institute{Department of Integrative Biology, University of California -- Berkeley}
\date{}

\begin{document}

\begin{frame}
  \maketitle
\end{frame}

\begin{frame}
  \frametitle{Foundational assertion of conservation paleobiology }

  \begin{center}
    \begin{LARGE}
      By studying the \alert{past}, \\we can better predict the \alert{future}.
    \end{LARGE}
  \end{center}

\end{frame}

\begin{frame}
  \frametitle{What are we predicting?}

  \begin{center}
    \begin{LARGE}
      Extinction is \underline{hard} to predict, \\but is extremely important to conservation.
    \end{LARGE}
  \end{center}

\end{frame}

\begin{frame}
  \frametitle{Predicting extinction}

  \begin{itemize}[<+->]
    \item A taxon with a \alert{greater than average} global geographic range is likely to survive for longer than a taxon with \alert{less than average} global geographic range.
    \item A taxon's global geographic range can change over time.
    \item What happens to extinction risk as a taxon changes geographic range? How is extinction risk impacted if that taxon's global geographic range has recently \alert{increased} or \alert{decreased}?
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Encoding the past}

  \begin{itemize}
    \item Change in geographic range between current observation and previous observation.
    \item Average global temperature at time of previous observation (Mg/Ca isotope).
    \item Age in millions of years at time of observation.
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Data being analyzed}

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/occ_time_label}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/age_label}
    \end{column}
  \end{columns}

\end{frame}

\begin{frame}
  \frametitle{How we're analyzing the data}

  \begin{itemize}[<+->]
    \item Predictors: geographic range, change in geographic range, global temperature, lag of global temperature.
    \item Bayesian discrete-time survival model.
      \begin{itemize}
        \item<.-> Bernoulli response distribution.
        \item<.-> Time/phylum varying intercepts and slopes.
        \item<.-> Taxon age as non-nested varying intercept.
      \end{itemize}
    \item \alert{Compare} models using WAIC/LOOIC.
    \item Explore model \alert{adequacy} using posterior predictive distribution.
    \item Estimate \alert{predictive performance} using \(k\)-fold cross-validation.
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{A conceptual model for predicting extinction}

\end{frame}

\begin{frame}
  \frametitle{A statistical model for predicting extinction}

\end{frame}

\begin{frame}
  \frametitle{Comparing our models}

  % table of waic/looic values
  %   se values come out second to demonstrate how all virtually identical

\end{frame}

\begin{frame}
  \frametitle{Cross-validation results}

  % figure comparing in-sample AUC to out-of-sample AUC
  %   they are similar

\end{frame}

\begin{frame}
  \frametitle{Tracking extinction risk over time}

\end{frame}

\begin{frame}
  \frametitle{Overall covariate effects}

\end{frame}

\begin{frame}
  \frametitle{Covariate effects over time}

\end{frame}

\begin{frame}
  \frametitle{Effects of age on extinction risk}

\end{frame}

\begin{frame}
  \frametitle{Summary}

\end{frame}

\begin{frame}
  \frametitle{Conclusions}

\end{frame}

\begin{frame}
  \frametitle{Acknowledgements}

\end{frame}

\end{document}

### 2018-10-08 doc/results.tex
restructured dir. gsa2018 talk skeleton

@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/cut_plot}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cut_plot}
   \caption{Posterior estimates of the optimal cutpoints for each of the three models.
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/roc_hist}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/roc_hist}
   \caption{Posterior predictive AUC estimates for each of the four models being compared.
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/roc_ts}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/roc_ts}
   \caption{Comparison between the posterior predictive AUC estimates for each of the time intervals for each of the four models.
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/fold_auc}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc}
   \caption{Results from our five-fold cross-validation of the time-series.
@@
 %  \centering
-%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/fold_auc_time}
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
 %  \caption{Approximate out-of-sample AUC values calculated for each of the My intervals using five-fold cross-validation of the time series.
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/effect_est}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/effect_est}
   \caption{Posterior estimates for the top-level covariate effects.
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/eff_time_group}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/eff_time_group}
   \caption{Effects of the four covariates as estimated for all time intervals and for each phyla.
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_baseline}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/hazard_baseline}
   \caption{Average hazard estimates based on all all observed species ages in millions of years.
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_bygroup}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/hazard_bygroup}
   \caption{Hazard estimates for each of the phyla analyzed here. The red line is the median hazard estimate along with 50\%, 80\%, and 95\% credible intervals.}
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/variance_components}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/variance_components}
   \caption{Contribution of multi-level components to unmodeled variance.
@@
   \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/relrisk_range}
+    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
     \caption{A}
@@
   \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/relrisk_ext}
+    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
     \caption{B}

### 2018-10-08 R/00_inspect_data.r

@@ -16,6 +16,8 @@ library(ggridges)
 library(pROC)
 source('../R/helper01_process_foo.r')
 
+theme_set(theme_bw())
+
 # important constants
 options(mc.cores = parallel::detectCores())
 
@@ -26,25 +28,33 @@ counti <- read_rds('../data/counting.rds')
 
 # form of data that was analyzed
 counti_trans <- prepare_analysis(counti)
-
-counti_trans %>% dplyr::select(fullname, mybin, relage, maxgcd, diff_maxgcd)
+#counti_trans %>% dplyr::select(fullname, mybin, relage, maxgcd, diff_maxgcd)
 
 
 # occurrences through time, labeled if LAD
-octg <- counti %>%
-  ggplot(aes(x = mybin, fill = factor(event))) +
+octg <-  counti %>%
+  mutate(state = case_when(event == 0 ~ 'Standard',
+                           event == 1 ~ 'Last'),
+         fossil_group = case_when(fossil_group == 'D' ~ 'Dinoflagellates',
+                                  fossil_group == 'R' ~ 'Radiolaria',
+                                  fossil_group == 'F' ~ 'Foraminifera',
+                                  fossil_group == 'N' ~ 'Calc. nanno.')) %>%
+  ggplot(aes(x = mybin, fill = state)) +
   stat_bin() +
-  facet_grid(fossil_group ~ ., switch = 'y')
-ggsave(filename = '../doc/figure/occ_time_label.png',
+  facet_grid(fossil_group ~ ., switch = 'y') +
+  scale_fill_manual(name = 'Occurrence type',
+                    values = c('goldenrod', 'skyblue')) +
+  theme(legend.position = 'bottom') +
+  labs(title = 'Occurrences', x = 'Time (My before present)', y = 'Count')
+ggsave(filename = '../results/figure/occ_time_label.png',
        plot = octg, width = 4, height = 6)
-
 
 
 # relative "abundance" of microfossil_groups over time
 ocag <- counti %>%
   ggplot(aes(x = mybin, fill = fossil_group)) +
   geom_histogram(position = 'fill')
-ggsave(filename = '../doc/figure/abn_time_stack.png',
+ggsave(filename = '../results/figure/abn_time_stack.png',
        plot = ocag, width = 6, height = 6)
 
 # occurrences by relage
@@ -53,11 +63,23 @@ ocrg <- counti %>%
[…]

### 2018-10-08 R/03_compare_models.r

@@ -67,7 +67,7 @@ rownames(compare_loo_tab) <- plyr::mapvalues(rownames(compare_loo_tab),
                                              sort(rownames(compare_loo_tab)),
                                              model_key)
 xtable::print.xtable(xtable::xtable(compare_loo_tab), 
-                     file = '../doc/loo_tab_draft.tex')
+                     file = '../results/loo_tab_draft.tex')
 
 wl <- map(disc_fit, ~ future::future(waic(.x)))
 wl <- map(wl, ~ future::value(.x))
@@ -76,7 +76,7 @@ rownames(compare_waic_tab) <- plyr::mapvalues(rownames(compare_waic_tab),
                                               sort(rownames(compare_waic_tab)),
                                               model_key)
 xtable::print.xtable(xtable::xtable(compare_waic_tab), 
-                     file = '../doc/waic_tab_draft.tex')
+                     file = '../results/waic_tab_draft.tex')
 # i should find a way to output these nicely
 
 
@@ -99,7 +99,7 @@ cut_plot <- list_cutpoint %>%
   dplyr::select(-draw) %>%
   ggplot(., aes(x = value, y = model)) + 
   geom_halfeyeh(.width = c(0.5, 0.8))
-ggsave(filename = '../doc/figure/cut_plot.png', plot = cut_plot,
+ggsave(filename = '../results/figure/cut_plot.png', plot = cut_plot,
        width = 6, height = 6)
 
 
@@ -118,10 +118,10 @@ roc_hist <- bind_rows(imap(pp_auc, ~ data.frame(model = .y, roc = .x))) %>%
          model = fct_relevel(model, !!model_key[2], after = 1)) %>%
   ggplot(aes(x = roc, y = model)) +
   geom_halfeyeh(.width = c(0.5, 0.8))
-ggsave(filename = '../doc/figure/roc_hist.png', plot = roc_hist,
+ggsave(filename = '../results/figure/roc_hist.png', plot = roc_hist,
        width = 6, height = 6)
 
 # roc as timeseries to see best and worst times
 roc_ts <- plot_roc_series(counti_trans, pp_est_new, model_key)
-ggsave(filename = '../doc/figure/roc_ts.png', plot = roc_ts,
+ggsave(filename = '../results/figure/roc_ts.png', plot = roc_ts,
        width = 8, height = 8)

### 2018-10-08 R/05_inspect_model.r

@@ -91,7 +91,7 @@ effect_eye <- disc_best %>%
                lag1_temp) %>%
 ggplot(aes(y = .variable, x = .value)) +
 geom_halfeyeh(.width = c(0.8, 0.5))
-ggsave(filename = '../doc/figure/effect_est.png',
+ggsave(filename = '../results/figure/effect_est.png',
        plot = effect_eye, width = 4, height = 6)
 
 
@@ -121,7 +121,7 @@ vary_eye <- disc_best %>%
   ggplot(aes(y = term, x = .value)) +
   geom_halfeyeh(.width = c(0.8, 0.5)) +
   labs(x = 'Estimate', y = 'Variance component')
-ggsave(filename = '../doc/figure/variance_components.png',
+ggsave(filename = '../results/figure/variance_components.png',
        plot = vary_eye, width = 6, height = 6)
 
 
@@ -151,13 +151,13 @@ hazard_plot <- disc_best %>%
   stat_lineribbon() +
   scale_fill_brewer() +
   labs(x = 'Age (My)', y = 'P(T = t | T >= t, x)')
-ggsave(filename = '../doc/figure/hazard_baseline.png',
+ggsave(filename = '../results/figure/hazard_baseline.png',
        plot = hazard_plot, width = 6, height = 4)
 
 # can also then compare between taxonomic groups because i have that calculated
 # faceted baseline hazard plot by fossil_group
 db <- plot_taxon_hazard(disc_best)
-ggsave(filename = '../doc/figure/hazard_bygroup.png', plot = db,
+ggsave(filename = '../results/figure/hazard_bygroup.png', plot = db,
        width = 6, height = 8)
 
 
@@ -165,7 +165,7 @@ ggsave(filename = '../doc/figure/hazard_bygroup.png', plot = db,
 #   3 covariates
 #   4 taxonomic groups
 by_taxon <- plot_taxon_covariate_time(disc_best)
-ggsave(filename = '../doc/figure/eff_time_group.png', plot = by_taxon,
+ggsave(filename = '../results/figure/eff_time_group.png', plot = by_taxon,
        height = 6, width = 8)
 
 
@@ -173,9 +173,7 @@ ggsave(filename = '../doc/figure/eff_time_group.png', plot = by_taxon,
 # grab a random selection of species
 set.seed(100)
 risk_plots <- plot_risk_time(counti_trans, disc_best, nsp = 4)
[…]

### 2018-10-08 R/06_inspect_predict_model.r

@@ -68,7 +68,7 @@ oos_auc <- bind_cols(pred_auc) %>%
   ggplot(aes(x = value, fill = key)) + 
   stat_density(alpha = 0.5) +
   labs(x = 'AUC', y = 'density')
-ggsave(filename = '../doc/figure/fold_auc.png', plot = oos_auc,
+ggsave(filename = '../results/figure/fold_auc.png', plot = oos_auc,
        width = 6, height = 6)
 
 # then do it for time
@@ -81,5 +81,5 @@ ta <- reshape2::melt(time_auc) %>%
          time = parse_double(L2)) %>%
   ggplot(aes(x = time, y = value, colour = fold)) +
   stat_interval(alpha = 0.5, .width = c(0.5, 0.8))
-ggsave(filename = '../doc/figure/fold_auc_time.png', plot = ta,
+ggsave(filename = '../results/figure/fold_auc_time.png', plot = ta,
        width = 8, height = 6)

### 2018-10-08 R/helper01_process_foo.r

@@ -100,11 +100,13 @@ prepare_analysis <- function(x, fg = NULL) {
 break_my <- function(x, by = NULL, number = NULL) {
   top <- ceiling(max(x))
   bot <- floor(min(x))
+  
   if(!is.null(by)) {
     unt <- seq(from = bot, to = top, by = by)
   } else if(!is.null(number)) {
     unt <- seq(from = bot, to = top, length.out = number + 1)
   }
+
   unt1 <- unt[-length(unt)]
   unt2 <- unt[-1]
   uu <- map2(unt1, unt2, ~ which(between(x, left = .x, right = .y)))

### 2018-10-08 R/helper02_plot_foo.r

@@ -144,14 +144,7 @@ plot_risk_time <- function(data, model, nsp = 4) {
     facet_grid(~ fullname) + 
     labs(x = 'Age (My)', y = 'Geographic range (standardized)')
 
-  full_plot <- full_est %>%
-    gather(key, value, -fullname, -iterations, -relage, -row) %>%
-    ggplot(aes(x = relage, y = value)) +
-    geom_point(alpha = 0.01) + 
-    facet_grid(key ~ fullname, scales = 'free_y') +
-    labs(x = 'Age (My)', y = 'P(T = t | T >= t, x)')
-
-  out <- list(ext_plot, range_plot, full_plot)
+  out <- list(ext_plot, range_plot)
   out
 }
 
@@ -165,6 +158,7 @@ plot_risk_time <- function(data, model, nsp = 4) {
 #' @param model rstanarm object
 #' @return ggplot2 object
 plot_taxon_hazard <- function(model) {
+
   # the average
   haz_avg <- model %>%
     spread_draws(b[i, f], `(Intercept)`) %>%
@@ -223,6 +217,8 @@ plot_roc_series <- function(data, model_pp, model_key) {
   out <- list()
   for(kk in seq(length(model_pp))) {
     est <- list()
+
+    # gather the predicted vs observed event
     for(ii in seq(max(data$mybin))) {
       tw <- model_pp[[kk]][, data$fact_mybin == ii]
       ew <- data %>%
@@ -230,12 +226,16 @@ plot_roc_series <- function(data, model_pp, model_key) {
         dplyr::select(event)
       ew <- as.data.frame(ew)[, 1]
       oo <- list()
+   
+      # calculate roc for each posterior draw
       for(jj in seq(nrow(tw))) {
         # see here for the big source of exceptions...
         oo[[jj]] <- safe_roc(ew, tw[jj, ])
       }
+
       est[[ii]] <- oo
     }
+
     est <- set_names(est, seq(length(est)))
     out[[kk]] <- est
   }

### 2018-10-08 src/run_analysis.sh

@@ -1,9 +1,0 @@
-Rscript ../R/data_process.r &
-Rscript ../R/inspect_data.r &
-Rscript ../R/fit_model.r &
-Rscript ../R/compare_model.r &
-Rscript ../R/inspect_model.r &
-#pdflatex ../doc/main.txt
-#bibtex ../doc/main.aux
-#pdflatex ../doc/main.txt
-#pdflatex ../doc/main.txt

### 2018-10-11 doc/gsa2018_talk.tex
ic table. a few graphs. summary notes.

@@
 \usepackage{attrib}
+\usepackage{array}
 
@@
 
-\title{The changing functional composition of the North American species pool}
-\subtitle{ }
+\title{How predictable is extinction?} 
+\subtitle{Forecasting species survival at million-year timescales}
 \author{Peter D Smits, Seth Finnegan}
@@
     \begin{LARGE}
-      Extinction is \underline{hard} to predict, \\but is extremely important to conservation.
+      Extinction is \alert{hard} to predict, \\but is extremely \alert{important} to conservation decisions.
     \end{LARGE}
@@
         \item<.-> Bernoulli response distribution.
-        \item<.-> Time/phylum varying intercepts and slopes.
-        \item<.-> Taxon age as non-nested varying intercept.
+        \item<.-> Time varying intercepts and slopes; varies by phylum.
+        \item<.-> Taxon age as non-nested varying intercept; varies by phylum.
       \end{itemize}
@@
     \item Explore model \alert{adequacy} using posterior predictive distribution.
-    \item Estimate \alert{predictive performance} using \(k\)-fold cross-validation.
+    \item Estimate out-of-sample \alert{predictive performance} using \(k\)-fold cross-validation.
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{Comparing our models}
-
-  % table of waic/looic values
-  %   se values come out second to demonstrate how all virtually identical
+  \frametitle{In-sample predictive performance}
+  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/roc_hist}
 
@@
 \begin{frame}
-  \frametitle{Cross-validation results}
+  \frametitle{Comparing our models}
 
-  % figure comparing in-sample AUC to out-of-sample AUC
-  %   they are similar
+  \begin{tabular}{ r c c c c }
+    Model & LOOIC & SE LOOIC & WAIC & SE WAIC \\ 
+    \hline
+    Past and vary & 12790.39 & \alert<2>{178.83} & 12786.06 & \alert<2>{178.77} \\ 
+    No past but vary & 12818.43 & \alert<2>{178.76} & 12815.40 & \alert<2>{178.71} \\ 
+    Past but no vary & 12850.45 & \alert<2>{179.42} & 12848.12 & \alert<2>{179.38} \\ 
+    No past or vary & 12850.87 & \alert<2>{179.46} & 12848.50 & \alert<2>{179.42} \\ 
+    \hline
+  \end{tabular}
 
@@
 \begin{frame}
-  \frametitle{Tracking extinction risk over time}
+  \frametitle{Cross-validation results, full dataset}
+
+  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/fold_auc}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Cross-validation results, by time}
+
+  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/fold_ts}
 
@@
 \begin{frame}
-  \frametitle{Effects of age on extinction risk}
+  \frametitle{Effects of age on extinction risk, by phylum}
+
+  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/hazard_bygroup}
 
@@
 
+  \begin{itemize}
+    \item extinction is very random and our estimates aren't near perfect
+    \item historical covariates and varying effects seem important
+    \item increasing model complexity has only 
+  \end{itemize}
+
 \end{frame}
@@
 
+  \begin{itemize}
+    \item consider non-linear effects of historical covariates -- thresholds
+    \item at MILLION YEAR timescales past kind of matters, but very marginally
+    \item extinction is hard to predict at million year timescales -- rarity
+  \end{itemize}
 \end{frame}

### 2018-10-11 doc/gsa2018_talk.tex
some name reshuffles. growing talk.

@@
     \begin{LARGE}
-      Extinction is \alert{hard} to predict, \\but is extremely \alert{important} to conservation decisions.
+      Extinction is \alert{hard} to predict, but is \alert{important} to conservation decisions.
     \end{LARGE}
@@
 \begin{frame}
-  \frametitle{In-sample predictive performance}
+  \frametitle{In-sample predictive performance, full dataset}
+
   \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/roc_hist}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{In-sample predictive performance, by time}
+
+  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/roc_ts}
 
@@
 
-  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/fold_ts}
+  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
 

### 2018-10-11 R/00_data_process.r

@@ -62,6 +62,7 @@ nano <- nano %>%
   dplyr::mutate(fullname = str_c(genus, '_', species)) %>% 
   arrange(fullname)
 
+
 # assign everything a geographic cell using paleocoordinates
 eq <- CRS("+proj=cea +lat_0=0 +lon_0=0 +lat_ts=30 +a=6371228.0 +units=m")
 globe.map <- readOGR('../data/ne_10m_coastline.shp')  # from natural earth
@@ -125,7 +126,7 @@ sprange <- nano %>%
 mgca <- read_tsv('../data/cramer/cramer_temp.txt')
 mgca <- mgca %>% clean_names
 mgca <- mgca %>%
-  mutate(mybin = break_my(age)) %>%
+  mutate(mybin = break_my(age, by = 1)) %>%
   group_by(mybin) %>%
   summarize(temp = mean(temperature, na.rm = TRUE),
             temp_sd = sd(temperature, na.rm = TRUE)) %>%
@@ -205,6 +206,9 @@ counti$cc.rescale <- plyr::mapvalues(counti$cc,
 counti$cc.rescale <- with(counti, {
                             factor(cc.rescale, 
                                    levels = sort(unique(as.numeric(cc.rescale))))})
+
+
+counti
 
 # write to file
 write_rds(survi, path = '../data/survival.rds')

### 2018-10-11 R/01_inspect_data.r

@@ -32,7 +32,8 @@ counti_trans <- prepare_analysis(counti)
 
 
 # occurrences through time, labeled if LAD
-octg <-  counti %>%
+octg <-  
+  counti %>%
   mutate(state = case_when(event == 0 ~ 'Standard',
                            event == 1 ~ 'Last'),
          fossil_group = case_when(fossil_group == 'D' ~ 'Dinoflagellates',
@@ -43,7 +44,7 @@ octg <-  counti %>%
   stat_bin() +
   facet_grid(fossil_group ~ ., switch = 'y') +
   scale_fill_manual(name = 'Occurrence type',
-                    values = c('goldenrod', 'skyblue')) +
+                    values = c('skyblue', 'goldenrod')) +
   theme(legend.position = 'bottom') +
   labs(title = 'Occurrences', x = 'Time (My before present)', y = 'Count')
 ggsave(filename = '../results/figure/occ_time_label.png',
@@ -73,7 +74,7 @@ ocrg <- counti %>%
   stat_bin() +
   facet_grid(fossil_group ~ ., switch = 'y') +
   scale_fill_manual(name = 'State', 
-                    values = c('goldenrod', 'skyblue')) +
+                    values = c('skyblue', 'goldenrod')) +
   theme(legend.position = 'bottom') +
   labs(title = 'Age distribution', x = 'Age (My)', y = 'Count')
 ggsave(filename = '../results/figure/age_label.png',

### 2018-10-11 R/helper01_process_foo.r

@@ -98,6 +98,7 @@ prepare_analysis <- function(x, fg = NULL) {
 #' @param by bin width
 #' @return vector of bin memberships
 break_my <- function(x, by = NULL, number = NULL) {
+
   top <- ceiling(max(x))
   bot <- floor(min(x))
   

### 2018-10-12 doc/gsa2018_talk.tex
formula details
growing talk more and more. notes from seth

--- 2018-10-08 doc/gsa2018_talk.tex
@@
     \begin{LARGE}
-      Extinction is \underline{hard} to predict, \\but is extremely important to conservation.
+      Extinction is \alert{hard} to predict, but is \alert{important} to conservation decisions.
     \end{LARGE}
@@
   \begin{itemize}[<+->]
-    \item A taxon with a \alert{greater than average} global geographic range is likely to survive for longer than a taxon with \alert{less than average} global geographic range.
+    \item A taxon with a \alert{greater than average} global geographic range is likely to \alert{survive for longer} than a taxon with \alert{less than average} global geographic range.
     \item A taxon's global geographic range can change over time.
@@
 How is extinction risk impacted if that taxon's global geographic range has recently \alert{increased} or \alert{decreased}?
-  \end{itemize}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Encoding the past}
-
-  \begin{itemize}
-    \item Change in geographic range between current observation and previous observation.
-    \item Average global temperature at time of previous observation (Mg/Ca isotope).
-    \item Age in millions of years at time of observation.
   \end{itemize}
@@
     \begin{column}{0.5\textwidth}
-      \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/occ_time_label}
+      \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/occ_time_label}
     \end{column}
     \begin{column}{0.5\textwidth}
-      \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/age_label}
+      \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/age_label}
     \end{column}
@@
 
-  \begin{itemize}[<+->]
-    \item Predictors: geographic range, change in geographic range, global temperature, lag of global temperature.
-    \item Bayesian discrete-time survival model.
+  \begin{itemize}%[<+->]
+    \item Encoding the past
       \begin{itemize}
-        \item<.-> Bernoulli response distribution.
-        \item<.-> Time/phylum varying intercepts and slopes.
-        \item<.-> Taxon age as non-nested varying intercept.
+        \item Change in geographic range between current observation and previous observation.
+        \item Average global temperature at time of previous observation (Mg/Ca isotope).
+        \item Age in millions of years at time of observation.
       \end{itemize}
-    \item \alert{Compare} models using WAIC/LOOIC.
-    \item Explore model \alert{adequacy} using posterior predictive distribution.
-    \item Estimate \alert{predictive performance} using \(k\)-fold cross-validation.
+    %\item \alert{Compare} models using WAIC/LOOIC.
+    \item Explore model adequacy using posterior predictive distribution.
+    \item Estimate out-of-sample predictive performance using \(k\)-fold cross-validation.
   \end{itemize}
+
+\end{frame}
+
+% explain ROC
+% TYPES of errors
+\begin{frame}
+  \frametitle{Measuring performance: Reciever Operating Characteristic}
+
+  \begin{center}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{figure/wiki_ROC_space-2}
+  \end{center}
+
+  \attrib{\footnotesize{wikimedia}}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Measuring performance: Reciever Operating Characteristic}
+
+  \begin{center}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{figure/wiki_709px-ROC_curves}
+  \end{center}
+
+  \attrib{\footnotesize{wikimedia}}
+
+  % want to balance two rates: 
+  %   true-positive rate -- predicting extinct when actual is extinct
+  %   false-positive rate -- predicting extinct when actual is extant
+
+  % curve describes these metrics at various thresholds
+
+  % AUC is integral of ROC -- 0.5 is random classification, 1 is perfect classification
 
@@
 
+% demonstrate as ROC curve
 \begin{frame}
-  \frametitle{A statistical model for predicting extinction}
+  \frametitle{In-sample predictive performance, full dataset}
+
+  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/roc_hist}
 
@@
 \begin{frame}
-  \frametitle{Comparing our models}
+  \frametitle{In-sample predictive performance, by time}
 
-  % table of waic/looic values
-  %   se values come out second to demonstrate how all virtually identical
+  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/roc_ts}
+
+\end{frame}
+
+%% figure out way to present this graphically
+%\begin{frame}
+%  \frametitle{Comparing our models}
+%
+%  \begin{tabular}{ r c c c c }
+%    Model & LOOIC & SE LOOIC & WAIC & SE WAIC \\ 
+%    \hline
+%    Past and vary & 12790.39 & \alert<2>{178.83} & 12786.06 & \alert<2>{178.77} \\ 
+%    No past but vary & 12818.43 & \alert<2>{178.76} & 12815.40 & \alert<2>{178.71} \\ 
+%    Past but no vary & 12850.45 & \alert<2>{179.42} & 12848.12 & \alert<2>{179.38} \\ 
+%    No past or vary & 12850.87 & \alert<2>{179.46} & 12848.50 & \alert<2>{179.42} \\ 
+%    \hline
+%  \end{tabular}
+%
+%\end{frame}
+
+\begin{frame}
+  \frametitle{Cross-validation results, full dataset}
+
+  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/fold_auc}
 
@@
 \begin{frame}
-  \frametitle{Cross-validation results}
+  \frametitle{Cross-validation results, by time}
 
-  % figure comparing in-sample AUC to out-of-sample AUC
-  %   they are similar
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Tracking extinction risk over time}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Overall covariate effects}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Covariate effects over time}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Effects of age on extinction risk}
+  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
 
@@
 
+\appendix
+
+% move to back
+\begin{frame}
+  \frametitle{A statistical model for predicting extinction}
+
+  \begin{equation}
+    \begin{aligned}
+      t_{i} &\sim \text{Bernoulli}(\Theta) \\
+      \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{d[i], p[i]}) \\
+      B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
+      \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
+    % double check this\dots is A MVN dist?
+      A_{w, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
+      \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
+      \mu_{d} &\sim 
+      \begin{cases}
+        N(0, 10) & \text{if } d = 1 \\
+        N(-1, 3) & \text{if } d = 2 \\
+        N(0, 3) & \text{if } d > 2 \\
+      \end{cases}
+      \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
+      \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
+      \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
+      \tau_{B} &\sim C^{+}(5) \\
+      \tau_{\alpha} &\sim C^{+}(5) \\
+      \tau_{A} &\sim C^{+}(5) \\
+      \Omega_{B} &\sim LKJ(1) \\
+      \Omega_{\alpha} &\sim LKJ(1) \\
+      \Omega_{A} &\sim LKJ(1) \\
+    \end{aligned}
+  \end{equation}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Effects of age on extinction risk, by phylum}
+
+  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/hazard_bygroup}
+
+\end{frame}
+
 \end{document}

### 2018-10-15 R/02_fit_model.r
roc curve plot

@@ -73,7 +73,7 @@ part_glmer <- partial(stan_glmer,
                       prior_intercept = normal(-2, 10, autoscale = FALSE),
                       prior_aux = cauchy(0, 3, autoscale = FALSE),
                       thin = 4, 
-                      adapt_delta = 0.999999)
+                      adapt_delta = 0.99999999)
 
 # some models have differen't priors
 list_part_glmer <- 

### 2018-10-15 R/03_compare_models.r

@@ -111,7 +111,48 @@ pp_est_new <- cut_newpoint(pp_prob, list_cutpoint)
 # these new, rescaled 0-1 results are then fed through the ROC/AUC machine.
 partial_post_roc <- purrr::partial(post_roc, y = counti_trans)
 pp_roc <- map(pp_est_new, partial_post_roc)
+
+# extract the parts of the ROC plot
+roc_df <- map(pp_roc, function(x) 
+           imap(x, ~ tibble(sim = .y, 
+                            fpr = 1 - .x$specificities,
+                            tpr = .x$sensitivities))) %>%
+  map(., ~ reduce(.x, rbind)) %>%
+  imap(., ~ cbind(mod = .y, .x)) %>%
+  reduce(., rbind) %>%
+  as.tibble(.) %>%
+  mutate(model = case_when(mod == 1 ~ model_key[1],
+                           mod == 2 ~ model_key[2],
+                           mod == 3 ~ model_key[3],
+                           mod == 4 ~ model_key[4]))
+
+roc_df_back <- roc_df                  # extra for background
+
+# plot the curves from just the binaries
+cur <- roc_df %>%
+  ggplot(aes(x = fpr, 
+             y = tpr, 
+             group = sim)) +
+#  geom_line(data = roc_df_back,
+#            colour = 'grey10') +
+  geom_line(data = roc_df,
+            mapping = aes(colour = model),
+            alpha = 0.01) +
+  facet_wrap(model ~ ., nrow = 2) +
+  coord_equal(ratio = 1) +
+  labs(x = 'False Positive Rate',
+       y = 'True Positive Rate') +
+  theme(legend.position = 'none')
+ggsave(filename = '../results/figure/roc_curve.png', plot = cur,
+       width = 5, height = 8)
+
+
+# get AUC values for the above
 pp_auc <- map(pp_roc, function(y) map_dbl(y, ~ auc(.x)))
+
+
+
+
 
 roc_hist <- bind_rows(imap(pp_auc, ~ data.frame(model = .y, roc = .x))) %>%
   mutate(model = recode(model, !!!model_key),

### 2018-10-15 R/04_fit_predict_model.r

@@ -74,7 +74,7 @@ part_glmer <- partial(stan_glmer,
                       prior_intercept = normal(-2, 10, autoscale = FALSE),
                       prior_aux = cauchy(0, 3, autoscale = FALSE),
                       thin = 4, 
-                      adapt_delta = 0.999999)
+                      adapt_delta = 0.99999999)
 
 # some models have differen't priors
 priors <- list(prior = normal(c(0, 0, -1, 0), 
@@ -89,6 +89,7 @@ priors <- list(prior = normal(c(0, 0, -1, 0),
                prior = normal(c(0, -1), 
                               rep(1, 2), 
                               autoscale = FALSE))
+
 # combine formula, prior, and model
 by_formula <- map2(forms, priors, ~ partial(part_glmer, 
                                             formula = .x, 
@@ -109,8 +110,8 @@ fit <- map(to_fit, ~ .x[[1]](data = .x[[2]]))
 # help us remember this fact
 #   folds numbered from oldest (1) to youngest (4)
 nn <- c('mod1_', 'mod2_', 'mod3_', 'mod_4') %>%
-  map(., function(a) map(1:4, paste0(a, 'fold', .x)))
-fit <- set_names(fit, nn)
+  map(., function(a) map(1:4, ~ paste0(a, 'fold', .x)))
+fit <- set_names(fit, flatten(nn))
 
 
 write_rds(fit, path = '../data/training_fit.rds')

### 2018-10-19 doc/gsa2018_talk.tex
more desc of ROC
organizing end of talk

@@
 
-% explain ROC
-% TYPES of errors
+\begin{frame}
+  \frametitle{A conceptual model for predicting extinction}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Measuring performance: Reciever Operating Characteristic}
+
+  \begin{center}
+    \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
+      \cline{3-4} 
+      & & \multicolumn{2}{ c |}{Actual class} \\ \cline{3-4}
+      & & 1 & 0 \\ \hline
+      \multicolumn{1}{| c |}{\multirow{2}{*}{Predicted class}}
+      & 1 & TRUE \newline POSITIVE & FALSE \newline POSITIVE \\ \cline{2-4}
+      \multicolumn{1}{| c |}{} & 0 & FALSE \newline NEGATIVE & TRUE \newline NEGATIVE \\
+      \hline
+    \end{tabular}
+  \end{center}
+
+\end{frame}
+
 \begin{frame}
@@
 
-  % want to balance two rates: 
-  %   true-positive rate -- predicting extinct when actual is extinct
-  %   false-positive rate -- predicting extinct when actual is extant
-
-  % curve describes these metrics at various thresholds
-
-  % AUC is integral of ROC -- 0.5 is random classification, 1 is perfect classification
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{A conceptual model for predicting extinction}
-
 \end{frame}
@@
 
-%% figure out way to present this graphically
-%\begin{frame}
-%  \frametitle{Comparing our models}
-%
-%  \begin{tabular}{ r c c c c }
-%    Model & LOOIC & SE LOOIC & WAIC & SE WAIC \\ 
-%    \hline
-%    Past and vary & 12790.39 & \alert<2>{178.83} & 12786.06 & \alert<2>{178.77} \\ 
-%    No past but vary & 12818.43 & \alert<2>{178.76} & 12815.40 & \alert<2>{178.71} \\ 
-%    Past but no vary & 12850.45 & \alert<2>{179.42} & 12848.12 & \alert<2>{179.38} \\ 
-%    No past or vary & 12850.87 & \alert<2>{179.46} & 12848.50 & \alert<2>{179.42} \\ 
-%    \hline
-%  \end{tabular}
-%
-%\end{frame}
-
 \begin{frame}
@@
 \begin{frame}
-  \frametitle{Summary}
+  \frametitle{Next steps}
+
+  \begin{itemize}
+    \item Consider non-linear/threshold effects of covariates (e.g. GAMM, etc.).
+    \item What are the possible effects of other, unconsidered covariates (e.g. foram ecology)?
+  \end{itemize}
 
@@
 \begin{frame}
-  \frametitle{Conclusions}
+  \frametitle{Summary}
+
+  \begin{itemize}
+    \item \alert{The past matters.} Our best supported model includes our historical covariates and allows all effects to vary over time.
+    \item 
+  \end{itemize}
 
@@
 \end{frame}
-
-\appendix
-
-% move to back
-\begin{frame}
-  \frametitle{A statistical model for predicting extinction}
-
-  \begin{equation}
-    \begin{aligned}
-      t_{i} &\sim \text{Bernoulli}(\Theta) \\
-      \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{d[i], p[i]}) \\
-      B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
-      \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
-    % double check this\dots is A MVN dist?
-      A_{w, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
-      \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
-      \mu_{d} &\sim 
-      \begin{cases}
-        N(0, 10) & \text{if } d = 1 \\
-        N(-1, 3) & \text{if } d = 2 \\
-        N(0, 3) & \text{if } d > 2 \\
-      \end{cases}
-      \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
-      \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
-      \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
-      \tau_{B} &\sim C^{+}(5) \\
-      \tau_{\alpha} &\sim C^{+}(5) \\
-      \tau_{A} &\sim C^{+}(5) \\
-      \Omega_{B} &\sim LKJ(1) \\
-      \Omega_{\alpha} &\sim LKJ(1) \\
-      \Omega_{A} &\sim LKJ(1) \\
-    \end{aligned}
-  \end{equation}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Effects of age on extinction risk, by phylum}
-
-  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/hazard_bygroup}
-
-\end{frame}
-
-\end{document}

### 2018-10-24 doc/gsa2018_talk.tex
conceptual diagram made
cross validation diagram

@@
 
+  \begin{center}
+    \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{figure/conceptual_diagram}
+  \end{center}
+
 \end{frame}
@@
   \begin{center}
-    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{figure/wiki_709px-ROC_curves}
+    \includegraphics[width=\textwidth,height=0.75\textheight,keepaspectratio=true]{figure/wiki_709px-ROC_curves}
   \end{center}
@@
 
+\end{frame}
+
+\begin{frame}
+  \frametitle{\textit{k}-fold cross-validation}
+
+  \begin{center}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{figure/ts_cv}
+  \end{center}
+
+  \attrib{\footnotesize{Ken Williams, https://goo.gl/qLcfL8}}
 \end{frame}
@@
 \end{frame}
+
+\end{document}

### 2018-10-26 doc/gsa2018_talk.tex
redoing auc aspects of in-sample comparisons

@@
 \begin{frame}
-  \frametitle{Measuring performance: Reciever Operating Characteristic}
+  \frametitle{Measuring performance: confusion matrix}
 
@@
 \begin{frame}
-  \frametitle{Measuring performance: Reciever Operating Characteristic}
+  \frametitle{Measuring performance: Receiver Operating Characteristic}
 
@@
 \begin{frame}
-  \frametitle{\textit{k}-fold cross-validation}
+  \frametitle{Measuring performance: \textit{k}-fold cross-validation}
 
@@
 \begin{frame}
-  \frametitle{Next steps}
-
-  \begin{itemize}
-    \item Consider non-linear/threshold effects of covariates (e.g. GAMM, etc.).
-    \item What are the possible effects of other, unconsidered covariates (e.g. foram ecology)?
-  \end{itemize}
-
-\end{frame}
-
-\begin{frame}
   \frametitle{Summary}
@@
   \begin{itemize}
-    \item \alert{The past matters.} Our best supported model includes our historical covariates and allows all effects to vary over time.
-    \item 
+    \item \alert{The past matters\dots} 
+      \begin{itemize}
+        \item Our best supported model includes our historical covariates and allows all effects to vary over time.
+      \end{itemize}
+    \item \alert{But not that much\dots}
+      \begin{itemize}
+        \item None of our models are good at predicting extinction.
+      \end{itemize}
+    \item<2-> Mechanisms behind changes to geographic range operate at sub-million year scales. Perhaps their effects are weak/masked at million (or greater) year scales.
   \end{itemize}

### 2018-10-26 R/03_compare_models.r

@@ -56,6 +56,15 @@ energy <- map2(np, ll, ~ mcmc_nuts_energy(.x, .y))
 
 # bayes R2
 br2 <- map(disc_fit, bayes_R2)
+br2_gg <- reshape2::melt(br2) %>%
+  rename(model = L1) %>%
+  mutate(model_name = plyr::mapvalues(model, unique(model), model_key),
+         model_name = factor(model_name, levels = model_key)) %>%
+  ggplot(aes(x = value, y = model_name)) +
+  geom_halfeyeh(width = c(0.5, 0.8)) +
+  labs(x = 'Bayesian R^2', y = 'Model')
+ggsave(filename = '../results/figure/bayes_r2.png', plot = br2_gg,
+       width = 6, height = 6)
 
 
 # estimate of out-of-sample performance
@@ -86,34 +95,13 @@ pp_est <- map(pe, ~ future::value(.x))
 pp <- map(disc_fit, ~ future::future(posterior_linpred(.x, transform = TRUE)))
 pp_prob <- map(pp, ~ future::value(.x))
 
-# adequacy of fit ROC
-pgc <- partial(get_cutpoints, target = counti_trans$event)
-list_cutpoint <- map(map(pp_prob, pgc), ~ map(.x, ~ .x[3])) %>%
-  purrr::set_names(., model_key)
 
-
-cut_plot <- list_cutpoint %>%
-  reshape2::melt(list_cutpoint) %>%
-  as.tibble(.) %>%
-  magrittr::set_names(., c('value', 'draw', 'model')) %>%
-  dplyr::select(-draw) %>%
-  ggplot(., aes(x = value, y = model)) + 
-  geom_halfeyeh(.width = c(0.5, 0.8))
-ggsave(filename = '../results/figure/cut_plot.png', plot = cut_plot,
-       width = 6, height = 6)
-
-
-pp_est_new <- cut_newpoint(pp_prob, list_cutpoint)
-# this resets the cutpoint for determining 0 vs 1 by setting it very low.
-# this is a product of really high class imbalance. 
-# cutpoint based on maximizing both sensitivity and specificity.
-# this is calculated for every posterior predictive simulation for each model.
-# these new, rescaled 0-1 results are then fed through the ROC/AUC machine.
-partial_post_roc <- purrr::partial(post_roc, y = counti_trans)
-pp_roc <- map(pp_est_new, partial_post_roc)
+# use linpred to get ROC curve
+# probability gives finer resolution
[…]

### 2018-10-26 R/06_inspect_predict_model.r

@@ -54,7 +54,7 @@ fit <- read_rds('../data/training_fit.rds')
 counti_accum_match <- rep(counti_accum, 4)
 
 # get estimated linear predictor for each 
-plin <- map2(fit, counti_match, 
+plin <- map2(fit, counti_accum_match, 
              ~ posterior_linpred(object = .x, 
                                  newdata = .y, 
                                  draws = 1000))
@@ -73,29 +73,28 @@ pred <- map2(fit, counti_fold_match,
   cut_newpoint(., cut_points)
 
 
-
-
-# third: calculate ROC/AUC for the predictions of test data
+# calculate ROC/AUC for the predictions of test data
 # to determine how good our out of sample predictions are
 pred_auc <- map2(pred, counti_fold_match, post_roc) %>%
   map(., function(x) map_dbl(x, ~ auc(.x))) %>%
   set_names(., names(fit))
+
 
 # now to make a plot about out-of-sample predictive accuracy
 # because 1000s of numbers are hard to visualize
 
 oos_auc <- bind_cols(pred_auc) %>% 
   gather() %>%
-  separate(key, into = c('mod', 'fold'), sep = '\\_',
-           mod = plyr::mapvalues(mod, unique(mod), model_key)) %>%
+  separate(key, into = c('mod', 'fold'), sep = '\\_') %>%
+  mutate(mod = plyr::mapvalues(mod, unique(mod), model_key)) %>%
   ggplot(aes(x = value, y = mod)) +
   geom_density_ridges(rel_min_height = 0.01) +
-  labs(x = 'AUC', y = 'density')
+  labs(x = 'AUC', y = 'density') +
+  scale_colour_brewer()
 ggsave(filename = '../results/figure/fold_auc.png', plot = oos_auc,
        width = 6, height = 6)
 
 
-# below is untested
 # then do it for time
 time_auc <- map2(pred, counti_fold_match, get_auc_time) %>%
   set_names(., names(fit))
@@ -105,8 +104,10 @@ ta <- reshape2::melt(time_auc) %>%
   as.tibble %>%
   mutate(time = parse_double(L2)) %>%
   separate(L1, into = c('mod', 'fold'), sep = '\\_') %>%
+  mutate(mod = plyr::mapvalues(mod, unique(mod), model_key)) %>%
   ggplot(aes(x = time, y = value)) +
   stat_interval(alpha = 0.5, .width = c(0.5, 0.8)) +
[…]

### 2018-10-26 R/helper01_process_foo.r

@@ -99,6 +99,14 @@ prepare_analysis <- function(x, fg = NULL) {
 #' @return vector of bin memberships
 break_my <- function(x, by = NULL, number = NULL) {
 
+  if(is.null(by) & is.null(number)) {
+    return('no scheme given. specify either bin width or number of bins.')
+  }
+
+  if(!is.null(by) & !is.null(number)) {
+    return('too much information. specify either bin width OR number of bins, not both.')
+  }
+  
   top <- ceiling(max(x))
   bot <- floor(min(x))
   

### 2018-10-26 R/03_compare_models.r
redid in-sample performance graphs

@@ -90,9 +90,12 @@ xtable::print.xtable(xtable::xtable(compare_waic_tab),
 
 
 # posterior probability of observation surviving
-pe <- map(disc_fit, ~ future::future(posterior_predict(.x)))
+pe <- map(disc_fit, ~ future::future(posterior_predict(.x, 
+                                                       draws = 100)))
 pp_est <- map(pe, ~ future::value(.x))
-pp <- map(disc_fit, ~ future::future(posterior_linpred(.x, transform = TRUE)))
+pp <- map(disc_fit, ~ future::future(posterior_linpred(.x, 
+                                                       transform = TRUE, 
+                                                       draws = 100)))
 pp_prob <- map(pp, ~ future::value(.x))
 
 
@@ -116,7 +119,8 @@ roc_df <- map(eroc, function(x)
 
 roc_df_back <- roc_df                  # extra for background
 
-# plot the curves from just the binaries
+
+# plot the curves from the probabilities
 cur <- roc_df %>%
   ggplot(aes(x = fpr, 
              y = tpr, 
@@ -134,7 +138,9 @@ cur <- roc_df %>%
 ggsave(filename = '../results/figure/roc_curve.png', plot = cur,
        width = 5, height = 8)
 
+
 # get AUC values for the above
+# plot as histogram
 auc_hist <- map(eroc, ~ map(.x, function(y) auc(y)[[1]])) %>%
   reshape2::melt(.) %>%
   as.tibble %>%
@@ -149,13 +155,8 @@ ggsave(filename = '../results/figure/auc_hist.png', plot = auc_hist,
        width = 6, height = 6)
 
 
+# auc roc as timeseries to see best and worst times
 
-
-
-
-# roc as timeseries to see best and worst times
-roc_ts <- plot_roc_series(counti_trans, pp_est_new, model_key)
-ggsave(filename = '../results/figure/roc_ts.png', plot = roc_ts,
+auc_ts <- plot_roc_series(counti_trans, pp_prob, model_key)
+ggsave(filename = '../results/figure/auc_ts.png', plot = auc_ts,
        width = 8, height = 8)
-eauc <- map(eroc, ~ map(.x, function(y) auc(y)[[1]]))
-eauc <- map(eroc, ~ map(.x, function(y) auc(y)[[1]]))

### 2018-10-26 R/helper02_plot_foo.r

@@ -207,9 +207,7 @@ plot_taxon_hazard <- function(model) {
 #' @param model_key vector characters naming each model
 #' @return ggplot object
 plot_roc_series <- function(data, model_pp, model_key) {
-#data <- counti_trans
-#model_pp <- pp_est_new
-#model_key <- model_key
+  
   # important exception functions
   safe_roc <- safely(roc)
   safe_auc <- safely(auc)

### 2018-10-31 doc/gsa2018_talk.tex
full talk draft to send to SETH

@@
   \begin{center}
-    \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
-      \cline{3-4} 
-      & & \multicolumn{2}{ c |}{Actual class} \\ \cline{3-4}
-      & & 1 & 0 \\ \hline
-      \multicolumn{1}{| c |}{\multirow{2}{*}{Predicted class}}
-      & 1 & TRUE \newline POSITIVE & FALSE \newline POSITIVE \\ \cline{2-4}
-      \multicolumn{1}{| c |}{} & 0 & FALSE \newline NEGATIVE & TRUE \newline NEGATIVE \\
-      \hline
-    \end{tabular}
-  \end{center}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Measuring performance: Receiver Operating Characteristic}
-
-  \begin{center}
-    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{figure/wiki_ROC_space-2}
+    \includegraphics[width=\textwidth,height=0.75\textheight,keepaspectratio=true]{figure/confusion_matrix_wiki}
   \end{center}
@@
 \end{frame}
+%
+%
+%\begin{frame}
+%  \frametitle{Measuring performance: Receiver Operating Characteristic}
+%
+%  \begin{center}
+%    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{figure/wiki_ROC_space-2}
+%  \end{center}
+%  
+%  \attrib{\footnotesize{wikimedia}}
+%
+%\end{frame}
 
@@
 \begin{frame}
+  \frametitle{Measuring performance: AUC ROC}
+
+  \begin{columns}
+    \begin{column}{0.4\textwidth}
+      \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{figure/AUC}
+
+      \attrib{\footnotesize{https://goo.gl/91nEpM}}
+
+    \end{column}
+    \begin{column}{0.6\textwidth}
+      \[
+        \text{AUC} = 
+        \begin{cases} 
+          0.5 & \text{non discrimination}\\
+          0.6-0.7 & \text{poor} \\
+          0.7-0.8 & \text{acceptable/fair} \\
+          0.8-0.9 & \text{excellent/good} \\
+          > 0.9 & \text{outstanding} \\
+        \end{cases}
+      \]
+
+      The area represents the probability of correct ranking of a random ``extinct''-``extant'' pair.
+    \end{column}
+  \end{columns}
+
+\end{frame}
+
+\begin{frame}
   \frametitle{Measuring performance: \textit{k}-fold cross-validation}
@@
 
-% demonstrate as ROC curve
+% demonstrate as ROC curve?
 \begin{frame}
@@
 
-  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/roc_hist}
+  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_hist}
 
@@
 
-  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/roc_ts}
+  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_ts}
 
@@
       \end{itemize}
-    \item<2-> Mechanisms behind changes to geographic range operate at sub-million year scales. Perhaps their effects are weak/masked at million (or greater) year scales.
+    \item Perhaps mechanisms behind changes to geographic range operate at sub-million year scales. Perhaps their effects are weak/masked at million (or greater) year scales.
   \end{itemize}
@@
   \frametitle{Acknowledgements}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \begin{itemize}
+        \item \textbf{Seth Finnegan}
+        \item Adiel Klompmaker 
+        \item Emily Orzechowski
+        \item Larry Taylor
+        \item Sara Kahanamoku
+        \item Josh Zimmt
+        \item Franziska Franeck \\(University of Oslo)
+      \end{itemize}
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \begin{center}
+        \includegraphics[height=0.15\textheight,width=\textwidth,keepaspectratio=true]{figure/github-logo}
+
+        psmits.github.io/ \hspace*{0.05\textwidth} \textbf{trident}
+      \end{center}
+      \vspace*{0.02\textheight}
+      \begin{center}
+        \includegraphics[height=0.1\textheight,width=0.5\textwidth,keepaspectratio=true]{figure/twitter}
+
+        @PeterDSmits
+      \end{center}
+      \vspace*{0.02\textheight}
+      \begin{center}
+        \includegraphics[height=0.25\textheight,width=\textwidth,keepaspectratio=true]{figure/INAlogo}
+      \end{center}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\end{document}
 

### 2018-10-31 R/01_inspect_data.r

@@ -44,7 +44,7 @@ octg <-
   stat_bin() +
   facet_grid(fossil_group ~ ., switch = 'y') +
   scale_fill_manual(name = 'Occurrence type',
-                    values = c('skyblue', 'goldenrod')) +
+                    values = c('goldenrod', 'skyblue')) +
   theme(legend.position = 'bottom') +
   labs(title = 'Occurrences', x = 'Time (My before present)', y = 'Count')
 ggsave(filename = '../results/figure/occ_time_label.png',
@@ -74,7 +74,7 @@ ocrg <- counti %>%
   stat_bin() +
   facet_grid(fossil_group ~ ., switch = 'y') +
   scale_fill_manual(name = 'State', 
-                    values = c('skyblue', 'goldenrod')) +
+                    values = c('goldenrod', 'skyblue')) +
   theme(legend.position = 'bottom') +
   labs(title = 'Age distribution', x = 'Age (My)', y = 'Count')
 ggsave(filename = '../results/figure/age_label.png',

### 2018-10-31 R/03_compare_models.r

@@ -156,4 +156,5 @@ ggsave(filename = '../results/figure/auc_hist.png', plot = auc_hist,
 # roc as timeseries to see best and worst times
 roc_ts <- plot_roc_series(counti_trans, pp_prob, model_key) +
   scale_x_reverse()
-ggsave(filename = '../results/figure/roc_ts.png', plot = roc_ts,
+ggsave(filename = '../results/figure/auc_ts.png', plot = roc_ts,
+       width = 8, height = 6)

### 2018-10-31 R/06_inspect_predict_model.r

@@ -75,7 +75,7 @@ oos_auc <- bind_cols(pred_auc) %>%
          mod = factor(mod, levels = model_key)) %>%
   ggplot(aes(x = value, y = mod)) +
   geom_density_ridges(rel_min_height = 0.01) +
-  labs(x = 'AUC', y = 'density') +
+  labs(x = 'AUC', y = 'NULL') +
   scale_colour_brewer() +
   NULL
 ggsave(filename = '../results/figure/fold_auc.png', plot = oos_auc,

### 2018-10-31 R/helper02_plot_foo.r

@@ -207,10 +207,6 @@ plot_taxon_hazard <- function(model) {
 #' @param model_key vector characters naming each model
 #' @return ggplot object
 plot_roc_series <- function(data, model_pp, model_key) {
-<<<<<<< HEAD
-=======
-  
->>>>>>> f7140e3c43c52dfb0387d33f9afb017cd13b7d51
   # important exception functions
   safe_roc <- safely(roc)
   safe_auc <- safely(auc)

### 2018-10-31 R/03_compare_models.r
various typographic/font size changes

@@ -149,12 +149,16 @@ auc_hist <- map(eroc, ~ map(.x, function(y) auc(y)[[1]])) %>%
          model_name = factor(model_name, levels = model_key)) %>%
   ggplot(aes(x = value, y = model_name)) +
   geom_halfeyeh(.width = c(0.5, 0.8)) +
-  labs(x = 'AUC ROC', y = NULL)
+  labs(x = 'AUC ROC', y = NULL) +
+  theme(axis.text.y = element_text(size = 15),
+        axis.text.x = element_text(size = 15)) +
+  NULL
 ggsave(filename = '../results/figure/auc_hist.png', plot = auc_hist,
        width = 6, height = 6)
 
 # roc as timeseries to see best and worst times
 roc_ts <- plot_roc_series(counti_trans, pp_prob, model_key) +
-  scale_x_reverse()
+  scale_x_reverse() +
+  labs(y = 'AUC ROC', x = 'Time (Mya)')
 ggsave(filename = '../results/figure/auc_ts.png', plot = roc_ts,
        width = 8, height = 6)

### 2018-10-31 R/06_inspect_predict_model.r

@@ -74,9 +74,11 @@ oos_auc <- bind_cols(pred_auc) %>%
   mutate(mod = plyr::mapvalues(mod, unique(mod), model_key),
          mod = factor(mod, levels = model_key)) %>%
   ggplot(aes(x = value, y = mod)) +
-  geom_density_ridges(rel_min_height = 0.01) +
-  labs(x = 'AUC', y = 'NULL') +
+  geom_halfeyeh(.width = c(0.5, 0.8)) +
+  labs(x = 'AUC ROC', y = NULL) +
   scale_colour_brewer() +
+  theme(axis.text.y = element_text(size = 15),
+        axis.text.x = element_text(size = 15)) +
   NULL
 ggsave(filename = '../results/figure/fold_auc.png', plot = oos_auc,
        width = 6, height = 6)
@@ -97,6 +99,7 @@ ta <- reshape2::melt(time_auc) %>%
   stat_lineribbon() +
   facet_grid(mod ~ .) +
   scale_fill_brewer() +
-  scale_x_reverse()
+  scale_x_reverse() +
+  labs(y = 'AUC ROC', x = 'Time (Mya)')
 ggsave(filename = '../results/figure/fold_auc_time.png', plot = ta,
        width = 8, height = 6)

### 2018-10-31 R/helper02_plot_foo.r

@@ -258,7 +258,6 @@ plot_roc_series <- function(data, model_pp, model_key) {
     stat_lineribbon() +
     scale_fill_brewer() +
     facet_grid(model ~ .) +
-    labs(x = 'Time (My)', y = 'AUC') +
     NULL
 
   roc_ts

### 2018-11-01 doc/gsa2018_talk.tex
various typographic/font size changes
updated plots with seth comments

@@
 \begin{frame}
-  \frametitle{Data being analyzed}
+  \frametitle{Data being analyzed: Neptune database}
 
   \begin{columns}
-    \begin{column}{0.5\textwidth}
+    \begin{column}{0.1\textwidth}
+      \begin{center}
+        \includegraphics[width=\textwidth,height=0.15\textheight,keepaspectratio=true]{figure/calc}
+
+        \includegraphics[width=\textwidth,height=0.15\textheight,keepaspectratio=true]{figure/dino}
+
+        \includegraphics[width=\textwidth,height=0.15\textheight,keepaspectratio=true]{figure/foram}
+
+        \includegraphics[width=\textwidth,height=0.15\textheight,keepaspectratio=true]{figure/rad}
+
+        \attrib{\footnotesize{UCL}}
+      \end{center}
+    \end{column}
+    \begin{column}{0.4\textwidth}
       \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/occ_time_label}
     \end{column}
-    \begin{column}{0.5\textwidth}
+    \begin{column}{0.4\textwidth}
       \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/age_label}
@@
   \end{columns}
+
+  \footnotesize{Global occurrences from Deep Sea Drilling Program and Ocean Drilling Project. -- Lazarus. 1994. Math. Geo.; Spencer-Cervato. 1999. Palaeo. Elec.}
 
@@
         \item Change in geographic range between current observation and previous observation.
-        \item Average global temperature at time of previous observation (Mg/Ca isotope).
+        \item Average global temperature at time of previous observation (Mg/Ca elemental ratio).
         \item Age in millions of years at time of observation.
@@
 
+  \begin{center}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_hist_zoom}
+
+    \footnotesize{AUC = 0.7-0.8 acceptable/fair}
+  \end{center}
+
+\end{frame}
+
+% demonstrate as ROC curve?
+\begin{frame}
+  \frametitle{In-sample predictive performance, full dataset}
+
+  \begin{center}
     \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_hist}
+
+    \footnotesize{AUC = 0.7-0.8 acceptable/fair}
+  \end{center}
 
@@
 
+  \begin{center}
     \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_ts}
+  \end{center}
 
@@
 
-  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/fold_auc}
+  \begin{center}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/fold_auc_zoom}
+
+    \footnotesize{AUC = 0.7-0.8 acceptable/fair}
+  \end{center}
 
@@
 
+  \begin{center}
     \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
+  \end{center}
 
@@
       \begin{itemize}
-        \item None of our models are good at predicting extinction.
+        \item Models only average/fair expected out-of-sample performance.
       \end{itemize}
-    \item Perhaps mechanisms behind changes to geographic range operate at sub-million year scales. Perhaps their effects are weak/masked at million (or greater) year scales.
+    \item Mechanisms behind changes to geographic range operate at sub-million year scales. Perhaps their effects are weak/masked at million (or greater) year scales.
   \end{itemize}
@@
       \begin{itemize}
-        \item \textbf{Seth Finnegan}
+        \item UC Berkeley
+          \begin{itemize}
             \item Adiel Klompmaker 
@@
             \item Josh Zimmt
-        \item Franziska Franeck \\(University of Oslo)
+          \end{itemize}
+        \item University of Oslo
+          \begin{itemize}
+            \item Franziska Franeck
+          \end{itemize}
+        \item Neptune DB/MFN - Berlin
+          \begin{itemize}
+            \item David Lazarus
+          \end{itemize}
       \end{itemize}
@@
       \end{center}
+
       \vspace*{0.02\textheight}
+
       \begin{center}
-        \includegraphics[height=0.25\textheight,width=\textwidth,keepaspectratio=true]{figure/INAlogo}
+        \includegraphics[height=0.25\textheight,width=\textwidth,keepaspectratio=true]{figure/nsb_logo}
       \end{center}

### 2018-11-01 R/03_compare_models.r

@@ -4,6 +4,7 @@ library(janitor)
 library(ggridges)
 #devtools::install_github("mjskay/tidybayes")
 library(tidybayes)
+library(deeptime)
 
 # parallel processing
 library(parallel)
@@ -155,10 +156,14 @@ auc_hist <- map(eroc, ~ map(.x, function(y) auc(y)[[1]])) %>%
   NULL
 ggsave(filename = '../results/figure/auc_hist.png', plot = auc_hist,
        width = 6, height = 6)
+ggsave(filename = '../results/figure/auc_hist_zoom.png', 
+       plot = auc_hist + 
+         xlim(0.5, 1) +
+         geom_vline(xintercept = 0.5, colour = 'red'),
+       width = 6, height = 6)
 
 # roc as timeseries to see best and worst times
 roc_ts <- plot_roc_series(counti_trans, pp_prob, model_key) +
-  scale_x_reverse() +
   labs(y = 'AUC ROC', x = 'Time (Mya)')
 ggsave(filename = '../results/figure/auc_ts.png', plot = roc_ts,
        width = 8, height = 6)

### 2018-11-01 R/06_inspect_predict_model.r

@@ -82,6 +82,11 @@ oos_auc <- bind_cols(pred_auc) %>%
   NULL
 ggsave(filename = '../results/figure/fold_auc.png', plot = oos_auc,
        width = 6, height = 6)
+ggsave(filename = '../results/figure/fold_auc_zoom.png', 
+       plot = oos_auc + 
+         xlim(0.5, 1) +
+         geom_vline(xintercept = 0.5, colour = 'red'),
+       width = 6, height = 6)
 
 
 # then do it for time

### 2018-11-01 R/helper02_plot_foo.r

@@ -257,6 +257,7 @@ plot_roc_series <- function(data, model_pp, model_key) {
     ggplot(aes(x = key, y = value)) +
     stat_lineribbon() +
     scale_fill_brewer() +
+    scale_x_reverse() +
     facet_grid(model ~ .) +
     NULL
 

### 2018-11-01 R/03_compare_models.r
seth wanted geological time scales

@@ -164,6 +164,18 @@ ggsave(filename = '../results/figure/auc_hist_zoom.png',
 
 # roc as timeseries to see best and worst times
 roc_ts <- plot_roc_series(counti_trans, pp_prob, model_key) +
-  labs(y = 'AUC ROC', x = 'Time (Mya)')
+  coord_cartesian(ylim = c(0.4, 1), xlim = c(0, 62))
+ggsave(filename = '../results/figure/auc_ts_tiny.png', 
+       plot = gggeo_scale(roc_ts, 
+                          dat = 'epochs', 
+                          size = 3, 
+                          rot = 90,
+                          height = 0.2),
+       width = 8, height = 6)
+roc_ts <- roc_ts +
+  facet_grid(model ~ .) +
+  labs(y = 'AUC ROC', x = 'Time (Mya)') +
+  geom_hline(yintercept = 0.5, colour = 'red', linetype = 'dashed') +
+  NULL
 ggsave(filename = '../results/figure/auc_ts.png', plot = roc_ts,
        width = 8, height = 6)

### 2018-11-01 R/06_inspect_predict_model.r

@@ -3,6 +3,7 @@ library(tidyverse)
 library(janitor)
 #devtools::install_github("mjskay/tidybayes")
 library(tidybayes)
+library(deeptime)
 
 # parallel processing
 library(parallel)
@@ -102,9 +103,22 @@ ta <- reshape2::melt(time_auc) %>%
          mod = factor(mod, levels = rev(model_key))) %>%
   ggplot(aes(x = time, y = value)) +
   stat_lineribbon() +
-  facet_grid(mod ~ .) +
   scale_fill_brewer() +
   scale_x_reverse() +
-  labs(y = 'AUC ROC', x = 'Time (Mya)')
-ggsave(filename = '../results/figure/fold_auc_time.png', plot = ta,
+  coord_cartesian(ylim = c(0.4, 1), xlim = c(0, 50)) +
+  labs(y = 'AUC ROC', x = 'Time (Mya)') +
+  NULL
+ggsave(filename = '../results/figure/fold_auc_time_tiny.png', 
+       plot = gggeo_scale(ta, 
+                          dat = 'epochs', 
+                          size = 3, 
+                          rot = 90,
+                          height = 0.2),
+       width = 8, height = 6)
+ta <- ta +
+  facet_grid(mod ~ .) +
+  geom_hline(yintercept = 0.5, colour = 'red', linetype = 'dashed') +
+  NULL
+ggsave(filename = '../results/figure/fold_auc_time.png', 
+       plot = ta,
        width = 8, height = 6)

### 2018-11-01 R/helper02_plot_foo.r

@@ -258,8 +258,8 @@ plot_roc_series <- function(data, model_pp, model_key) {
     stat_lineribbon() +
     scale_fill_brewer() +
     scale_x_reverse() +
-    facet_grid(model ~ .) +
     NULL
+    
 
   roc_ts
 }

### 2018-11-28 doc/methods.tex
restructured dir. gsa2018 talk skeleton
readying updates for methods and results

@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/cramer_temp}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
   \caption{Comparison of initial temperature estimates from Cramer et al.
@@
 \end{equation}
-
-%The complement of the hazard function is called the survival function, which describes the probability of a species going extinct after time \(t\).
-Just as extinction is defined as when the current occurrence is the last occurrence, or \(T = t\), survival is defined as when the current occurrence is less than (in age) than the final occurrence, or \(T > t\).
-The discrete time survival function is expressed as
-%\begin{equation}
-%  S(t | X) = P(T > t | X) = \prod{i = 1}{t}(1 - \lambda(i | X).
-%  \label{eq:surv}
-%\end{equation}
 
@@
       \begin{cases}
-        N(0, 10) & \text{if } d = 1 \\
-        N(-1, 3) & \text{if } d = 2 \\
-        N(0, 3) & \text{if } d > 2 \\
+        N(-2, 5) & \text{if } d = \text{intercept} \\
+        N(-1, 1) & \text{if } d = \text{geo. range} \\
+        N(0, 1) & \text{else } \\
       \end{cases}
@@
              family = 'binomial',
-             prior = normal(c(-1, 0, 0, 0), rep(3, 4), autoscale = FALSE), 
+             prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
              prior_intercept = normal(0, 10, autoscale = FALSE), 
@@
              thin = 4,
-             adapt_delta = 0.9999).
+             adapt_delta = 0.999999).
 \end{verbatim}
@@
 
-One of the issues with our data set is extreme class imbalance; this means that there are many more survival observations than extinction observations.
-When the classes are not approximately balanced, the base probability of 1 versus 0 is below 0.5.
-When the intercept term of the logistic regression is sufficiently negative (-2 or less), the only way to obtain a linear predictor value of 0.5 or greater is if the covariate effects are large or there the covariates are at extreme values.
-The 0.5 default cutpoint for binary prediction is much to high when events are rare.
-A method for determining the new ``optimal'' cutpoint involves calculating the receiver operating characteristic (ROC) curve from the linear predictor estimates and then identifying the point on the ROC curve which maximizes both true positive rate and the false positive rate.
-The new cutpoint is then back-calculated from the true positive and false positive rates.
-Class membership is then predicted given this new cutpoint.
-This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cutpoint.
-
-%\section{Variance Partioning and Intraclass Correlation Coeficient}
-%
-%There are multiple levels and sources of variance in this model.
-Each of the grouping factors account for some amount of variance in the data.
-Normalizing that variance by the total variance from all sources, we can see the percent contribution to variance.
-%
-%``level 1'' is the ``data level.'' in a regular linear regression problem, this would be the residual variance.
-that's not calculated in logistic regression, which poses a problem.
-how do we get this value?
-simulation?
-%
-%``level 2'' is the ``time level.'' the logistic regression coefficients \textbf{vary} through time, we are actually estimating that variance in \(\Sigma_{B}\) and \(\tau_{B}\).
-there is also the non-nested age-based varying-intercept.
-this source of variance is in \(\Sigma_{a}\) and \(\tau_{a}\).
-%
-%``level 3'' is the ``phyla level.'' the over-time averages of the regression coefficients further vary between phyla.
-that variance is part of \(\Sigma_{\alpha}\) and \(\tau_{\alpha}\).
-there is also the average age effect on extinction probability which is part of \(\Sigma_{\delta}\) and \(\tau_{\delta}\).
-%
-%that's a lot of variance. 
-%
-%we want to figure out what the total variance is, so we can determine the percent contribution for each level and segment of variance.
-the level 1 issue and the nonnestedness pose potential challenges.
-
 See our code repository LINK for full code details.

### 2018-11-28 doc/results.tex
readying updates for methods and results

@@
 
-Our best model, as selected via LOOIC and WAIC, is the parameter rich ``past and vary'' model which includes our historical covariantes and allows all regression parameters to vary through time (Table \ref{tab:selection}).
+%Our best model, as selected via LOOIC and WAIC, is the parameter rich ``past and vary'' model which includes our historical covariantes and allows all regression parameters to vary through time (Table \ref{tab:selection}).
 Our best model is an improvement over the next best model by approximately 13 LOOIC and 24 WAIC points.
@@
 These results mean that while our ``past and vary'' model could be considered the best of the four, all four of the models are very close and only offer incremental improvement but not major differences.
-\begin{table}[ht]
-  \centering
-  \caption{Model comparison using WAIC and LOOIC}
-  \begin{tabular}{ r r r r r }
-    \hline
-    Model & looic & looic se & waic & waic se \\
-    \hline
-    Past and vary & 12784.04 & 178.71 & 12784.04 & 178.71 \\
-    No past but vary & 12819.01 & 178.80 & 12819.01 & 178.80 \\ 
-    Past but no vary & 12847.00 & 179.38 & 12847.00 & 179.38 \\ 
-    No past or vary & 12851.53 & 179.48 & 12851.53 & 179.48 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:selection}
-\end{table}
-
-As described above, our data is extremely unbalanced with respect to the number of extinction events relative to the total number of observations; this reality means that 0.5 is an inappropriate cutpoint for estimating if a species observation is their last (i.e.
+%\begin{table}[ht]
+%  \centering
+%  \caption{Model comparison using WAIC and LOOIC}
+%  \begin{tabular}{ r r r r r }
+%    \hline
+%    Model & looic & looic se & waic & waic se \\
+%    \hline
+%    Past and vary & 12784.04 & 178.71 & 12784.04 & 178.71 \\
+%    No past but vary & 12819.01 & 178.80 & 12819.01 & 178.80 \\ 
+%    Past but no vary & 12847.00 & 179.38 & 12847.00 & 179.38 \\ 
+%    No past or vary & 12851.53 & 179.48 & 12851.53 & 179.48 \\ 
+%    \hline
+%  \end{tabular}
+%  \label{tab:selection}
+%\end{table}
+%
+%As described above, our data is extremely unbalanced with respect to the number of extinction events relative to the total number of observations; this reality means that 0.5 is an inappropriate cutpoint for estimating if a species observation is their last (i.e.
 goes extinct).
 New, optimal cutpoint were calculated from the ROC curve as the level which maximized the true and false positive rates.
-
-The distributions of optimal cut-off point estimates for each of the four models are broadly similar, with all 50\% credible intervals overlapping.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cut_plot}
-  \caption{Posterior estimates of the optimal cutpoints for each of the three models.
+%
+%The distributions of optimal cut-off point estimates for each of the four models are broadly similar, with all 50\% credible intervals overlapping.
+%
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cut_plot}
+%  \caption{Posterior estimates of the optimal cutpoints for each of the three models.
 The optimal cutpoint corresponds to the point which maximizes the true and false positive rates; this new cutpoint is used instead of 0.5 when estimating if an observation is the last observation of that species (i.e.
 goes extinct).}
-  \label{fig:cut_point}
-\end{figure}
+%  \label{fig:cut_point}
+%\end{figure}
 
@@
 
-%When our cross-validation estimates are presented for each time interval, 
-%% ROC OOS estimate time series
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
-%  \caption{Approximate out-of-sample AUC values calculated for each of the My intervals using five-fold cross-validation of the time series.
+When our cross-validation estimates are presented for each time interval,
+
+% ROC OOS estimate time series
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
+  \caption{Approximate out-of-sample AUC values calculated for each of the My intervals using five-fold cross-validation of the time series.
 The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
 The AUC estimates from each fold are labeled and are numbered from oldest to youngest.}
-%  \label{fig:fold_auc_time}
-%\end{figure}
+  \label{fig:fold_auc_time}
+\end{figure}
+
+We can also compare expected out-of-sample performance based on taxonomic group
+
+Finally, expected out-of-sample performance is compared between taxonomic groups over time.
 

### 2018-11-28 doc/results.tex
added more to methods and results...tried at least

@@
 
-Because of the parameter rich ``past and vary'' model displays some modest improvements over the other models we choose this model for all further analysis.
+We can also compare the by-taxon in-sample performances of the four models.
+
+Finally, we can also examine how the by-taxon in-sample performances of the four models vary over time.
 
@@
 
-The approximate out-of-sample predictive performance for our selected model, as measured by AUC, is approximately identical to our in-sample performance (Fig.
+The approximate out-of-sample predictive performance for each of our four models, as measured by AUC, are approximately identical to each other and to our in-sample performance (Fig.
 \ref{fig:fold_auc}).
-The quality of performance, however, is not great, with average out-of-sample AUC estimated to be just above 0.7 which is far from perfect.
+The quality of performance, however, is not great, with average out-of-sample AUC for each of our models estimated to be about 0.77 which is far from perfect.
 This result means that while we expect our model to yield consistent results when provided with new data, we do not expect that our predictions will be very accurate.
@@
 
-When our cross-validation estimates are presented for each time interval,
+When our cross-validation estimates are presented for each time interval, the similarity in out-of-sample performance between the models becomes even more obvious -- while there are minor differences in the posterior estimates for individual time points, the overall structures are nearly identical.
 

### 2018-11-30 doc/intro.tex
adding more details to methods

--- 2018-09-04 doc/intro.tex
@@
 
-Species global geographic range changes over time due to both local and global processes.
+Instead of modeling geographic range through time, we instead model extinction probability as a function of change in geographic range.
+This formulation means that, if there if geographic range trajectory is deterministic and not entirely Markovian, that past geographic range, specifically the change in geographic range, and not just current geographic range predicts the probability of extinction.
+For example, is extinction probability affected by a species has been decreasing in geographic range over time or does it just matter if that species has below average geographic range regardless of how it got there?
 
-An old assumption in paleobiology was that species quickly obtained their maximum global geographic range and then maintained that range till their decline eventual extinction.
-Observation of species distribution records does not support this assumption.
-Instead, changes to a species' global geographic range is currently theorized to follow one of two general patterns: hat or hump PERFECT TIME FOR A FIGURE.
-Both are symmetric patterns but have different shapes.
-The hat pattern is characterized by an increase in geographic range for the first half a species' duration followed by a brief peak in geographic range, and the a decrease in geographic range for the second half of that species' duration.
-The hump pattern is instead characterized by a brief period of increasing species geographic range, an extended period of that species' duration at its maximum geographic range, and finally a quick decline and extinction of the species.
-
-Measuring this pattern is extremely difficult for multiple reasons.
-Species are not observed continuously through time, species geographic ranges are based on what samples are avaliable, and species are harder to observe at the tails of their duration because they are rare (small geographic range, small population sizes).
-
-Additionally, neither of these patterns are necessarily different from a Markovian process with a lower bound.
-A random walk beginning at 0 will most likely be unimodal with a period of increase followed by a period of decrease and then eventually reaching 0.
-PERFECT PLACE FOR A SIMULATION.
-
-\end{document}
+Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale.
+Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales.
+Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years.
+The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record.
+By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

### 2018-11-30 doc/methods.tex
added more to methods and results...tried at least
adding more details to methods

@@
 Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from Cramer et al CITATION.
-These are considered a more accurate estimate of global temperature than the standard Oxygen isotope-based estimates because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
-meteoric water).
+These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
+meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself.
 This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
@@
 
+We considered four variants of the model described above: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect).
+The second and fourth models modify the number of columns in \(X\) by removing two of the covariates.
+The third and fourth model simplify the first model to just varying intercepts without varying slopes.
+
 \section{Model Parameter Estimation}
@@
 
-An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as
+An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
 \begin{verbatim}
@@
              thin = 4,
-             adapt_delta = 0.999999).
+             adapt_delta = 0.999999)
 \end{verbatim}
 
-Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} as
+Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} CITATION as:
 \begin{verbatim}
@@
               thin = 4,
-              control = list(adapt_delta = 0.999999).
+              control = list(adapt_delta = 0.999999)
 \end{verbatim}
@@
 
-We considered four variants of the model described above: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect).
-The second and fourth models modify the number of columns in \(X\) by removing two of the covariates.
-The third and fourth model simplify the first model to just varying intercepts without varying slopes.
-
-The four variant models were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior CITATION.
-WAIC and LOOIC are interpreted similarly to AIC, with lower values indicating greater expected out-of-sample performance.
-Additionally, the Bayesian nature of WAIC and LOOIC mean that they are calculated with standard errors allowing for a more nuanced understanding the actual amount of difference between the models.
-We selected the model with the lowest WAIC and LOOIC, which happened to be are most parameter rich model.
-
-Model adequacy was measured using the area under the receiver operating characteristic curve (AUC).
-This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION.
-This value ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.
-AUC was calculated for the dataset as a whole, and for each of time bins.
-This measure was used in two contexts: measuring in-sample and out-of-sample predictive performance.
+We are interested in model adequacy and performance into two contexts: in-sample and out-of-sample predictive performance.
 ``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
@@
 
-We are also interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data).
+Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve (AUC).
+This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives.
+AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.
+
+The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time.
+These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.
+
+We are particularly interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data).
 To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation.

### 2018-11-30 doc/results.tex
added more flesh to some results

@@
 
+Our first model comparison is based on in-sample AUC estiamtes.
+These estimates describe the relative and absolute adequacy of our models to describe the data to which they were fit.
 Comparison of the in-sample AUC estimates between the four models reveals similar results as the selection criteria (Fig.
 \ref{fig:roc_hist}).
-While the parameter rich ``past and vary'' model has the greatest mean AUC when compared to the other three models, the other three models are quite similar.
-We chose to base our conclusions on the parameter rich ``past and vary'' model but these comparisons of model selection and adequacy illustrate the very small gain in performance that is achieved by allowing parameter effects to vary through time and including the historical covariates, though it appears that each of these choices on their own yield approximately equivalent models.
+While the parameter rich ``past and vary'' model has the greatest mean AUC when compared to the other three models, there is substantial overlap in their posterior distributions.
+Additionally, all of our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance.
+It is hard, then, to conclude that there is one ``best'' model when all of them are approximately equivalent.
+However, our mode complex models have relatively higher performance than our simplest models -- in particular, allowing covariate effects to vary over time appears to improve in-sample model performance as much as including the historical covariates.
 % ROC model comparison
@@
 
-Additional comparison of in-sample model performance at each of the time intervals reveals just how similar our four models are to each other (Fig.
+Additional comparison of in-sample model performance at each of the time intervals reveals just how similar in adequacy our models performance are to each other (Fig.
 \ref{fig:roc_ts}).
@@
 \section{Parameter estimates}
-
-The overall average probability of a species going extinct during any interval is very low (Fig.
-\ref{fig:effect_est}).
-This result means that even if covariate effects are very large on the log-odds scale, when compared on the probability scale will only produce a minor change in probability of extinction.
-This is one of the hardest parts of logistic regression for people to understand.
-Effect sizes are on the log-odds scale but prediction is on the probability scale.
-When the intercept term is very negative or positive (greater than 2), covariate effects are fighting over an increasingly small amount of probability.
-Logistic regressions are easiest to interpret when the intercept term is approximately 0; this is a fundamental difficulty of dealing with very uneven data sets.
 

### 2018-12-07 doc/intro.tex
growing introduction

@@
 
-How predictable is extinction? Forecasting species survival at million-year timescales.
+Paleobiology is concerned with understanding why certain species go extinct while others do not CITATIOn.
+Being able to predict which species are more likely to go extinct than others is critical for making good conservation decisions to limit the impact of the current biodiversity crisis.
+We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable.
+One way of approaching this problem is the analyze the past and use that to predict the future.
+The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant.
+By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions.
 
-One of the great promises of paleobiology is that by studying the past we can better predict the future.
-This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
-Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)?
-(2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?
+By studying how species vary in their extinction risk over time and we can assess which species are at greater risk under unobserved conditions.
+We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa).
 
+By including and modeling this natural variation in extinction risk, we are able to improve our overall predictions because we incorporate and explictly model differences between observations from across a range of possible intensities and selectivies
+
+or the intrisic risk of extinction -- all species can go extinct at any ``moment,'' what are the relative risks between groups and over time
+if we gather the kinds of predictors that apparently matter, and use them to make specific predictions of extinction 
+If we gather the kinds of predictors that have been found to correlate strongly with extinction risk, how well can we predict species extinction?
+
+Past studies have focused on identifying and measuring the effect of various predictors on extinction risk e.g. does difference in diet or geographic range affect differences in extinction risk?
+Here we focus on the quality of our predictions e.g. how accurately can we predict if one species is more likely to go extinct than another?
+increasing focus on prediction -- how accurate are we?
+understanding versus prediction
+long history focused on what determinates are important, but not on how accurate our predictions are given these factors.
+the evaluate how well we make predictions in the present
+trajectories studied before as phenonmenon, but not as tool -- we are using this information as tool!
+
+has been suggested CITE seth wolfgang michael
+incorporating extinction survival data from the fossil record can aide in predicting extinction risk of extant species -- the present must at some level be a function of the past
+what has not been evaluated is given selectivity changes and intensity changes, how accurate are assessements based on the past likely to be in the future.
+
+for this kind of excercise the best resolved fossil record of skeletonihze marine plantonic microorganisms (names). high temporal resolution, lots of samples, internally consistent taxonomy. 
+Rich fossil record -- we might be able to better understand predicting extinction 
+trajectories studied before as phenonmenon, but not as tool -- we are using this information as tool!
 To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
 Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature.
-Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.
+cite the papers that evalutae the importance of the predcitors.
+60 million year record lets us ask the 3 main things -- accurate, how much variation in accuracy,
 
-The effects of the historical covariates are extremely small and vary considerably over time.
-For example, the effect of change in geographic range size can either increase or decrease probability of extinction depending on when in the Cenozoic the observation takes place.
-The improvement in predictive power by including these historical covariates is modest at best and reflects the rarity of extinction events (i.e.
-class-imbalance) and the extremely stochastic nature of species survival.
-Correcting for class imbalance, we find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation.
-These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.
-This result implies that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
-The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.
+there have been major climate and oceanographic events that have prfoundly influenced the macroevolution of these groups CITATIONS.
+intervals of time that are particularly 
+things have changed over time -- these moments of big change are ``different'' so predicting them well is porentially important
 
-Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups.
-This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton.
-The greatest source of variance in survival probability is the timing of observation.
-Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age.
+the amoung of predictive power that the past has at any given 
+just the last million years may matter -- so change to range matters
+predictions to be better when similar stress over long period of time -- better on the average
+a more protracted event/transition (Eo-Mi) -- might expect there to be more predictive power then because past might matter more (key is protraction)
 
-Ultimately, we find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales.
-However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
-The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
-
-Instead of modeling geographic range through time, we instead model extinction probability as a function of change in geographic range.
-This formulation means that, if there if geographic range trajectory is deterministic and not entirely Markovian, that past geographic range, specifically the change in geographic range, and not just current geographic range predicts the probability of extinction.
-For example, is extinction probability affected by a species has been decreasing in geographic range over time or does it just matter if that species has below average geographic range regardless of how it got there?
-
-Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale.
-Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales.
-Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years.
-The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record.
-By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.
+\end{document}

### 2018-12-07 doc/intro.tex
passing intro

@@
 We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa).
+Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that varies over time is an important dynamic which shapes the rate and structure of extinction.
+By including the kinds of biological and abiological predictors that have been shown to affect differences in extinction risk, we can improve our predictions of species extinction risk for a given set of conditions.
+Additionaly, by specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explictly model differences between observations from across a range of possible intensities and selectivies
 
-By including and modeling this natural variation in extinction risk, we are able to improve our overall predictions because we incorporate and explictly model differences between observations from across a range of possible intensities and selectivies
-
-or the intrisic risk of extinction -- all species can go extinct at any ``moment,'' what are the relative risks between groups and over time
-if we gather the kinds of predictors that apparently matter, and use them to make specific predictions of extinction 
-If we gather the kinds of predictors that have been found to correlate strongly with extinction risk, how well can we predict species extinction?
+Extinction studies have normally been focused on understanding how biological and abiological factors impact differences in extinction risk over time and between taxa CITATION.
 

### 2018-12-10 doc/results.tex
added in taxon figures. filled out basic captions

@@
 
-\section{Model selection and adequacy}
-
-%Our best model, as selected via LOOIC and WAIC, is the parameter rich ``past and vary'' model which includes our historical covariantes and allows all regression parameters to vary through time (Table \ref{tab:selection}).
-Our best model is an improvement over the next best model by approximately 13 LOOIC and 24 WAIC points.
-However, the actual differences between these models are minor as revealed by the wide the standard errors.
-These results mean that while our ``past and vary'' model could be considered the best of the four, all four of the models are very close and only offer incremental improvement but not major differences.
-%\begin{table}[ht]
-%  \centering
-%  \caption{Model comparison using WAIC and LOOIC}
-%  \begin{tabular}{ r r r r r }
-%    \hline
-%    Model & looic & looic se & waic & waic se \\
-%    \hline
-%    Past and vary & 12784.04 & 178.71 & 12784.04 & 178.71 \\
-%    No past but vary & 12819.01 & 178.80 & 12819.01 & 178.80 \\ 
-%    Past but no vary & 12847.00 & 179.38 & 12847.00 & 179.38 \\ 
-%    No past or vary & 12851.53 & 179.48 & 12851.53 & 179.48 \\ 
-%    \hline
-%  \end{tabular}
-%  \label{tab:selection}
-%\end{table}
-%
-%As described above, our data is extremely unbalanced with respect to the number of extinction events relative to the total number of observations; this reality means that 0.5 is an inappropriate cutpoint for estimating if a species observation is their last (i.e.
-goes extinct).
-New, optimal cutpoint were calculated from the ROC curve as the level which maximized the true and false positive rates.
-%
-%The distributions of optimal cut-off point estimates for each of the four models are broadly similar, with all 50\% credible intervals overlapping.
-%
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cut_plot}
-%  \caption{Posterior estimates of the optimal cutpoints for each of the three models.
-The optimal cutpoint corresponds to the point which maximizes the true and false positive rates; this new cutpoint is used instead of 0.5 when estimating if an observation is the last observation of that species (i.e.
-goes extinct).}
-%  \label{fig:cut_point}
-%\end{figure}
+\section{Model adequacy}
 
@@
 Comparison of the in-sample AUC estimates between the four models reveals similar results as the selection criteria (Fig.
-\ref{fig:roc_hist}).
+\ref{fig:auc_hist}).
 While the parameter rich ``past and vary'' model has the greatest mean AUC when compared to the other three models, there is substantial overlap in their posterior distributions.
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/roc_hist}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist}
   \caption{Posterior predictive AUC estimates for each of the four models being compared.
@@
 AUC is bounded between 0.5 and 1.}
-  \label{fig:roc_hist}
+  \label{fig:auc_hist}
 \end{figure}
@@
 Additional comparison of in-sample model performance at each of the time intervals reveals just how similar in adequacy our models performance are to each other (Fig.
-\ref{fig:roc_ts}).
+\ref{fig:auc_ts}).
 There are few if any obvious differences between the models in their performances.
@@
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/roc_ts}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_ts}
   \caption{Comparison between the posterior predictive AUC estimates for each of the time intervals for each of the four models.
-These estimates are reflections of each model's ability to fit the various time intervals.
-The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals.
-In all cases, higher values AUC indicate greater performance versus lower AUC values.}
-  \label{fig:roc_ts}
+These estimates are reflections of each model's fit to the various time intervals.
+The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend.
+In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
+  \label{fig:auc_ts}
 \end{figure}
@@
 We can also compare the by-taxon in-sample performances of the four models. 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon}
+  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group.
+These estimates reflect each model's fit to the various taxonomic groups present in this analysis.
+The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals.
+In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
+  \label{fig:auc_taxon}
+\end{figure}
 
 Finally, we can also examine how the by-taxon in-sample performances of the four models vary over time.
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time}
+  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged over time and by taxonomic group.
+These estimates reflect each model's fit to the various taxonomic groups over time.
+The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend.
+In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
+  \label{fig:auc_taxon_time}
+\end{figure}
 
@@
 Each fold represents a section of data being predicted from a model fit to all data before the start of that fold.
-Given that there are only five folds, performance is measured from model predictions to 4 of the folds.
-Folds are labeled from oldest to youngest, with the oldest fold only being predicted by a single previous fold and the youngest fold being predicted by the four previous folds.}
+Given that there are only five folds, performance is measured from predictions for four of the folds.}
   \label{fig:fold_auc}
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
-  \caption{Approximate out-of-sample AUC values calculated for each of the My intervals using five-fold cross-validation of the time series.
+  \caption{Comparison of out-of-sample AUC values calculated for each of the My intervals for each of the four models.
 The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
-The AUC estimates from each fold are labeled and are numbered from oldest to youngest.}
+This presentation decomposes each of the 12-million year folds (Fig.
+\ref{fig:fold_auc}) into the predictions made for each of the million-year intervals.
+The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
   \label{fig:fold_auc_time}
@@
 We can also compare expected out-of-sample performance based on taxonomic group
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon}
+  \caption{Comparison of out-of-sample AUC values aggregated by taxonomic group for each of the four models.
+Depicted for each taxon-model combination is an aggregate density of all posterior predictive estimates for each of four folds -- cross-validation estimates are commonly multi-modal as each fold presents its own challenges for prediction.
+Beneath these densities is marked the median estimate along with 50\% and 80\% credible intervals.}
+  \label{fig:fold_auc_taxon}
+\end{figure}
 
 Finally, expected out-of-sample performance is compared between taxonomic groups over time.
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time}
+  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
+The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
+This presentation decomposes each of the 12-million year folds by each of the taxonomic groups (Fig.
+\ref{fig:fold_auc_taxon}) into the predictions made for each of the million-year intervals.
+The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
+  \label{fig:fold_auc_taxon_time}
+\end{figure}
 
-\section{Parameter estimates}
-
-As expected, a species with greater than average geographic range is expected to have a greater probability of surviving than a species with average or less than average geographic range (Fig.
+%\section{Parameter estimates}
+%
+%As expected, a species with greater than average geographic range is expected to have a greater probability of surviving than a species with average or less than average geographic range (Fig.
 \ref{fig:effect_est})
-
-We estimated an overall positive effect of the change in geographic range on extinction probability; this means that a species gaining in geographic range between that observation and their previous observed geographic range is associated with a greater extinction risk than a species which decreased in geographic range (Fig.
+%
+%We estimated an overall positive effect of the change in geographic range on extinction probability; this means that a species gaining in geographic range between that observation and their previous observed geographic range is associated with a greater extinction risk than a species which decreased in geographic range (Fig.
 \ref{fig:effect_est}).
 This sign of this result is consistent with virtually all previous analyses of the relation between geographic range and extinction.
-
-We estimated an overall positive effect of both global temperature and the lag of global temperature with species extinction probability (Fig.
+%
+%We estimated an overall positive effect of both global temperature and the lag of global temperature with species extinction probability (Fig.
 \ref{fig:effect_est}.
@@
 These results imply that, on average, the species during Paleogene had a greater extinction probability that species during the Neogene.
-
-% effects, average
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/effect_est}
-  \caption{Posterior estimates for the top-level covariate effects.
+%
+%% effects, average
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/effect_est}
+%  \caption{Posterior estimates for the top-level covariate effects.
 These effect estimates are effectively the weighted average of the estimates for each of the time intervals, for each of the phyla analyzed.
 Presented are posterior densities with the median estimate labeled along 80\% credible intervals at the bottom of the density.}
-  \label{fig:effect_est}
-\end{figure}
-
-The overall effect averages described above do not reflect the between time-interval variance of the effects.
+%  \label{fig:effect_est}
+%\end{figure}
+%
+%
+%The overall effect averages described above do not reflect the between time-interval variance of the effects.
 While the effect of geographic range has a consistent sign for the entire Cenozoic, the effects of the other covariates do not have a consistent sign (Fig.
 \ref{fig:effect_time_group}).
-
-% effects, time series
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/eff_time_group}
-  \caption{Effects of the four covariates as estimated for all time intervals and for each phyla.
+%
+%% effects, time series
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/eff_time_group}
+%  \caption{Effects of the four covariates as estimated for all time intervals and for each phyla.
 For each time series, the red line corresponds to the median estimate for each interval, while the envelopes represent various credible intervals as labeled.}
-  \label{fig:effect_time_group}
-\end{figure}
-
-% effect of age at observation
-
-% average hazard
-The hazard function describes how age of observation impacts probability of extinction.
+%  \label{fig:effect_time_group}
+%\end{figure}
+%
+%
+%% effect of age at observation
+%
+%% average hazard
+%The hazard function describes how age of observation impacts probability of extinction.
 Hazard is estimated for each phylum for the range of ages observed within the phylum.
@@
 The phylum-level averages of the effect of age of observation on extinction \(\delta\) are themselves drawn from a distribution centered around 0.
-
-Figure \ref{fig:hazard_baseline} is a graph of the average hazard associated with age of observation.
+%
+%
+%Figure \ref{fig:hazard_baseline} is a graph of the average hazard associated with age of observation.
 This graph illustrates how average extinction risk increases with age for the first few million years before leveling off.
@@
 The flat-ness of the median estimate and the homogeneity of variance durations between approximately 30 million years to under 60 million years is due to the lack of species having survived over 30 million years, which inherently means that these estimates are based on less information and are shrunk towards the overall average while reflecting more and more of our prior uncertainty over time.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/hazard_baseline}
-  \caption{Average hazard estimates based on all all observed species ages in millions of years.
+%
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/hazard_baseline}
+%  \caption{Average hazard estimates based on all all observed species ages in millions of years.
 Hazard in discrete-survival analysis is the probability of going extinct at a given age having survived up to that age.
@@
 The red line is the median hazard estimate along with 50\%, 80\%, and 95\% credible intervals.}
-  \label{fig:hazard_baseline}
-\end{figure}
-
-The hazard associated with the individual phyla have broad similarities with the overall hazard estimates (Fig.
+%  \label{fig:hazard_baseline}
+%\end{figure}
+%
+%The hazard associated with the individual phyla have broad similarities with the overall hazard estimates (Fig.
 \ref{fig:hazard_bygroup}).
@@
 In contrast, diatoms have the highest initial hazard of the phyla while having the smallest change hazard with age.
-
-Optimal cutpoint between 0.1 and 0.11 percent. We estimate that the effect of age of observation on extinction can reach this threshold after XX My at YY\% posterior probability.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/hazard_bygroup}
-  \caption{Hazard estimates for each of the phyla analyzed here. The red line is the median hazard estimate along with 50\%, 80\%, and 95\% credible intervals.}
-  \label{fig:hazard_bygroup}
-\end{figure}
-
-% variance components
-There are multiple levels and sources of variance in our model.
+%
+%Optimal cutpoint between 0.1 and 0.11 percent. We estimate that the effect of age of observation on extinction can reach this threshold after XX My at YY\% posterior probability.
+%
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/hazard_bygroup}
+%  \caption{Hazard estimates for each of the phyla analyzed here. The red line is the median hazard estimate along with 50\%, 80\%, and 95\% credible intervals.}
+%  \label{fig:hazard_bygroup}
+%\end{figure}
+%
+%
+%% variance components
+%There are multiple levels and sources of variance in our model.
 Each of the grouping factors account for some amount of variance in the data.
@@
 In the context of logistic regression, however, this value is not an inherent part of the model.
-
-There are then two non-nested sources of the second ``level'' of variance: intercept and slopes varying by time of observation, and a mean-zero age at observation varying intercept.
+%
+%There are then two non-nested sources of the second ``level'' of variance: intercept and slopes varying by time of observation, and a mean-zero age at observation varying intercept.
 Each of the elements of the vector of scale terms \(\tau_{B}\) describe the within-phyla variance for the intercept and the slope terms for the covariates.
 Similarly, the scale term \(\sigma_{A}\) is a scalar that describes the within-phyla variance of the effect of age at observation.
-
-The final ``level'' of variance the phyla-level: the between phyla variance for the intercepts and regression coefficients \(\tau_{\alpha}\), and the between phyla variance for the effect of age at observation \(\sigma_{\delta}\).
-
-The greatest source of variance accounted for by the grouping factors is the between phylum variance in the model intercept (Fig.
+%
+%The final ``level'' of variance the phyla-level: the between phyla variance for the intercepts and regression coefficients \(\tau_{\alpha}\), and the between phyla variance for the effect of age at observation \(\sigma_{\delta}\).
+%
+%The greatest source of variance accounted for by the grouping factors is the between phylum variance in the model intercept (Fig.
 \ref{fig:variance_components}); this is the first element of \(\tau_{B}\).
@@
 For example, the within-phyla effect of geographic range on extinction probability is associated with approximately as much variance as there is between the phyla themselves.
-
-This result means that, given this model, the greatest contributor to differences in extinction risk over time is when a species is present.
+%
+%This result means that, given this model, the greatest contributor to differences in extinction risk over time is when a species is present.
 In comparison, the covariates we studied individually contribute only a limited amount to the variation in extinction.
 Similarly, while age at observation does contribute to some differences in extinction risk (Fig XXX age effect), this effect is most likely overwhelmed by the importance of when that observation takes place.
-
-The variance in effects contributed by the phyla is approximately equal to the variance in those effects due to differences over time.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/variance_components}
-  \caption{Contribution of multi-level components to unmodeled variance.
+%
+%The variance in effects contributed by the phyla is approximately equal to the variance in those effects due to differences over time.
+%
+%
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/variance_components}
+%  \caption{Contribution of multi-level components to unmodeled variance.
 Larger values indicate a greater contribution to overall variance.
@@
 If the within component is greater than the overall component, then the overall variance is structured by the groupings.}
-  \label{fig:variance_components}
-\end{figure}
-
-Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of
-
-% risk estimate compared to change in geo-range
-\begin{figure}[ht]
-  \centering
-  \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
-    \caption{A}
-    \label{fig:relrisk_range}
-  \end{subfigure}
-
-  \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
-    \caption{B}
-    \label{fig:relrisk_ext}
-  \end{subfigure}
-  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation.
+%  \label{fig:variance_components}
+%\end{figure}
+%
+%
+%
+%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
+%
+%
+%% risk estimate compared to change in geo-range
+%\begin{figure}[ht]
+%  \centering
+%  \begin{subfigure}{\textwidth}
+%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
+%    \caption{A}
+%    \label{fig:relrisk_range}
+%  \end{subfigure}
+%
+%  \begin{subfigure}{\textwidth}
+%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
+%    \caption{B}
+%    \label{fig:relrisk_ext}
+%  \end{subfigure}
+%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation.
 Geographic range is assumed observed without error.
 Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
-  \label{fig:relrisk_compare}
-\end{figure}
+%  \label{fig:relrisk_compare}
+%\end{figure}
 

### 2018-12-11 doc/results.tex
adding text to results

@@
 
-Our first model comparison is based on in-sample AUC estiamtes.
-These estimates describe the relative and absolute adequacy of our models to describe the data to which they were fit.
-Comparison of the in-sample AUC estimates between the four models reveals similar results as the selection criteria (Fig.
+The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to.
+Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig.
 \ref{fig:auc_hist}).
-While the parameter rich ``past and vary'' model has the greatest mean AUC when compared to the other three models, there is substantial overlap in their posterior distributions.
-Additionally, all of our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance.
-It is hard, then, to conclude that there is one ``best'' model when all of them are approximately equivalent.
-However, our mode complex models have relatively higher performance than our simplest models -- in particular, allowing covariate effects to vary over time appears to improve in-sample model performance as much as including the historical covariates.
+The parameter rich ``past and vary'' model has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions.
+Additionally, while our parameter rich ``past and vary'' model is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions.
+All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance.
+It is then hard to conclude that there is one ``best'' model which we can rely upon.
 % ROC model comparison
@@
 
-Additional comparison of in-sample model performance at each of the time intervals reveals just how similar in adequacy our models performance are to each other (Fig.
+When the posterior predictive distributions of the in-sample AUC estimates are presented over time, the similarity in adequacy between the models becomes more apparent (Fig.
 \ref{fig:auc_ts}).
-There are few if any obvious differences between the models in their performances.
+There are few major or obvious differences in model adequacy between the four models.
 % ROC model comparison time series
@@
 
-We can also compare the by-taxon in-sample performances of the four models. 
+When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}).
 \begin{figure}[ht]
@@
 
-Finally, we can also examine how the by-taxon in-sample performances of the four models vary over time.
+Finally, when the posterior predictive distributions of the in-sample AUC estimates are presented over time and by taxonomic group, SOMETHING (Fig. \ref{fig:auc_taxon_time}).
 \begin{figure}[ht]

### 2018-12-11 doc/results.tex
describing figures in results

@@
 
-When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}).
+When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig.
+\ref{fig:auc_taxon}).
+While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig.
+\ref{fig:auc_hist}).
+For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.
+
 \begin{figure}[ht]

### 2018-12-12 doc/results.tex
filled out prediction results

@@
 For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.
-
 \begin{figure}[ht]
@@
 
-Finally, when the posterior predictive distributions of the in-sample AUC estimates are presented over time and by taxonomic group, SOMETHING (Fig. \ref{fig:auc_taxon_time}). 
+Finally, when the posterior predictive distributions of the in-sample AUC estimates are presented over time and by taxonomic group, the reasons for the differences in model adequacy between the taxonomic groups becomes more apparent (Fig.
+\ref{fig:auc_taxon_time}), in particular the marginally better performance of the ``past and vary'' model over the other three.
+
+For many taxon/model combinations there are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 -- AUC value of 0.5 indicates that the model's predictions are no better than random (Fig.
+\ref{fig:auc_taxon_time}).
+However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the ``past and vary'' model.
+Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the ``past and vary'' model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
 \begin{figure}[ht]
@@
 
-The approximate out-of-sample predictive performance for each of our four models, as measured by AUC, are approximately identical to each other and to our in-sample performance (Fig.
-\ref{fig:fold_auc}).
-The quality of performance, however, is not great, with average out-of-sample AUC for each of our models estimated to be about 0.77 which is far from perfect.
-This result means that while we expect our model to yield consistent results when provided with new data, we do not expect that our predictions will be very accurate.
-% ROC OOS estimate
+Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data CITATION.
+This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data.
+These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data CITATION.
+Additionally, multimodality increases with model complexity (Fig.
+\ref{fig:fold_auc}) -- this makes sense as the more complex models allow for predictor effects to vary with time, allowing for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
+
+Comparison between the posterior predictive distributions of expected out-of-sample AUC (Fig.
+\ref{fig:fold_auc}) reveals a similar range in plausible values for all models as the in-sample AUC posterior predictive distributions (Fig.
+\ref{fig:auc_hist}).
+Interestingly, the differences between the posterior predictive distributions for the models have decreased.
+For example, the ``past and vary'' model not clearly better than either of the ``no past or vary'' and ``past and no vary'' models (Fig.
+\ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig.
+\ref{fig:auc_hist}).
+These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
+However, the shapes of the posterior distributions means interepting from the median values is incorrect -- the models are effectively indistinguishable in their expected out-of-sample AUC values.
+
+Additionally, the quality of expected out-of-sample performance is not great, with average out-of-sample AUC for each of our models estimated to be between 0.7 and 0.8 which is far from perfect.
+This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time.
+However, this expected out-of-sample performance is approximately the same as the in-sample performance results (Fig.
+\ref{fig:auc_hist}), indicating that our models would yield consistent results when generalized to future extinctions.
 \begin{figure}[ht]
@@
 
-When our cross-validation estimates are presented for each time interval, the similarity in out-of-sample performance between the models becomes even more obvious -- while there are minor differences in the posterior estimates for individual time points, the overall structures are nearly identical.
-
-% ROC OOS estimate time series
+When the posterior predictive distribution of expected out-of-sample AUC is presented as a time series, the similarity between the models is even more apparent (Fig.
+\ref{fig:fold_auc_time}).
+While the width of the credible intervals at various time points varies between the models, the overall picture of expected out-of-sample AUC is almost identical when you compare the models -- periods of relatively better or worse performance map identically between the time series.
 \begin{figure}[ht]
@@
 
-We can also compare expected out-of-sample performance based on taxonomic group
+We can also compare expected out-of-sample AUC by taxonomic group for each of the models (Fig.
+\ref{fig:fold_auc_taxon}).
+These comparisons reveal a lot about the differences in predictive potential of the taxonomic groups.
+For example, the posterior predictive distributions of out-of-sample AUC for Foraminifera from all four models are approximately identical.
+In contrast, expected out-of-sample AUC for Radiolaria exhibits the same or similar pattern in relative model performance to the pooled comparisons (Fig.
+\ref{fig:fold_auc}).
+Additionally, we can state that our out-of-sample predictions for calcareous nannoplankton and dinoflagellates are not necessarily as precise as our estimates for Foraminifera or even Radiolaria.
+These results indicate that out-of-sample predictions may be easier for some taxonomic groups than others (e.g.
+Foraminifera versus Dinoflagellates.
 \begin{figure}[ht]
@@
 
-Finally, expected out-of-sample performance is compared between taxonomic groups over time.
+Finally, we can present the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig.
+\ref{fig:fold_auc_taxon_time}).
+For each taxonomic group, the time-series of posterior predictive values for each model are broadly congruent.
+
+In the analysis of the posterior predictive distributions of the in-sample AUC values for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig.
+\ref{fig:auc_taxon_time}).
+This occurrence is generally much rarer for the posterior predictive distribution of out-of-sample AUC values -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random.
+In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the ``no past or vary'' model.
 \begin{figure}[ht]

### 2018-12-20 doc/methods.tex
added in taxon figures. filled out basic captions
full set of cv figures for all datasets

@@
 We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax}.
-We downloaded all occurrence information for calcareous nannofossils, diatoms, foraminifera, and radiolarians from the entire globe and having occurred between 120 and 0 Mya.
-This selection of occurrences was then culled to just those species which have their first occurrence at most 63 My; this avoids those taxa which survived the K/Pg boundary, those taxa arising just after the K/Pg boundary, and lines up the occurrences with the temperature time-series (discussed below).
+All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
+This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya.
+This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction (discussed below).
 
-All occurrences were binned into 1 My bins based on the estimated age of the fossil occurrence.
+All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence.
 Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources.
-We then determined for each species the million year bins when it had occurred.
+The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species.
 For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0.
-The last occurrence of that species is considered the bin in which the taxon has gone extinct.
+The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1.
 This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species CITATIONS.
 However, this practice is common with marine microfossil data CITATIONS.
+In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.
 
@@
 
-\section{Model Selection and Adequacy}
+\section{Model adequacy}
 

### 2018-12-20 R/05_fit_predict_model.r

@@ -79,6 +79,7 @@ fit_models <- function(.data) {
   to_fit <- cross2(by_formula, counti_accum)
 
   # all that is missing is the data
+  # i have to admit that this is really cool -- beginning me would never have figured this out!
   fit <- map(to_fit, ~ .x[[1]](data = .x[[2]]))
   # everything is multiples of 4
   #   1:4 first form applied to each fold

### 2018-12-20 R/06_inspect_predict_model.r

@@ -63,6 +63,7 @@ fit <- read_rds('../data/training_fit.rds')
 fit_rt <- read_rds('../data/training_fit_rt.rds')
 fit_rl <- read_rds('../data/training_fit_rl.rds')
 
+# the following is called entirely for its side effects
 # general summary
 cv_model(fit, counti_fold_match, 
          model_key, 'full', '../results/figure/')

### 2018-12-20 R/helper02_plot_foo.r

@@ -693,7 +693,7 @@ get_pred <- function(fit, .data) {
 #' @return NULL
 cv_model <- function(fit, .data, key, name, path) {
   # predict the test data
-  pred <- get_pred(fig, .data)
+  pred <- get_pred(fit, .data)
   
   # calculate ROC/AUC for the predictions of test data
   # to determine how good our out of sample predictions are
@@ -742,7 +742,7 @@ cv_model <- function(fit, .data, key, name, path) {
 #' @return NULL
 cv_model_time <- function(fit, .data, key, name, path) {
   # predict the test data
-  pred <- get_pred(fig, .data)
+  pred <- get_pred(fit, .data)
   
   time_auc <- map2(pred, .data, get_auc_time) %>%
     set_names(., names(fit))
@@ -768,7 +768,7 @@ cv_model_time <- function(fit, .data, key, name, path) {
     coord_cartesian(ylim = c(0.4, 1), xlim = c(0, 50)) +
     labs(y = 'AUC ROC', x = 'Time (Mya)') +
     NULL
-  fn <- pasteo(path, 'fold_auc_time_tiny_', name, '.png')
+  fn <- paste0(path, 'fold_auc_time_tiny_', name, '.png')
   ggsave(filename = fn,
          plot = gggeo_scale(ta, 
                             dat = 'epochs', 
@@ -780,7 +780,7 @@ cv_model_time <- function(fit, .data, key, name, path) {
     facet_grid(mod ~ .) +
     geom_hline(yintercept = 0.5, colour = 'red', linetype = 'dashed') +
     NULL
-  fn <- pasteo(path, 'fold_auc_time_', name, '.png')
+  fn <- paste0(path, 'fold_auc_time_', name, '.png')
   ggsave(filename = fn,
          plot = ta,
          width = 11, height = 8.5)
@@ -800,9 +800,9 @@ cv_model_time <- function(fit, .data, key, name, path) {
 #' @return NULL
 cv_model_taxon <- function(fit, .data, key, name, path) {
   # predict the test data
-  pred <- get_pred(fig, .data)
-  
-  pp_taxon <- map2(pred, .data
+  pred <- get_pred(fit, .data)
+ 
+  pp_taxon <- map2(pred, .data, 
                    ~ split(data.frame(t(.x)), 
                            .y$fossil_group)) %>%
     map(., function(x) map(x, ~ t(.x)))
[…]

### 2018-12-20 doc/intro.tex

--- 2018-12-07 doc/intro.tex
@@
 We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa).
+Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that varies over time is an important dynamic which shapes the rate and structure of extinction CITATION.
+What has not been evaluated is as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
+By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range of possible intensities and selectivies
 
-By including and modeling this natural variation in extinction risk, we are able to improve our overall predictions because we incorporate and explictly model differences between observations from across a range of possible intensities and selectivies
+By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past.
+Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk CITATION or on how to identify or measure these effects CITATION.
+This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
+For example, while a predictor may be considered a ``significant'' factor in comparing the odds of extinction risk between groups, the difference that predictor makes in the probability of extinction can be minimal CITATION STATS TEXTBOOK LOGISTIC REGRESSION.
+By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.
 
-or the intrisic risk of extinction -- all species can go extinct at any ``moment,'' what are the relative risks between groups and over time
-if we gather the kinds of predictors that apparently matter, and use them to make specific predictions of extinction 
-If we gather the kinds of predictors that have been found to correlate strongly with extinction risk, how well can we predict species extinction?
+A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction.
+For example, we know that a species' global geographic range changes over its duration CITATION.
+We also know that a species' geographic range size is a good predictor of differences extinction risk CITATION.
+This begs the question: how does species' extinction risk change over their duration?
+While the phenomenon of species' geographic range change over time has been studied CITATION FOOTE LIOW WOLFGANG, the potential predictive impact of this change has been under-evaluated CITATION SEE WOLFGANG.
+For example, does a species' extinction risk increase if they decreased in global geographic range size over 1 million years?
+Here, we explicitly model and quantify the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction.
+Similarly, we include species geologic age at time of observation as a potential predictor of extinction -- a factor that may or may not contribute to differences in species extinction risk over time CITATION.
+Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.
 
-Past studies have focused on identifying and measuring the effect of various predictors on extinction risk e.g. does difference in diet or geographic range affect differences in extinction risk?
-Here we focus on the quality of our predictions e.g. how accurately can we predict if one species is more likely to go extinct than another?
-increasing focus on prediction -- how accurate are we?
-understanding versus prediction
-long history focused on what determinates are important, but not on how accurate our predictions are given these factors.
-the evaluate how well we make predictions in the present
-trajectories studied before as phenonmenon, but not as tool -- we are using this information as tool!
+For this kind of exercise, we chose to analyze what is the longest continuous and best resolved fossil record -- that of skeletonized marine plantonic microorganisms from the Cenozoic such as Foraminifera, Radiolarians, Diatoms, and calcareous nannofossils (e.g.
+coccolithophores).
+This data is available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project CITATION.
+This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.
+Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts.
 
-has been suggested CITE seth wolfgang michael
-incorporating extinction survival data from the fossil record can aide in predicting extinction risk of extant species -- the present must at some level be a function of the past
-what has not been evaluated is given selectivity changes and intensity changes, how accurate are assessements based on the past likely to be in the future.
+In particular, there have been multiple major climatic and oceanographic events, both short and protracted in duration, over the course of the Cenozoic.
+For example, the Paleocene-Eocene Thermal Maximum (PETM) was an geologically close-to instantaneous global warming event believed to have been caused by a massive injection of carbon into the atmosphere over a 20,000 year period which rapidly warmed and is associated with substantial changes to the global distribution of species CITATION.
+In contrast, the Eocene-Oligocene boundary is associated with a protracted global cooling event caused by the slow opening of Drake's passage and the development of the Antarctic ice sheet CITATION.
+These moments of large scale global change are inherently ``different,'' so predicting extinction before, during, and after these is important -- especially if those events are analogous to future climate change.
 
-for this kind of excercise the best resolved fossil record of skeletonihze marine plantonic microorganisms (names). high temporal resolution, lots of samples, internally consistent taxonomy. 
-Rich fossil record -- we might be able to better understand predicting extinction 
-trajectories studied before as phenonmenon, but not as tool -- we are using this information as tool!
-To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
-Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature.
-cite the papers that evalutae the importance of the predcitors.
-60 million year record lets us ask the 3 main things -- accurate, how much variation in accuracy,
+Additionally, we would expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions.
+Alternatively, we would expect our model based predictions of extinction surrounding the PETM may be less accurate because there are inherently fewer samples from the climatic event.
 
-there have been major climate and oceanographic events that have prfoundly influenced the macroevolution of these groups CITATIONS.
-intervals of time that are particularly 
-things have changed over time -- these moments of big change are ``different'' so predicting them well is porentially important
-
-the amoung of predictive power that the past has at any given 
-just the last million years may matter -- so change to range matters
-predictions to be better when similar stress over long period of time -- better on the average
-a more protracted event/transition (Eo-Mi) -- might expect there to be more predictive power then because past might matter more (key is protraction)
-
-\end{document}
+Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale.
+Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales.
+Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years.
+The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record.
+By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

### 2018-12-20 doc/results.tex
full set of cv figures for all datasets

@@
 It is then hard to conclude that there is one ``best'' model which we can rely upon.
+
 % ROC model comparison
@@
 For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.
+
+We can also compare the by-taxon in-sample performances of the four models.
+
+Finally, we can also examine how the by-taxon in-sample performances of the four models vary over time.
+
+\section{Cross-validation}
+
+The approximate out-of-sample predictive performance for each of our four models, as measured by AUC, are approximately identical to each other and to our in-sample performance (Fig.
+\ref{fig:fold_auc}).
+The quality of performance, however, is not great, with average out-of-sample AUC for each of our models estimated to be about 0.77 which is far from perfect.
+This result means that while we expect our model to yield consistent results when provided with new data, we do not expect that our predictions will be very accurate.
+% ROC OOS estimate
+
 \begin{figure}[ht]
@@
 \end{figure}
-
-Finally, when the posterior predictive distributions of the in-sample AUC estimates are presented over time and by taxonomic group, the reasons for the differences in model adequacy between the taxonomic groups becomes more apparent (Fig.
-\ref{fig:auc_taxon_time}), in particular the marginally better performance of the ``past and vary'' model over the other three.
 
@@
 \end{figure}
+
+We can also compare expected out-of-sample performance based on taxonomic group
+
+Finally, expected out-of-sample performance is compared between taxonomic groups over time.
 
@@
 \end{figure}
+
+%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
+%
+%
+%% risk estimate compared to change in geo-range
+%\begin{figure}[ht]
+%  \centering
+%  \begin{subfigure}{\textwidth}
+%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
+%    \caption{A}
+%    \label{fig:relrisk_range}
+%  \end{subfigure}
+%
+%  \begin{subfigure}{\textwidth}
+%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
+%    \caption{B}
+%    \label{fig:relrisk_ext}
+%  \end{subfigure}
+%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation.
+Geographic range is assumed observed without error.
+Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
+%  \label{fig:relrisk_compare}
+%\end{figure}
+
+\end{document}
 
@@
 %
-%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
-%
-%
-%% risk estimate compared to change in geo-range
-%\begin{figure}[ht]
-%  \centering
-%  \begin{subfigure}{\textwidth}
-%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
-%    \caption{A}
-%    \label{fig:relrisk_range}
-%  \end{subfigure}
-%
-%  \begin{subfigure}{\textwidth}
-%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
-%    \caption{B}
-%    \label{fig:relrisk_ext}
-%  \end{subfigure}
-%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation.
-Geographic range is assumed observed without error.
-Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
-%  \label{fig:relrisk_compare}
-%\end{figure}
-
-\end{document}

### 2019-01-03 doc/intro.tex
adding citations to intro draft

@@
 
-Paleobiology is concerned with understanding why certain species go extinct while others do not CITATIOn.
 Being able to predict which species are more likely to go extinct than others is critical for making good conservation decisions to limit the impact of the current biodiversity crisis.
 We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable.
-One way of approaching this problem is the analyze the past and use that to predict the future.
-The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant.
+We approach this this problem by analyzing the past in order to predict the future.
+The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012,Finnegan2015}.
 By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions.
@@
 By studying how species vary in their extinction risk over time and we can assess which species are at greater risk under unobserved conditions.
-We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa).
-Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that varies over time is an important dynamic which shapes the rate and structure of extinction CITATION.
+We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016k,Ezard2011}.
+Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that varies over time is an important dynamic which shapes the rate and structure of extinction.
 What has not been evaluated is as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
@@
 By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past.
-Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk CITATION or on how to identify or measure these effects CITATION.
+Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Exard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010,Alroy2014,Alroy2001,Alroy2000,Alroy2000a,Foote2001}.
 This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
-For example, while a predictor may be considered a ``significant'' factor in comparing the odds of extinction risk between groups, the difference that predictor makes in the probability of extinction can be minimal CITATION STATS TEXTBOOK LOGISTIC REGRESSION.
+For example, while a predictor may be a ``significant'' factor in comparing the odds of extinction risk between groups, the practical difference that predictor makes on extinction can be minimal \citep{ARM}.
 By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.
@@
 A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction.
-For example, we know that a species' global geographic range changes over its duration CITATION.
-We also know that a species' geographic range size is a good predictor of differences extinction risk CITATION.
+For example, we know that a species' global geographic range changes over its duration \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}.
+We also know that a species' geographic range size is a good predictor of differences extinction risk \citep{Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
 This begs the question: how does species' extinction risk change over their duration?
-While the phenomenon of species' geographic range change over time has been studied CITATION FOOTE LIOW WOLFGANG, the potential predictive impact of this change has been under-evaluated CITATION SEE WOLFGANG.
+While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}.
 For example, does a species' extinction risk increase if they decreased in global geographic range size over 1 million years?
 Here, we explicitly model and quantify the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction.
-Similarly, we include species geologic age at time of observation as a potential predictor of extinction -- a factor that may or may not contribute to differences in species extinction risk over time CITATION.
+Similarly, we include species geologic age at time of observation as a potential predictor of extinction -- a factor that may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016a}.
 Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.
@@
 coccolithophores).
-This data is available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project CITATION.
+This data is available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}.
 This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.
@@
 
-In particular, there have been multiple major climatic and oceanographic events, both short and protracted in duration, over the course of the Cenozoic.
+%In particular, there have been multiple major climatic and oceanographic events, both short and protracted in duration, over the course of the Cenozoic.
 For example, the Paleocene-Eocene Thermal Maximum (PETM) was an geologically close-to instantaneous global warming event believed to have been caused by a massive injection of carbon into the atmosphere over a 20,000 year period which rapidly warmed and is associated with substantial changes to the global distribution of species CITATION.
@@
 
-Additionally, we would expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions.
+Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions.
 Alternatively, we would expect our model based predictions of extinction surrounding the PETM may be less accurate because there are inherently fewer samples from the climatic event.
@@
 By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.
+
+\end{document}

### 2019-01-04 doc/methods.tex
citations to methods

@@
 The geodesic distance was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1.
-This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly comparable in magnitude.
+This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly comparable in magnitude \citep{ARM}.
 
-Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from Cramer et al CITATION.
+Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}.
 These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
@@
 This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
-Our data source, Cramer et al., provides temperature estimates for every 0.1 My from 0 to 63 Mya.
+Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya.
 We binned these estimates into 1 My intervals as we did with the fossil occurrences.
@@
 \(h(\Theta)\) is then modeled as with any regression as it is defined for all real-values.
-In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes CITATION.
+In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes \citep{ARM}.
 
@@
 These priors are also slightly more interpretable than other common prior distributions for covariance matrices such as the inverse-Wishart distribution.
-This approach to assigning priors to a covariance matrix is recommended by CITATION.
+This approach to assigning priors to a covariance matrix is recommended by the Stan Manual \citep{StanManual}.
 
@@
 
-We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language CITATION.
+We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}.
 This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R.
 Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling.
-In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.
+In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.
 
@@
 
-Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} CITATION as:
+Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} \citep{brms2017,brms2018} as:
 \begin{verbatim}
@@
 Posterior convergence was determined using the general and HMC-specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)).
-For further explanation of these diagnostic criteria, see the Stan Manual CITATION.
+For further explanation of these diagnostic criteria, see the Stan Manual \citep{StanManual}.
 
@@
 See our code repository LINK for full code details.
-Our code uses ``tidyverse'' tools such as dplyr CITATION, purr CITATION, and tidybayes CITATION, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
+Our code uses ``tidyverse'' tools such as dplyr \citep{dplyr}, purr \citep{purrr}, and tidybayes \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
 

### 2019-01-08 doc/methods.tex
condensed sections into paper.tex -- lots of references added and in bib file

@@
 
-We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax}.
+We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
 All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
@@
 The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1.
-This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species CITATIONS.
-However, this practice is common with marine microfossil data CITATIONS.
+This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang1996}.
+However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}.
 In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
-  \caption{Comparison of initial temperature estimates from Cramer et al.
-CITATION (goldenrod) versus the binned values used in this analysis (blue).
+  \caption{Comparison of initial temperature estimates from \citet{Cramer2011} (goldenrod) versus the binned values used in this analysis (blue).
 The initial values are for every 0.1 My while our bins are defined for every 1 My.}

### 2019-01-08 doc/paper.tex
condensed sections into paper.tex -- lots of references added and in bib file

\documentclass[12pt,letterpaper]{article}

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{docmute}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}

\frenchspacing

\title{How predictable is extinction? Forecasting species survival at million-year timescales}
\author{
  Smits, Peter\\
  %Department of Integrative Biology, University of California Berkeley\\
  \texttt{psmits@berkeley.edu} 
  \and
  Finnegan, Seth\\
  %Department of Integrative Biology, University of California Berkeley\\
  \texttt{sethf@berkeley.edu}
}

\begin{document}

\maketitle

\begin{abstract}
  One of the great promises of paleobiology is that by studying the past we can better predict the future.  This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.  Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?

  To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.

  The effects of the historical covariates are extremely small and vary considerably over time. For example, the effect of change in geographic range size can either increase or decrease probability of extinction depending on when in the Cenozoic the observation takes place. The improvement in predictive power by including these historical covariates is modest at best and reflects the rarity of extinction events (i.e. class-imbalance) and the extremely stochastic nature of species survival. Correcting for class imbalance, we find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation. These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.  This result implies that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.

  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. The greatest source of variance in survival probability is the timing of observation. Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age.

  Ultimately, we find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

\section{Introduction}

Being able to predict which species are more likely to go extinct than others is critical for making good conservation decisions to limit the impact of the current biodiversity crisis. We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable. We approach this this problem by analyzing the past in order to predict the future. The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012,Finnegan2015}. By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions.

By studying how species vary in their extinction risk over time and we can assess which species are at greater risk under unobserved conditions. We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}. Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that varies over time is an important dynamic which shapes the rate and structure of extinction. What has not been evaluated is as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range of possible intensities and selectivies

By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past. Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are. For example, while a predictor may be a ``significant'' factor in comparing the odds of extinction risk between groups, the practical difference that predictor makes on extinction can be minimal \citep{ARM}. By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.

A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction. For example, we know that a species' global geographic range changes over its duration \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}. We also know that a species' geographic range size is a good predictor of differences extinction risk \citep{Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. This begs the question: how does species' extinction risk change over their duration? While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}. For example, does a species' extinction risk increase if they decreased in global geographic range size over 1 million years? Here, we explicitly model and quantify the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction. Similarly, we include species geologic age at time of observation as a potential predictor of extinction -- a factor that may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}. Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.

  For this kind of exercise, we chose to analyze what is the longest continuous and best resolved fossil record -- that of skeletonized marine plantonic microorganisms from the Cenozoic such as Foraminifera, Radiolarians, Diatoms, and calcareous nannofossils (e.g. coccolithophores). This data is available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible. Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts.

%In particular, there have been multiple major climatic and oceanographic events, both short and protracted in duration, over the course of the Cenozoic. For example, the Paleocene-Eocene Thermal Maximum (PETM) was an geologically close-to instantaneous global warming event believed to have been caused by a massive injection of carbon into the atmosphere over a 20,000 year period which rapidly warmed and is associated with substantial changes to the global distribution of species CITATION. In contrast, the Eocene-Oligocene boundary is associated with a protracted global cooling event caused by the slow opening of Drake's passage and the development of the Antarctic ice sheet CITATION. These moments of large scale global change are inherently ``different,'' so predicting extinction before, during, and after these is important -- especially if those events are analogous to future climate change.

Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions. Alternatively, we would expect our model based predictions of extinction surrounding the PETM may be less accurate because there are inherently fewer samples from the climatic event.

Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record. By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

\section{Data Specifications}

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction (discussed below).

All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence. Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources. The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species. For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0. The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}. However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}. In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.

A taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the globe) between any two occurrences; this measure is also called a geodesic. This distance was measured in kilometers. The geodesic distance was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly comparable in magnitude \citep{ARM}.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig. \ref{fig:temp_curve}). Temperature was then transformed and standardized the in the same manner as geographic range (above).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
  \caption{Comparison of initial temperature estimates from \citet{Cramer2011} (goldenrod) versus the binned values used in this analysis (blue). The initial values are for every 0.1 My while our bins are defined for every 1 My.}
  \label{fig:temp_curve}
\end{figure}

\section{Model Specifications}

In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and relevant covariates. The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals. For the discrete time intervals \(T = 1, \cdots, k\), extinction is defined as \(T = t\). The discrete time hazard function is defined as
\begin{equation}
  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k. \\
  \label{eq:hazard}
\end{equation}

The hazard function (Eq. \ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\). \(h(\Theta)\) is then modeled as with any regression as it is defined for all real-values. In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes \citep{ARM}.

Our covariates matrix \(X\) is a \(N \times D\) matrix where \(N\) is the total number of observations and \(D\) is the total number of covariates. The first column of \(X\) is entirely 1's as it corresponds to the intercept term in the regression model. The next two columns of \(X\) are two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\). The difference in geographic range was calculated from the transformed and standardized geographic range values; this means that change in geographic range is in units of changes in standard deviations. The final two columns are two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e. mean temperature during interval \(t - 1\).) As with change to geographic range, the lag of temperature is based on the transformed and standardized temperature estimates.

The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\). These regression coefficients are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\). These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept. \(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.

This logistic regression model, minus the final selection of priors, is expressed as 
\begin{equation}
  \begin{aligned}
    t_{i} &\sim \text{Bernoulli}(\Theta) \\
    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{d[i], p[i]}) \\
    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
    % double check this\dots is A MVN dist?
    A_{w, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.

To complete the generative model, we need to assign final priors to the ``top level'' parameters. In general we favored weakly informative priors which help regularize our estimates. In the case of a regression coefficient, this means a Normal distribution with mean 0 and a standard deviation of 3. For our scale parameters (e.g. standard deviations), we used half-Cauchy distributed priors with heavy tails but the majority of probability density near 0.

Our top-level intercept was given a more diffuse prior than our regression coefficients, which reflects our greater degree of uncertainty about its value. Our top-level regression coefficient for the effect of geographic range was given an informative prior reflecting the overwhelming amount of evidence that species with a larger than average geographic range have a lower risk of extinction than species with an average or less than average geographic range. In the context of this analysis, this means that we are again using a weakly informative prior but instead of centering the density around -1 (i.e. larger than average geographic range decreases extinction risk).

Instead of assigning a prior distribution for each of the covariance matrices in the model, we instead decomposed the covariance matrices (e.g. \(\Sigma_{B}\)) which allows us to assign independent priors for the scale and correlation aspects of covariance. The scale parameters were assigned half-Cauchy priors as described above in the context of all other scale parameters. The correlation matrices were assigned LKJ priors each with shape parameter set to 1. This choice of shape parameter produces a uniform distribution over possible correlation matrices. These priors are also slightly more interpretable than other common prior distributions for covariance matrices such as the inverse-Wishart distribution. This approach to assigning priors to a covariance matrix is recommended by the Stan Manual \citep{StanManual}.

In total, these prior choices can be expressed as
\begin{equation}
  \begin{aligned}
    \mu_{d} &\sim 
      \begin{cases}
        N(-2, 5) & \text{if } d = \text{intercept} \\
        N(-1, 1) & \text{if } d = \text{geo. range} \\
        N(0, 1) & \text{else } \\
      \end{cases} \\
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{A} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{A} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:priors}
\end{equation}

We considered four variants of the model described above: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect). The second and fourth models modify the number of columns in \(X\) by removing two of the covariates. The third and fourth model simplify the first model to just varying intercepts without varying slopes.

\section{Model Parameter Estimation}

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.

An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
\begin{verbatim}
form <- event ~ range + range_diff + temp + temp_lag + 
                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                (1 | age/phylum), 
stan_glmer(formula = form,
           data = data, 
           family = 'binomial',
           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
           prior_intercept = normal(0, 10, autoscale = FALSE), 
           prior_aux = cauchy(0, 3, autoscale = FALSE), 
           chains = 4,
           thin = 4,
           adapt_delta = 0.999999)
\end{verbatim}

Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} \citep{brms2017,brms2018} as:
\begin{verbatim}
priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 3)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
form <- bf(event ~ range + range_diff + temp + temp_lag +
           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
           (1 | age/phylum))
brmfit <- brm(formula = form,
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999)
\end{verbatim}

Posterior convergence was determined using the general and HMC-specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)). For further explanation of these diagnostic criteria, see the Stan Manual \citep{StanManual}.

\section{Model adequacy}

We are interested in model adequacy and performance into two contexts: in-sample and out-of-sample predictive performance. ``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. ``Out-of-sample'' is defined below.

Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve (AUC). This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.

The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time. These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.

We are particularly interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data). To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

See our code repository LINK for full code details. Our code uses ``tidyverse'' tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.

\section{Model adequacy}

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich ``past and vary'' model has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich ``past and vary'' model is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance. It is then hard to conclude that there is one ``best'' model which we can rely upon.

% ROC model comparison
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist}
  \caption{Posterior predictive AUC estimates for each of the four models being compared. These estimates are calculated from each of the models posterior predictive distribution compared to the empirical values. Models with a higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1.}
  \label{fig:auc_hist}
\end{figure}

When the posterior predictive distributions of the in-sample AUC estimates are presented over time, the similarity in adequacy between the models becomes more apparent (Fig. \ref{fig:auc_ts}). There are few major or obvious differences in model adequacy between the four models.
% ROC model comparison time series
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_ts}
  \caption{Comparison between the posterior predictive AUC estimates for each of the time intervals for each of the four models. These estimates are reflections of each model's fit to the various time intervals. The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
  \label{fig:auc_ts}
\end{figure}

When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}). While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig. \ref{fig:auc_hist}). For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.

We can also compare the by-taxon in-sample performances of the four models.

Finally, we can also examine how the by-taxon in-sample performances of the four models vary over time.

\section{Cross-validation}

The approximate out-of-sample predictive performance for each of our four models, as measured by AUC, are approximately identical to each other and to our in-sample performance (Fig. \ref{fig:fold_auc}). The quality of performance, however, is not great, with average out-of-sample AUC for each of our models estimated to be about 0.77 which is far from perfect. This result means that while we expect our model to yield consistent results when provided with new data, we do not expect that our predictions will be very accurate.
% ROC OOS estimate

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon}
  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups present in this analysis. The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
  \label{fig:auc_taxon}
\end{figure}

For many taxon/model combinations there are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 -- AUC value of 0.5 indicates that the model's predictions are no better than random (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the ``past and vary'' model. Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the ``past and vary'' model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time}
  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
  \label{fig:auc_taxon_time}
\end{figure}

We can also compare expected out-of-sample performance based on taxonomic group

Finally, expected out-of-sample performance is compared between taxonomic groups over time.

\section{Cross-validation}

Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data CITATION. This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data. These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data CITATION. Additionally, multimodality increases with model complexity (Fig. \ref{fig:fold_auc}) -- this makes sense as the more complex models allow for predictor effects to vary with time, allowing for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

Comparison between the posterior predictive distributions of expected out-of-sample AUC (Fig. \ref{fig:fold_auc}) reveals a similar range in plausible values for all models as the in-sample AUC posterior predictive distributions (Fig. \ref{fig:auc_hist}). Interestingly, the differences between the posterior predictive distributions for the models have decreased. For example, the ``past and vary'' model not clearly better than either of the ``no past or vary'' and ``past and no vary'' models (Fig. \ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig. \ref{fig:auc_hist}). These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, the shapes of the posterior distributions means interepting from the median values is incorrect -- the models are effectively indistinguishable in their expected out-of-sample AUC values.

Additionally, the quality of expected out-of-sample performance is not great, with average out-of-sample AUC for each of our models estimated to be between 0.7 and 0.8 which is far from perfect. This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time. However, this expected out-of-sample performance is approximately the same as the in-sample performance results (Fig. \ref{fig:auc_hist}), indicating that our models would yield consistent results when generalized to future extinctions.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc}
  \caption{Results from our five-fold cross-validation of the time-series. Each labeled distribution of AUC values correspond to expected out-of-sample performance as estimated from that fold. Each fold represents a section of data being predicted from a model fit to all data before the start of that fold. Given that there are only five folds, performance is measured from predictions for four of the folds.}
  \label{fig:fold_auc}
\end{figure}

When the posterior predictive distribution of expected out-of-sample AUC is presented as a time series, the similarity between the models is even more apparent (Fig. \ref{fig:fold_auc_time}). While the width of the credible intervals at various time points varies between the models, the overall picture of expected out-of-sample AUC is almost identical when you compare the models -- periods of relatively better or worse performance map identically between the time series.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
  \caption{Comparison of out-of-sample AUC values calculated for each of the My intervals for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. This presentation decomposes each of the 12-million year folds (Fig. \ref{fig:fold_auc}) into the predictions made for each of the million-year intervals. The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
  \label{fig:fold_auc_time}
\end{figure}

We can also compare expected out-of-sample AUC by taxonomic group for each of the models (Fig. \ref{fig:fold_auc_taxon}). These comparisons reveal a lot about the differences in predictive potential of the taxonomic groups. For example, the posterior predictive distributions of out-of-sample AUC for Foraminifera from all four models are approximately identical. In contrast, expected out-of-sample AUC for Radiolaria exhibits the same or similar pattern in relative model performance to the pooled comparisons (Fig. \ref{fig:fold_auc}). Additionally, we can state that our out-of-sample predictions for calcareous nannoplankton and dinoflagellates are not necessarily as precise as our estimates for Foraminifera or even Radiolaria. These results indicate that out-of-sample predictions may be easier for some taxonomic groups than others (e.g. Foraminifera versus Dinoflagellates.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon}
  \caption{Comparison of out-of-sample AUC values aggregated by taxonomic group for each of the four models. Depicted for each taxon-model combination is an aggregate density of all posterior predictive estimates for each of four folds -- cross-validation estimates are commonly multi-modal as each fold presents its own challenges for prediction. Beneath these densities is marked the median estimate along with 50\% and 80\% credible intervals.}
  \label{fig:fold_auc_taxon}
\end{figure}

Finally, we can present the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}). For each taxonomic group, the time-series of posterior predictive values for each model are broadly congruent.

In the analysis of the posterior predictive distributions of the in-sample AUC values for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior predictive distribution of out-of-sample AUC values -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random. In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the ``no past or vary'' model.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time}
  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups (Fig. \ref{fig:fold_auc_taxon}) into the predictions made for each of the million-year intervals. The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
  \label{fig:fold_auc_taxon_time}
\end{figure}

%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
%
%
%% risk estimate compared to change in geo-range
%\begin{figure}[ht]
%  \centering
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
%    \caption{A}
%    \label{fig:relrisk_range}
%  \end{subfigure}
%
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
%    \caption{B}
%    \label{fig:relrisk_ext}
%  \end{subfigure}
%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation. Geographic range is assumed observed without error. Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
%  \label{fig:relrisk_compare}
%\end{figure}

\bibliographystyle{abbrvnat}
\bibliography{citations}

\end{document}

### 2019-01-08 doc/paper.tex
more citations and improving intro

@@
 class-imbalance) and the extremely stochastic nature of species survival.
-Correcting for class imbalance, we find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation.
+We find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation.
 These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.
@@
 
-  Ultimately, we find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales.
+  Ultimately, our results demonstrate that, given two random species, there is an approximately 77\% probability of correctly ranking their relative risks of extinction.
+We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales.
 However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
@@
 The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012,Finnegan2015}.
-By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions.
+By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions -- we ask the probability that, given two random species, we correctly rank their relative risks of extinction.
 
@@
 
-Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions.
-Alternatively, we would expect our model based predictions of extinction surrounding the PETM may be less accurate because there are inherently fewer samples from the climatic event.
+Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition \citep{Zachos2008} -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions.
+Alternatively, we would expect our model based predictions of extinction surrounding the Paleocene-Eocene Thermal Maximum may be less accurate because there are inherently fewer samples from the rapid climatic event \citep{Zachos2008}.
 
@@
 
-\section{Data Specifications}
+\section{Materials and Methods}
+
+\subsection{Data Specifications}
 
@@
 
-\section{Model Specifications}
+\subsection{Model Specifications}
 
@@
 
-\section{Model Parameter Estimation}
+\subsection{Model Parameter Estimation}
 
@@
 
-\section{Model adequacy}
+\subsection{Model adequacy}
 
@@
 
-Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve (AUC).
+Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}.
 This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives.
 AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.
+AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of any two randomly selected species \citep{Fawcett2006,Mason2002}.
 
@@
 
-\section{Model adequacy}
+\section{Results}
+
+\subsection{Model adequacy}
 
@@
 For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.
-
-We can also compare the by-taxon in-sample performances of the four models.
-
-Finally, we can also examine how the by-taxon in-sample performances of the four models vary over time.
-
-\section{Cross-validation}
-
-The approximate out-of-sample predictive performance for each of our four models, as measured by AUC, are approximately identical to each other and to our in-sample performance (Fig.
-\ref{fig:fold_auc}).
-The quality of performance, however, is not great, with average out-of-sample AUC for each of our models estimated to be about 0.77 which is far from perfect.
-This result means that while we expect our model to yield consistent results when provided with new data, we do not expect that our predictions will be very accurate.
-% ROC OOS estimate
-
 \begin{figure}[ht]
@@
 
-We can also compare expected out-of-sample performance based on taxonomic group
+\subsection{Cross-validation}
 
-Finally, expected out-of-sample performance is compared between taxonomic groups over time.
-
-\section{Cross-validation}
-
-Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data CITATION.
+Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data \citep{ESL}.
 This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data.
-These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data CITATION.
+These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data \citep{ESL}.
 Additionally, multimodality increases with model complexity (Fig.

### 2019-01-09 doc/paper.tex
continuing edits to introduction and abstract

@@
 
-  The effects of the historical covariates are extremely small and vary considerably over time.
-For example, the effect of change in geographic range size can either increase or decrease probability of extinction depending on when in the Cenozoic the observation takes place.
-The improvement in predictive power by including these historical covariates is modest at best and reflects the rarity of extinction events (i.e.
-class-imbalance) and the extremely stochastic nature of species survival.
-We find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation.
-These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.
-This result implies that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
+  Our results show that our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species.
+Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them.
+However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction.
+Using \textit{k}-fold cross-validation, we found that in-sample model performance measures are approximately equal to out-of-sample performance, meaning that we can have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past.
+These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
 The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.
@@
 
-  Ultimately, our results demonstrate that, given two random species, there is an approximately 77\% probability of correctly ranking their relative risks of extinction.
   We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales.
 However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
+These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.
 The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
@@
 By studying how species vary in their extinction risk over time and we can assess which species are at greater risk under unobserved conditions.
-We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}.
-Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that varies over time is an important dynamic which shapes the rate and structure of extinction.
-What has not been evaluated is as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
-By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range of possible intensities and selectivies
+We know that a species' risk of extinction varies over time in both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}.
+Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction.
+What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
+By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range extinction intensities and selectivities.
 
@@
 This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
-For example, while a predictor may be a ``significant'' factor in comparing the odds of extinction risk between groups, the practical difference that predictor makes on extinction can be minimal \citep{ARM}.
+For example, while a predictor may be ``significant'' when comparing the odds of extinction risk between groups, the practical difference that predictor makes on prediction can be minimal \citep{ARM}.
 By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.
@@
 We also know that a species' geographic range size is a good predictor of differences extinction risk \citep{Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
-This begs the question: how does species' extinction risk change over their duration?
-While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}.
-For example, does a species' extinction risk increase if they decreased in global geographic range size over 1 million years?
+This begs the question: how does a species' extinction risk change over its duration?
+While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}).
+For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years?
 Here, we explicitly model and quantify the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction.
@@
 This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.
-Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts.
+
+Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale.
+Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales.
+Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years.
+The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record.
+By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.
 
@@
 
+Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts.
 Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition \citep{Zachos2008} -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions.
 Alternatively, we would expect our model based predictions of extinction surrounding the Paleocene-Eocene Thermal Maximum may be less accurate because there are inherently fewer samples from the rapid climatic event \citep{Zachos2008}.
-
-Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale.
-Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales.
-Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years.
-The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record.
-By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.
 
@@
 In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and relevant covariates.
-The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals.
+The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals \citep{Tutz2016}.
 For the discrete time intervals \(T = 1, \cdots, k\), extinction is defined as \(T = t\).
@@
 The hazard function (Eq.
-\ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\).
+\ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\) \citep{Tutz2016}.
 \(h(\Theta)\) is then modeled as with any regression as it is defined for all real-values.
@@
 
+\section{Discussion}
+
 \bibliographystyle{abbrvnat}

### 2019-02-12 doc/paper.tex
large chunk of discussion vomited on page

@@
 \maketitle
+
+\linenumbers{}
+\modulolinenumbers[3]
 
@@
 
+The results of this paper are a measure of baseline ability to predict relative differences in extinction risk.
+By focusing on two aspects of species macroecology -- geographic range and global temperature -- we developed a rather naive model for estimating the probability of a species extinction based on their occurrence records.
+
+We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations.
+Similarly, these models are expected to correctly rank the extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time.
+A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original).
+The similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
+
+The striking aspects of our in-sample results are the statistical differences in AUC between the models, and the incredible narrow band of estimated AUC values.
+For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible.
+This means that even though one model has a statistically greater AUC, the practical difference between the models is not scientifically significant.
+
+This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses.
+By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored.
+For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the inverse-logit of the intercept is close to 0 or 1 CITATION.
+
+The success of model is partially driven by the size of our dataset and the structure of our model.
+Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping.
+Instead of relying on large amounts of ecological information to shape individual differences, our model leverages the entire Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.
+
+The reason we were not able to include more biological information in our models about the species studied is that we lack most any life history or ecological information on most marine micro- and nano-plankton.
+Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species.
+
+If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a subset of the fossil occurrences present the Neptune Database.
+This means ignoring the vast trove of occurrence information present in the rest of the database.
+
+This presents an interesting conundrum about how to improve upon our results.
+A simple hypothesis for how to improve upon our results is that if we were to include more biological information in our model, our estimates of species relative extinction risk would be improved.
+However, if we decrease the amount of data in our model, our results by definition decrease in their quality.
+For example, compare the results from our models fit using full in-sample dataset, and the results from the cross-validation where the fit to each fold is by definition based on less than full information.)
+
+In conclusion, our results provide a promising picture of our ability to predict the relative extinction risk of two randomly selected species.
+Considering that conservation decisions are made based on a continuum of risk, from most to least, this means that our results are in the same language as how conservation resources are allocated.
+
+% bibliographic information
 \bibliographystyle{abbrvnat}

### 2019-02-22 doc/paper.tex
small paper changes

@@
 (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?
-
   To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
-Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature.
-Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.
-
+We analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature.
+%Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.
   Our results show that our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species.
@@
 However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction.
-Using \textit{k}-fold cross-validation, we found that in-sample model performance measures are approximately equal to out-of-sample performance, meaning that we can have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past.
-These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
-The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.
-
+We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance.
   %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups.
@@
 Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age.
-
   We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales.
-However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
+The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
+These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
+The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.
 These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.
@@
 Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e.
-the globe) between any two occurrences; this measure is also called a geodesic.
+the Earth) between any two occurrences of that species; this measure is also called a geodesic.
 This distance was measured in kilometers.
-The geodesic distance was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1.
-This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly comparable in magnitude \citep{ARM}.
+Geographic range was then log-plus-one transformed, and standardized by zero-centering the data and then dividing by the standard deviation of the distribution of geographic ranges.
+This standardization means that the analyzed values of geographic range had mean 0 and standard deviation 1.
+This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the interecpt term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}.
+Change in geographic range between observations was measured from the transformed and standardized geographic range values and not standardized seperately.
 
@@
 Temperature was then transformed and standardized the in the same manner as geographic range (above).
+Difference in temperature between between observations was measured from the transformed and standardized temperature values and not standardized seperately.
 
@@
 
-The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\).
-These regression coefficients are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\).
+The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a  \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla.
+The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\).
 These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept.
@@
 
-This logistic regression model, minus the final selection of priors, is expressed as 
-\begin{equation}
-  \begin{aligned}
-    t_{i} &\sim \text{Bernoulli}(\Theta) \\
-    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{d[i], p[i]}) \\
-    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
-    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
-    % double check this\dots is A MVN dist?
-    A_{w, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
-    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
-  \end{aligned}
-  \label{eq:core}
-\end{equation}
-with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.
+The effect of species age on the log-odds of species extinction is modeled as a non-nested random intercept \(A\) \citep{Tutz2016}.
+This term describes how the log-odds of extinction varies along a species duration, and how this effect can differ between the phyla.
+\(A\) is a \(l\) by \(p\) matrix, where \(l\) is the age at observation of a species and \(p\) is its phylum.
+\(A\) is modeled as following a multivariate normal distribution with phylum means being vector \(\delta\) and covariance matrix \(\Sigma_{A}\).
+The covariation between the elements of vector \(\delta\) are modeled as a multivariate normal distribution with a mean vector of all 0s and covariance matrix \(\Sigma_{\delta}\).
 
@@
 
-In total, these prior choices can be expressed as
+In total, our model can be expressed as: 
 \begin{equation}
   \begin{aligned}
+    t_{i} &\sim \text{Bernoulli}(\Theta) \\
+    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{l[i], p[i]}) \\
+    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
+    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
+    % double check this\dots is A MVN dist?
+    A_{l, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
+    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
     \mu_{d} &\sim 
@@
 \end{equation}
+with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.
 
@@
 
-An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
-\begin{verbatim}
-form <- event ~ range + range_diff + temp + temp_lag + 
-                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
-                (1 | age/phylum), 
-stan_glmer(formula = form,
-           data = data, 
-           family = 'binomial',
-           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
-           prior_intercept = normal(0, 10, autoscale = FALSE), 
-           prior_aux = cauchy(0, 3, autoscale = FALSE), 
-           chains = 4,
-           thin = 4,
-           adapt_delta = 0.999999)
-\end{verbatim}
-
-Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} \citep{brms2017,brms2018} as:
-\begin{verbatim}
-priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
-            set_prior('normal(0, 1)', class = 'b'),
-            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
-            set_prior('cauchy(0, 3)', class = 'sd'),
-            set_prior('lkj(1)', class = 'cor'))
-form <- bf(event ~ range + range_diff + temp + temp_lag +
-           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
-           (1 | age/phylum))
-brmfit <- brm(formula = form,
-              data = data, 
-              family = bernoulli(), 
-              prior = priors,
-              chains = 4, 
-              thin = 4,
-              control = list(adapt_delta = 0.999999)
-\end{verbatim}
+%An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
+%\begin{verbatim}
+%form <- event ~ range + range_diff + temp + temp_lag + 
+%                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
+%                (1 | age/phylum), 
+%stan_glmer(formula = form,
+%           data = data, 
+%           family = 'binomial',
+%           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
+%           prior_intercept = normal(0, 10, autoscale = FALSE), 
+%           prior_aux = cauchy(0, 3, autoscale = FALSE), 
+%           chains = 4,
+%           thin = 4,
+%           adapt_delta = 0.999999)
+%\end{verbatim}
+%
+%Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} \citep{brms2017,brms2018} as:
+%\begin{verbatim}
+%priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
+%            set_prior('normal(0, 1)', class = 'b'),
+%            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
+%            set_prior('cauchy(0, 3)', class = 'sd'),
+%            set_prior('lkj(1)', class = 'cor'))
+%form <- bf(event ~ range + range_diff + temp + temp_lag +
+%           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
+%           (1 | age/phylum))
+%brmfit <- brm(formula = form,
+%              data = data, 
+%              family = bernoulli(), 
+%              prior = priors,
+%              chains = 4, 
+%              thin = 4,
+%              control = list(adapt_delta = 0.999999)
+%\end{verbatim}
 
@@
 
-The results of this paper are a measure of baseline ability to predict relative differences in extinction risk.
-By focusing on two aspects of species macroecology -- geographic range and global temperature -- we developed a rather naive model for estimating the probability of a species extinction based on their occurrence records.
-
+The results of this paper set out our baseline ability to predict relative differences in extinction risk.
 We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations.
@@
 
-The striking aspects of our in-sample results are the statistical differences in AUC between the models, and the incredible narrow band of estimated AUC values.
-For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible.
-This means that even though one model has a statistically greater AUC, the practical difference between the models is not scientifically significant.
+One of the most striking results of this analysis is that the in-sample and out-of-sample comparisons between our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig.
+\ref{fig:auc_hist}, \ref{fig:auc_ts}, \ref{fig:auc_taxon}, \ref{fig:auc_taxon_time}), all of them generalize to out-of-sample data with a similar degree of success (Fig.
+\ref{fig:fold_auc}, \ref{fig:fold_auc_time}, \ref{fig:fold_auc_taxon}).
+
+Other noteworthy aspects of our in-sample results are that AUC estimates for our models models differ significantly between models and that all of these estimates are in a narrow range of possible AUC values (Fig.
+\ref{fig:auc_hist}).
+For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig.
+\ref{fig:fold_auc}).
+This means that even though our ``past and vary'' model has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.
 
@@
 By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored.
-For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the inverse-logit of the intercept is close to 0 or 1 CITATION.
+For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the intercept is greater than 2 or less than -2 because the inverse-logit of those values are close to 0 or 1, respectively \citep{ARM}.
 
-The success of model is partially driven by the size of our dataset and the structure of our model.
+The success of model is partially driven by the size of our dataset and the hierarchical structure of our model.
 Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping.
-Instead of relying on large amounts of ecological information to shape individual differences, our model leverages the entire Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.
+Instead of relying on large amounts of ecological information to shape individual differences, our model leverages most of the Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.
 
-The reason we were not able to include more biological information in our models about the species studied is that we lack most any life history or ecological information on most marine micro- and nano-plankton.
-Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species.
-
-If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a subset of the fossil occurrences present the Neptune Database.
-This means ignoring the vast trove of occurrence information present in the rest of the database.
+The principal reason we were not able to include more biological information in our models is that we lack most any life history or ecological information on most marine micro- and nano-plankton.
+Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}.
+However, not this information does not exist all foram species.
+If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species.
+This means ignoring the majority of occurrence information present in the Neptune database.
 

### 2019-02-22 R/helper06_geotime.r

@@ -2,7 +2,7 @@
 
 #' Highlight timeseries using geological time scale
 get_geotime_box <- function(time_range, type = 'age', alternate = 'even') {
-  gts <- read_csv(here('data', 'geologic_time_scale.csv')
+  gts <- read_csv(here('data', 'geologic_time_scale.csv'))
 
   lts <- 
     gts %>%

### 2019-02-23 doc/model_descibe.tex
updating graphs. everything runs headless. table of model desc

\documentclass[12pt,letterpaper]{article}

\usepackage[margin=1cm]{geometry}

\begin{document}

\begin{tabular}[ht]{ l l l l }
  Code & Description & Covariates & Formula \\
  \hline
  C & Constant effects, no historical covariates & Geographic range, global temperature & $ event \sim range + temp + (1 | age/phylum) $ \\
  V & Varying effects, no historical covariates & Geographic range, global temperature  & $ event \sim range + temp + (1 + range + temp | mybin/phylum) + (1 | age/phylum) $ \\
  CP & Constant effects, historical covariates & Geographic range, change in geographic range, global temperature, previous temperature & $ event \sim range + range\_diff + temp + temp\_lag + (1 | age/phylum) $ \\
  VP & Varying effects, historical covariates & Geographic range, change in geographic range, global temperature, previous temperature & $ event \sim range + range\_diff + temp + temp\_lag + (1 + range + range\_diff + temp + temp\_lag | mybin/phylum) + (1 | age/phylum) $ \\
\end{tabular}

{
  \def\arraystretch{1.5}
  \begin{tabular}[ht]{ l p{3cm} l l }
    Code & Description & Covariates & Formula \\
    \hline
    C & Constant effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 $|$ age/phylum)} \\
    V & Varying effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ phylum) +\\(1 $|$ age/phylum)} \\ 
    CP & Constant effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range,\\temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff +\\temp + temp\_lag +\\(1 $|$ age/phylum)} \\
    VP & Varying effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range,\\temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff +\\temp + temp\_lag +\\(1 + range + range\_diff + temp + temp\_lag $|$ phylum) +\\(1 $|$ age/phylum)} \\
  \end{tabular}
}

\end{document}

### 2019-02-23 R/00_data_process.r

@@ -1,23 +1,16 @@
-# because of course
-library(tidyverse)
-library(janitor)
+library(pacman)
 
-# geolocation stuff
-library(sp)
-library(dismo)
-library(raster)
-library(rgdal)
-library(XML)
-library(maptools)
-library(geosphere)
-
-# plotting stuff
-library(scales)
+p_load(tidyverse, janitor, here, furrr,
+       sp, dismo, raster, rgdal, XML, maptools, geosphere, 
+       scales)
 
 # misc
-library(here)
 source(here('R', 'helper01_process_foo.r'))
 source(here('R', 'helper03_misc_foo.r'))
+
+# parallel set up
+plan(multiprocess)
+
 
 # process timescale information
 gts <- read_csv("https://raw.githubusercontent.com/japhir/stratPlot/master/GTS_colours.csv") %>%
@@ -74,45 +67,48 @@ partial_short_bin <- partial(raw_to_clean,
 
 prep_short <- function(width = bin_width) {
 
-  varname <- paste0("bin_", width)
+  varname <- paste0("bin_", width)     # tidy eval stuff
 
   partial_short_bin(bin_width = width)  %>%
   dplyr::select(fullname, mybin, latext, nocc, ncell, maxgcd) %>%
-  rename(!!varname := mybin) %>%
-  gather(key = 'key', value = 'value', -fullname, -maxgcd, -latext, -nocc, -ncell) %>%
+  rename(!!varname := mybin) %>%       # tidy eval stuff
+  gather(key = 'key', value = 'value', 
+         -fullname, -maxgcd, -latext, -nocc, -ncell) %>%
   separate(key, c('bin', 'width'), sep = '_') %>%
   mutate(width = parse_number(width)) %>%
   dplyr::select(maxgcd, latext, nocc, ncell, width)
 
 }
 
-counti_01 <- prep_short(0.1) 
-
-counti_025 <- prep_short(0.25) 
-
-counti_05 <- prep_short(0.5)
-
-counti_1 <- prep_short(1)
-
-counti_15 <- prep_short(1.5)
-
-counti_2 <- prep_short(2)
-
-counti_25 <- prep_short(2.5)
-
-counti_3 <- prep_short(3)
-
-counti_5 <- prep_short(5)
-
-vary_bin <- bind_rows(counti_01, 
-                      counti_025, 
-                      counti_05, 
-                      counti_1, 
-                      counti_15, 
-                      counti_2, 
[…]

### 2019-02-23 R/01_inspect_data.r

@@ -1,29 +1,19 @@
-# data manipulation
-library(tidyverse)
+library(pacman)
 
-# parallel processing
-library(parallel)
-
-# bayes
-library(arm)
-library(rstanarm)
-library(bayesplot)
-
-# misc
-library(scales)
-library(ggridges)
-library(pROC)
-library(here)
+p_load(tidyverse, furrr, here,
+       arm, rstanarm, bayesplot,
+       scales, ggridges,
+       pROC)
 
 source(here('R', 'helper01_process_foo.r'))
 source(here('R', 'helper02_plot_foo.r'))
 source(here('R', 'helper03_misc_foo.r'))
 source(here('R', 'helper04_stan_utility.r'))
 
-theme_set(theme_bw())
-
 # important constants
-options(mc.cores = parallel::detectCores())
+theme_set(theme_bw())
+plan(multiprocess)
+
 
 # get data in
 counti <- read_rds(here('data', 'counting.rds'))
@@ -88,10 +78,11 @@ ggsave(filename = here('results', 'figure', 'cramer_temp.png'),
 #  geographic range of species
 vary_width <- read_rds(here('data', 'counting_vary_binwidth.rds'))
 
-
-inplace <- compose(~ .x / length(.x), log)
-
-
+inplace <- function(x) {
+  xx <- log(x)
+  out <- xx / length(xx)
+  out
+}
 
 width_effect <- 
   vary_width %>%

### 2019-02-23 R/06_inspect_predict_model.r

@@ -1,25 +1,7 @@
-# data manipulation
-library(tidyverse)
-library(janitor)
-#devtools::install_github("mjskay/tidybayes")
-library(tidybayes)
-library(deeptime)
+library(pacman)
 
-# parallel processing
-library(parallel)
-library(future)
-library(furrr)
-
-# bayes
-library(arm)
-library(rstanarm)
-library(bayesplot)
-
-# misc
-library(pROC)
-library(ROCR)
-library(ggridges)
-library(here)
+p_load(tidyverse, readr, janitor, tidybayes, deeptime, furrr, arm, rstanarm, 
+       bayesplot, pROC, ROCR, ggridges, here)
 
 # my function set
 source(here('R', 'helper01_process_foo.r'))
@@ -30,7 +12,7 @@ source(here('R', 'helper05_roc_utility.r'))
 source(here('R', 'helper06_geotime.r'))
 
 # important constants
-future::plan(strategy = multicore)
+plan(multiprocess)
 
 # get data in
 counti_fold <- read_rds(here('data', 'counting.rds')) %>%

### 2019-02-23 R/helper02_plot_foo.r

@@ -126,7 +126,7 @@ plot_taxon_covariate_time <- function(disc_best) {
 
   out <- by_taxon %>%
     ggplot(aes(x = age, y = value)) +
-    stat_lineribbon() +
+    stat_lineribbon(size = 0.5) +
     geom_hline(yintercept = 0, linetype = 'dashed', alpha = 0.5) +
     scale_fill_brewer() +
     facet_grid(type ~ key, scales = 'free_y')
@@ -156,7 +156,7 @@ plot_risk_time <- function(data, model, nsp = 4) {
                                 transform = TRUE)
   temp_est <- reshape2::melt(temp_est)
   names(temp_est) <- c('iterations', 'row', 'value')
-  temp_est <- as.tibble(temp_est) %>% 
+  temp_est <- as_tibble(temp_est) %>% 
     mutate(fullname = row,
            relage = row)
   temp_est$fullname <- plyr::mapvalues(temp_est$fullname, 
@@ -230,7 +230,7 @@ plot_taxon_hazard <- function(model) {
     bind_rows %>%
     mutate(effect_prob = invlogit(effect)) %>%  # put on prob scale
     ggplot(aes(x = age, y = effect_prob)) + 
-      stat_lineribbon() +
+      stat_lineribbon(size = 0.5) +
       scale_fill_brewer() +
       facet_grid(type ~ .) +
       labs(x = 'Age (My)', y = 'P(T = t | T >= t, x)')
@@ -303,9 +303,10 @@ plot_roc_series <- function(data, model_pp, model_key) {
     geom_rect(data = rects,
               aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
               fill = 'gray80', alpha = 0.8) +
-    stat_lineribbon(aes(x = key, y = value)) +
+    stat_lineribbon(aes(x = key, y = value), size = 0.5) +
     scale_fill_brewer(aes(x = key, y = value)) + 
     scale_x_reverse() +
+    theme(legend.position = 'bottom') +
     NULL
     
   roc_ts
@@ -338,7 +339,7 @@ view_neptune <- function(.data, name = 'full', path = '../results/figure/') {
     theme(legend.position = 'bottom') +
     labs(title = 'Occurrences', x = 'Time (My before present)', y = 'Count')
 
-  filename <- paste0(path, 'occ_time_label_', name, '.png')
+  filename <- paste0(path, '/occ_time_label_', name, '.png')
[…]

### 2019-02-23 R/misc01_deeptime_error.r

@@ -1,0 +1,23 @@
+library(pacman)
+
+p_load(tidyverse, tidybayes, reprex)
+p_load_gh('willgearty/deeptime')
+
+tb <- 
+  tibble(time = 0:100) %>%
+  group_by_all() %>%
+  do(tibble(value = rnorm(100, .$time, 10),
+            group = sample(1:4, 100, TRUE)))
+
+gg <- 
+  tb %>%
+  ggplot() +
+  stat_lineribbon(aes(x = time, y = value)) +
+  scale_x_reverse() +
+  coord_cartesian(xlim = c(0, 100), 
+                  ylim = range(tb$value),
+                  expand = FALSE) +
+  facet_wrap(~ group) +
+  scale_fill_brewer()
+
+gggeo_scale(gg)

### 2019-02-25 doc/model_descibe.tex
seth comments on manuscript filtered through. supp info. fewer figures. regen figures.

@@
 \begin{document}
-
-\begin{tabular}[ht]{ l l l l }
-  Code & Description & Covariates & Formula \\
-  \hline
-  C & Constant effects, no historical covariates & Geographic range, global temperature & $ event \sim range + temp + (1 | age/phylum) $ \\
-  V & Varying effects, no historical covariates & Geographic range, global temperature  & $ event \sim range + temp + (1 + range + temp | mybin/phylum) + (1 | age/phylum) $ \\
-  CP & Constant effects, historical covariates & Geographic range, change in geographic range, global temperature, previous temperature & $ event \sim range + range\_diff + temp + temp\_lag + (1 | age/phylum) $ \\
-  VP & Varying effects, historical covariates & Geographic range, change in geographic range, global temperature, previous temperature & $ event \sim range + range\_diff + temp + temp\_lag + (1 + range + range\_diff + temp + temp\_lag | mybin/phylum) + (1 | age/phylum) $ \\
-\end{tabular}
 

### 2019-02-25 doc/paper.tex
updating graphs. everything runs headless. table of model desc
seth comments on manuscript filtered through. supp info. fewer figures. regen figures.

@@
 \usepackage[comma,numbers,sort&compress]{natbib}
+\usepackage[margin=1in]{geometry}
 \usepackage{lineno}
 \usepackage{longtable}
-\usepackage{docmute}
 \usepackage{caption, subcaption, multirow, morefloats, rotating}
@@
 \usepackage{hyperref}
+\usepackage{threeparttable}
 
 \frenchspacing
+
+\newcommand{\beginsupplement}{
+  \setcounter{table}{0}
+  \renewcommand{\thetable}{S\arabic{table}}
+  \setcounter{figure}{0}
+  \renewcommand{\thefigure}{S\arabic{figure}}
+  \setcounter{equation}{0}
+  \renewcommand{\theequation}{S\arabic{equation}}
+}
 
@@
   Our results show that our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. 
+  We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance.
   Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
   However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction.
-We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance.
   %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
-This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton.
-The greatest source of variance in survival probability is the timing of observation.
-Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age.
-  We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales.
-The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
+  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
+  %The greatest source of variance in survival probability is the timing of observation. 
+  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
+  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
+  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
   These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
-The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.
+  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.
   These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
@@
 We binned these estimates into 1 My intervals as we did with the fossil occurrences.
-The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig.
-\ref{fig:temp_curve}).
+The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.
 Temperature was then transformed and standardized the in the same manner as geographic range (above).
@@
 
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
+%  \caption{Comparison of initial temperature estimates from \citet{Cramer2011} (goldenrod) versus the binned values used in this analysis (blue).
+The initial values are for every 0.1 My while our bins are defined for every 1 My.}
+%  \label{fig:temp_curve}
+%\end{figure}
+
+\subsection{Model Specifications}
+
+We developed a discrete-time survival model to analyze our data in order to answer our question of how wee can we predict extinction risk at million year time scales.
+We considered four model variations: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but we include none of our historical covariates (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and we include our historical covariates (Model VP).
+See Table \ref{tab:model_def} for further explanation of how these models differ from each other.
+For a complete description of the statistical model used in this analysis, please see Section \ref{sec:model_desc}.
+For a description of how our models were implemented, please see Section \ref{sec:model_est}.
+
+\begin{table}[ht]
+  \caption{Models and their definitions}
+  \begin{threeparttable}
+    {
+      \def\arraystretch{1.5}
+      \begin{tabular}{ l p{3cm} l l }
+        Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
+        \hline
+        C & Constant effects, no historical cov.
+& \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ age\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}})} \\
+        V & Varying effects, no historical cov.
+& \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ phylum) + (1 $|$ age/phylum)} \\
+        CP & Constant effects, historical cov.
+& \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff\tnote{g}\phantom{\textsuperscript{g}} +\\temp + temp\_lag\tnote{h}\phantom{\textsuperscript{h}} +\\(1 $|$ age/phylum)} \\
+        VP & Varying effects, historical cov.
+& \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff +\\temp + temp\_lag +\\(1 + range + range\_diff +\\temp + temp\_lag $|$ phylum) +\\(1 $|$ age/phylum)} \\
+        \hline
+      \end{tabular}
+    }
+    \begin{tablenotes}
+    \item[a] See Equation \ref{eq:model} for full statistical model definition.
+    \item[b] Species observation where 1 if time of last observation, otherwise 0.
+    \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
+    \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
+    \item[e] Species are at observation in millions of years.
+    \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcaeous nannoplankton).
+    \item[g] Change in geographic range since last observation.
+    \item[h] Temperature at previous observation.
+    \end{tablenotes}
+  \end{threeparttable}
+  \label{tab:model_def}
+\end{table}
+
+\subsection{Model adequacy}
+
+We are interested in model adequacy and performance into two contexts: in-sample and out-of-sample predictive performance.
+``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
+``Out-of-sample'' is defined below.
+
+Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}.
+This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives.
+AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.
+AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of any two randomly selected species \citep{Fawcett2006,Mason2002}.
+
+The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time.
+These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.
+
+We are particularly interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data).
+To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation.
+For time-series data, the folds (data partitions) are approximately equal segments of time.
+The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold.
+With 63 time points, each of the five folds represents approximately 13 time points.
+Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
+
+See our code repository LINK for full code details.
+Our code uses ``tidyverse'' tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
+
+\section{Results}
+
+% ROC model comparison
+\begin{figure}[ht]
+  \begin{subfigure}[ht]{0.45\textwidth}
+    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
+    \caption{In-sample}
+    \label{fig:auc_hist}
+  \end{subfigure}
+  \begin{subfigure}[ht]{0.45\textwidth}
+    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
+    \caption{Out-of-sample}
+    \label{fig:fold_auc}
+  \end{subfigure}
+  \caption{In-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) AUC estimates for each of our four models.
+These estimates are calculated from the models posterior predictive distribution or from predictions made to new data, respectively.
+Models with a higher AUC values indicate better performance over models with lower AUC values.
+AUC is bounded between 0.5 and 1.
+See Table \ref{tab:model_def} for a description of each of the four models.}
+  \label{fig:auc_compare}
+\end{figure}
+
+\subsection{Model adequacy}
+
+The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to.
+Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig.
+\ref{fig:auc_hist}).
+The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions.
+Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions.
+All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance.
+It is then hard to conclude that there is one ``best'' model which we can rely upon.
+
+%When the posterior predictive distributions of the in-sample AUC estimates are presented over time, the similarity in adequacy between the models becomes more apparent (Fig.
+\ref{fig:auc_ts}).
+There are few major or obvious differences in model adequacy between the four models.
+%% ROC model comparison time series
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_ts_full}
+%  \caption{Comparison between the posterior predictive AUC estimates for each of the time intervals for each of the four models.
+These estimates are reflections of each model's fit to the various time intervals.
+The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend.
+In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
+%  \label{fig:auc_ts}
+%\end{figure}
+
+%When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig.
+\ref{fig:auc_taxon}).
+While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig.
+\ref{fig:auc_hist}).
+For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_full}
+%  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group.
+These estimates reflect each model's fit to the various taxonomic groups present in this analysis.
+The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals.
+In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
+%  \label{fig:auc_taxon}
+%\end{figure}
+
+For many taxon/model combinations there are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 -- AUC value of 0.5 indicates that the model's predictions are no better than random (Fig.
+\ref{fig:auc_taxon_time}).
+However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the VP model.
+Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the VP model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
 \begin{figure}[ht]
   \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
-  \caption{Comparison of initial temperature estimates from \citet{Cramer2011} (goldenrod) versus the binned values used in this analysis (blue).
-The initial values are for every 0.1 My while our bins are defined for every 1 My.}
-  \label{fig:temp_curve}
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
+  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged over time and by taxonomic group.
+These estimates reflect each model's fit to the various taxonomic groups over time.
+The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend.
+In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.
+See Table \ref{tab:model_def} for a description of each of the four models.}
+  \label{fig:auc_taxon_time}
 \end{figure}
 
-\subsection{Model Specifications}
+\subsection{Cross-validation}
+
+Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data \citep{ESL}.
+This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data.
+These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data \citep{ESL}.
+Additionally, multimodality increases with model complexity (Fig.
+\ref{fig:fold_auc}) -- this makes sense as the more complex models allow for predictor effects to vary with time, allowing for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
+
+Comparison between the posterior predictive distributions of expected out-of-sample AUC (Fig.
+\ref{fig:fold_auc}) reveals a similar range in plausible values for all models as the in-sample AUC posterior predictive distributions (Fig.
+\ref{fig:auc_hist}).
+Interestingly, the differences between the posterior predictive distributions for the models have decreased.
+For example, model VP not clearly better than either models V or CP (Fig.
+\ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig.
+\ref{fig:auc_hist}).
+These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
+However, the shapes of the posterior distributions means interepting from the median values is incorrect -- the models are effectively indistinguishable in their expected out-of-sample AUC values.
+
+Additionally, the quality of expected out-of-sample performance is not great, with average out-of-sample AUC for each of our models estimated to be between 0.7 and 0.8 which is far from perfect.
+This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time.
+However, this expected out-of-sample performance is approximately the same as the in-sample performance results (Fig.
+\ref{fig:auc_hist}), indicating that our models would yield consistent results when generalized to future extinctions.
+
+%When the posterior predictive distribution of expected out-of-sample AUC is presented as a time series, the similarity between the models is even more apparent (Fig.
+\ref{fig:fold_auc_time}).
+While the width of the credible intervals at various time points varies between the models, the overall picture of expected out-of-sample AUC is almost identical when you compare the models -- periods of relatively better or worse performance map identically between the time series.
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time_full}
+%  \caption{Comparison of out-of-sample AUC values calculated for each of the My intervals for each of the four models.
+The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
+This presentation decomposes each of the 12-million year folds (Fig.
+\ref{fig:fold_auc}) into the predictions made for each of the million-year intervals.
+The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
+%  \label{fig:fold_auc_time}
+%\end{figure}
+%We can also compare expected out-of-sample AUC by taxonomic group for each of the models (Fig.
+\ref{fig:fold_auc_taxon}).
+These comparisons reveal a lot about the differences in predictive potential of the taxonomic groups.
+For example, the posterior predictive distributions of out-of-sample AUC for Foraminifera from all four models are approximately identical.
+In contrast, expected out-of-sample AUC for Radiolaria exhibits the same or similar pattern in relative model performance to the pooled comparisons (Fig.
+\ref{fig:fold_auc}).
+Additionally, we can state that our out-of-sample predictions for calcareous nannoplankton and dinoflagellates are not necessarily as precise as our estimates for Foraminifera or even Radiolaria.
+These results indicate that out-of-sample predictions may be easier for some taxonomic groups than others (e.g.
+Foraminifera versus Dinoflagellates.
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_full}
+%  \caption{Comparison of out-of-sample AUC values aggregated by taxonomic group for each of the four models.
+Depicted for each taxon-model combination is an aggregate density of all posterior predictive estimates for each of four folds -- cross-validation estimates are commonly multi-modal as each fold presents its own challenges for prediction.
+Beneath these densities is marked the median estimate along with 50\% and 80\% credible intervals.}
+%  \label{fig:fold_auc_taxon}
+%\end{figure}
+%
+
+We can also compare the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig.
+\ref{fig:fold_auc_taxon_time}).
+For each taxonomic group, the time-series of posterior predictive values for each model are broadly congruent.
+
+In the analysis of the posterior predictive distributions of the in-sample AUC values for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig.
+\ref{fig:auc_taxon_time}).
+This occurrence is generally much rarer for the posterior predictive distribution of out-of-sample AUC values -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random.
+In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
+  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
+The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
+This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals.
+The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.
+See Table \ref{tab:model_def} for a description of each of the four models.}
+  \label{fig:fold_auc_taxon_time}
+\end{figure}
+
+%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
+%
+%
+%% risk estimate compared to change in geo-range
+%\begin{figure}[ht]
+%  \centering
+%  \begin{subfigure}{\textwidth}
+%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
+%    \caption{A}
+%    \label{fig:relrisk_range}
+%  \end{subfigure}
+%
+%  \begin{subfigure}{\textwidth}
+%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
+%    \caption{B}
+%    \label{fig:relrisk_ext}
+%  \end{subfigure}
+%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation.
+Geographic range is assumed observed without error.
+Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
+%  \label{fig:relrisk_compare}
+%\end{figure}
+
+\section{Discussion}
+
+The results of this paper set out our baseline ability to predict relative differences in extinction risk.
+We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations.
+Similarly, these models are expected to correctly rank the extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time.
+A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original).
+The similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
+
+One of the most striking results of this analysis is that the in-sample and out-of-sample comparisons between our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig.
+\ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them generalize to out-of-sample data with a similar degree of success (Fig.
+\ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).
+
+Other noteworthy aspects of our in-sample results are that AUC estimates for our models models differ significantly between models and that all of these estimates are in a narrow range of possible AUC values (Fig.
+\ref{fig:auc_hist}).
+For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig.
+\ref{fig:fold_auc}).
+This means that even though model VP has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.
+
+This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses.
+By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored.
+For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the intercept is greater than 2 or less than -2 because the inverse-logit of those values are close to 0 or 1, respectively \citep{ARM}.
+
+The success of model is partially driven by the size of our dataset and the hierarchical structure of our model.
+Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping.
+Instead of relying on large amounts of ecological information to shape individual differences, our model leverages most of the Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.
+
+The principal reason we were not able to include more biological information in our models is that we lack most any life history or ecological information on most marine micro- and nano-plankton.
+Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}.
+However, not this information does not exist all foram species.
+If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species.
+This means ignoring the majority of occurrence information present in the Neptune database.
+
+This presents an interesting conundrum about how to improve upon our results.
+A simple hypothesis for how to improve upon our results is that if we were to include more biological information in our model, our estimates of species relative extinction risk would be improved.
+However, if we decrease the amount of data in our model, our results by definition decrease in their quality.
+For example, compare the results from our models fit using full in-sample dataset, and the results from the cross-validation where the fit to each fold is by definition based on less than full information.)
+
+In conclusion, our results provide a promising picture of our ability to predict the relative extinction risk of two randomly selected species.
+Considering that conservation decisions are made based on a continuum of risk, from most to least, this means that our results are in the same language as how conservation resources are allocated.
+
+% bibliographic information
+\bibliographystyle{abbrvnat}
+\bibliography{citations}
+
+\section{Supplementary Material}
+\beginsupplement
+
+\subsection{Model Specifications} \label{sec:model_desc}
 
@@
     \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
-    \tau_{B} &\sim C^{+}(5) \\
-    \tau_{\alpha} &\sim C^{+}(5) \\
-    \tau_{A} &\sim C^{+}(5) \\
+    \tau_{B} &\sim C^{+}(1) \\
+    \tau_{\alpha} &\sim C^{+}(1) \\
+    \tau_{A} &\sim C^{+}(1) \\
     \Omega_{B} &\sim LKJ(1) \\
@@
   \end{aligned}
-  \label{eq:priors}
+  \label{eq:model}
 \end{equation}
@@
 
-We considered four variants of the model described above: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect).
-The second and fourth models modify the number of columns in \(X\) by removing two of the covariates.
-The third and fourth model simplify the first model to just varying intercepts without varying slopes.
+\subsection{Model Parameter Estimation} \label{sec:model_est}
 
-\subsection{Model Parameter Estimation}
-
-We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}.
+We implemented our model (Eq.
+\ref{eq:model} using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}.
 This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R.
@@
 
-%An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
-%\begin{verbatim}
-%form <- event ~ range + range_diff + temp + temp_lag + 
-%                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
-%                (1 | age/phylum), 
-%stan_glmer(formula = form,
-%           data = data, 
-%           family = 'binomial',
-%           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
-%           prior_intercept = normal(0, 10, autoscale = FALSE), 
-%           prior_aux = cauchy(0, 3, autoscale = FALSE), 
-%           chains = 4,
-%           thin = 4,
-%           adapt_delta = 0.999999)
-%\end{verbatim}
-%
-%Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} \citep{brms2017,brms2018} as:
-%\begin{verbatim}
-%priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
-%            set_prior('normal(0, 1)', class = 'b'),
-%            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
-%            set_prior('cauchy(0, 3)', class = 'sd'),
-%            set_prior('lkj(1)', class = 'cor'))
-%form <- bf(event ~ range + range_diff + temp + temp_lag +
-%           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
-%           (1 | age/phylum))
-%brmfit <- brm(formula = form,
-%              data = data, 
-%              family = bernoulli(), 
-%              prior = priors,
-%              chains = 4, 
-%              thin = 4,
-%              control = list(adapt_delta = 0.999999)
-%\end{verbatim}
+An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
+\begin{verbatim}
+form <- event ~ range + range_diff + temp + temp_lag + 
+                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
+                (1 | age/phylum), 
+stan_glmer(formula = form,
+           data = data, 
+           family = 'binomial',
+           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
+           prior_intercept = normal(-2, 5, autoscale = FALSE), 
+           prior_aux = cauchy(0, 1, autoscale = FALSE), 
+           chains = 4,
+           thin = 4,
+           adapt_delta = 0.999999)
+\end{verbatim}
+
+Similarly, our full model can also be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
+\begin{verbatim}
+priors <- c(set_prior('normal(-2, 5)', class= 'Intercept'),
+            set_prior('normal(0, 1)', class = 'b'),
+            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
+            set_prior('cauchy(0, 1)', class = 'sd'),
+            set_prior('lkj(1)', class = 'cor'))
+form <- bf(event ~ range + range_diff + temp + temp_lag +
+           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
+           (1 | age/phylum))
+brmfit <- brm(formula = form,
+              data = data, 
+              family = bernoulli(), 
+              prior = priors,
+              chains = 4, 
+              thin = 4,
+              control = list(adapt_delta = 0.999999)
+\end{verbatim}
 
@@
 
-\subsection{Model adequacy}
-
-We are interested in model adequacy and performance into two contexts: in-sample and out-of-sample predictive performance.
-``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
-``Out-of-sample'' is defined below.
-
-Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}.
-This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives.
-AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.
-AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of any two randomly selected species \citep{Fawcett2006,Mason2002}.
-
-The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time.
-These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.
-
-We are particularly interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data).
-To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation.
-For time-series data, the folds (data partitions) are approximately equal segments of time.
-The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold.
-With 63 time points, each of the five folds represents approximately 13 time points.
-Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
-
-See our code repository LINK for full code details.
-Our code uses ``tidyverse'' tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
-
-\section{Results}
-
-\subsection{Model adequacy}
-
-The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to.
-Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig.
-\ref{fig:auc_hist}).
-The parameter rich ``past and vary'' model has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions.
-Additionally, while our parameter rich ``past and vary'' model is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions.
-All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance.
-It is then hard to conclude that there is one ``best'' model which we can rely upon.
-
-% ROC model comparison
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist}
-  \caption{Posterior predictive AUC estimates for each of the four models being compared.
-These estimates are calculated from each of the models posterior predictive distribution compared to the empirical values.
-Models with a higher AUC values indicate better performance over models with lower AUC values.
-AUC is bounded between 0.5 and 1.}
-  \label{fig:auc_hist}
-\end{figure}
-
-When the posterior predictive distributions of the in-sample AUC estimates are presented over time, the similarity in adequacy between the models becomes more apparent (Fig.
-\ref{fig:auc_ts}).
-There are few major or obvious differences in model adequacy between the four models.
-% ROC model comparison time series
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_ts}
-  \caption{Comparison between the posterior predictive AUC estimates for each of the time intervals for each of the four models.
-These estimates are reflections of each model's fit to the various time intervals.
-The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend.
-In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
-  \label{fig:auc_ts}
-\end{figure}
-
-When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig.
-\ref{fig:auc_taxon}).
-While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig.
-\ref{fig:auc_hist}).
-For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon}
-  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group.
-These estimates reflect each model's fit to the various taxonomic groups present in this analysis.
-The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals.
-In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
-  \label{fig:auc_taxon}
-\end{figure}
-
-For many taxon/model combinations there are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 -- AUC value of 0.5 indicates that the model's predictions are no better than random (Fig.
-\ref{fig:auc_taxon_time}).
-However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the ``past and vary'' model.
-Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the ``past and vary'' model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time}
-  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged over time and by taxonomic group.
-These estimates reflect each model's fit to the various taxonomic groups over time.
-The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend.
-In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
-  \label{fig:auc_taxon_time}
-\end{figure}
-
-\subsection{Cross-validation}
-
-Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data \citep{ESL}.
-This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data.
-These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data \citep{ESL}.
-Additionally, multimodality increases with model complexity (Fig.
-\ref{fig:fold_auc}) -- this makes sense as the more complex models allow for predictor effects to vary with time, allowing for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
-
-Comparison between the posterior predictive distributions of expected out-of-sample AUC (Fig.
-\ref{fig:fold_auc}) reveals a similar range in plausible values for all models as the in-sample AUC posterior predictive distributions (Fig.
-\ref{fig:auc_hist}).
-Interestingly, the differences between the posterior predictive distributions for the models have decreased.
-For example, the ``past and vary'' model not clearly better than either of the ``no past or vary'' and ``past and no vary'' models (Fig.
-\ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig.
-\ref{fig:auc_hist}).
-These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
-However, the shapes of the posterior distributions means interepting from the median values is incorrect -- the models are effectively indistinguishable in their expected out-of-sample AUC values.
-
-Additionally, the quality of expected out-of-sample performance is not great, with average out-of-sample AUC for each of our models estimated to be between 0.7 and 0.8 which is far from perfect.
-This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time.
-However, this expected out-of-sample performance is approximately the same as the in-sample performance results (Fig.
-\ref{fig:auc_hist}), indicating that our models would yield consistent results when generalized to future extinctions.
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc}
-  \caption{Results from our five-fold cross-validation of the time-series.
-Each labeled distribution of AUC values correspond to expected out-of-sample performance as estimated from that fold.
-Each fold represents a section of data being predicted from a model fit to all data before the start of that fold.
-Given that there are only five folds, performance is measured from predictions for four of the folds.}
-  \label{fig:fold_auc}
-\end{figure}
-
-When the posterior predictive distribution of expected out-of-sample AUC is presented as a time series, the similarity between the models is even more apparent (Fig.
-\ref{fig:fold_auc_time}).
-While the width of the credible intervals at various time points varies between the models, the overall picture of expected out-of-sample AUC is almost identical when you compare the models -- periods of relatively better or worse performance map identically between the time series.
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
-  \caption{Comparison of out-of-sample AUC values calculated for each of the My intervals for each of the four models.
-The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
-This presentation decomposes each of the 12-million year folds (Fig.
-\ref{fig:fold_auc}) into the predictions made for each of the million-year intervals.
-The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
-  \label{fig:fold_auc_time}
-\end{figure}
-
-We can also compare expected out-of-sample AUC by taxonomic group for each of the models (Fig.
-\ref{fig:fold_auc_taxon}).
-These comparisons reveal a lot about the differences in predictive potential of the taxonomic groups.
-For example, the posterior predictive distributions of out-of-sample AUC for Foraminifera from all four models are approximately identical.
-In contrast, expected out-of-sample AUC for Radiolaria exhibits the same or similar pattern in relative model performance to the pooled comparisons (Fig.
-\ref{fig:fold_auc}).
-Additionally, we can state that our out-of-sample predictions for calcareous nannoplankton and dinoflagellates are not necessarily as precise as our estimates for Foraminifera or even Radiolaria.
-These results indicate that out-of-sample predictions may be easier for some taxonomic groups than others (e.g.
-Foraminifera versus Dinoflagellates.
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon}
-  \caption{Comparison of out-of-sample AUC values aggregated by taxonomic group for each of the four models.
-Depicted for each taxon-model combination is an aggregate density of all posterior predictive estimates for each of four folds -- cross-validation estimates are commonly multi-modal as each fold presents its own challenges for prediction.
-Beneath these densities is marked the median estimate along with 50\% and 80\% credible intervals.}
-  \label{fig:fold_auc_taxon}
-\end{figure}
-
-Finally, we can present the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig.
-\ref{fig:fold_auc_taxon_time}).
-For each taxonomic group, the time-series of posterior predictive values for each model are broadly congruent.
-
-In the analysis of the posterior predictive distributions of the in-sample AUC values for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig.
-\ref{fig:auc_taxon_time}).
-This occurrence is generally much rarer for the posterior predictive distribution of out-of-sample AUC values -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random.
-In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the ``no past or vary'' model.
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time}
-  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
-The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
-This presentation decomposes each of the 12-million year folds by each of the taxonomic groups (Fig.
-\ref{fig:fold_auc_taxon}) into the predictions made for each of the million-year intervals.
-The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
-  \label{fig:fold_auc_taxon_time}
-\end{figure}
-
-%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
-%
-%
-%% risk estimate compared to change in geo-range
-%\begin{figure}[ht]
-%  \centering
-%  \begin{subfigure}{\textwidth}
-%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
-%    \caption{A}
-%    \label{fig:relrisk_range}
-%  \end{subfigure}
-%
-%  \begin{subfigure}{\textwidth}
-%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
-%    \caption{B}
-%    \label{fig:relrisk_ext}
-%  \end{subfigure}
-%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation.
-Geographic range is assumed observed without error.
-Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
-%  \label{fig:relrisk_compare}
-%\end{figure}
-
-\section{Discussion}
-
-The results of this paper set out our baseline ability to predict relative differences in extinction risk.
-We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations.
-Similarly, these models are expected to correctly rank the extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time.
-A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original).
-The similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
-
-One of the most striking results of this analysis is that the in-sample and out-of-sample comparisons between our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig.
-\ref{fig:auc_hist}, \ref{fig:auc_ts}, \ref{fig:auc_taxon}, \ref{fig:auc_taxon_time}), all of them generalize to out-of-sample data with a similar degree of success (Fig.
-\ref{fig:fold_auc}, \ref{fig:fold_auc_time}, \ref{fig:fold_auc_taxon}).
-
-Other noteworthy aspects of our in-sample results are that AUC estimates for our models models differ significantly between models and that all of these estimates are in a narrow range of possible AUC values (Fig.
-\ref{fig:auc_hist}).
-For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig.
-\ref{fig:fold_auc}).
-This means that even though our ``past and vary'' model has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.
-
-This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses.
-By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored.
-For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the intercept is greater than 2 or less than -2 because the inverse-logit of those values are close to 0 or 1, respectively \citep{ARM}.
-
-The success of model is partially driven by the size of our dataset and the hierarchical structure of our model.
-Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping.
-Instead of relying on large amounts of ecological information to shape individual differences, our model leverages most of the Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.
-
-The principal reason we were not able to include more biological information in our models is that we lack most any life history or ecological information on most marine micro- and nano-plankton.
-Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}.
-However, not this information does not exist all foram species.
-If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species.
-This means ignoring the majority of occurrence information present in the Neptune database.
-
-This presents an interesting conundrum about how to improve upon our results.
-A simple hypothesis for how to improve upon our results is that if we were to include more biological information in our model, our estimates of species relative extinction risk would be improved.
-However, if we decrease the amount of data in our model, our results by definition decrease in their quality.
-For example, compare the results from our models fit using full in-sample dataset, and the results from the cross-validation where the fit to each fold is by definition based on less than full information.)
-
-In conclusion, our results provide a promising picture of our ability to predict the relative extinction risk of two randomly selected species.
-Considering that conservation decisions are made based on a continuum of risk, from most to least, this means that our results are in the same language as how conservation resources are allocated.
-
-% bibliographic information
-\bibliographystyle{abbrvnat}
-\bibliography{citations}
-
 \end{document}

### 2019-02-25 R/06_inspect_predict_model.r

@@ -38,10 +38,11 @@ counti_fold_match_rt <- rep(counti_fold_rt[-1], 4)
 counti_fold_match_rl <- rep(counti_fold_rl[-1], 4)
 
 # get in fits and posterior work
-model_key <- c('Past and vary', 
-               'Past but no vary',
-               'No past but vary', 
-               'No past or vary')
+#model_key <- c('Past and vary', 
+#               'Past but no vary',
+#               'No past but vary', 
+#               'No past or vary')
+model_key <- c('VP', 'CP', 'V', 'C')
 fit <- read_rds(here('results', 'training_fit.rds'))
 fit_rt <- read_rds(here('results', 'training_fit_rt.rds'))
 fit_rl <- read_rds(here('results', 'training_fit_rl.rds'))

### 2019-02-25 R/helper02_plot_foo.r

@@ -491,7 +491,7 @@ p_model <- function(fit_list, .data, key, name, path) {
            model_name = factor(model_name, levels = key)) %>%
     ggplot(aes(x = value, y = model_name)) +
     geom_halfeyeh(.width = c(0.5, 0.8)) +
-    labs(x = 'AUC ROC', y = NULL) +
+    labs(x = 'AUC', y = NULL) +
     theme(axis.text.y = element_text(size = 15),
           axis.text.x = element_text(size = 15)) +
     NULL
@@ -542,7 +542,7 @@ p_model_time <- function(fit_list, .data, key, name, path) {
   # facet
   roc_ts <- roc_ts +
     facet_grid(model ~ .) +
-    labs(y = 'AUC ROC', x = 'Time (Mya)') +
+    labs(y = 'AUC', x = 'Time (Mya)') +
     geom_hline(yintercept = 0.5, colour = 'red', linetype = 'dashed') +
     NULL
 
@@ -594,7 +594,7 @@ p_model_taxon <- function(fit_list, .data, key, name, path) {
     ggplot(aes(x = value, y = model)) +
     geom_halfeyeh(.width = c(0.5, 0.8)) +
     facet_wrap(~ taxon) +
-    labs(x = 'AUC ROC', y = NULL) 
+    labs(x = 'AUC', y = NULL) 
   fn <- paste0(path, '/auc_taxon_', name, '.png')
   ggsave(filename = fn,
          plot = auc_taxon,
@@ -662,10 +662,11 @@ p_model_taxon_time <- function(fit_list, .data, key, name, path) {
               fill = 'gray80', alpha = 0.8) +
     geom_hline(yintercept = 0.5, linetype = 'dashed') +
     stat_lineribbon(aes(x = time, y = value), size = 0.5) + 
-    scale_fill_brewer(aes(x = time, y = value)) + 
+    scale_fill_brewer() +
     scale_x_reverse() +
     facet_grid(fossil_group ~ model) +
     theme(legend.position = 'bottom') +
+    labs(x = 'Time (Mya)', y = 'AUC') +
     NULL
   fn <- paste0(path, '/auc_taxon_time_', name, '.png')
   ggsave(filename = fn,
@@ -719,7 +720,7 @@ cv_model <- function(fit, .data, key, name, path) {
            mod = factor(mod, levels = key)) %>%
     ggplot(aes(x = value, y = mod)) +
[…]

### 2019-03-07 doc/paper.tex
pandoc-d some alternate formats for paper
seth comments entered

@@
 \begin{abstract}
-  One of the great promises of paleobiology is that by studying the past we can better predict the future.
-This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
-Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)?
-(2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?
-  To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). 
-  We analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature.
+  One of the central tenants of conservation paleobiology is that by studying the past we can better predict the future.
+It has been suggested that risk assessments for extant species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
+Such suggestions prompt two key questions: (1) At a given timescale, to what extent are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)?
+(2) At a given point in time, how accurate do extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns  turn out to be?
+To address these questions we analyze the exceptionally well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
+We examine how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, climate state, and the change in climate state.
+Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species.
+In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic.
+Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor.
+Our results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
+However, models based on present geographic ranges and past range-risk relationships are likely to be modernly successful at predicting future extinction/survival outcomes.
+AN important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
+
   %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.
-  Our results show that our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. 
-  We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance.
-  Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
-  However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction.
+  %We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance.
+  %Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
+  %However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction.
   %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
@@
   %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
-  These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
+  %These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
   %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.
-  These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
-  The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
+  %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
+  %The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
 \end{abstract}
@@
 
-Being able to predict which species are more likely to go extinct than others is critical for making good conservation decisions to limit the impact of the current biodiversity crisis.
+Being able to predict which species are more likely than others to go extinct over a given time frame is crucial for planning decisions policies to ameliorate the current biodiversity crisis.
+Some species have been the focus of intense study to define short-term (e.g.
+decadal to centennial scale) abundance and geographic range trajectories.
+
 We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable.
@@
 
-For this kind of exercise, we chose to analyze what is the longest continuous and best resolved fossil record -- that of skeletonized marine plantonic microorganisms from the Cenozoic such as Foraminifera, Radiolarians, Diatoms, and calcareous nannofossils (e.g.
-coccolithophores).
-This data is available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}.
+For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms F(oraminifera, Radiolaria, Diatoms, and Coccolithophores).
+These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}.
 This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.
 
-Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale.
+Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale.
 Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales.
@@
 This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya.
-This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction (discussed below).
+This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction as discussed below.
 
@@
 Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e.
-the Earth) between any two occurrences of that species; this measure is also called a geodesic.
-This distance was measured in kilometers.
-Geographic range was then log-plus-one transformed, and standardized by zero-centering the data and then dividing by the standard deviation of the distribution of geographic ranges.
-This standardization means that the analyzed values of geographic range had mean 0 and standard deviation 1.
-This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the interecpt term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}.
-Change in geographic range between observations was measured from the transformed and standardized geographic range values and not standardized seperately.
+the Earth) between any two occurrences of that species; this distance was measured in kilometers.
+Geographic range was then log-plus-one transformed, and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges.
+This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}.
+Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.
 
@@
 The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.
-Temperature was then transformed and standardized the in the same manner as geographic range (above).
-Difference in temperature between between observations was measured from the transformed and standardized temperature values and not standardized seperately.
+Temperature was then transformed and standardized the in the same manner as geographic range.
+Difference in temperature between between observations was measured from the standardized temperature values and was not standardized separately.
 
@@
 All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance.
-It is then hard to conclude that there is one ``best'' model which we can rely upon.
+It is therefore hard to conclude that there is one ``best'' model which we can rely upon.
 
@@
 \begin{equation}
-  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k. \\
+  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k.
   \label{eq:hazard}

### 2019-03-07 doc/paper.tex
citation on auc quality. trying to shorten paper

@@
 AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of any two randomly selected species \citep{Fawcett2006,Mason2002}.
+AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, which means that values between 0.7 and 0.8 could then be considered ``fair.''
 
-The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time.
+%The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time.
 These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.
@@
 Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions.
-All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance.
-It is therefore hard to conclude that there is one ``best'' model which we can rely upon.
+All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77.
+It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.
 
@@
 
-Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data \citep{ESL}.
+Expected out-of-sample predictive performance was estimated using five-fold cross-validation \citep{ESL}.
 This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data.
-These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data \citep{ESL}.
-Additionally, multimodality increases with model complexity (Fig.
-\ref{fig:fold_auc}) -- this makes sense as the more complex models allow for predictor effects to vary with time, allowing for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
+The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}.
+We find the visual degree of this multimodality increases with model complexity (Fig.
+\ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time which allows for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
 
@@
 
-Additionally, the quality of expected out-of-sample performance is not great, with average out-of-sample AUC for each of our models estimated to be between 0.7 and 0.8 which is far from perfect.
-This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time.
+%This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time.
 However, this expected out-of-sample performance is approximately the same as the in-sample performance results (Fig.
@@
 
-This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses.
+%This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses.
 By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored.
@@
 
-The success of model is partially driven by the size of our dataset and the hierarchical structure of our model.
+The success of our models is partially driven by the size of our dataset and the hierarchical structure of our model.
 Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping.

### 2019-03-08 doc/paper.tex
moved to biber. moved more to supplement

@@
 \usepackage{microtype, parskip, graphicx}
-\usepackage[comma,numbers,sort&compress]{natbib}
+%\usepackage[comma,numbers,sort&compress]{natbib}
+\usepackage[style=authoryear-icomp,
+            backend=biber,
+            bibencoding=utf8,
+            natbib=true,
+            url=false,
+            doi=false,
+            isbn=false]{biblatex}
 \usepackage[margin=1in]{geometry}
@@
 }
+
+% bibliographic information
+%\bibliographystyle{abbrvnat}
+%\bibliography{citations}
+\addbibresource{citations.bib}
 
@@
 Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species.
-In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic.
 Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor.
@@
 AN important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
-
+  %In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic.
   %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.
@@
 
-By studying how species vary in their extinction risk over time and we can assess which species are at greater risk under unobserved conditions.
+Studying how species vary in their extinction risk over time means we can assess which species are at greater risk under unobserved conditions.
 We know that a species' risk of extinction varies over time in both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}.
@@
 While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}).
-For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years?
-Here, we explicitly model and quantify the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction.
-Similarly, we include species geologic age at time of observation as a potential predictor of extinction -- a factor that may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}.
+%For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years? 
+Here, we explicitly estimate the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction.
 Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.
@@
 This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.
-
-Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale.
+%Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale.
 Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales.
@@
 
-Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts.
+%Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts.
 Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition \citep{Zachos2008} -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions.
@@
 All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence. 
-Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources.
-The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species.
-For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0.
-The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1.
-This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}.
-However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}.
-In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.
-
-A taxon's geographic range was calculated for each of the 1 My bins in which it occurred.
+After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred.
 Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e.
 the Earth) between any two occurrences of that species; this distance was measured in kilometers.
-Geographic range was then log-plus-one transformed, and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges.
-This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}.
-Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.
 
@@
 The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.
-Temperature was then transformed and standardized the in the same manner as geographic range.
-Difference in temperature between between observations was measured from the standardized temperature values and was not standardized separately.
+
+See Section \ref{sec:data_desc} for a further explanation on how observations were temporally binned and how our covariates were standardized and transformed prior to analysis.
 
@@
 For a description of how our models were implemented, please see Section \ref{sec:model_est}.
+
+Importantly, we include species geologic age at time of observation as a non-nested varying intercept term.
+This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.
+Again, see Section \ref{sec:model_desc} for how this term was specified as part of our model.
 
@@
 
-See our code repository LINK for full code details.
+See our code repository https://github.com/psmits/trident for full code details.
 Our code uses ``tidyverse'' tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
@@
 
-This presents an interesting conundrum about how to improve upon our results.
+%This presents an interesting conundrum about how to improve upon our results.
 A simple hypothesis for how to improve upon our results is that if we were to include more biological information in our model, our estimates of species relative extinction risk would be improved.
@@
 
-% bibliographic information
-\bibliographystyle{abbrvnat}
-\bibliography{citations}
+\printbibliography
 
@@
 \beginsupplement
+
+\subsection{Data Specifications} \label{sec:data_desc}
+
+\subsubsection{Binning fossil occurrences}
+
+Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources.
+The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species.
+For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0.
+The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1.
+This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}.
+However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}.
+In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.
+
+\subsubsection{Covariate transforms and standardizations} \label{sec:data_trans}
+
+Prior to analysis, geographic range was then log-plus-one transformed and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges.
+This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}.
+Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.
+
+Temperature was also transformed and standardized the in the same manner as geographic range.
+Difference in temperature between between observations was measured from the standardized temperature values and was not standardized separately.
 

### 2019-03-08 doc/paper.tex
messing with citation format to make it more proc b like

@@
 %\usepackage[comma,numbers,sort&compress]{natbib}
-\usepackage[style=authoryear-icomp,
+\usepackage[style=numeric-comp,
             backend=biber,
             bibencoding=utf8,
+            %maxcitenames=2,
+            uniquename=init,
+            giveninits=true,
+            sorting=none,
             natbib=true,
@@
 \newcommand{\beginsupplement}{
+  \setcounter{section}{0}
+  \renewcommand{\thesection}{S\arabic{section}}
   \setcounter{table}{0}
@@
 However, models based on present geographic ranges and past range-risk relationships are likely to be modernly successful at predicting future extinction/survival outcomes.
-AN important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
+An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
   %In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic.
@@
 We approach this this problem by analyzing the past in order to predict the future.
-The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012,Finnegan2015}.
+The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012a,Finnegan2015}.
 By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions -- we ask the probability that, given two random species, we correctly rank their relative risks of extinction.
@@
 By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past.
-Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
+Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
 This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
@@
 Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales.
-Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years.
-The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record.
-By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.
+%Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. 
+The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores from which the occurrences are recorded.
+Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.
 
@@
 We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
-All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
+All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
 This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya.
@@
 
+\beginsupplement
 \section{Supplementary Material}
-\beginsupplement
 

### 2019-03-08 doc/paper.tex
seperate main text and supplement references
cleaning supple ref
cleaning up citation formatting

@@
             bibencoding=utf8,
-            %maxcitenames=2,
+            maxbibnames=10,
             uniquename=init,
@@
 
+% https://3d.bk.tudelft.nl/hledoux/blog/fiddling-biblatex/
+%-- formatting hell for biblatex
+%-- remove "In:"
+\renewbibmacro{in:}{}
+%-- no "quotes" around titles of chapters/article titles
+\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]
+{title}{#1}
+%-- no punctuation after volume
+\DeclareFieldFormat[article]{volume}{ {#1} } 
+%-- puts number/issue between brackets
+\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]{number}{\mkbibparens{#1}} 
+%-- and then for articles directly the pages w/o any "pages" or "pp." 
+\DeclareFieldFormat[article]{pages}{#1}
+%-- for some types replace "pages" by "p."
+\DeclareFieldFormat[inproceedings, incollection, inbook]{pages}{p. #1}
+
+% https://tex.stackexchange.com/questions/81569/biblatex-parentheses-around-the-volume-number-of-an-article
+\renewbibmacro*{volume+number+eid}{%
+  \printfield{volume}%
+  \printfield{number}%
+  \printfield{eid}}
+\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
+
 % bibliographic information
@@
 }
+\date{}
 
 \begin{document}
+
+\begin{refsection}
 
@@
 \end{abstract}
+
+{\bf Keywords:} conservation, paleobiology, extinction
 
@@
 We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
-All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
+All occurrence information was downloaded for calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
 This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya.
@@
     \item[e] Species are at observation in millions of years.
-    \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcaeous nannoplankton).
+    \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcareous nannoplankton).
     \item[g] Change in geographic range since last observation.
@@
 
-We are interested in model adequacy and performance into two contexts: in-sample and out-of-sample predictive performance.
-``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
-``Out-of-sample'' is defined below.
+We are interested in model adequacy into two contexts: in-sample and out-of-sample predictive performance.
+In-sample means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
+Out-of-sample is defined in context below.
+
+We are particularly interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data).
+To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation.
+For time-series data, the folds (data partitions) are approximately equal segments of time.
+The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold.
+With 63 time points, each of the five folds represents approximately 13 time points.
+Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
 
@@
 
-We are particularly interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data).
-To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation.
-For time-series data, the folds (data partitions) are approximately equal segments of time.
-The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold.
-With 63 time points, each of the five folds represents approximately 13 time points.
-Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
-
-See our code repository https://github.com/psmits/trident for full code details.
-Our code uses ``tidyverse'' tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
+See our code repository at https://github.com/psmits/trident for full code details.
+The entire analysis was coded in R and uses tidyverse-style tools such as \texttt{dplyr} \citep{dplyr}, and \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}.
+Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.
 
@@
 \printbibliography
+\end{refsection}
+\clearpage
 
 \beginsupplement
+\begin{refsection}
 \section{Supplementary Material}
@@
 
+\printbibliography[title={Supplementary References}]
+\end{refsection}
+
 \end{document}

### 2019-03-08 doc/paper.tex
kept hacking away at abstract

@@
 \begin{abstract}
-  One of the central tenants of conservation paleobiology is that by studying the past we can better predict the future.
-It has been suggested that risk assessments for extant species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
-Such suggestions prompt two key questions: (1) At a given timescale, to what extent are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)?
-(2) At a given point in time, how accurate do extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns  turn out to be?
-To address these questions we analyze the exceptionally well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
-We examine how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, climate state, and the change in climate state.
+  One of the central tenants of conservation paleobiology is that risk assessments for extant species can be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
+  We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
+  To address these questions we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
+We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, recent change in geographic range, climate state, and the change in climate state.
 Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species.
 Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor.
-Our results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
-However, models based on present geographic ranges and past range-risk relationships are likely to be modernly successful at predicting future extinction/survival outcomes.
 An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
+  %Such suggestions prompt two key questions: (1) At a given timescale, to what extent are extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)?
+(2)
+  %Our results imply that at million-year timescales extinction risk trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
+  %However, models based on present geographic ranges and past range-risk relationships are likely to be moderately successful at predicting future extinction/survival outcomes. 
   %In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic.

### 2019-03-12 doc/paper.tex
added in seth comments round 2. made out of sample all about forecasting to make it clearer to audience

@@
 \begin{abstract}
-  One of the central tenants of conservation paleobiology is that risk assessments for extant species can be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
-  We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
-  To address these questions we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
-We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, recent change in geographic range, climate state, and the change in climate state.
-Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species.
+  A tenants of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions.
+Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past.
+To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) For each group we examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
 Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor.
+This model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species.
+It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time.
 An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
+
+  %risk assessments for extant species can be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
+  %We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
   %Such suggestions prompt two key questions: (1) At a given timescale, to what extent are extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)?
@@
 This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
-For example, while a predictor may be ``significant'' when comparing the odds of extinction risk between groups, the practical difference that predictor makes on prediction can be minimal \citep{ARM}.
+
+%For example, while a predictor may be ``significant'' when comparing the odds of extinction risk between groups, the practical difference that predictor makes on prediction can be minimal \citep{ARM}.
 By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.
@@
 %For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years?
-Here, we explicitly estimate the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction.
+
+%Here, we explicitly estimate the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction.
 Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.
@@
 
-\subsection{Model adequacy}
+\subsection{In-sample and out-of-sample forecasting}
 
-We are interested in model adequacy into two contexts: in-sample and out-of-sample predictive performance.
-In-sample means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
-Out-of-sample is defined in context below.
+We are interested in model performance into two contexts: in-sample and out-of-sample predictive performance i.e. forecasting.
 
-We are particularly interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data).
-To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation.
+In-sample forecasting means we are estimating how well our model predicts extinction probability for observations that that model was fit to.
+This is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
+
+We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data).
+These in-sample measures, however, are not necessarily good estimates of the model's ability to predict data from the future.
+To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation.
 For time-series data, the folds (data partitions) are approximately equal segments of time.
-The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold.
+Each fold represents a sequence of time points.
 With 63 time points, each of the five folds represents approximately 13 time points.
 Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
+
+Here is a description of how k-fold cross-validation for time series precedes \citep{Arlot2010,Bergmeir2016}.
+The model is fit first to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e.
+the future).
+Then the model is fit to the first and second golds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold.
+This process continues with the model fit to the first three folds used to forecast extinction probabilities for the fourth fold, and then the model fit to the first four folds being used to forecast forecast the fifth fold.
 
@@
 
-\subsection{Model adequacy}
+\subsection{In-sample forecasting adequacy}
 
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
-  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged over time and by taxonomic group.
+  \caption{Comparison of in-sample forecasting performance measured by AUC for each of the four models, arranged over time and by taxonomic group.
 These estimates reflect each model's fit to the various taxonomic groups over time.
@@
 
-\subsection{Cross-validation}
+\subsection{Out-of-sample forecasting performance}
 
-Expected out-of-sample predictive performance was estimated using five-fold cross-validation \citep{ESL}.
-This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data.
+Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}.
 The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}.
@@
 
-Comparison between the posterior predictive distributions of expected out-of-sample AUC (Fig.
-\ref{fig:fold_auc}) reveals a similar range in plausible values for all models as the in-sample AUC posterior predictive distributions (Fig.
-\ref{fig:auc_hist}).
-Interestingly, the differences between the posterior predictive distributions for the models have decreased.
+Comparison between the in-sample forecasting performance distributions of expected out-of-sample AUC reveals a similar range in plausible values for all models (Fig.
+\ref{fig:auc_hist}, \ref{fig:fold_auc}).
+Interestingly, the differences in the posterior predictive distributions of AUC for the models have decreased.
 For example, model VP not clearly better than either models V or CP (Fig.
@@
 These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
-However, the shapes of the posterior distributions means interepting from the median values is incorrect -- the models are effectively indistinguishable in their expected out-of-sample AUC values.
+However, the shapes of the posterior distributions means interpretations based only on the median values is incorrect -- the models are effectively indistinguishable in their out-of-sample forecast performance.
 
@@
 \ref{fig:fold_auc_taxon_time}).
-For each taxonomic group, the time-series of posterior predictive values for each model are broadly congruent.
+The time-series of posterior predictive values for each model are broadly congruent between each taxonomic group.
 
-In the analysis of the posterior predictive distributions of the in-sample AUC values for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig.
+In the analysis of the posterior predictive distributions of AUC from the in-sample forecasts for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig.
 \ref{fig:auc_taxon_time}).
-This occurrence is generally much rarer for the posterior predictive distribution of out-of-sample AUC values -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random.
+This occurrence is generally much rarer for the posterior predictive distribution of AUC from the out-of-sample forecasts -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random.
 In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
@@
 We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations.
-Similarly, these models are expected to correctly rank the extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time.
+Similarly, these models are expected to correctly forecast the rank order in extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time.
 A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original).
-The similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
+Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past.
+However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
 
-One of the most striking results of this analysis is that the in-sample and out-of-sample comparisons between our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig.
-\ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them generalize to out-of-sample data with a similar degree of success (Fig.
+One of the most striking results of this analysis is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig.
+\ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig.
 \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).
 
-Other noteworthy aspects of our in-sample results are that AUC estimates for our models models differ significantly between models and that all of these estimates are in a narrow range of possible AUC values (Fig.
+Other noteworthy aspects of the results of our in-sample forecasts are that although AUC estimates for our models models differ significantly between models, all of these estimates are in a narrow range of possible AUC values (Fig.
 \ref{fig:auc_hist}).
@@
 \begin{refsection}
-\section{Supplementary Material}
+
+%\section{Supplemental Figures}
+
+\section{Supplement to Materials and Methods}
 
@@
 
-\subsubsection{Covariate transforms and standardizations} \label{sec:data_trans}
+\subsubsection{Covariate transforms and standardizations}
 

### 2019-03-13 doc/paper.tex
new figure showing congruence between ins and oos

@@
 \begin{abstract}
-  A tenants of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions.
+  A tenant of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions.
 Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past.
@@
 
-{\bf Keywords:} conservation, paleobiology, extinction
+{\bf Keywords:} conservation, paleobiology, extinction, forecasting
 
@@
 
+We quantitatively compared our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig.
+\ref{fig:oos_ins_diff}).
+This calculation means that differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts.
+A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast.
+Divergences between our out-of-sample and in-sample forecasts are rare and do not form multimillion year patters, quantifying broad congruence that visual comparisons of the two AUC time series suggests (Fig.
+\ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
+  \caption{Comparison between between our out-of-sample forecasts and in-sample forecasts for all models.
+This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figre \ref{fig:auc_taxon_time}.
+A differences close to 0 indicate complete congruence betwene in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast.
+See Table \ref{tab:model_def} for a description of each of the four models.}
+  \label{fig:oos_ins_diff}
+\end{figure}
+
 %Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 

### 2019-03-13 doc/paper.tex
grammar checks

@@
 We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable.
-We approach this this problem by analyzing the past in order to predict the future.
+We approach this problem by analyzing the past in order to predict the future.
 The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012a,Finnegan2015}.
@@
 Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
-This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
+This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
 
@@
 
-For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms F(oraminifera, Radiolaria, Diatoms, and Coccolithophores).
+For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores).
 These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}.
@@
 Then the model is fit to the first and second golds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold.
-This process continues with the model fit to the first three folds used to forecast extinction probabilities for the fourth fold, and then the model fit to the first four folds being used to forecast forecast the fifth fold.
+This process continues with the model fit to the first three folds used to forecast extinction probabilities for the fourth fold, and then the model fit to the first four folds being used to forecast the fifth fold.
 
@@
 See our code repository at https://github.com/psmits/trident for full code details.
-The entire analysis was coded in R and uses tidyverse-style tools such as \texttt{dplyr} \citep{dplyr}, and \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}.
+The entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, and \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}.
 Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.
@@
 The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to.
-Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig.
+Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all the models have approximately equal in-sample performance (Fig.
 \ref{fig:auc_hist}).
@@
 Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions.
-All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77.
+All the in-sample AUC estimates from our models are concentrated on an AUC value of 0.77.
 It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.
@@
 \ref{fig:auc_hist}).
-For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.
+For example, the difference between the ``past and vary'' model and the others is more pronounced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.
 %\begin{figure}[ht]
@@
 
-For many taxon/model combinations there are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 -- AUC value of 0.5 indicates that the model's predictions are no better than random (Fig.
+There are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 for the various taxon/model combinations (Fig.
 \ref{fig:auc_taxon_time}).
@@
 \ref{fig:auc_hist}).
-These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
+These differences mean that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
 However, the shapes of the posterior distributions means interpretations based only on the median values is incorrect -- the models are effectively indistinguishable in their out-of-sample forecast performance.
@@
 %  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time_full}
-%  \caption{Comparison of out-of-sample AUC values calculated for each of the My intervals for each of the four models.
-The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
+%  \caption{Comparison of out-of-sample AUC values calculated for each of the million year intervals for each of the four models.
+The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds.
 This presentation decomposes each of the 12-million year folds (Fig.
@@
   \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
-The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds.
+The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds.
 This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals.
@@
 A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast.
-Divergences between our out-of-sample and in-sample forecasts are rare and do not form multimillion year patters, quantifying broad congruence that visual comparisons of the two AUC time series suggests (Fig.
+Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, quantifying the broad congruence that the visual comparison of the two types of AUC time series suggest (Fig.
 \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).
+The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig.
+\ref{fig:oos_ins_diff}).
 
@@
   \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
-  \caption{Comparison between between our out-of-sample forecasts and in-sample forecasts for all models.
-This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figre \ref{fig:auc_taxon_time}.
-A differences close to 0 indicate complete congruence betwene in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast.
+  \caption{Comparison between our out-of-sample forecasts and in-sample forecasts for all models.
+This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}.
+A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast.
 See Table \ref{tab:model_def} for a description of each of the four models.}
@@
 %  \end{subfigure}
-%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation.
+%  \caption{Comparison between species geographic range histories and our estimate of their probability of extinction for each time of observation.
 Geographic range is assumed observed without error.
@@
 
-One of the most striking results of this analysis is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig.
+One of the most striking results of this analysis is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical covariates are predictors of species extinction have better in-sample performance (Fig.
 \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig.
@@
 
-Other noteworthy aspects of the results of our in-sample forecasts are that although AUC estimates for our models models differ significantly between models, all of these estimates are in a narrow range of possible AUC values (Fig.
+Other noteworthy aspects of the results of our in-sample forecasts are that although AUC estimates for our models differ significantly between models, all of these estimates are in a narrow range of possible AUC values (Fig.
 \ref{fig:auc_hist}).
@@
 
-The principal reason we were not able to include more biological information in our models is that we lack most any life history or ecological information on most marine micro- and nano-plankton.
+The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nanno-plankton.
 Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}.
@@
 The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1.
-This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}.
+This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is an overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}.
 However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}.
@@
 Temperature was also transformed and standardized the in the same manner as geographic range.
-Difference in temperature between between observations was measured from the standardized temperature values and was not standardized separately.
+Difference in temperature between observations was measured from the standardized temperature values and was not standardized separately.
 
@@
 The next two columns of \(X\) are two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\).
-The difference in geographic range was calculated from the transformed and standardized geographic range values; this means that change in geographic range is in units of changes in standard deviations.
+Change in geographic range was calculated from the transformed and standardized geographic range values; this means that change in geographic range is in units of changes in standard deviations.
 The final two columns are two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e.
@@
 The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a  \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla.
-The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\).
+The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each Coefficients for each phylum \(p\).
 These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept.
@@
 \(A\) is a \(l\) by \(p\) matrix, where \(l\) is the age at observation of a species and \(p\) is its phylum.
-\(A\) is modeled as following a multivariate normal distribution with phylum means being vector \(\delta\) and covariance matrix \(\Sigma_{A}\).
+\(A\) is modeled as following a multivariate normal distribution with phylum means being the vector \(\delta\) and covariance matrix \(\Sigma_{A}\).
 The covariation between the elements of vector \(\delta\) are modeled as a multivariate normal distribution with a mean vector of all 0s and covariance matrix \(\Sigma_{\delta}\).
 
-To complete the generative model, we need to assign final priors to the ``top level'' parameters.
-In general we favored weakly informative priors which help regularize our estimates.
+To complete the generative model, we need to assign final priors to the ``top-level'' parameters.
+In general, we favored weakly informative priors which help regularize our estimates.
 In the case of a regression coefficient, this means a Normal distribution with mean 0 and a standard deviation of 3.
@@
 Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling.
-In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.
+In order to prevent divergent transitions adapt delta was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.
 

### 2019-04-03 doc/norcalpaleo2019_talk.tex
norcalpaleo talk if needed

--- 2018-11-01 doc/gsa2018_talk.tex
@@
     \begin{column}{0.4\textwidth}
-      \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/occ_time_label}
+      \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/occ_time_label_full}
     \end{column}
     \begin{column}{0.4\textwidth}
-      \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/age_label}
+      \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/age_label_full}
     \end{column}
@@
     \begin{column}{0.6\textwidth}
+      The area represents the probability of correct ranking of a random ``extant''-``extinct'' pair.
+
       \[
@@
 
-      The area represents the probability of correct ranking of a random ``extinct''-``extant'' pair.
     \end{column}
@@
 
-% demonstrate as ROC curve?
-\begin{frame}
-  \frametitle{In-sample predictive performance, full dataset}
-
-  \begin{center}
-    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_hist_zoom}
-
-    \footnotesize{AUC = 0.7-0.8 acceptable/fair}
-  \end{center}
-
-\end{frame}
+%\begin{frame}
+%  \frametitle{In-sample predictive performance, full dataset}
+%
+%  \begin{center}
+%    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_hist_zoom}
+%
+%    \footnotesize{AUC = 0.7-0.8 acceptable/fair}
+%  \end{center}
+%
+%\end{frame}
 
@@
   \begin{center}
-    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_hist}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
 
@@
 \begin{frame}
-  \frametitle{In-sample predictive performance, by time}
+  \frametitle{In-sample predictive performance, by time and taxa}
 
   \begin{center}
-    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_ts}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/fold_auc_zoom}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/fold_auc_zoom_full}
 
@@
 \begin{frame}
-  \frametitle{Cross-validation results, by time}
+  \frametitle{Cross-validation results, by time and taxa}
 
   \begin{center}
-    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
   \end{center}
@@
 \end{document}
-
-\end{frame}
-
-\end{document}

### 2019-04-08 doc/paper.tex
worksmithing methods + results

@@
 We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
-All occurrence information was downloaded for calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
+Occurrence information was downloaded for calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
 This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya.
@@
 
-All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence.
+All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence as listed in the Neptune Database.
 After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred.
@@
 
-Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}.
+We also calculated the average global temperature of each 1 My bin as estimated from Magnesium/Calcium isotope ratios \citet{Cramer2011}.
 These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
@@
 Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya.
-We binned these estimates into 1 My intervals as we did with the fossil occurrences.
 The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.
@@
 
-We developed a discrete-time survival model to analyze our data in order to answer our question of how wee can we predict extinction risk at million year time scales.
-We considered four model variations: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but we include none of our historical covariates (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and we include our historical covariates (Model VP).
+We used a discrete-time survival modelling framework in order to estimate how well we can predict extinction risk at one million year time scales.
+At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}.
+We considered four model variations of the predictors included in our model: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but none of our historical covariates are included (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and historical covariates are included (Model VP).
 See Table \ref{tab:model_def} for further explanation of how these models differ from each other.
 For a complete description of the statistical model used in this analysis, please see Section \ref{sec:model_desc}.
-For a description of how our models were implemented, please see Section \ref{sec:model_est}.
+For a full description of how our models were implemented, please see Section \ref{sec:model_est}.
 
@@
 
-We are interested in model performance into two contexts: in-sample and out-of-sample predictive performance i.e. forecasting.
+We are interested in our models' performance into two contexts: in-sample and out-of-sample predictive performance i.e. forecasting.
 
@@
 We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data).
-These in-sample measures, however, are not necessarily good estimates of the model's ability to predict data from the future.
+In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}.
 To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation.
@@
 
-Here is a description of how k-fold cross-validation for time series precedes \citep{Arlot2010,Bergmeir2016}.
-The model is fit first to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e.
+k-fold cross-validation for time series precedes as follows \citep{Arlot2010,Bergmeir2016}.
+First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e.
 the future).
-Then the model is fit to the first and second golds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold.
-This process continues with the model fit to the first three folds used to forecast extinction probabilities for the fourth fold, and then the model fit to the first four folds being used to forecast the fifth fold.
+Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold.
+This process continues with the model fit to the first three folds combined is then used to forecast extinction probabilities for the fourth fold.
+Finally, the model is fit to the first four folds combined and then is used to forecast the fifth fold.
+The results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.
 
-Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}.
-This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives.
-AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.
-AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of any two randomly selected species \citep{Fawcett2006,Mason2002}.
-AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, which means that values between 0.7 and 0.8 could then be considered ``fair.''
+The relative adequacy of our four model variants were compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}.
+This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives.
+AUC ranges between 0.5 and 1, with 0.5 indicating no difference in classification from random and 1 indicating perfect classification.
+AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of a randomly selected extinct-extant species pair \citep{Fawcett2006,Mason2002}.
+AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we interpret values between 0.7 and 0.8 could then be considered ``fair,'' and values between 0.6 and 0.7 as ``poor.''
 
@@
 See our code repository at https://github.com/psmits/trident for full code details.
-The entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, and \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}.
+This entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}.
 Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.
@@
 
-The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to.
-Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all the models have approximately equal in-sample performance (Fig.
+The in-sample model comparisons are useful comparing the relative ability of our models to represent the data it was fit to.
+Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that all of our models have approximately equal in-sample forecasting performance (Fig.
 \ref{fig:auc_hist}).
@@
 
-There are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 for the various taxon/model combinations (Fig.
+There are one or more time periods where our posterior distribution of in-sample AUC has a median value less than or equal to 0.5 (Fig.
 \ref{fig:auc_taxon_time}).
-However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the VP model.
-Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the VP model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
+However, this pattern is absent for the posterior estimates of in-sample AUC for Foraminfera and Radiolaria as fit by the VP model.
+In contrast, there are fewer periods of low model performance for calcareous nannoplankton and Dinoflagellates as estimated from our VP model than in those estimates from the other three model variations.
+
 \begin{figure}[ht]
@@
 The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}.
-We find the visual degree of this multimodality increases with model complexity (Fig.
+We find that this multimodality increases with model complexity (Fig.
 \ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time which allows for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
 
-Comparison between the in-sample forecasting performance distributions of expected out-of-sample AUC reveals a similar range in plausible values for all models (Fig.
+Comparison between the in-sample forecasting performance estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig.
 \ref{fig:auc_hist}, \ref{fig:fold_auc}).
-Interestingly, the differences in the posterior predictive distributions of AUC for the models have decreased.
-For example, model VP not clearly better than either models V or CP (Fig.
-\ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig.
-\ref{fig:auc_hist}).
+Interestingly, the differences in posterior predictive distributions of AUC between the four model variations has decreased.
+For example, the VP model no longer has the greatest median AUC of the four models (Fig.
+\ref{fig:fold_auc}).
 These differences mean that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
-However, the shapes of the posterior distributions means interpretations based only on the median values is incorrect -- the models are effectively indistinguishable in their out-of-sample forecast performance.
+However, making interpretations bsaed only incorrect -- our estimates are the full posterior, not individual points.
+Because of this, the four models are effectively indistinguishable in their out-of-sample forecast performance (Fig.
+\ref{fig:fold_auc}).
 
@@
 \ref{fig:fold_auc_taxon_time}).
-The time-series of posterior predictive values for each model are broadly congruent between each taxonomic group.
+In all cases, our four models are nearly identical in their out-of-sample forecasting performance over time and between each of the taxonomic groups.
 
-In the analysis of the posterior predictive distributions of AUC from the in-sample forecasts for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig.
+In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig.
 \ref{fig:auc_taxon_time}).
-This occurrence is generally much rarer for the posterior predictive distribution of AUC from the out-of-sample forecasts -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random.
+This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts.
+The major exception to this pattern are our estimates for the Dinoflagellates, which has at least one time interval for all four models where the median AUC of the out-of-sample forecasts were no better random.
 In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
@@
 
-We quantitatively compared our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig.
+We compared the difference in our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig.
 \ref{fig:oos_ins_diff}).
-This calculation means that differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts.
+A difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts.
 A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast.
-Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, quantifying the broad congruence that the visual comparison of the two types of AUC time series suggest (Fig.
+Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, something that is consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig.
 \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).
@@
 Similarly, these models are expected to correctly forecast the rank order in extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time.
-A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original).
+A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original).
 Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past.

### 2019-04-11 doc/paper.tex
revised supplement. added some citations about core mixing

@@
 
-Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources.
-The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species.
-For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0.
-The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1.
-This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is an overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}.
+The estimated age of each occurrence is based on the core-specific age-model that observation is from and can be overly precise.
+To alleviate this overprecision, we coarsened our temporal information in an effort to limit the effects of between-core heterogeneity in age.
+The occurrence histories of each species was then summarized as a series of binary codes indicating the presence or last occurrence of that species.
+For every occurrence of a species, except the last, that species existence and survival is recorded as a 0.
+The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is recorded as 1.
+This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is an overconfident statement of preservation and may be shortening the actual durations of the studied species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}.
 However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}.
-In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.
+In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores \citep{Mekik2018,Broecker1999,Mekik2014,Peng1984}.
 
-\subsubsection{Covariate transforms and standardizations}
+\subsubsection{Covariate transformation and standardization}
 
 Prior to analysis, geographic range was then log-plus-one transformed and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges.
-This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}.
+This standardization means that a regression coefficient associated with each covariate describes the change in extinction probability per change in standard deviation of that covariate, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}.
 Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.
@@
 The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a  \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla.
-The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each Coefficients for each phylum \(p\).
+The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each phylum \(p\).
 These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept.

### 2019-04-11 doc/paper.tex
integrated seth round 3 comments. shortened discussion

@@
 \begin{abstract}
-  A tenant of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions.
+  A tenet of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions.
 Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past.
@@
 
-Being able to predict which species are more likely than others to go extinct over a given time frame is crucial for planning decisions policies to ameliorate the current biodiversity crisis.
-Some species have been the focus of intense study to define short-term (e.g.
-decadal to centennial scale) abundance and geographic range trajectories.
+The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the nea future.
+Such predictions are difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented.
+This has led to the suggestions that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}.
+The fossil record preserves information about the full histories, include ultimate extinction, of thousands of lineages, and this information can help to augment the shorter-term higher-resolution data used to make risk assessments of extant taxa.
 
-We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable.
-We approach this problem by analyzing the past in order to predict the future.
-The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012a,Finnegan2015}.
-By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions -- we ask the probability that, given two random species, we correctly rank their relative risks of extinction.
+Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
 
-Studying how species vary in their extinction risk over time means we can assess which species are at greater risk under unobserved conditions.
-We know that a species' risk of extinction varies over time in both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}.
+This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
+
+Here we ask how precise risk predictions based on fossil data might be.
+Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions.
+However, we can address the question by taking a given point in the geological past, developing predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g.
+``future'', from the point of view of the model) extinction/survival events.
+
+Both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}.
+Given this variation, how accurate are our assessments based on past events likely to be when applied to the future?
+Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.
+
 Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction.
@@
 
-By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past.
-Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
-This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
-
-%For example, while a predictor may be ``significant'' when comparing the odds of extinction risk between groups, the practical difference that predictor makes on prediction can be minimal \citep{ARM}.
-By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.
-
-A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction.
-For example, we know that a species' global geographic range changes over its duration \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}.
-We also know that a species' geographic range size is a good predictor of differences extinction risk \citep{Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
-This begs the question: how does a species' extinction risk change over its duration?
-While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}).
-%For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years?
-
-%Here, we explicitly estimate the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction.
-Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.
+Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that geographic range trajectories are highly variable \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
+The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian.
+In the former case, knowledge of the specific past trajectory of a species -whether its range has expanded or contracted over a given timespan- might help to improve assessments of its current risk.
+In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past.
+Discriminating among these alternative models is thus very important for determining how best to incorporate fossil data in present risk assessments.
 
@@
 This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.
-%Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale.
-Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales.
-%Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. 
 The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores from which the occurrences are recorded.
 Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.
-
-%In particular, there have been multiple major climatic and oceanographic events, both short and protracted in duration, over the course of the Cenozoic.
-For example, the Paleocene-Eocene Thermal Maximum (PETM) was an geologically close-to instantaneous global warming event believed to have been caused by a massive injection of carbon into the atmosphere over a 20,000 year period which rapidly warmed and is associated with substantial changes to the global distribution of species CITATION.
-In contrast, the Eocene-Oligocene boundary is associated with a protracted global cooling event caused by the slow opening of Drake's passage and the development of the Antarctic ice sheet CITATION.
-These moments of large scale global change are inherently ``different,'' so predicting extinction before, during, and after these is important -- especially if those events are analogous to future climate change.
-
-%Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts.
-Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition \citep{Zachos2008} -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions.
-Alternatively, we would expect our model based predictions of extinction surrounding the Paleocene-Eocene Thermal Maximum may be less accurate because there are inherently fewer samples from the rapid climatic event \citep{Zachos2008}.
 
@@
 
+%A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original).
+Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past.
+However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
+
 %This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time.
@@
 
-The results of this paper set out our baseline ability to predict relative differences in extinction risk.
-We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations.
-Similarly, these models are expected to correctly forecast the rank order in extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time.
-A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original).
-Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past.
-However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
-
-One of the most striking results of this analysis is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical covariates are predictors of species extinction have better in-sample performance (Fig.
+We find that all of our models are expected to correctly forecasting the rank order in extinction risk of two randomly selected future observations approximately 70\% to 80\% of the time.
+One of the most striking aspects of this results is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical covariates are predictors of species extinction have better in-sample performance (Fig.
 \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig.
@@
 
-Other noteworthy aspects of the results of our in-sample forecasts are that although AUC estimates for our models differ significantly between models, all of these estimates are in a narrow range of possible AUC values (Fig.
+We also find that our four models are practically identical in their ability to make forecasts both in-sample and out-of-sample.
+Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig.
 \ref{fig:auc_hist}).
-For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig.
+While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig.
 \ref{fig:fold_auc}).
-This means that even though model VP has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.
+This means that even though model VP has a statistically greater AUC than the other three models, this result is not practically or scientifically significant.
 
@@
 
-The success of our models is partially driven by the size of our dataset and the hierarchical structure of our model.
-Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping.
-Instead of relying on large amounts of ecological information to shape individual differences, our model leverages most of the Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.
-
+The quality and consistency of our models out-of-sample forecasting performance is suprising given that its estimates are based on rather limited information about the taxa themselves.
+Our most complex models are only taking into account some aspects of species geographic range and their taxonomic grouping.
 The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nanno-plankton.

### 2019-04-15 doc/paper.tex
additional word smithing to methods and supplement

@@
 \begin{abstract}
-  A tenet of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions.
+  A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions.
 Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past.
@@
 
-  %risk assessments for extant species can be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.
+  %risk assessments for extant species can be improved by examining past extinction patterns and by using palaeontological records to establish geographic range and abundance trajectories on geological timescales.
   %We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
@@
   %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
-  %The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
+  %The results of this study reinforce the importance of the promise of palaeontology and using the past to predict the future.
 \end{abstract}
 
-{\bf Keywords:} conservation, paleobiology, extinction, forecasting
+{\bf Keywords:} conservation, palaeobiology, extinction, forecasting
 
@@
 
-Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
+Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
 
@@
 
-We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
-Occurrence information was downloaded for calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
-This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya.
-This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction as discussed below.
+We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
+This dataset of occurrences covered calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
+These occurrences were then filtered to include only those species who's first occurrence was no more than 63 Mya.
+This filtering criterion means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and ensures that our occurrence histories are congruent with the temperature time-series used as a predictor of extinction (see below).
 
@@
 
-We also calculated the average global temperature of each 1 My bin as estimated from Magnesium/Calcium isotope ratios \citet{Cramer2011}.
-These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
+Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}.
+Elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
 meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself.
 This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
-Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya.
+Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My from 0 to 63 Mya.
 The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.
 
-See Section \ref{sec:data_desc} for a further explanation on how observations were temporally binned and how our covariates were standardized and transformed prior to analysis.
+See Section \ref{sec:data_desc} for a further explanation on how observations were temporally binned, and how our covariates were standardized and transformed prior to analysis.
 
@@
 At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}.
-We considered four model variations of the predictors included in our model: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but none of our historical covariates are included (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and historical covariates are included (Model VP).
-See Table \ref{tab:model_def} for further explanation of how these models differ from each other.
-For a complete description of the statistical model used in this analysis, please see Section \ref{sec:model_desc}.
-For a full description of how our models were implemented, please see Section \ref{sec:model_est}.
+We considered four variations of the predictors included in our model: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but none of our historical covariates are included (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and historical covariates are included (Model VP).
+We always included species geologic age at time of observation as a non-nested varying intercept term.
+This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.
 
-Importantly, we include species geologic age at time of observation as a non-nested varying intercept term.
-This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.
-Again, see Section \ref{sec:model_desc} for how this term was specified as part of our model.
+See Table \ref{tab:model_def} for further explanation of how the four models we considered differ from each other.
+A complete description of the statistical model used in this analysis is available in Section \ref{sec:model_desc}.
+Additionally, the full description of how these models were implemented and coded, including choice of priors, is available Section \ref{sec:model_est}.
 
@@
 This is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
+In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}.
 
 We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data).
-In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}.
 To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation.
@@
 
-k-fold cross-validation for time series precedes as follows \citep{Arlot2010,Bergmeir2016}.
+k-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}.
 First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e.
@@
 Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold.
-This process continues with the model fit to the first three folds combined is then used to forecast extinction probabilities for the fourth fold.
+Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold.
 Finally, the model is fit to the first four folds combined and then is used to forecast the fifth fold.
@@
 
-The relative adequacy of our four model variants were compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}.
-This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives.
+The relative adequacy of the four model variants was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}.
+This measure is commonly used to measure the performance of classification models as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives.
 AUC ranges between 0.5 and 1, with 0.5 indicating no difference in classification from random and 1 indicating perfect classification.
@@
 
-We find that all of our models are expected to correctly forecasting the rank order in extinction risk of two randomly selected future observations approximately 70\% to 80\% of the time.
-One of the most striking aspects of this results is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical covariates are predictors of species extinction have better in-sample performance (Fig.
+We find that all of our models are expected to correctly forecasting the rank order in extinction risk of two randomly selected future observations between 70\% to 80\% of the time.
+One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations.
+A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original).
+Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past.
+However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
+An important caveat to our results is that human impacts may substantially alter extinction risk dynamics away from the average Cenozoic ``state'' so that the future could become less predictable than it has been in the past.
+
+%This similarity demonstrates that while models with historical covariates are predictors of species extinction have better in-sample performance (Fig.
 \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig.
@@
 \ref{fig:fold_auc}).
-This means that even though model VP has a statistically greater AUC than the other three models, this result is not practically or scientifically significant.
+This means that even though model VP has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant.
+The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance.
 
@@
 
-The quality and consistency of our models out-of-sample forecasting performance is suprising given that its estimates are based on rather limited information about the taxa themselves.
+The quality and consistency of our models out-of-sample forecasting performance is surprising given that its estimates are based on rather limited biological and environmental information about the taxa.
 Our most complex models are only taking into account some aspects of species geographic range and their taxonomic grouping.
-The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nanno-plankton.
-Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}.
-However, not this information does not exist all foram species.
-If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species.
-This means ignoring the majority of occurrence information present in the Neptune database.
+The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nannoplankton.
+Forams are an exception to this problem as aspects of life history, ecology, and physiology for many foram species \citep{Ezard2011}.
+However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here.
+Including this type of information in a predictive model of extinction risk we require analyzing a single, limited taxonomic subset of the fossil occurrences present the Neptune Database.
+This would mean ignoring the majority of occurrence information present in the Neptune database.
+
+%It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time.
+An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
 
@@
 Temperature was also transformed and standardized the in the same manner as geographic range.
-Difference in temperature between observations was measured from the standardized temperature values and was not standardized separately.
+The change in temperature between an observation and its previous observation was measured from the standardized temperature values and was not standardized separately.
 
@@
 
-In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and relevant covariates.
+In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and covariate information.
 The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals \citep{Tutz2016}.
@@
 \ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\) \citep{Tutz2016}.
-\(h(\Theta)\) is then modeled as with any regression as it is defined for all real-values.
+\(h(\Theta)\) is then modeled as with any regression.
 In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes \citep{ARM}.
@@
 Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling.
-In order to prevent divergent transitions adapt delta was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.
+In order to prevent divergent transitions adapt delta was increased to 0.999999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.
 
-An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
+To implement our VP model in \begin{texttt}rstanarm\end{texttt}, where ``data'' is a data.frame object of all necessary data (response, covariates), is coded as:
 \begin{verbatim}
@@
 
-Similarly, our full model can also be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
+Similarly, our VP model can be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
 \begin{verbatim}

### 2019-04-15 doc/paper.tex
changes to introduction to improve clarity

@@
 
-The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the nea future.
-Such predictions are difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented.
+The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the near future.
+Predicting which species will go extinct is difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented.
 This has led to the suggestions that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}.
@@
 Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
-
 This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
@@
 Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions.
-However, we can address the question by taking a given point in the geological past, developing predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g.
+However, we can address the question by taking a given point in the geological past, developing a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g.
 ``future'', from the point of view of the model) extinction/survival events.
 
-Both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}.
+Extinction intensity (average rate) and selectivity (difference in risk between taxa) vary through time and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}.
+What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
 Given this variation, how accurate are our assessments based on past events likely to be when applied to the future?
@@
 
-Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction.
-What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
-By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range extinction intensities and selectivities.
-
-Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that geographic range trajectories are highly variable \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
+Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
 The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian.
-In the former case, knowledge of the specific past trajectory of a species -whether its range has expanded or contracted over a given timespan- might help to improve assessments of its current risk.
+In the former case, knowledge of the specific past trajectory of a species -- whether its range has expanded or contracted over a given time span -- might help to improve assessments of its current risk.
 In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past.
@@
 
-We also find that our four models are practically identical in their ability to make forecasts both in-sample and out-of-sample.
+We also find that our four models are practically identical in their ability to make in-sample and out-of-sample forecasts.
 Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig.

### 2019-04-15 doc/paper.tex
work smithing before sending to seth

@@
 
-We find that all of our models are expected to correctly forecasting the rank order in extinction risk of two randomly selected future observations between 70\% to 80\% of the time.
+We find that all of our models are expected to correctly forecasting the rank order in extinction probabilities for a randomly selected extinct-extant pair of future observations between 70\% to 80\% of the time (Fig.
+\ref{fig:fold_auc}).
 One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations.

### 2019-04-26 doc/paper.tex
seth 4th rnd comments in

@@
 Extinction intensity (average rate) and selectivity (difference in risk between taxa) vary through time and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}.
-What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
-Given this variation, how accurate are our assessments based on past events likely to be when applied to the future?
+This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
 Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.
@@
 Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
-The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian.
+The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}.
 In the former case, knowledge of the specific past trajectory of a species -- whether its range has expanded or contracted over a given time span -- might help to improve assessments of its current risk.
@@
 We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
-This dataset of occurrences covered calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
-These occurrences were then filtered to include only those species who's first occurrence was no more than 63 Mya.
-This filtering criterion means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and ensures that our occurrence histories are congruent with the temperature time-series used as a predictor of extinction (see below).
+This occurrence-based dataset includes calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
+Occurrences were filtered to include only those species with first occurrence was no more than 63 Mya.
+This filtering criterion excludes taxa that survived the K/Pg extinction or arose during its ecologically unusual recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).
 
@@
 Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}.
-Elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g.
-meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself.
-This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
+We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
 Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My from 0 to 63 Mya.
@@
 
-We used a discrete-time survival modelling framework in order to estimate how well we can predict extinction risk at one million year time scales.
+We used a discrete-time survival modelling framework to estimate how well we can predict extinction risk at one million year time scales.
 At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}.
-We considered four variations of the predictors included in our model: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but none of our historical covariates are included (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and historical covariates are included (Model VP).
-We always included species geologic age at time of observation as a non-nested varying intercept term.
+We considered four different models involving different permutations of covariate effects (fixed or time-varying) and historical covariates: covariate effects constant over time and no historical covariates included (Model C), covariate effects allowed to vary over time but no historical covariates included (Model V), covariate effects constant over time and historical covariates included (Model CP), and covariate effects allowed to vary over time and historical covariates are included (Model VP).
+The C and P models atempt to predict based only on present state, wheras the CP and VP models allow for the possibility of non-Markovian behaviour by including change in state from the previous time increment.
+
+We always included species age at time of observation (.e.
+observed prior duration) as a non-nested varying intercept term.
 This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.
@@
 
-We are interested in our models' performance into two contexts: in-sample and out-of-sample predictive performance i.e. forecasting.
+We are interested in our models' performance in two contexts: in-sample performance, and out-of-sample predictive performance (i.e. forecasting).
 
@@
 Each fold represents a sequence of time points.
-With 63 time points, each of the five folds represents approximately 13 time points.
-Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
+With 63 time points, each of the five folds represents approximately 13 million-year time increments.
+It is important to bear in mind, however, that each time increment includes many (100s-1000s) individual observations.
 
@@
 
-The in-sample model comparisons are useful comparing the relative ability of our models to represent the data it was fit to.
+The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to.
 Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that all of our models have approximately equal in-sample forecasting performance (Fig.
@@
 
-There are one or more time periods where our posterior distribution of in-sample AUC has a median value less than or equal to 0.5 (Fig.
+Depending on the taxon-model combination, there are between zero and 4 time intervals where our posterior distribution of in-sample AUC has a median value less than or equal to 0.5 (Fig.
 \ref{fig:auc_taxon_time}).
@@
 Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}.
-The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}.
-We find that this multimodality increases with model complexity (Fig.
-\ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time which allows for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
+The resulting distribution when all folds are combined is highly multimodal as expected given that they are fit to and estimated from different data sets with different numbers of observations \citep{ESL}.
+This multimodality increases with model complexity (Fig.
+\ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time, allowing a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
 
-Comparison between the in-sample forecasting performance estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig.
+Comparison the in-sample AUC estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig.
 \ref{fig:auc_hist}, \ref{fig:fold_auc}).
-Interestingly, the differences in posterior predictive distributions of AUC between the four model variations has decreased.
+Interestingly, the differences in posterior predictive distributions of AUC between the four model variations is reduced in out-of-sample prediction.
 For example, the VP model no longer has the greatest median AUC of the four models (Fig.
 \ref{fig:fold_auc}).
-These differences mean that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
-However, making interpretations bsaed only incorrect -- our estimates are the full posterior, not individual points.
+For this reason the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
+However, making interpretations based only on the median AUC estimates is incorrect -- our estimates are the full posterior, not individual points.
 Because of this, the four models are effectively indistinguishable in their out-of-sample forecast performance (Fig.
@@
 
-We can also compare the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig.
-\ref{fig:fold_auc_taxon_time}).
-In all cases, our four models are nearly identical in their out-of-sample forecasting performance over time and between each of the taxonomic groups.
+The posterior predictive distribution of expected out-of-sample AUC over time and across taxonomic groups are nearly identical for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}).
 
@@
 This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts.
-The major exception to this pattern are our estimates for the Dinoflagellates, which has at least one time interval for all four models where the median AUC of the out-of-sample forecasts were no better random.
-In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
+The major exception to this pattern are our estimates for the Dinoflagellates, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random.
+The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
 \begin{figure}[ht]
@@
 A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast.
-Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, something that is consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig.
+Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig.
 \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).
@@
 A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original).
-Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past.
-However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
-An important caveat to our results is that human impacts may substantially alter extinction risk dynamics away from the average Cenozoic ``state'' so that the future could become less predictable than it has been in the past.
+Additionally, a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect.
+However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to variation in extinction intensity over the Cenozoic.
+A extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
 
@@
 \ref{fig:fold_auc}).
-This means that even though model VP has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant.
+Thus even though the VP model has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant.
 The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance.
+This including geographic range trajectory results in only very minor improvements in forecasting accuracy.
+This contrasts with findings from coral genera \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.
 
@@
 
-The quality and consistency of our models out-of-sample forecasting performance is surprising given that its estimates are based on rather limited biological and environmental information about the taxa.
-Our most complex models are only taking into account some aspects of species geographic range and their taxonomic grouping.
-The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nannoplankton.
-Forams are an exception to this problem as aspects of life history, ecology, and physiology for many foram species \citep{Ezard2011}.
+The relative quality and consistency of our models out-of-sample forecasting performance is encouraging given that its estimates are based on very limited biological and environmental information about the taxa.
+Even our most complex models only account for a few simple aspects of geographic range, prior history, and taxonomy.
+The principal reason we were not able to include more biological information in our models is that we lack suitably detailed life history or ecological information for many marine micro- and nannoplankton.
+Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}.
 However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here.
-Including this type of information in a predictive model of extinction risk we require analyzing a single, limited taxonomic subset of the fossil occurrences present the Neptune Database.
-This would mean ignoring the majority of occurrence information present in the Neptune database.
+Further analyses including this type of information and focused more narrowly on the foraminifera may be informative.
 
@@
 
-In conclusion, our results provide a promising picture of our ability to predict the relative extinction risk of two randomly selected species.
+In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting the relative extinction risk of two randomly selected species based on a small number of taxonomic, geographic, and historical predictors.
 Considering that conservation decisions are made based on a continuum of risk, from most to least, this means that our results are in the same language as how conservation resources are allocated.
+These results present an encouraging baseline for using the fossil record to predict future extinction.
+This suggests, that all else being equal, extending this model to include the information in modern risk assessments might actually do decently well at forecasting future extinctions.
 

### 2019-04-29 doc/paper.tex
paper ready for submission process

@@
 Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past.
-To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) For each group we examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
-Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor.
-This model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species.
-It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time.
+To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
+Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate) the improvement in predictive power over models with constant effects and no historical covariates is very minor.
+Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample pair, implying that determinants of extinction have varied only modestly through time.
 An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
@@
 We considered four different models involving different permutations of covariate effects (fixed or time-varying) and historical covariates: covariate effects constant over time and no historical covariates included (Model C), covariate effects allowed to vary over time but no historical covariates included (Model V), covariate effects constant over time and historical covariates included (Model CP), and covariate effects allowed to vary over time and historical covariates are included (Model VP).
-The C and P models atempt to predict based only on present state, wheras the CP and VP models allow for the possibility of non-Markovian behaviour by including change in state from the previous time increment.
+The C and P models attempt to predict based only on present state, whereas the CP and VP models allow for the possibility of non-Markovian behaviour by including change in state from the previous time increment.
 
-We always included species age at time of observation (.e.
+We always included species age at time of observation (i.e.
 observed prior duration) as a non-nested varying intercept term.
@@
 
-k-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}.
+\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}.
 First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e.
@@
 
-We find that all of our models are expected to correctly forecasting the rank order in extinction probabilities for a randomly selected extinct-extant pair of future observations between 70\% to 80\% of the time (Fig.
+We find that all of our models are expected to correctly forecasting the rank order of extinction probability for a randomly selected extinct-extant pair of future observations between 70\% to 80\% of the time (Fig.
 \ref{fig:fold_auc}).
@@
 However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here.
-Further analyses including this type of information and focused more narrowly on the foraminifera may be informative.
+Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.
 
@@
 For example, compare the results from our models fit using full in-sample dataset, and the results from the cross-validation where the fit to each fold is by definition based on less than full information.)
-
-In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting the relative extinction risk of two randomly selected species based on a small number of taxonomic, geographic, and historical predictors.
-Considering that conservation decisions are made based on a continuum of risk, from most to least, this means that our results are in the same language as how conservation resources are allocated.
-These results present an encouraging baseline for using the fossil record to predict future extinction.
-This suggests, that all else being equal, extending this model to include the information in modern risk assessments might actually do decently well at forecasting future extinctions.
 
@@
 
+\setcounter{page}{1}
 \section{Supplement to Materials and Methods}

### 2019-05-01 doc/letter_editor.tex
submission letter written

\documentclass{letter}
\usepackage{microtype}
\frenchspacing

\signature{Peter D. Smits}
\address{Department of Integrative Biology \\ University of California Berkeley \\ 3040 Valley Life Sciences Building Rm. 5151 \\ Berkeley, CA 94720 \\ psmits@berkeley.edu}

\begin{document}
\begin{letter}{Editor \\ \textit{Philosophical Transactions of the Royal Society B: Biological Sciences}}
  \opening{Dear Editor,}

  Please find enclosed our manuscript entitled ``How predictable is extinction? Forecasting species survival at million-year timescales'' by myself (Peter D. Smits) and Seth Finnegan. In this study we asked a relatively simple yet fundamental question in paleobiology: how well do models conditioned on past observations predict future extinction events. To answer this question we analyzed the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) We examined how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.

  Using cross-validation, we were able to estimate the expected forecast performance of our model to future observations. Our models were found to have a 70-80\% probability of correctly forecasting the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time. The relative quality and consistency of our models' out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.

  These results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of taxonomic, geographic, and historical predictors. Importantly, our results are directly comparable to conservation determinations because both are expressed in terms of a continuum of risk, from most to least. The results of this simple exercise suggest that conservation decisions would be bolstered by including fossil data.

  We believe that this study we will be of general interest to paleontologists, conservation biologists, ecologists, and evolutionary biologists. Thank you for considering our work. Please send all correspondence regarding this manuscript to me via my email address (psmits@berkeley.edu) or that of my co-author, Seth Finnegan (sethf@berkeley.edu).

  \closing{Sincerely,}

  \encl{Article, supplementary text}.

\end{letter}
\end{document}

### 2019-05-01 doc/paper.tex
submission letter written

@@
 Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate) the improvement in predictive power over models with constant effects and no historical covariates is very minor.
-Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample pair, implying that determinants of extinction have varied only modestly through time.
+Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time.
 An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
@@
 The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance.
-This including geographic range trajectory results in only very minor improvements in forecasting accuracy.
+Thus, including geographic range trajectory results in only very minor improvements in forecasting accuracy.
 This contrasts with findings from coral genera \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.
@@
 
-The relative quality and consistency of our models out-of-sample forecasting performance is encouraging given that its estimates are based on very limited biological and environmental information about the taxa.
-Even our most complex models only account for a few simple aspects of geographic range, prior history, and taxonomy.
+The relative quality and consistency of our models' out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
+Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity.
 The principal reason we were not able to include more biological information in our models is that we lack suitably detailed life history or ecological information for many marine micro- and nannoplankton.
@@
 Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.
+
+In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of taxonomic, geographic, and historical predictors.
+Our results are directly comparable to conservation determinations because both are expressed in terms of a continuum of risk, from most to least.
+The results of this simple exercise suggest that conservation decisions would be bolstered by including fossil data.
+
+%are encouraging for efforts to incorporate fossil data into current and future extinction risk assessments, and
 

### 2019-05-01 doc/letter_editor.tex
paper submitted

@@
 Smits) and Seth Finnegan.
-In this study we asked a relatively simple yet fundamental question in paleobiology: how well do models conditioned on past observations predict future extinction events.
+In this study we asked a relatively simple yet fundamental question for the emerging field of conservation paleobiology: how well do models conditioned on past observations predict future extinction events?
 To answer this question we analyzed the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) We examined how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
 
-  Using cross-validation, we were able to estimate the expected forecast performance of our model to future observations.
-Our models were found to have a 70-80\% probability of correctly forecasting the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time.
-The relative quality and consistency of our models' out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
+We used cross-validation to estimate the expected forecast performance of our model for future observations.
+Our models were found to have a 70-80\% probability of correctly forecasting the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction risk have varied only modestly through time.
+Including specific information about individual species trajectories resulted in only extremely marginal improvement in fit, suggesting that, at least at the million-year timescale, extinction risk is best described by dominantly Markovian (e.g.
+memoryless) models.
 
-  These results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of taxonomic, geographic, and historical predictors.
-Importantly, our results are directly comparable to conservation determinations because both are expressed in terms of a continuum of risk, from most to least.
-The results of this simple exercise suggest that conservation decisions would be bolstered by including fossil data.
-
-  We believe that this study we will be of general interest to paleontologists, conservation biologists, ecologists, and evolutionary biologists.
-Thank you for considering our work.
+The relative quality and consistency of our models’ out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
+The results of this simple exercise bolster the case that fossil data can meaningfully inform present and future conservation decisions.
+We therefore believe that our study will be of general interest to paleontologists, conservation biologists, ecologists, and evolutionary biologists.
+We look forward to your decision.
 Please send all correspondence regarding this manuscript to me via my email address (psmits@berkeley.edu) or that of my co-author, Seth Finnegan (sethf@berkeley.edu).

### 2019-06-01 R/01_inspect_data.r
added explore graph + table of temporal obs for all species

@@ -2,7 +2,7 @@ library(pacman)
 
 p_load(tidyverse, furrr, here,
        arm, rstanarm, bayesplot,
-       scales, ggridges,
+       scales, ggridges, viridis,
        pROC)
 
 source(here('R', 'helper01_process_foo.r'))
@@ -34,6 +34,48 @@ view_neptune(.data = counti_rt_trans,
 view_neptune(.data = counti_rl_trans, 
              name = 'restrict_local',
              path = here('results', 'figure'))
+
+
+
+# for each species, how many temporal windows do we observe it?
+# what percent of taxa with in a phylum are observed that many times?
+appear_count <- 
+  counti_trans %>%
+  group_by(fullname, fossil_group) %>%
+  summarize(n_appear = n_distinct(mybin)) %>%
+  ungroup()
+
+appear_count %>%
+  group_by(fossil_group) %>%
+  summarize(p_2 = sum(n_appear >=2) / n(),
+            p_3 = sum(n_appear >= 3) / n(),
+            p_4 = sum(n_appear >= 4) / n(),
+            p_5 = sum(n_appear >= 5) / n(),
+            p_6 = sum(n_appear >= 6) / n(),
+            p_7 = sum(n_appear >= 7) / n(),
+            p_8 = sum(n_appear >= 8) / n(),
+            p_9 = sum(n_appear >= 9) / n()) %>%
+  xtable::xtable(.) %>%
+  xtable::print.xtable(., file = here::here('results', 'appear_percent.tex'),
+                       include.rownames = FALSE)
+
+appear_graph <- 
+  appear_count %>%
+  mutate(fossil_name = case_when(fossil_group == 'D' ~ 'Dinoflagellates',
+                                 fossil_group == 'R' ~ 'Radiolaria',
+                                 fossil_group == 'F' ~ 'Foraminifera',
+                                 fossil_group == 'N' ~ 'Calc. nanno.')) %>%
+  ggplot(aes(x = n_appear, fill = fossil_name)) +
+  geom_bar() +
+  scale_fill_viridis(discrete = TRUE) +
+  theme(legend.position = 'bottom') +
+  labs(x = 'Time bin appearances',
+       y = 'Species',
+       fill = 'Taxon')
+ggsave(filename = here::here('results', 'figure', 'appear_time_bin.png'),
+       plot = appear_graph, width = 10, height = 6)
+
+
 
 
 # looking at the temperature data

### 2019-06-03 R/01_inspect_data.r
handling more lags/diffs for predictors

@@ -82,7 +82,8 @@ ggsave(filename = here::here('results', 'figure', 'appear_time_bin.png'),
 mgca <- read_tsv(here('data', 'cramer', 'cramer_temp.txt'))
 names(mgca) <- str_to_lower(names(mgca))
 names(mgca) <- str_remove_all(names(mgca), '[^[[:alnum:]]]')
-mgca <- mgca %>%
+mgca <- 
+  mgca %>%
   mutate(bin = break_my(age, by = 1)) %>%
   group_by(bin) %>%
   mutate(temp_bin = mean(temperature, na.rm = TRUE)) %>%
@@ -130,7 +131,7 @@ width_effect <-
   vary_width %>%
   dplyr::select(-latext) %>%
   group_by(width) %>%
-  mutate_at(vars(-width), funs(inplace(.))) %>%
+  mutate_at(vars(-width), .funs = list(~ inplace(.))) %>%
   gather(key = 'key', value = 'value', -width) %>%
   ggplot(aes(x = width, y = value)) +
   geom_jitter(height = 0, width = 0.05, alpha = 0.01) +

### 2019-06-03 R/02_fit_model.r

@@ -89,6 +89,7 @@ counti <- read_rds(here('data', 'counting.rds'))
 counti_restrict_time <- read_rds(here('data', 'counting_restrict_time.rds'))
 counti_restrict_local <- read_rds(here('data', 'counting_restrict_local.rds'))
 
+
 # transform the data for analysis
 counti_trans <- prepare_analysis(counti)
 counti_rt_trans <- prepare_analysis(counti_restrict_time)

### 2019-06-03 R/helper01_process_foo.r

@@ -38,29 +38,79 @@ prepare_analysis <- function(x, fg = NULL) {
   # prep the data
   tb <- tb %>%
     arrange(fullname, desc(mybin)) %>%
-    mutate_at(.vars = vars(ncell, latext, maxgcd, area, temp),
+    mutate_at(.vars = vars(mst, ncell, latext, maxgcd, area, temp),
               .funs = ~ scale(log1p(.x))) %>%
     group_by(fullname) %>%
-    mutate(diff_maxgcd = c(0, diff(maxgcd)),
-           lag1_maxgcd = lag(maxgcd, default = 0),
-           lag2_maxgcd = lag(maxgcd, n = 2, default = 0),
-           lead1_maxgcd = lead(maxgcd, default = 0),
-           lead2_maxgcd = lead(maxgcd, n = 2, default = 0),
-           diff_latext = c(0, diff(latext)),
-           lag1_latext = lag(latext, default = 0),
-           lag2_latext = lag(latext, n = 2, default = 0),
-           lead1_latext = lead(latext, default = 0),
-           lead2_latext = lead(latext, n = 2, default = 0),
-           diff_area = c(0, diff(area)),
-           lag1_area = lag(area, default = 0),
-           lag2_area = lag(area, n = 2, default = 0),
-           lead1_area = lead(area, default = 0),
-           lead2_area = lead(area, n = 2, default = 0),
-           diff_temp = c(0, diff(temp)),
-           lag1_temp = lag(temp, default = 0),
-           lag2_temp = lag(temp, n = 2, default = 0),
-           lead1_temp = lead(temp, default = 0),
-           lead2_temp = lead(temp, n = 2, default = 0)) %>%
+    mutate(diff1_mst = c(0, diff(mst)),
+           diff2_mst = ifelse(n() > 1,
+                              c(rep(0, 2), diff(mst, lag = 2)),
+                              rep(0, n())),
+           diff3_mst = ifelse(n() > 2,
+                              c(rep(0, 3), diff(mst, lag = 3)),
+                              rep(0, n())),
+           diff4_mst = ifelse(n() > 3,
+                              c(rep(0, 4), diff(mst, lag = 4)),
+                              rep(0, n())),
+           lag1_mst = dplyr::lag(mst, n = 1, default = 0),
+           lag2_mst = dplyr::lag(mst, n = 2, default = 0),
+           lag3_mst = dplyr::lag(mst, n = 3, default = 0),
+           lag4_mst = dplyr::lag(mst, n = 4, default = 0),
[…]

### 2019-06-05 R/02_fit_model.r
moved to mst over maxgcd. issue with rank deficiency

@@ -18,6 +18,31 @@ source(here('R', 'helper04_stan_utility.r'))
 # important constants
 options(mc.cores = parallel::detectCores())
 
+
+# get data in
+counti <- read_rds(here('data', 'counting.rds'))
+counti_restrict_time <- read_rds(here('data', 'counting_restrict_time.rds'))
+counti_restrict_local <- read_rds(here('data', 'counting_restrict_local.rds'))
+
+# transform the data for analysis
+counti_trans <- prepare_analysis(counti)
+counti_rt_trans <- prepare_analysis(counti_restrict_time)
+counti_rl_trans <- prepare_analysis(counti_restrict_local)
+
+## test eff
+#test_maxgcd <- 
+#  stan_glmer(formula = event ~ maxgcd + (1 | fact_mybin / fossil_group),
+#             data = counti_trans,
+#             family = 'binomial')
+#
+#test_mst <- 
+#  stan_glmer(formula = event ~ mst + (1 | fact_mybin / fossil_group),
+#             data = counti_trans,
+#             family = 'binomial')
+#
+#loo(test_maxgcd)
+#loo(test_mst)
+
 #priors <- c(set_prior('normal(-2, 1)', class= 'Intercept'),
 #            set_prior('normal(0, 1)', class = 'b'),
 #            set_prior('normal(-1, 1)', class = 'b', coef = 'maxgcd'),
@@ -37,30 +62,37 @@ options(mc.cores = parallel::detectCores())
 #              prior = priors)
 
 # past and vary
-form <- formula(event ~ temp + lag1_temp + maxgcd + diff_maxgcd 
-                + (1 | fact_relage/fossil_group)
-                + (1 + temp + lag1_temp + maxgcd + diff_maxgcd | 
-                   fact_mybin/fossil_group))
+form <- 
+  formula(event ~ 
+          temp + lag1_temp + mst + diff1_mst + 
+          (temp + lag1_temp + mst + diff1_mst | fact_mybin/fossil_group) +
+          (1 | fact_relage/fossil_group))
 
 # past but no vary
-form2 <- update(form, ~ . - (1 + temp + lag1_temp + maxgcd + diff_maxgcd | 
-                             fact_mybin/fossil_group) 
-                      + (1 | fact_mybin/fossil_group))
+form2 <- 
+  formula(event ~ 
+          temp + lag1_temp + mst + diff1_mst +
+          (1 | fact_mybin/fossil_group) +
[…]

### 2019-06-05 R/01_inspect_data.r
safe diff function

@@ -1,6 +1,6 @@
 library(pacman)
 
-p_load(tidyverse, furrr, here,
+p_load(tidyverse, furrr, here, corrr,
        arm, rstanarm, bayesplot,
        scales, ggridges, viridis,
        pROC)
@@ -9,6 +9,7 @@ source(here('R', 'helper01_process_foo.r'))
 source(here('R', 'helper02_plot_foo.r'))
 source(here('R', 'helper03_misc_foo.r'))
 source(here('R', 'helper04_stan_utility.r'))
+source(here('R', 'helper07_diffsafe.r'))
 
 # important constants
 theme_set(theme_bw())
@@ -75,7 +76,14 @@ appear_graph <-
 ggsave(filename = here::here('results', 'figure', 'appear_time_bin.png'),
        plot = appear_graph, width = 10, height = 6)
 
-
+# correlations 
+mst_corr <- 
+  counti_trans %>%
+  dplyr::select(mst, diff1_mst:diff4_mst) %>%
+  correlate() %>%
+  shave() %>%
+  rplot(print_cor = TRUE)
+ggsave(filename = here::here('results', 'figure', 'mst_correlation.png'))
 
 
 # looking at the temperature data

### 2019-06-05 R/helper01_process_foo.r

@@ -41,72 +41,43 @@ prepare_analysis <- function(x, fg = NULL) {
     mutate_at(.vars = vars(mst, ncell, latext, maxgcd, area, temp),
               .funs = ~ scale(log1p(.x))) %>%
     group_by(fullname) %>%
-    mutate(diff1_mst = c(0, diff(mst)),
-           diff2_mst = ifelse(n() > 1,
-                              c(rep(0, 2), diff(mst, lag = 2)),
-                              rep(0, n())),
-           diff3_mst = ifelse(n() > 2,
-                              c(rep(0, 3), diff(mst, lag = 3)),
-                              rep(0, n())),
-           diff4_mst = ifelse(n() > 3,
-                              c(rep(0, 4), diff(mst, lag = 4)),
-                              rep(0, n())),
+    mutate(
+           diff1_mst = diff_safe(x = mst, lag = 1),
+           diff2_mst = diff_safe(x = mst, lag = 2),
+           diff3_mst = diff_safe(x = mst, lag = 3),
+           diff4_mst = diff_safe(x = mst, lag = 4),
            lag1_mst = dplyr::lag(mst, n = 1, default = 0),
            lag2_mst = dplyr::lag(mst, n = 2, default = 0),
            lag3_mst = dplyr::lag(mst, n = 3, default = 0),
            lag4_mst = dplyr::lag(mst, n = 4, default = 0),
-           diff1_maxgcd = c(0, diff(maxgcd)),
-           diff2_maxgcd = ifelse(n() > 1,
-                                 c(rep(0, 2), diff(maxgcd, lag = 2)),
-                                 rep(0, n())),
-           diff3_maxgcd = ifelse(n() > 2,
-                                 c(rep(0, 3), diff(maxgcd, lag = 3)),
-                                 rep(0, n())),
-           diff4_maxgcd = ifelse(n() > 3,
-                                 c(rep(0, 4), diff(maxgcd, lag = 4)),
-                                 rep(0, n())),
+           diff1_maxgcd = diff_safe(x = maxgcd, lag = 1),
+           diff2_maxgcd = diff_safe(x = maxgcd, lag = 2),
+           diff3_maxgcd = diff_safe(x = maxgcd, lag = 3),
+           diff4_maxgcd = diff_safe(x = maxgcd, lag = 4),
            lag1_maxgcd = dplyr::lag(maxgcd, default = 0),
[…]

### 2019-06-05 R/helper07_diffsafe.r

@@ -1,0 +1,18 @@
+#' Safe differencing with uncertain vector length
+#'
+#' What it says on the tin.
+#'
+#' @param x numerical vector
+#' @param lag difference lag
+#' @return vector of differences of appropriate length for tibble
+diff_safe <- function(x, lag) {
+  n <- length(x)
+  if(n < lag) {
+    out <- rep(0, n)
+  } else {
+    out <- c(rep(0, lag), diff(x, lag = lag))
+  }
+  out
+}
+
+

### 2019-06-05 R/02_fit_model.r
mst and diffs to diff3

@@ -14,6 +14,7 @@ library(here)
 source(here('R', 'helper01_process_foo.r'))
 source(here('R', 'helper03_misc_foo.r'))
 source(here('R', 'helper04_stan_utility.r'))
+source(here('R', 'helper07_diffsafe.r'))
 
 # important constants
 options(mc.cores = parallel::detectCores())
@@ -64,14 +65,14 @@ counti_rl_trans <- prepare_analysis(counti_restrict_local)
 # past and vary
 form <- 
   formula(event ~ 
-          temp + lag1_temp + mst + diff1_mst + 
-          (temp + lag1_temp + mst + diff1_mst | fact_mybin/fossil_group) +
+          temp + lag1_temp + mst + diff1_mst + diff2_mst + diff3_mst +
+          (temp + lag1_temp + mst + diff1_mst + diff2_mst + diff3_mst | fact_mybin/fossil_group) +
           (1 | fact_relage/fossil_group))
 
 # past but no vary
 form2 <- 
   formula(event ~ 
-          temp + lag1_temp + mst + diff1_mst +
+          temp + lag1_temp + mst + diff1_mst + diff2_mst + diff3_mst +
           (1 | fact_mybin/fossil_group) +
           (1 | fact_relage/fossil_group))
 
@@ -103,11 +104,11 @@ part_glmer <- partial(stan_glmer,
 
 # some models have differen't priors
 list_part_glmer <- 
-  list(partial(part_glmer, prior = normal(location = c(0, 0, -1, 0),
-                                          scale = rep(1, 4),
+  list(partial(part_glmer, prior = normal(location = c(0, 0, -1, 0, 0, 0),
+                                          scale = rep(1, 6),
                                           autoscale = FALSE)),
-       partial(part_glmer, prior = normal(location = c(0, 0, -1, 0),
-                                          scale = rep(1, 4),
+       partial(part_glmer, prior = normal(location = c(0, 0, -1, 0, 0, 0),
+                                          scale = rep(1, 6),
                                           autoscale = FALSE)),
        partial(part_glmer, prior = normal(location = c(0, -1),
                                           scale = rep(1, 2),

### 2019-06-10 doc/paper.tex
working through reviewer comments

@@
 Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past.
-To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
-Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate) the improvement in predictive power over models with constant effects and no historical covariates is very minor.
+To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
+We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
+Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate), the improvement in predictive power over models with constant effects and no historical covariates is very minor.
 Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time.
@@
 Predicting which species will go extinct is difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented.
-This has led to the suggestions that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}.
+This has led to the suggestion that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}.
 The fossil record preserves information about the full histories, include ultimate extinction, of thousands of lineages, and this information can help to augment the shorter-term higher-resolution data used to make risk assessments of extant taxa.
@@
 Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
-This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.
+This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how capable we are at predicting future extinction events.
 
-Here we ask how precise risk predictions based on fossil data might be.
+We ask how accurate risk predictions based on fossil data are when estimating future extinctions.
 Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions.
@@
 
-Extinction intensity (average rate) and selectivity (difference in risk between taxa) vary through time and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}.
+Extinction intensity (average rate) and selectivity (difference in risk among taxa) vary through time, and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}.
 This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
@@
 These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}.
-This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.
-The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores from which the occurrences are recorded.
+This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomy synonymization strategy for delaing wiht 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible.
 Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.
@@
 We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
-This occurrence-based dataset includes calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya).
-Occurrences were filtered to include only those species with first occurrence was no more than 63 Mya.
-This filtering criterion excludes taxa that survived the K/Pg extinction or arose during its ecologically unusual recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).
+This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic foraminifera, and radiolarians.
+Occurrences were filtered to include only those species who's first occurrence is no earlier than 63 Mya (millions of years ago).
+This filtering criterion excludes taxa that survived the K/Pg extinction or arose during this recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).
 
-All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence as listed in the Neptune Database.
+All fossil occurrences were assigned to 1 My (million year) bins based on the estimated age of the fossil occurrence as listed in the Neptune Database.
 After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred.
-Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e.
-the Earth) between any two occurrences of that species; this distance was measured in kilometers.
+Geographic range was calculated as the minimum spanning tree distance between all observations of that taxon during that temporal bin; this distance was measured in kilometers.
+Minimum spanning tree distance was caluclated using the \texttt{GeoRange} package for R \citep{GeoRange}.
 
@@
 We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
-Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My from 0 to 63 Mya.
+Estimating temperature over long periods of time from Mg/Ca ratios also suffers from complications because Mg/Ca based temperature estimates requires benthic foram and seawater Mg/Ca isotope ratio information.
+Because seawater Mg/Ca ratio has changed over time, the method to estimate temperature used \citet{Cramer2011} attempts to account for this unknown in order to obtain accurate, albeit uncertain, temperature estimates.
+Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My interval between 0 to 63 Mya.
 The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.
+
+These temperature estimates are measured from benthic forams, and is an estimate of deep water temperature.
+The organisms included in this study are all planktonic, so these temperature estimates do not necessarily represent the actual environment these organisms inhabit.
+However, detailed ocean surface temperature estimates do not exist for the entirety of the Cenozoic.
+So, while deepwater temperature estimates are not ideal for understanding the extinction risk of planktonic taxa, we interpret our estimate of temperature as reflecting global climate state as opposed to a taxon-specific ecological descriptor.
 
@@
 A complete description of the statistical model used in this analysis is available in Section \ref{sec:model_desc}.
-Additionally, the full description of how these models were implemented and coded, including choice of priors, is available Section \ref{sec:model_est}.
+Additionally, the full description of how these models were implemented and coded, including choice of priors, is available in Section \ref{sec:model_est}.
 
@@
     C & Constant effects, no historical cov.
-& \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ age\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}})} \\
+& \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ time\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}}) +\\(1 $|$ age\tnote{g}\phantom{\textsuperscript{g}}/phylum)} \\
     V & Varying effects, no historical cov.
-& \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ phylum) + (1 $|$ age/phylum)} \\
+& \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ time/phylum) + (1 $|$ age/phylum)} \\
     CP & Constant effects, historical cov.
-& \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff\tnote{g}\phantom{\textsuperscript{g}} +\\temp + temp\_lag\tnote{h}\phantom{\textsuperscript{h}} +\\(1 $|$ age/phylum)} \\
+& \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ + range\_diff1\tnote{g}\phantom{\textsuperscript{h}} + range\_diff2\tnote{h}\phantom{\textsuperscript{h}} + range\_diff3\tnote{h}\phantom{\textsuperscript{h}} + \\temp + temp\_lag\tnote{i}\phantom{\textsuperscript{i}} +\\(1 $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
     VP & Varying effects, historical cov.
-& \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff +\\temp + temp\_lag +\\(1 + range + range\_diff +\\temp + temp\_lag $|$ phylum) +\\(1 $|$ age/phylum)} \\
+& \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff1 + range\_diff2 + range\_diff3 + \\temp + temp\_lag +\\(1 + range + range\_diff1 + range\_diff2 + range\_diff3 +\\temp + temp\_lag $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
     \hline
@@
   \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
-  \item[e] Species are at observation in millions of years.
+  \item[e] Time of observation.
   \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcareous nannoplankton).
-  \item[g] Change in geographic range since last observation.
-  \item[h] Temperature at previous observation.
+  \item[g] Age at observation.
+  \item[h] Change in geographic range since last observation (number indicates how lags).
+  \item[i] Temperature at previous observation.
   \end{tablenotes}
@@
 
-In-sample forecasting means we are estimating how well our model predicts extinction probability for observations that that model was fit to.
-This is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
+In-sample forecasting is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
+This is a type of sensitivity analysis for understanding the quality of model fit to the data, and how well our model predicts extinction probability for observations that we used to fit out model initially.
 In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}.
@@
 Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold.
-Finally, the model is fit to the first four folds combined and then is used to forecast the fifth fold.
-The results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.
+Next, the model is fit to the first four folds combined and then is used to forecast the fifth fold.
+This process continues until $k - 1$ folds are included in the fitting the model and the final fold is predicted from this model.
+When combined, the results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.
+
+Some readers may be familiar with metrics such as AIC, WAIC, or LOOIC for comparing model fit.
+These metrics are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL}.
+Because we've estimated out-of-sample performance using cross-validation, these measures would be redundant, and in the case on single value measures such as AIC, much less informative than an entire posterior distributions.
+When out-of-sample estimates from a complex model are noticeably worse than out-of-sample estimates from a simpler model, than this is evidence for the complex model being overfit to the data \citep{ESL}.
+In contrast a high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).
 
@@
 AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of a randomly selected extinct-extant species pair \citep{Fawcett2006,Mason2002}.
-AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we interpret values between 0.7 and 0.8 could then be considered ``fair,'' and values between 0.6 and 0.7 as ``poor.''
+AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we consider values between between 0.7 and 0.8 as ``fair,'' and values between 0.6 and 0.7 as ``poor.''
 
@@
  \end{subfigure}
- \caption{In-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) AUC estimates for each of our four models.
+ \caption{In-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) area under the receiver operating charactered curve (AUC) estimates for each of our four models.
 These estimates are calculated from the models posterior predictive distribution or from predictions made to new data, respectively.
-Models with a higher AUC values indicate better performance over models with lower AUC values.
+Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in our dataset.
+Models with higher AUC values indicate better performance over models with lower AUC values.
 AUC is bounded between 0.5 and 1.
-See Table \ref{tab:model_def} for a description of each of the four models.}
+See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
  \label{fig:auc_compare}
@@
 
-Depending on the taxon-model combination, there are between zero and 4 time intervals where our posterior distribution of in-sample AUC has a median value less than or equal to 0.5 (Fig.
+The in-sample forecasts from our four models over time are virtually identical within taxa (Fig.
 \ref{fig:auc_taxon_time}).
-However, this pattern is absent for the posterior estimates of in-sample AUC for Foraminfera and Radiolaria as fit by the VP model.
-In contrast, there are fewer periods of low model performance for calcareous nannoplankton and Dinoflagellates as estimated from our VP model than in those estimates from the other three model variations.
+Our in-sample forecasts for Diatoms are the weakest of the taxonomic groups as all four models have an equal number of intervals with no predictive power (AUC approximately 0.5).
+In contrast, our best in-sample forecast results are for Radiolarians where there is at most 1 interval with almost no predictive power.
+Ultimately, our in-sample forecasts are virtually identical across models.
 
@@
 In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.
+The grey intervals mark the geologic ages of the Cenozoic.
 See Table \ref{tab:model_def} for a description of each of the four models.}
@@
 Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}.
-The resulting distribution when all folds are combined is highly multimodal as expected given that they are fit to and estimated from different data sets with different numbers of observations \citep{ESL}.
-This multimodality increases with model complexity (Fig.
-\ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time, allowing a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
-
-Comparison the in-sample AUC estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig.
+Comparing the in-sample AUC estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig.
 \ref{fig:auc_hist}, \ref{fig:fold_auc}).
@@
 This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts.
-The major exception to this pattern are our estimates for the Dinoflagellates, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random.
+The major exception to this pattern are our estimates for the diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random.
 The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
@@
 The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.
+The grey intervals mark the geologic ages of the Cenozoic.
 See Table \ref{tab:model_def} for a description of each of the four models.}
@@
 
-We find that all of our models are expected to correctly forecasting the rank order of extinction probability for a randomly selected extinct-extant pair of future observations between 70\% to 80\% of the time (Fig.
+We find that all of our models are expected to correctly forecast which species of randomly selected extinct-extant pair is more likely to go extinct between 70\% to 80\% of the time (Fig.
 \ref{fig:fold_auc}).
@@
 However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to variation in extinction intensity over the Cenozoic.
-A extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
+An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
 
@@
 \ref{fig:fold_auc}).
-Thus even though the VP model has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant.
+Thus even though the VP model has a statistically greater AUC for in-sample forecasts than the other three models, this result has no practical significance because the predictions are functionally identical.
 The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance.
 Thus, including geographic range trajectory results in only very minor improvements in forecasting accuracy.
-This contrasts with findings from coral genera \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.
+This contrasts with previous findings using geological stages as their time unit \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.
 
@@
 Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity.
-The principal reason we were not able to include more biological information in our models is that we lack suitably detailed life history or ecological information for many marine micro- and nannoplankton.
+The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study.
 Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}.
-However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here.
+However, comparable information does not exist for all foram species, nor does this type of data exist for the other three taxonomic groups studied here.
 Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.
@@
 In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of taxonomic, geographic, and historical predictors.
-Our results are directly comparable to conservation determinations because both are expressed in terms of a continuum of risk, from most to least.
+While change to in geographic range and past temperature information is not found to greatly improve our ability to estimate future extinction risk, we do find that the immediate past is a modest predict of the immediate future.
+These results are comparable to conservation determinations because both are expressed in terms of an relative risk.
 The results of this simple exercise suggest that conservation decisions would be bolstered by including fossil data.

### 2019-06-12 doc/paper.tex
export foram genera

@@
  \end{subfigure}
- \caption{In-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) area under the receiver operating charactered curve (AUC) estimates for each of our four models.
+ \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation.
+We used area under the receiver operating charactered curve (AUC) estimates for each of our four models as our measure of performance.
 These estimates are calculated from the models posterior predictive distribution or from predictions made to new data, respectively.
@@
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
- \caption{Comparison of in-sample forecasting performance measured by AUC for each of the four models, arranged over time and by taxonomic group.
+ \caption{Understanding model adequacy over time and taxonomic group by comparing in-sample forecasting performance measured by AUC for each of the four models.
 These estimates reflect each model's fit to the various taxonomic groups over time.
@@
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
- \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
+ \caption{Comparison of our models ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
 The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds.
@@
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
- \caption{Comparison between our out-of-sample forecasts and in-sample forecasts for all models.
+ \caption{Comparing our models' ability to forecast out-of-sample extinction compared to our models' in-sample forecasts.
+Congruence between these measures indicates that our models are not necessarily overfit to the data.
 This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}.

### 2019-06-12 R/01_inspect_data.r

@@ -86,6 +86,22 @@ mst_corr <-
 ggsave(filename = here::here('results', 'figure', 'mst_correlation.png'))
 
 
+# foram genera to check for planktonic vs benthic
+foram_genera <- 
+  counti_trans %>%
+  filter(fossil_group == 'F') %>%
+  dplyr::select(fullname) %>%
+  distinct()
+write_csv(x = foram_genera,
+          path = here::here('results', 'foram_genera.csv'))
+foram_genera_time <- 
+  counti_trans %>%
+  filter(fossil_group == 'F') %>%
+  dplyr::select(fullname, mybin) %>%
+  distinct()
+write_csv(x = foram_genera_time,
+          path = here::here('results', 'foram_genera_time.csv'))
+
 # looking at the temperature data
 mgca <- read_tsv(here('data', 'cramer', 'cramer_temp.txt'))
 names(mgca) <- str_to_lower(names(mgca))

### 2019-06-12 doc/paper.tex
trying to understand data

@@
 
+We also included how a taxon's geographic range has changed since its last three observation times.
+We measured this change in geographic range by calculating the difference in geographic range between an observation and that taxon's three previous occurrences.
+If there are not enough previous observations to calculate, then that value is recorded as a 0.
+These differences were calculated after minimum spanning tree distance was transformed and standardized (see Section \ref{sec:data_desc}).
+
 Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}.
@@
 
-These temperature estimates are measured from benthic forams, and is an estimate of deep water temperature.
-The organisms included in this study are all planktonic, so these temperature estimates do not necessarily represent the actual environment these organisms inhabit.
-However, detailed ocean surface temperature estimates do not exist for the entirety of the Cenozoic.
-So, while deepwater temperature estimates are not ideal for understanding the extinction risk of planktonic taxa, we interpret our estimate of temperature as reflecting global climate state as opposed to a taxon-specific ecological descriptor.
+We also included the global temperature from the previous time that taxon was observed.
+If there are not enough previous observations to calculate, then that value is recorded as a 0.
+This lag was calculated after global temperature was transformed and standardized (see Section \ref{sec:data_desc}).
+
+Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature.
+The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit.
+Ideally, we would have detailed ocean surface water temperature estimates for the entire globe for the entire Cenozoic.
+Unfortunately, that type of data does not exist.
+So, we interpret our temperature estimates as reflecting the global climate state that an organisms experiences, and not as a descriptor of that taxon's local environmental ecology.
 
@@
 We always included species age at time of observation (i.e.
-observed prior duration) as a non-nested varying intercept term.
+observed prior duration) as a varying-intercept term.
 This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.
+The effect of species age is allowed to vary by taxonomic group.
+
+Similarly, we included time of obseration as an additional varying-intercept term to account for changes average global extinction risk over time that are not related to the covariates included in this model.
+This varying-intercept is further allowed to vary by taxonomic group.
+This varying-intercept term allows us to tease apart the differences in extinction risk associated with time of observation versus age since first observation.
+An important note is that for our V and VP models, the covariation between this varying-intercept and the varying-slopes of our covariates is explicitly modeled (see Section \ref{sec:model_desc}).
 
@@
 
-In-sample forecasting is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.
-This is a type of sensitivity analysis for understanding the quality of model fit to the data, and how well our model predicts extinction probability for observations that we used to fit out model initially.
-In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}.
+In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to recapitulate the data to which it was fit.
+Posterior predictive checks are a type of sensitivity analysis because we are checking the quality of model's fit to the data.
+If our models have poor in-sample forecasting performance, than our models are not adequate descriptors of the data and will most likely make poor out-of-sample predictions.
+In-sample forecasting measures, however, are not the same as understanding our models' ability to forecast future extinctions or if our models are overfit to our data and produce biased out-of-sample estimates \citep{ESL}.
 
@@
 
-\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}.
-First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e.
+\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016,ESL}.
+Prior to cross-valdiation, the data is divided into \textit{k} nearly even segments or folds -- for a time series, this means the data is divided into \textit{k} continuous sequences.
+Next, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e.
 the future).
@@
 When combined, the results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.
+In 5-fold cross validaition, the data is divided into five folds the cross-validation procedure yields predictions for four of the folds.
 
-Some readers may be familiar with metrics such as AIC, WAIC, or LOOIC for comparing model fit.
-These metrics are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL}.
-Because we've estimated out-of-sample performance using cross-validation, these measures would be redundant, and in the case on single value measures such as AIC, much less informative than an entire posterior distributions.
-When out-of-sample estimates from a complex model are noticeably worse than out-of-sample estimates from a simpler model, than this is evidence for the complex model being overfit to the data \citep{ESL}.
-In contrast a high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).
+Cross-validation is a procedure for estimating a models expected out-of-sample error, which is a similar goal to metrics such as AIC or WAIC \citep{ESL,Gelman2013d}.
+Measures like AIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}.
+Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, then out-of-sample predictions will be very biased and inaccurate \citep{ESL}.
+Similarly, a high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).
+Cross-validation is preferable to simple metrics such as AIC because instead of a single value, we have an entire posterior distribution of estimates.
 
@@
  \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation.
-We used area under the receiver operating charactered curve (AUC) estimates for each of our four models as our measure of performance.
-These estimates are calculated from the models posterior predictive distribution or from predictions made to new data, respectively.
+The area under the receiver operating charactered curve (AUC) was calculated for each model.
+these estimates are calculated from the models posterior predictive distribution (\ref{fig:auc_hist}) or from predictions made to new data (\ref{fig:fold_auc}), respectively.
 Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in our dataset.
@@
 The grey intervals mark the geologic ages of the Cenozoic.
-See Table \ref{tab:model_def} for a description of each of the four models.}
+See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
  \label{fig:auc_taxon_time}
@@
 The grey intervals mark the geologic ages of the Cenozoic.
-See Table \ref{tab:model_def} for a description of each of the four models.}
+See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
  \label{fig:fold_auc_taxon_time}
@@
 A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast.
-See Table \ref{tab:model_def} for a description of each of the four models.}
+See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
  \label{fig:oos_ins_diff}
@@
 \begin{verbatim}
-form <- event ~ range + range_diff + temp + temp_lag + 
+form <- event ~ 
+        range + range_diff1 + range_diff2 + range_diff3 + 
+        temp + temp_lag + 
         (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
@@
       family = 'binomial',
-      prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
+      prior = normal(c(-1, 0, 0, 0, 0, 0), rep(1, 6), autoscale = FALSE), 
       prior_intercept = normal(-2, 5, autoscale = FALSE), 
@@
       set_prior('lkj(1)', class = 'cor'))
-form <- bf(event ~ range + range_diff + temp + temp_lag +
+form <- bf(event ~ 
+           range + range_diff1 + range_diff2 + range_diff3 + 
+           temp + temp_lag +
            (1 + range + range_diff + temp + temp_lag | mybin/phylum) +

### 2019-06-12 R/01_inspect_data.r

@@ -86,21 +86,28 @@ mst_corr <-
 ggsave(filename = here::here('results', 'figure', 'mst_correlation.png'))
 
 
-# foram genera to check for planktonic vs benthic
+#foram genera to check for planktonic vs benthic
 foram_genera <- 
   counti_trans %>%
   filter(fossil_group == 'F') %>%
   dplyr::select(fullname) %>%
+  transmute(genus = str_extract(fullname, '^[^_]+(?=_)')) %>%
   distinct()
 write_csv(x = foram_genera,
           path = here::here('results', 'foram_genera.csv'))
+
 foram_genera_time <- 
   counti_trans %>%
-  filter(fossil_group == 'F') %>%
+  filter(fossil_group == 'F',
+         mybin == 1) %>%
   dplyr::select(fullname, mybin) %>%
+  mutate(genus = str_extract(fullname, '^[^_]+(?=_)')) %>%
+  dplyr::select(genus, mybin) %>%
   distinct()
 write_csv(x = foram_genera_time,
           path = here::here('results', 'foram_genera_time.csv'))
+
+
 
 # looking at the temperature data
 mgca <- read_tsv(here('data', 'cramer', 'cramer_temp.txt'))

### 2019-06-14 R/00_data_process.r
regenerated all figures with large labels

@@ -11,6 +11,9 @@ source(here('R', 'helper03_misc_foo.r'))
 # parallel set up
 plan(multiprocess)
 
+# useful graphical constants
+theme_set(theme_bw(base_size = 20))
+
 
 # process timescale information
 gts <- read_csv("https://raw.githubusercontent.com/japhir/stratPlot/master/GTS_colours.csv") %>%

### 2019-06-14 R/01_inspect_data.r

@@ -12,8 +12,10 @@ source(here('R', 'helper04_stan_utility.r'))
 source(here('R', 'helper07_diffsafe.r'))
 
 # important constants
-theme_set(theme_bw())
 plan(multiprocess)
+
+# useful graphical constants
+theme_set(theme_bw(base_size = 20))
 
 
 # get data in

### 2019-06-14 R/04_inspect_model.r

@@ -13,6 +13,9 @@ source(here('R', 'helper07_diffsafe.r'))
 # important constants
 plan(multiprocess)
 
+# useful graphical constants
+theme_set(theme_bw(base_size = 20))
+
 # get data in
 counti <- read_rds(here('data', 'counting.rds'))
 
@@ -23,96 +26,134 @@ counti_trans <- prepare_analysis(counti)
 disc_fit <- read_rds(here('results', 'disc_fit.rds'))
 disc_best <- disc_fit[[1]]
 
-# check how much information learned
-priors <- prior_summary(disc_best)
-prmn <- priors$prior$location
-prsc <- priors$prior$scale
-
-posc <- disc_best %>% 
-  spread_draws(maxgcd,
-               diff_maxgcd,
-               temp,
-               lag1_temp) %>% 
-clean_names %>% 
-dplyr::summarize_at(vars(-chain, -iteration), funs(sd(.)))
-
-# 1 = lot of shrinkage, 0 = no shrinkage
-posterior_shrinkage <- 1 - (posc / prsc) ^ 2
-
+## check how much information learned
+#priors <- prior_summary(disc_best)
+#prmn <- priors$prior$location
+#prsc <- priors$prior$scale
+#
+#posc <- disc_best %>% 
+#  spread_draws(maxgcd,
+#               diff_maxgcd,
+#               temp,
+#               lag1_temp) %>% 
+#clean_names %>% 
+#dplyr::summarize_at(vars(-chain, -iteration), funs(sd(.)))
+#
+## 1 = lot of shrinkage, 0 = no shrinkage
+#posterior_shrinkage <- 1 - (posc / prsc) ^ 2
 
 # posterior intervals
-interval_est <- 
+effect_group <- 
   disc_best %>%
-  spread_draws(`(Intercept)`, 
-               temp,
-               lag1_temp,
+  gather_draws(`(Intercept)`, 
                mst,
                diff1_mst,
                diff2_mst,
-               diff3_mst)
-%>%
-median_qi(.width = c(0.8, 0.5))
-
-effect_eye <- disc_best %>%
-  gather_draws(`(Intercept)`, 
-               maxgcd, 
-               diff_maxgcd,
-               #`maxgcd:diff_maxgcd`,
+               diff3_mst,
                temp,
                lag1_temp) %>%
-ggplot(aes(y = .variable, x = .value)) +
-geom_halfeyeh(.width = c(0.8, 0.5))
[…]

### 2019-06-14 R/06_inspect_predict_model.r

@@ -10,9 +10,13 @@ source(here('R', 'helper03_misc_foo.r'))
 source(here('R', 'helper04_stan_utility.r'))
 source(here('R', 'helper05_roc_utility.r'))
 source(here('R', 'helper06_geotime.r'))
+source(here('R', 'helper07_diffsafe.r'))
 
 # important constants
 plan(multiprocess)
+
+# useful graphical constants
+theme_set(theme_bw(base_size = 20))
 
 # get data in
 counti_fold <- read_rds(here('data', 'counting.rds')) %>%

### 2019-06-14 R/07_predict_quality.r

@@ -11,9 +11,13 @@ source(here('R', 'helper03_misc_foo.r'))
 source(here('R', 'helper04_stan_utility.r'))
 source(here('R', 'helper05_roc_utility.r'))
 source(here('R', 'helper06_geotime.r'))
+source(here('R', 'helper07_diffsafe.r'))
 
 # important constants
 plan(multiprocess)
+
+# useful graphical constants
+theme_set(theme_bw(base_size = 20))
 
 # get data in
 counti <- 

### 2019-06-14 R/helper02_plot_foo.r

@@ -65,7 +65,7 @@ plot_georange_compare <- function(data) {
 #' @return a ggplot object
 plot_taxon_covariate_time <- function(disc_best) {
   comp_const <- disc_best %>%
-    spread_draws(maxgcd, diff_maxgcd, temp, lag1_temp)
+    spread_draws(mst, diff1_mst, diff2_mst, diff3_mst, temp, lag1_temp)
 
   comp_var <- disc_best %>%
     spread_draws(b[i, f]) %>%
@@ -82,20 +82,24 @@ plot_taxon_covariate_time <- function(disc_best) {
   # first element is just temporal effect
   core <- full_join(comp_const, cv[[1]], 
                     by = c('.chain', '.iteration', '.draw')) %>%
-    mutate(eff_maxgcd = maxgcd.x + maxgcd.y,
-           eff_diff_maxgcd = diff_maxgcd.x + diff_maxgcd.y,
+    mutate(eff_mst = mst.x + mst.y,
+           eff_diff1_mst = diff1_mst.x + diff1_mst.y,
+           eff_diff2_mst = diff2_mst.x + diff2_mst.y,
+           eff_diff3_mst = diff3_mst.x + diff3_mst.y,
            eff_temp = temp.x + temp.y,
            eff_lag1_temp = lag1_temp.x + lag1_temp.y) %>%
     dplyr::select(.chain, .iteration, .draw, f, age, 
-                eff_maxgcd, eff_diff_maxgcd,
-                eff_temp, eff_lag1_temp) %>%
+                  eff_mst, eff_diff1_mst, eff_diff2_mst, eff_diff3_mst,
+                  eff_temp, eff_lag1_temp) %>%
     arrange(.chain, .iteration, .draw, age)
 
   # need to confirm correct line-up
   by_taxon <- map(cv[-1], ~ full_join(core, .x, by = c('.chain', '.iteration', 'age'))) %>%
     map(., ~ .x %>%
-        mutate(taxon_eff_maxgcd = eff_maxgcd + maxgcd,
-               taxon_eff_diff_maxgcd = eff_diff_maxgcd + diff_maxgcd,
+        mutate(taxon_eff_mst = eff_mst + mst,
+               taxon_eff_diff1_mst = eff_diff1_mst + diff1_mst,
+               taxon_eff_diff2_mst = eff_diff2_mst + diff2_mst,
+               taxon_eff_diff3_mst = eff_diff3_mst + diff3_mst,
[…]

### 2019-06-18 doc/paper.tex
more edits to paper. figures in supplement

@@
 We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
-Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate), the improvement in predictive power over models with constant effects and no historical covariates is very minor.
+We find that our model that includes our historical covariates (e.g.
+change in geographic range) and allows covariate effects to vary over time to be overfit to our data, yielding biased forecasts of future extinction events.
+In contrast, our models which include either historical covariates or which allow for covariate effects to vary over time yield equivalent forecasts.
 Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time.
@@
 
- %risk assessments for extant species can be improved by examining past extinction patterns and by using palaeontological records to establish geographic range and abundance trajectories on geological timescales.
- %We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
- %Such suggestions prompt two key questions: (1) At a given timescale, to what extent are extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)?
-(2)
- %Our results imply that at million-year timescales extinction risk trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
- %However, models based on present geographic ranges and past range-risk relationships are likely to be moderately successful at predicting future extinction/survival outcomes. 
- %In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic.
- %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.
- %We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance.
- %Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
- %However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction.
- %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
- %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
- %The greatest source of variance in survival probability is the timing of observation. 
- %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
- %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
- %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range.
- %These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales.
- %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.
- %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
- %The results of this study reinforce the importance of the promise of palaeontology and using the past to predict the future.
 \end{abstract}
@@
 
+The focus of this study is on understanding how well our models forecast future extinction events by comparing our in-sample and out-of-sample forecast estimates.
+A presentation of the posterior estimates for the regression coefficient estimates from our VP model (Table \ref{tab:model_def}) is available our Supplemental Materials (Section \ref{sec:supp_res}).
+
 % ROC model comparison
@@
 
-The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to.
-Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that all of our models have approximately equal in-sample forecasting performance (Fig.
+The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis.
+Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that the parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, while there is substantial overlap in the posterior distributions of the forecasts from the other three models (Fig.
 \ref{fig:auc_hist}).
-The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions.
-Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions.
-All the in-sample AUC estimates from our models are concentrated on an AUC value of 0.77.
-It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.
 
-%When the posterior predictive distributions of the in-sample AUC estimates are presented over time, the similarity in adequacy between the models becomes more apparent (Fig.
-\ref{fig:auc_ts}).
-There are few major or obvious differences in model adequacy between the four models.
-%% ROC model comparison time series
-%\begin{figure}[ht]
-% \centering
-% \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_ts_full}
-% \caption{Comparison between the posterior predictive AUC estimates for each of the time intervals for each of the four models.
-These estimates are reflections of each model's fit to the various time intervals.
-The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend.
-In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
-% \label{fig:auc_ts}
-%\end{figure}
+However, the actual difference in forecast AUC result between the VP model and the other three models is extremely small (0.01 AUC unit), and all the in-sample AUC estimates from our models are concentrated between an AUC value of 0.775 and 0.795 (Fig.
+\ref{fig:auc_hist}).
+This result indicates that the practical difference in performanc between these models might be so small that there is no practical benefit that the VP model over the other three.
+Ultimately, determining which of these models produces the best forecasts of future extinctions requires comparing these in-sample results to our out-of-sample results (see below).
 
-%When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig.
-\ref{fig:auc_taxon}).
-While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig.
-\ref{fig:auc_hist}).
-For example, the difference between the ``past and vary'' model and the others is more pronounced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.
-%\begin{figure}[ht]
-% \centering
-% \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_full}
-% \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group.
-These estimates reflect each model's fit to the various taxonomic groups present in this analysis.
-The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals.
-In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
-% \label{fig:auc_taxon}
-%\end{figure}
-
-The in-sample forecasts from our four models over time are virtually identical within taxa (Fig.
+The in-sample forecasts from our four models over time are broadly similar between taxonomic groups (Fig.
 \ref{fig:auc_taxon_time}).
 Our in-sample forecasts for Diatoms are the weakest of the taxonomic groups as all four models have an equal number of intervals with no predictive power (AUC approximately 0.5).
-In contrast, our best in-sample forecast results are for Radiolarians where there is at most 1 interval with almost no predictive power.
-Ultimately, our in-sample forecasts are virtually identical across models.
+In contrast, our best in-sample forecast results are for Radiolarians where for any of our models there is at most 1 interval with almost no predictive power.
+Ultimately, our in-sample forecasts over time by taxonimc group are broadly consistent between our four models.
 
@@
 Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}.
-Comparing the in-sample AUC estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig.
+Our out-of-sample forecast AUC estimates demonstrate a broader range of results, with AUC estimates ranging between approximately 0.7 and 0.85 (Fig.
 \ref{fig:auc_hist}, \ref{fig:fold_auc}).
-Interestingly, the differences in posterior predictive distributions of AUC between the four model variations is reduced in out-of-sample prediction.
-For example, the VP model no longer has the greatest median AUC of the four models (Fig.
+While our VP model was the model with greatest in-sample forecasting performance (Fig.
+\ref{fig:auc_hist}), this model has the greatest decrease in out-of-sample forecasting performance (Fig.
 \ref{fig:fold_auc}).
-For this reason the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC.
-However, making interpretations based only on the median AUC estimates is incorrect -- our estimates are the full posterior, not individual points.
-Because of this, the four models are effectively indistinguishable in their out-of-sample forecast performance (Fig.
-\ref{fig:fold_auc}).
-
-%A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original).
-Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past.
-However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
-
-%This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time.
-However, this expected out-of-sample performance is approximately the same as the in-sample performance results (Fig.
-\ref{fig:auc_hist}), indicating that our models would yield consistent results when generalized to future extinctions.
-
-%When the posterior predictive distribution of expected out-of-sample AUC is presented as a time series, the similarity between the models is even more apparent (Fig.
-\ref{fig:fold_auc_time}).
-While the width of the credible intervals at various time points varies between the models, the overall picture of expected out-of-sample AUC is almost identical when you compare the models -- periods of relatively better or worse performance map identically between the time series.
-%\begin{figure}[ht]
-% \centering
-% \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time_full}
-% \caption{Comparison of out-of-sample AUC values calculated for each of the million year intervals for each of the four models.
-The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds.
-This presentation decomposes each of the 12-million year folds (Fig.
-\ref{fig:fold_auc}) into the predictions made for each of the million-year intervals.
-The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
-% \label{fig:fold_auc_time}
-%\end{figure}
-%We can also compare expected out-of-sample AUC by taxonomic group for each of the models (Fig.
-\ref{fig:fold_auc_taxon}).
-These comparisons reveal a lot about the differences in predictive potential of the taxonomic groups.
-For example, the posterior predictive distributions of out-of-sample AUC for Foraminifera from all four models are approximately identical.
-In contrast, expected out-of-sample AUC for Radiolaria exhibits the same or similar pattern in relative model performance to the pooled comparisons (Fig.
-\ref{fig:fold_auc}).
-Additionally, we can state that our out-of-sample predictions for calcareous nannoplankton and dinoflagellates are not necessarily as precise as our estimates for Foraminifera or even Radiolaria.
-These results indicate that out-of-sample predictions may be easier for some taxonomic groups than others (e.g.
-Foraminifera versus Dinoflagellates.
-%\begin{figure}[ht]
-% \centering
-% \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_full}
-% \caption{Comparison of out-of-sample AUC values aggregated by taxonomic group for each of the four models.
-Depicted for each taxon-model combination is an aggregate density of all posterior predictive estimates for each of four folds -- cross-validation estimates are commonly multi-modal as each fold presents its own challenges for prediction.
-Beneath these densities is marked the median estimate along with 50\% and 80\% credible intervals.}
-% \label{fig:fold_auc_taxon}
-%\end{figure}
-%
-
-The posterior predictive distribution of expected out-of-sample AUC over time and across taxonomic groups are nearly identical for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}).
+In comparison, the other three models demontrstrate a much smaller difference between in-sample forecast and out-of-sample forecast AUC values which indicates that our VP model is overfit to our data, and that one of the simpler model would be preferable for predicting future extinctions.
+This result means that our model which includes the historical covariates (e.g.
+change in geographic range) and allowing these effects to vary over time produces biased extinciton forecasts.
+Interestingly, including either historical covariates like change in geographic range or allowing the effects of geographic range and other covariate effects to vary over time have approximately equal effect on forecasting future extinction events (Fig.
+\ref{fig:auc_compare}).
 
@@
 
-%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
-%
-%
-%% risk estimate compared to change in geo-range
-%\begin{figure}[ht]
-% \centering
-% \begin{subfigure}{\textwidth}
-%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
-%  \caption{A}
-%  \label{fig:relrisk_range}
-% \end{subfigure}
-%
-% \begin{subfigure}{\textwidth}
-%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
-%  \caption{B}
-%  \label{fig:relrisk_ext}
-% \end{subfigure}
-% \caption{Comparison between species geographic range histories and our estimate of their probability of extinction for each time of observation.
-Geographic range is assumed observed without error.
-Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
-% \label{fig:relrisk_compare}
-%\end{figure}
-
 \section{Discussion}
@@
 \ref{fig:fold_auc}).
-One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations.
+One of the most striking aspect of these results is that our results is that our model that includes historical covariates (e.g.
+change in geographic range) and allows for covariate effects to vary over time is overfit to our data and produces more biased out-of-sample predictions than the other three models (Fig.
+\ref{fig:auc_compare}).
+In comparison, the models that include either the historical covariates or allows covariate effects to vary over time produce less biased out-of-sample forecasts.
 A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original).
 Additionally, a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect.
-However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to variation in extinction intensity over the Cenozoic.
 An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
 
-%This similarity demonstrates that while models with historical covariates are predictors of species extinction have better in-sample performance (Fig.
-\ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig.
-\ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).
-
-We also find that our four models are practically identical in their ability to make in-sample and out-of-sample forecasts.
+We also find that our three of our four models are practically identical in their ability to make in-sample and out-of-sample forecasts.
 Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig.
 \ref{fig:auc_hist}).
-While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig.
-\ref{fig:fold_auc}).
-Thus even though the VP model has a statistically greater AUC for in-sample forecasts than the other three models, this result has no practical significance because the predictions are functionally identical.
-The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance.
-Thus, including geographic range trajectory results in only very minor improvements in forecasting accuracy.
+While the ``best'' model based on in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time, the out-of-sample forecasts from this mode indicate that this model is overfit to our data and produces less accurate forecasts than the other three models \ref{fig:fold_auc}).
+Including historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to a model that allows covariate effects to vary over time but does not include historical covariates.
 This contrasts with previous findings using geological stages as their time unit \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.
 
-%This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses.
-By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored.
-For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the intercept is greater than 2 or less than -2 because the inverse-logit of those values are close to 0 or 1, respectively \citep{ARM}.
+It is interesting to note that both historical covariates like change in geographic range and allowing covariate effects to vary over time can be thought of as different ways of encoding the past: explicitly or implicitly.
+Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time.
+Allowing covariate effects to vary over time, on the other hand, does not explicitly encode ``the past'' into our model but instead models how covariate effects vary over time which allows the past to be implicitly included in our model.
+By modeling this variation, forecasts made for future extinction events can allow for a wide range of potential effects of our covariates on predicting extinction.
+Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig.
+\ref{fig:fold_auc}).
 
-The relative quality and consistency of our models' out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
+The relative quality and consistency between three of our models' in-sample and out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
 Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity.
@@
 In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of taxonomic, geographic, and historical predictors.
-While change to in geographic range and past temperature information is not found to greatly improve our ability to estimate future extinction risk, we do find that the immediate past is a modest predict of the immediate future.
-These results are comparable to conservation determinations because both are expressed in terms of an relative risk.
+We find that our VP model chich includes the historical covariates such as change in geographic range between observations and past global temperature information while also allowing covariate effects to vary over time is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models.
+
+We find that three of our models yield functional equivalent out-of-sample forecasts: our model with historical covariates but assume covariate effects are constant over time (CP model), our model which allows for covariate effects to vary over time but does not include the historical covarites (V model), and our model without historical covariates and does not allow covariate effects to vary over time (C model).
+This result implies that explicit modeling of historical covariate effects on extinction yields similar forecasts to modeling how covariate effects have changed over time.
+We interpret this to mean that paleobiological information, such as variation in covariate effects on extinction risk over time, can improve our ability to forecast future extinctions, however including both historical coviarates and allowing covariate effects to vary over time yields an overfit model which produces biased forecasts.
+Importantly, while our model with no historical covariates and no variation on effects over time yielded similar out-of-sample forecasts to our model with historical covariates but constant effect and our model with varying effects but no historical covariates, the inclusion of this additional information does not harm these models' forecast ability.
 The results of this simple exercise suggest that conservation decisions would be bolstered by including fossil data.
-
-%are encouraging for efforts to incorporate fossil data into current and future extinction risk assessments, and
-
-%It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time.
-An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
-
-%This presents an interesting conundrum about how to improve upon our results.
-A simple hypothesis for how to improve upon our results is that if we were to include more biological information in our model, our estimates of species relative extinction risk would be improved.
-However, if we decrease the amount of data in our model, our results by definition decrease in their quality.
-For example, compare the results from our models fit using full in-sample dataset, and the results from the cross-validation where the fit to each fold is by definition based on less than full information.)
 
@@
 
+\section{Supplement to Results} \label{sec:supp_res}
+
+Here we present the parameter estimates for our from our VP model (Table \ref{tab:model_def}).
+We choose to present the estimates from this model because it is our most inclusive model and its parameter estimates are indicative of parameter estimates from other models with a subset of those covariates included in this model.
+
+We present the group-level estimates are the average effects of that covariate over time and across taxonomic groups.
+These estimates make up the vector $\mu$ described above.
+In addition to the average effect of our covariates, we also included the estimate of the group-level intercept, which is the average log-odds of extinction for an average observation over time and across taxonomic groups.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{../results/figure/effect_est}
+  \caption{Group-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
+The posterior distribution of our parameter estimates are presented as densities.
+Below each of these densities is marked the median estimate along with 50\% and 80\% credible intervals.
+Estimates are on the log-odds scale.}
+  \label{fig:param_est}
+\end{figure}
+
+Our VP model allows for these our regression coefficients to vary over time and between taxonomic groups.
+Here we present the population-level estimates for our regression coefficients along with the population-level intercept estimates.
+Population-level estimates are those for a specific time interval and taxonomic group.
+In this case, the intercept estimate describes the log-odds of extinction for an average observation at that point in time and taxonomic group.
+
+\begin{figure}[<+htpb+>]
+  \centering
+  \includegraphics{../results/figure/eff_time_group}
+  \caption{Population-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
+Posterior estimates are presented as a time series.
+The black line represents the median estimate of that parameter.
+In addition, the 50\%, 80\%, and 95\% credible intervals are indicated.
+Estimates are on the log-odds scale.}
+  \label{fig:param_est_time_group}
+\end{figure}
+
 \printbibliography[title={Supplementary References}]

### 2019-06-19 doc/paper.tex
paper has all figures
edits complete. pass to seth

@@
 We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
-We find that our model that includes our historical covariates (e.g.
-change in geographic range) and allows covariate effects to vary over time to be overfit to our data, yielding biased forecasts of future extinction events.
-In contrast, our models which include either historical covariates or which allow for covariate effects to vary over time yield equivalent forecasts.
 Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time.
+We find that models which include either historical covariates or account for variation in covariate effects over time yield equivalent forecasts, but a model including both is overfit and yields biased forecasts.
 An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
@@
 \ref{fig:fold_auc}).
-One of the most striking aspect of these results is that our results is that our model that includes historical covariates (e.g.
-change in geographic range) and allows for covariate effects to vary over time is overfit to our data and produces more biased out-of-sample predictions than the other three models (Fig.
+One of the most striking results is that the models that include either the historical covariates (CP model) or allows covariate effects to vary over time (V model) produce less biased out-of-sample forecasts than the model which includes both historical covariates and allows covariate effects to vary over time (VP model) (Fig.
 \ref{fig:auc_compare}).
-In comparison, the models that include either the historical covariates or allows covariate effects to vary over time produce less biased out-of-sample forecasts.
-A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original).
-Additionally, a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect.
+However, we might expect a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect.
 An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
 
-We also find that our three of our four models are practically identical in their ability to make in-sample and out-of-sample forecasts.
+We find that three of our four models are practically identical in their ability to make in-sample and out-of-sample forecasts.
 Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig.
 \ref{fig:auc_hist}).
-While the ``best'' model based on in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time, the out-of-sample forecasts from this mode indicate that this model is overfit to our data and produces less accurate forecasts than the other three models \ref{fig:fold_auc}).
+While the model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time, the out-of-sample forecasts from this mode demonstrate this model yields biased forecasts, which indicate that this model is overfit to our data \ref{fig:fold_auc}).
 Including historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to a model that allows covariate effects to vary over time but does not include historical covariates.
-This contrasts with previous findings using geological stages as their time unit \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.
 
@@
 \ref{fig:fold_auc}).
+Our results supplement those of \citet{Kiessling2016} which examined using differences in geographic range over time to predict extinction risk.
+
+It is imporatant to remember that our model with historical covariates assumes that extinction selection based on these covarites is constant over time.
+If human impacts substatially alter present and future extinction risk dynamics, the assumption the effects of covariates have not changed over time may be limit or bias our ability to predict extinction in truly novel environmental regimes.
+Thus, it might be preferable to model the variation in extinction risk and selectivity over time instead of relying solely on measures from a taxon's past such as change in geographic range over time.
+Thus, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.
 
@@
 
-We present the group-level estimates are the average effects of that covariate over time and across taxonomic groups.
+First, we present the group-level estimates of our regression coefficients and intercept (Fig.
+\ref{fig:param_est}).
+Group-level effects are the average effects of that covariate over time and across taxonomic groups.
 These estimates make up the vector $\mu$ described above.
@@
   \centering
-  \includegraphics{../results/figure/effect_est}
+  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/effect_est}
   \caption{Group-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
@@
 
+Second, we present the population-level estimates for our regression coefficients along with the population-level intercept estimates (Fig.
+\ref{fig:param_est_time_group}.
 Our VP model allows for these our regression coefficients to vary over time and between taxonomic groups.
-Here we present the population-level estimates for our regression coefficients along with the population-level intercept estimates.
 Population-level estimates are those for a specific time interval and taxonomic group.
@@
 
-\begin{figure}[<+htpb+>]
+\begin{figure}[ht]
   \centering
-  \includegraphics{../results/figure/eff_time_group}
+  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/eff_time_group}
   \caption{Population-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).

### 2019-06-20 R/01_inspect_data.r
regen figure

@@ -15,7 +15,7 @@ source(here('R', 'helper07_diffsafe.r'))
 plan(multiprocess)
 
 # useful graphical constants
-theme_set(theme_bw(base_size = 20))
+theme_set(theme_bw(base_size = 25))
 
 
 # get data in

### 2019-06-20 R/helper02_plot_foo.r

@@ -376,7 +376,7 @@ view_neptune <- function(.data, name = 'full', path = '../results/figure/') {
   filename <- paste0(path, '/occ_time_label_', name, '.png')
   ggsave(filename = filename,
          plot = octg, 
-         width = 4, height = 6)
+         width = 8, height = 10)
   
   
   # relative "abundance" of microfossil_groups over time
@@ -387,7 +387,7 @@ view_neptune <- function(.data, name = 'full', path = '../results/figure/') {
   filename <- paste0(path, '/abn_time_stack_', name, '.png')
   ggsave(filename = filename,
          plot = ocag, 
-         width = 6, height = 6)
+         width = 10, height = 10)
   
   # occurrences by relage
   ocrg <- .data %>% 
@@ -412,7 +412,7 @@ view_neptune <- function(.data, name = 'full', path = '../results/figure/') {
   filename <- paste0(path, '/age_label_', name, '.png')
   ggsave(filename = filename,
          plot = ocrg, 
-         width = 4, height = 6)
+         width = 6, height = 10)
   
   
   
@@ -430,7 +430,7 @@ view_neptune <- function(.data, name = 'full', path = '../results/figure/') {
   filename <- paste0(path, '/range_time_', name, '.png')
   ggsave(filename = filename,
          plot = srg, 
-         width = 8, height = 6)
+         width = 10, height = 8)
   
   # lots of little code here
   # for FAD/LAD accumulation curves
@@ -467,13 +467,13 @@ view_neptune <- function(.data, name = 'full', path = '../results/figure/') {
   filename <- paste0(path, '/fad_lad_count_wide_', name, '.png')
   ggsave(filename = filename,
          plot = ccg, 
-         width = 6, height = 3)
+         width = 10, height = 5)
   
   ccg2 <- ccg + facet_grid(fossil_group ~ ., switch = 'y', scales = 'free_y')
   filename <- paste0(path, '/fad_lad_count_tall_', name, '.png')
   ggsave(filename = filename,
          plot = ccg2, 
-         width = 4, height = 6)
+         width = 8, height = 10)
 }
 
 

### 2019-06-20 doc/napc2019_talk.tex
working on napc talk
talk ready for napc?

--- 2019-04-03 doc/norcalpaleo2019_talk.tex
@@
-\documentclass{beamer} 
+\documentclass[aspectratio=169]{beamer} 
 \usepackage{amsmath,amsthm}
@@
 
-        \includegraphics[width=\textwidth,height=0.15\textheight,keepaspectratio=true]{figure/dino}
+        \includegraphics[width=\textwidth,height=0.15\textheight,keepaspectratio=true]{figure/diatom}
 
@@
     \begin{column}{0.4\textwidth}
-      \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/occ_time_label_full}
+      \includegraphics[width=\textwidth,height=0.7\textheight,keepaspectratio=true]{../results/figure/occ_time_label_full}
     \end{column}
     \begin{column}{0.4\textwidth}
-      \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{../results/figure/age_label_full}
+      \includegraphics[width=\textwidth,height=0.7\textheight,keepaspectratio=true]{../results/figure/age_label_full}
     \end{column}
@@
       \begin{itemize}
-        \item Our best supported model includes our historical covariates and allows all effects to vary over time.
+        \item Our best supported model includes either our historical covariates or allows all effects to vary over time.
       \end{itemize}
@@
       \end{itemize}
+    \item Allowing effects to vary over time is probably preferable to historical covariates -- measures and accounts for variation which is important when predicting extinction in novel environments.
     \item Mechanisms behind changes to geographic range operate at sub-million year scales. Perhaps their effects are weak/masked at million (or greater) year scales.
   \end{itemize}
+
+\end{frame}
+
+\begin{frame}
+
+  \begin{center}
+    \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio=true]{figure/book}
+
+    \Large{https://psmits.github.io/paleo\_book/index.html}
+  \end{center}
 

### 2019-07-08 doc/response_review.tex
process for resubmit

\documentclass[12pt,letterpaper]{article}

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
%\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage[style=numeric-comp,
      backend=biber,
      bibencoding=utf8,
      maxbibnames=10,
      uniquename=init,
      giveninits=true,
      sorting=none,
      natbib=true,
      url=false,
      doi=false,
      isbn=false]{biblatex}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{threeparttable}

\frenchspacing

% https://3d.bk.tudelft.nl/hledoux/blog/fiddling-biblatex/
%-- formatting hell for biblatex
%-- remove "In:"
\renewbibmacro{in:}{}
%-- no "quotes" around titles of chapters/article titles
\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]
{title}{#1}
%-- no punctuation after volume
\DeclareFieldFormat[article]{volume}{ {#1} } 
%-- puts number/issue between brackets
\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]{number}{\mkbibparens{#1}} 
%-- and then for articles directly the pages w/o any "pages" or "pp." 
\DeclareFieldFormat[article]{pages}{#1}
%-- for some types replace "pages" by "p."
\DeclareFieldFormat[inproceedings, incollection, inbook]{pages}{p. #1}

% https://tex.stackexchange.com/questions/81569/biblatex-parentheses-around-the-volume-number-of-an-article
\renewbibmacro*{volume+number+eid}{%
 \printfield{volume}%
 \printfield{number}%
 \printfield{eid}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}

% bibliographic information
%\bibliographystyle{abbrvnat}
%\bibliography{citations}
\addbibresource{citations.bib}

\title{Responses to reviews of ``How predictable is extinction? Forecasting species survival at million-year timescales''}
\author{
 Smits, Peter\\
 %Department of Integrative Biology, University of California Berkeley\\
 \texttt{psmits@berkeley.edu} 
 \and
 Finnegan, Seth\\
 %Department of Integrative Biology, University of California Berkeley\\
 \texttt{sethf@berkeley.edu}
}
\date{}

\begin{document}

\begin{refsection}

\maketitle

\linenumbers{}
\modulolinenumbers[3]

Below are our responses to the editor and reviewer comments that we recieved on our original manuscrpt. We would like to thank the editor and the three reviewers for their thoughtful and insightful comments which have greatly improved the manuscript.

Editor and reviewer comments are presented in bold. Below these comments, we detail how we updated our manuscript in response to that comment.

\section{Editor comments}

\begin{bfseries}
  Line 5 (abstract): comma after ‘to answer this question’

  Line 7: period after parenthesis

  Line 10: comma before ‘the improvement in predictive power’

  Line 25: ‘the suggestion’ or ‘suggestions’
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\begin{bfseries}
  Line 31-33: I think this sentence needs to be clarified. Can you describe more clearly what is meant by ‘how accurate or strong our predictions about the differences in extinction risk are’, and how this differs from strong and general determinants of extinction risk?
\end{bfseries}

In our general reorganization of the introduction section of this paper we tried to improve the clarity of our language. We thank the editor for pointing out this sentence as it was redundant and confusing.

\begin{bfseries}
  Line 34: try to avoid ending a sentence with a preposition

  Line 40: among taxa, rather than between?

  Line 41: comma before ‘and the relative risk of…’

  Line 77: need space before ‘Occurrences were filtered…’

  Line 78: ‘was no more’…is that right?
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\begin{bfseries}
  Line 83: what about taxa with age ranges that were more uncertain than 1 mys? Did this ever occur? [Ah, I read further and now see your call to the SI]
\end{bfseries}

Thank you for pointing this issue out. We have tried to improve our references to the supplemental material by putting more references in the text instead of just a single reference at the end of the section.

\begin{bfseries}
  Line 86: I echo Wolfgang’s comment in wondering how congruent would be patterns if they would be measured using another range-based metric?
\end{bfseries}

Thank you both for the suggestion to move to minimum spanning tree measure of geographic range. We have updated our analysis using this new range metric. The change in range metric did not change our interpretation of the results.

\begin{bfseries}
  Line 115: insert ‘in’ prior to ‘Section S1.3’
\end{bfseries}

We have updated the relevant section of the text with this suggestion.

\begin{bfseries}
  Line 114(ish): have you considered testing change in geographic range over multiple intervals? I.e., maybe a stronger signal would be observed if there was a persistent decline in range size over multiple 1 my time bins? Similarly, species may be buffered from extinction if their range increases over multiple time bins. It would be interesting to tease whether this is more important than absolute size…
\end{bfseries}

Thank you for this suggestion. We have update our analysis to include change in geographic range from the last three observations. This change adds two more covariates to all models that include our ``historical covariates.'' This addition to our analysis did change our results, and we now consider our most complex model that includes historical covariates and variation in covariate effect over time to be overfit to our data. Our models that include either historical covariates or variation in effects over time, however, are found to have functionally equivalent out-of-sample performance.

\begin{bfseries}
  Line 119: try to avoid ending a sentence with a preposition

  Line 183: ‘Comparing’?

  Line 195: I am probably being dense, but how can the in-sample have a poorer performance than the out-of-sample? Isn’t this fairly counter-intuitive?

  Line 214: ‘forecast’ vs ‘forecasting’?

  Line 224: ‘an’ before ‘extremely important…’
\end{bfseries}

We have updated the relevant section of the text with this suggestion, in particular focusing on clarity.

\section{Reviewer comments}

\subsection{Reviewer 1}

\begin{bfseries}
  Smits and Finnegan use the Neptune database to check which combination of potentially influencing variables has the best predictive power of extinction risk in the marine plankton. They use sophisticated logistic modelling of the time-binned raw data to suggest that considering changes of parameters only marginally improves prediction accuracy. Overall, this is a very well written, although difficult to comprehend, account, which is to the point and rich in insights.

  While the main result is well founded by the sophisticated analyses (in fact too sophisticated to be understandable in all details by the reviewer), it is somewhat disappointing, as the relative importance of parameters in multivariate models remain vague. It is also disappointing to see that change of range had a negligible effect in the multivariate context, as this is one of the main arguments for using the fossil record in extinction risk assessments (besides taxon age and intrinsic turnover rates). I argue that, to really discard change of range, the authors would need to look at longer trajectories of range. Bin-to-bin changes at this temporal resolution (1 myr) are perhaps too noisy to inform models. I am not asking for additional analyses for this paper, but this point should be added to the discussion.
\end{bfseries}

Thank you for this suggestion. We have update our analysis to include change in geographic range from the last three observations. This change adds two more covariates to all models that include our ``historical covaraites.'' This addition to our analysis did change our results, and we now consider our most complex model that includes historical covariates and variation in covariate effect over time to be overfit to our data. Our models that include either historical covariates or variation in effects over time, however, are found to have functionaly equivalent out-of-sample performance.

\begin{bfseries}
  We are confronted with four models and even the simplest one is already quite complex. Is there a chance to provide for each model the relative importance of each variable for the model performance? Alternatively, as standing geographic range is among the best predictors of extinction risk in paleo studies, can the authors add a very simple model of just event ~ range? I’d be curious to see if the AUC is then “significantly” lower than in the other models. Of course, in a frequentist world, I would start with a complex model and then perform model selection with a step function. I’m not Bayesian enough to suggest a specific strategy for this paper, but somehow model performance needs to be balanced by model complexity. The authors seem to have compared just model performance.
\end{bfseries}

Thank you for pointing out this issue in our text and we apologize for the lack of clarity. We have updated our methods section to further explain cross-validation and how we have implicitly dealt with model complexity because of the linked definitions of cross-validation and AIC. From our updated text:

``Cross-validation is a procedure for estimating a models expected out-of-sample error. Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}. Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, out-of-sample predictions will be biased and inaccurate \citep{ESL}. A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data). Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces, an entire posterior distribution of estimates.''

\begin{bfseries}
  The methods are very well explained (especially in the supplement), although uncertainties in parameter estimates merit some further prose. Using maximum great circle distance (GCD) as a measure of geographic range is ok but prone to sampling bias. Better sampled intervals will inevitably have greater mean ranges than more poorly sampled intervals. While the scaling applied by the authors takes different means and variance into account it may not be sufficient. If GCD is used, I’d recommend normalizing to the maximum observed geographic range in an interval prior to scaling. Grid cell occupancy or minimum spanning trees are alternative measures, which are worthy of exploration. 
\end{bfseries}

Thank you both for the suggestion to move to minimum spanning tree measure of geographic range. We have updated our analysis using this new range metric. The change in range metric did not change our interpretation of the results.

We chose not to standardize geographic range to in-bin max because we believe this breaks ``coherence'' of the measure and the interpretation of the regression coefficients. With geographic range not standardized by max in-bin geographic range, the regression coefficients for the ``change in range'' covariates are the expected change in log-odds of extinction risk per standard deviation change in geographic range between that previous bin and the current bin. In contrast, if the variables are defined as percentages, the difference between bins gives a less obvious or interpretable regression coefficient. Additionally, max in-bin geographic range is nearly identical for the entire Cenozoic.

\begin{bfseries}
  Looking at change of range is good, but the authors could also look at changes beyond two intervals. Although models incorporating multiple bins may result in overfitting, the argument is that long-term decline will increase extinction risk. I’d argue that two bins (especially at the relatively fine temporal scale applied) can hardly measure long-term decline. Finally, the authors should provide a rationale why these and not other parameters were used in the study (see also below). Does latitude or skeletal mineralogy have little to add to model performance?
\end{bfseries}

Thank you for this suggestion. We have update our analysis to include change in geographic range from the last three observations. This change adds two more covariates to all models that include our ``historical covariates.'' This addition to our analysis did change our results, and we now consider our most complex model that includes historical covariates and variation in covariate effect over time to be overfit to our data. Our models that include either historical covariates or variation in effects over time, however, are found to have functionally equivalent out-of-sample performance.

We have modified our discussion sections to further explain the rational behind our covariate choices. Simply, our goal is to provide a baseline for extinction prediction that can be compared to more specific models which include these more specific covariates. Additionally, a lot of the suggested covariates are either not available for a lot of the taxa throughout the Cenozoic or are nearly constant for the entire taxonomic group. In many ways, by allowing our parameter estimates vary by taxonomic group we have attempted to implictly deal with these types of ecological differences.

\begin{bfseries}
  Presentation of results could be improved. Showing raw data of extinction rates in the four groups, and individual analyses how age and phyla actually influence extinction risk would be preferred over the wiggly lines in Figs. 2 and 3. The latter could be moved to the supplement. The text in the results section is very technical, describing too much detail. 
\end{bfseries}

Thank you for this suggestion. We have added a section to our supplemental material that includes figures presenting the parameter estimates for the group-level regression coefficients and the individual-level estimates for those regression coefficients over time.

\begin{bfseries}
  The discussion is good, but may have to be partly rewritten in light of my comments.

  Specific comments (line numbers refer to the smaller ones on the right):

  - “the relative risk of extinction exhibited by different taxonomic groups” could be abbreviated as “intrinsic extinction risk”

  - l. 68-69. Partial repetition of sentences before. “Incredible” should be omitted. If incredible why trust the age models?

  - l. 79. “Ecologically unusual” should be omitted from reasoning. Not needed. 
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\begin{bfseries}
  - l. 88-91. Authors may want to add that Mg/Ca has its own issues, especially the steep rise of Mg/Ca in seawater through the Cenozoic. Cramer et al. have discussed and accounted for this trend, but with some assumptions, which result in considerable uncertainties.
\end{bfseries}

Thank you for pointing this out. We have updated our Methods section with a discussion of the potential problems with Mg/Ca based temperature estimates.

\begin{bfseries}
  - l. 132: paragraph starts with k-fold but is specified for 5-fold. I recommend rewriting this paragraph for the general application and then specify.

  - l. 149-150: “so we interpret values between 0.7 and 0.8 could then be considered”  “so we consider values between   0.7 and 0.8  as”

  - Caption Fig. 1: Cut “a” from “with a higher AUC values”. It is not entirely clear what the error bars and bean plots represent. Are the AUC values derived from the 62 time bins or is there more to it? Please explain.

  - l. 173: “Dinoflagellates” should be “diatoms” unless I have missed something. Same in labels of Figs. 2-4 and some incidences in the text. Please correct throughout.

  - Xlab of Fig. 2 should be “Mya” or similar instead of “CI”

  - l. 176-177. Repetition from methods.

  - l. 183. “Comparison”  “Comparing”

  - l. 192-193. No single-sentence paragraphs

  - l. 197. “diatoms” instead of “Dinoflagellates”

  - l. 214-216. To my understanding AUC does not simply measure a rank-order analysis but model performance. That the AUC’s of the eight models are statistically indistinguishable (are they?) does not necessarily imply a correct forecasting of rank-order extinction probability.  Please rephrase!
\end{bfseries}

We have updated the relevant sections of the text and our figures with these suggestions.

\begin{bfseries}
  - l. 223-224. “Variation of extinction intensity”. Would be useful to either depict this variation or report some summary statistics (median, range, variance).
\end{bfseries}

Thank you for this suggestion. We have added a section to our supplmental material that includes figures presenting the parameter estimates for the group-level regression coefficients and the individual-level estimates for those regression coefficients over time. These estimates include average log-odds of extinction over time.

\begin{bfseries}
  - l. 233. What does “scientifically significant” mean is this context? I am aware that simple model selection is not applicable here but it would be good to have a comparable metric.
\end{bfseries}

Thank you for pointing out this lack of clarity. We have removed this section and attempted to improve the clarity of our discussion of our results.

\begin{bfseries}
  - l. 236-237. Kiessling and Kocsis and used more than just corals. Their coarser temporal resolution (geological stages) is an obvious suspect for the greater role of change of occupancy than reported here. 
\end{bfseries}

We have update our covarites to include additional lags. Please see the updated Methods, Results, and Discussion for how our results have changed.

\begin{bfseries}
  - l. 244-245. Poor reasoning. For example skeletal mineralogy (silica versus calcium carbonate), trophic level (phyto- vs. zooplankton), latitudinal preference (e.g., median paleolatitude) or additional physico-chemical parameters (e.g., sea-level, Mg/Ca, pCO2) are intuitive variables, whereas taxon-age is less intuitive in light of the recent literature. I’m not asking for additional analyses or more complex models but a sound rationale.
\end{bfseries}

We have modified our discussion sections to further explain the rational behind our covariate choices. Simply, our goal is to provide a baseline for extinction prediction that can be compared to more specific models which include these more specific covariates. Additionally, a lot of the suggested covariates are either not available for a lot of the taxa throughout the Cenozoic or are nearly constant for the entire taxonomic group. In many ways, by allowing our parameter estimates vary by taxonomic group we have attempted to implictly deal with these types of ecological differences.

\begin{bfseries}
  - l. 247. Add “for” between “not exist” and “all”. In any case, a poor defense given the obvious traits to look at (see above).

  - l. 254. Modern “conservation determinations” are not continuous but ordinal. The difference between “critically endangered” and “endangered” may be very different from the difference between “endangered” and “vulnerable”.

  - l. 255-256. I argue that the authors’ results actually show that fossil data are not very relevant. If this impression is wrong, the authors should perhaps rephrase some parts of the text.

  Also check use of “My” vs. “Mya” and cases in plankton groups. As a general rule these should be upper case in a formal context (e.g., Radiolaria) and lower case when used informally (e.g., radiolarians). 
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\begin{bfseries}
  The supplement is good but lacks sensitivity tests and a rationale of why the specific parameters and not others where chosen in the models. 
\end{bfseries}

Thank you for this suggestion. We have improved the language in our methods section surrounding in-sample verus out-of-sample forecasts. Our in-sample forecasts are a type of sensitivity test.

From the updated manuscript:

``In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to correctly classify the data to which it was fit. Posterior predictive checks are a type of sensitivity analysis because we are checking the quality of model's fit to the data. If our models have poor in-sample forecasting performance, than our models are not adequate descriptors of the data and will most likely make poor out-of-sample predictions. In-sample forecasting measures, however, are not the same as understanding our models' ability to forecast future extinctions or if our models are overfit to our data and produce biased out-of-sample estimates \citep{ESL}.''

And:

``The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis. Comparison between the posterior distributions \dots''

\subsection{Reviewer 2}

\begin{bfseries}
  This manuscript uses a high-resolution record of marine microfossils to test the predictive power of fossil record extinctions. A lot of paleontological research has focused on determining the importance of particular biological and environmental risk factors, so this study is novel in testing whether past history can predict future extinction. The methodology is explained well and the main result is well-supported.

  I only have two broader thoughts and two more specific notes:

  I find it interesting that the predictability of extinction doesn’t obviously differ between background intervals and intervals with more unusual environmental perturbations (such as the PETM or E-O transition). Perhaps that’s because those events weren’t mass extinctions in the vein of the P/T or K/Pg. There’s also a fair amount of volatility in the AUC time-series. Nevertheless, does this consistent predictive ability regardless of differences in conditions say anything interesting about extinction predictions?

  The caveat that human impacts may dramatically alter extinction risk seems like an important one and, as it stands, seems like it could undercut the (rather terse) final conclusion. Would conservation decisions really be bolstered by including fossil data? I would agree with that (and of course I would, as a paleobiologist), but it might be useful to expand a bit on this topic. Do you mean that a model such as yours, incorporating geographic range and other such parameters, would be helpful? In what way could it be used? Or do you mean using models based on past intervals with environmental changes inferred to be similar – the PETM as an analogue for warming and acidification, for example? I wonder if it would be helpful to draw a stronger connection linking your finding to conservation biology, given the different stressors now.
\end{bfseries}

WAITING FOR SETH

\begin{bfseries}
  I found the description on lines 171-174 to be a bit confusing. The phrase “this pattern” (line 171) seems to refer to the posterior distribution less than/equal to 0.5. The sentence states that the pattern is absent in absent in VP model for forams and radiolarians, but to me the radiolarian pattern doesn’t look too different from the V model (which also doesn’t dip below 0.5). The next sentence argue that there are fewer period of low performance for calcareous nannos and dinoflagellates in the VP model, but the differences seem minimal. Perhaps I’m not familiar with these graphs, but there only seems to be a difference in calcareous nannos around 50 Ma, and among dinoflagellates a small difference around 58 Ma and 30 Ma?
\end{bfseries}

Thank you for pointing out this lack of clarity in the Results section. We have revised this section for clarity.

\begin{bfseries}
  Figures 2-4. I guessed (and later confirmed) that the alternating gray and white bars represented stages of the geological time scale, but it would be helpful to indicate that in the captions.
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\subsection{Reviewer 3}

\begin{bfseries}
  This is an interesting paper on extinction prediction using microfossil plankton data from deep-sea cores.

  It’s very important that they found simple model with only a few parameters of geographic range, temperature etc can robustly predict extinction risk regardless of minor model difference.

  I enjoyed my reading and hope the comments below help to improve the ms.

  1. just in case, for foraminifera, benthic species are deleted from the database? Or they originally include planktonic foram only?
\end{bfseries}

We have confirmed using MicroTax that all foram species included our data set our planktonic species only.

\begin{bfseries}
  2. Neptune database has an internally consistent taxonomic identification strategy. Is this true? I guess it’s non-critical compilation of census data done by shipboard micropaleontologists over ~50 years (ie taxonomic sense of different generation micropaleontologists are very different). David Lazarus (in charge person of Neptune) is radiolarian person, and so radiolarian may be better in taxonomy? Planktonic foram diversity is low, so may be better straightforward. Anyway, I recommend deeper reference search and reading to see how is Neptune taxonomy and explain it in the method section more in details.
\end{bfseries}

We have added a statement in our introduction discussing that the Neptune database has an internal synonymization framework as detailed in \citet{Lazarus1994}. While the taxonomic identities may not be the most accepted by the community, they are consistent within the database.

\begin{bfseries}
  3. Fig 1, better to explain what are C, V, CP, VP, AUC in the caption

  4. Fig 2 and 3, What do gray bars mean? Gray bay = envelope?

  5. It will help our understanding if you explain the main point/message of each fig in the contest of discussion in each caption 
\end{bfseries}

Thank you for these suggestions. We have updated our figure captions to be more descriptive. Additionally, we have clarified the link between our figure elements and our model description table.

\begin{bfseries}
  6. I am not sure if it’s possible, but I am curious where geographic range or temperature are the key? Or both are essential? It is possible to try geographic range only model and temperature only model to see such?
\end{bfseries}

Thank you for your suggestion. We chose to always included geographic range and temperature because these represent basic measurements of a species and its environmental context. Additionally, if we were to use models with one of these covariates and not the other, then we are assuming that the ignored covariate is known with 100\% probability of having 0 effect on extinction risk. We are uncomfortable making this assumption and instead Instead we use weakly informative priors to limit spuriously strong parameter estimates and ``shrink'' effects towards 0 if that covariate has a weak outcome on the response.

We have also added a section to the supplement with the group-level and individual-level estimates for all of our regression coefficients to demonstrate the size of our covariate effects.

\begin{bfseries}
  7. The authors use deep-water temperature (using geochemistry of benthic foram from deep-sea cores) data. And their biotic data is plankton. Their plankton can be mostly from ocean surface or photic zone (eg diatoms), but some have good bunch of deep-water species (radioralians).

  The authors need some discussion and justification on this.
\end{bfseries}

We have added a section to the Methods discussing this data limitation

From the updated manuscript:

``Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature. The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit. Ideally, we would have detailed ocean surface water temperature estimates for the entire globe for the entire Cenozoic. Unfortunately, that type of data does not exist. So, we interpret our temperature estimates as reflecting the global climate state that an organisms experiences, and not as a descriptor of that taxon's local environmental ecology.''

\begin{bfseries}
  8. While geographical range consider spatial stuff, the authors use one temperature curve for all (I understand there will be no other way). Is there any difference in temperature sensitivity between narrow and wide distribution species; or low latitude and high latitude species? I understand this may be out of the scope. The authors may ignore.
\end{bfseries}

We thank the reviewer for this comment. We do agree that surface temperature estimates, especially geographically localized ones, would be a major improvement to our data. However, detailed sea-surface estimates for the entire Cenozoic are not readily available and generating those would be outside the scope of this study.

\printbibliography
\end{refsection}

\end{document}

### 2019-07-08 doc/article.tex
process for resubmit

--- 2019-06-19 doc/paper.tex
@@
 
-Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
-This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how capable we are at predicting future extinction events.
+Extinction intensity (average rate) and selectivity (difference in risk among taxa) have varied greatly through time, and the relative risk of extinction exhibited by different taxonomic and ecological groups can provide insights into the drivers of both background and mass extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019,Wang2008a,Knoll2007}.
+Many studies have examined the effects of various potential predictors on extinction risk through time \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Harnik2012a,Ezard2011,Foote2006} or refined methods for identifying and measuring these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
+These studies have produced a growing body of knowledge regarding which factors have been general determinates of extinction risk in the geological past.
 
-We ask how accurate risk predictions based on fossil data are when estimating future extinctions.
+A related question that has received much less attention is how successful we might expect to be when using this knowledge to attempt to predict future extinction events.
+
 Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions.
-However, we can address the question by taking a given point in the geological past, developing a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g.
+However, we can take a given point in the geological past, develop a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on subsequent (e.g.
 ``future'', from the point of view of the model) extinction/survival events.
+Putting aside the very important question of how human activities will alter the determinants of future extinction risk, such an approach provides a framework for evaluating the expected accuracy of future risk assessments based on past extinction events.
 
-Extinction intensity (average rate) and selectivity (difference in risk among taxa) vary through time, and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}.
-This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
+%This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
 Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.
 
-Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
-The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}.
-In the former case, knowledge of the specific past trajectory of a species -- whether its range has expanded or contracted over a given time span -- might help to improve assessments of its current risk.
-In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past.
-Discriminating among these alternative models is thus very important for determining how best to incorporate fossil data in present risk assessments.
+Here we take this approach, using as a model system the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores).
+This record has several key strengths for our purposes: planktonic microorganisms are widespread and abundant in pelagic habitats, have high preservation potential, and because of their utility for biostratigraphic, paleoclimatic, and oceanographic study they have been the focus of an extensive international coring and study effort \citep{Lazarus1994,SpencerCervato1999}.
+A compliation of these data is readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}.
+This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomic synonymization framework for dealing with 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible.
+Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.
 
-For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores).
-These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}.
-This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomy synonymization strategy for delaing wiht 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible.
-Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.
+The overall question of how well models based on past extinction patterns perform at forecasting future extinctions depends in part on model complexity.
+Simple models requiring only a few parameters are in general preferable because more complex models run a greater risk of being overfit to the observations on which they are trained.
+In addition, many traits that might influence extinction risk among extant species are difficult to assign confidently to extinct species.
+For these reasons we elect to focus on “baseline” models which include only a few parameters that have been shown to be important and/or consistent determinants of extinction risk in the marine fossil record.
+Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species’ geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
+In addition to geographic range, we also considered global climate state and change in climate state since previous observation in order to evaluate the influence of climate or climate change trajectory on extinction risk.
+Finally, we included species age, both because previous studies of planktonic taxa have found it to be a determinant of extinction risk and because its inclusion in our models is critical to their nature as survival models (see Model Specifications below).
+We reiterate that our primary objective is to evaluate the predictive performance of simple models that include only a few general parameters; more complex models including other likely determinants of extinction risk such as skeletal mineralogy, trophic ecology, and thermal tolerance range might well perform better.
+
+There are a number of ways in which past extinction patterns might be used to model present risk.
+The simplest case assumes that relationships between predictors (hereafter covariates) and extinction risk have been constant through time.
+A more complex but more realistic case allows relationships between covariates and extinction risk to vary through time, consistent with evidence for temporal variation in extinction selectivity regime.
+Finally, an important consideration is the degree to which species’ geographic range trajectories exhibit deterministic versus Markovian behavior \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}.
+In the former case, knowledge of the specific past trajectory of a species – whether its range has expanded or contracted from some point in the past to the present – might help to improve assessments of its current risk.
+In the latter case only the current geographic range of the species would convey useful information about current and future risk (although that assessment would still be based on the relative extinction risk of species that had similar ranges at different points in the geological past).
+
+Below, we evaluate four models along a spectrum from simplest (fixed covariate effects, Markovian range dynamics) to most complex (varying covariate effects, deterministic range dynamics).
+We ask (1) how well they perform at classifying species as extinctions or survivors in the data they were fit to, and (2) how well they perform at classifying species as extinctions or survivors in “future” data that were not used in fitting the models.
 
@@
 This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic foraminifera, and radiolarians.
-Occurrences were filtered to include only those species who's first occurrence is no earlier than 63 Mya (millions of years ago).
+Occurrences were filtered to include only those species with first occurrences no earlier than 63 Mya (millions of years ago).
 This filtering criterion excludes taxa that survived the K/Pg extinction or arose during this recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).
@@
 All fossil occurrences were assigned to 1 My (million year) bins based on the estimated age of the fossil occurrence as listed in the Neptune Database.
-After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred.
+After binning, each species' geographic range was calculated for each of the 1 My bins in which it occurred.
 Geographic range was calculated as the minimum spanning tree distance between all observations of that taxon during that temporal bin; this distance was measured in kilometers.
-Minimum spanning tree distance was caluclated using the \texttt{GeoRange} package for R \citep{GeoRange}.
+Minimum spanning tree distance was calculated using the \texttt{GeoRange} package for R \citep{GeoRange}.
 
@@
 We measured this change in geographic range by calculating the difference in geographic range between an observation and that taxon's three previous occurrences.
+Change between the most recent and the three previous occurrences was calculated individually for each of those lags.
 If there are not enough previous observations to calculate, then that value is recorded as a 0.
-These differences were calculated after minimum spanning tree distance was transformed and standardized (see Section \ref{sec:data_desc}).
+These differences were calculated after minimum spanning tree distance was transformed and standardized (see Supplement Section S1.1.2)
 
@@
 If there are not enough previous observations to calculate, then that value is recorded as a 0.
-This lag was calculated after global temperature was transformed and standardized (see Section \ref{sec:data_desc}).
+This lag was calculated after global temperature was transformed and standardized (see Supplement Section S1.1.2).
 
@@
 
-See Section \ref{sec:data_desc} for a further explanation on how observations were temporally binned, and how our covariates were standardized and transformed prior to analysis.
+See Supplemental Section S1.1 for a further explanation on how observations were temporally binned, and how our covariates were standardized and transformed prior to analysis.
 
@@
 This varying-intercept term allows us to tease apart the differences in extinction risk associated with time of observation versus age since first observation.
-An important note is that for our V and VP models, the covariation between this varying-intercept and the varying-slopes of our covariates is explicitly modeled (see Section \ref{sec:model_desc}).
+An important note is that for our V and VP models, the covariation between this varying-intercept and the varying-slopes of our covariates is explicitly modeled (see Supplement Section S1.2).
 
 See Table \ref{tab:model_def} for further explanation of how the four models we considered differ from each other.
-A complete description of the statistical model used in this analysis is available in Section \ref{sec:model_desc}.
-Additionally, the full description of how these models were implemented and coded, including choice of priors, is available in Section \ref{sec:model_est}.
+A complete description of the statistical model used in this analysis is available in Supplement Section S1.2.
+Additionally, the full description of how these models were implemented and coded, including choice of priors, is available in Supplement Section S1.2.
 
@@
   \begin{tablenotes}
-  \item[a] See Equation \ref{eq:model} for full statistical model definition.
+  \item[a] See Supplemental Equation S2 for full statistical model definition.
   \item[b] Species observation where 1 if time of last observation, otherwise 0.
@@
 
-We are interested in our models' performance in two contexts: in-sample performance, and out-of-sample predictive performance (i.e. forecasting).
+We are interested in our models' performance in two distinct contexts: in-sample performance, and out-of-sample predictive performance (i.e. forecasting).
 
-In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to recapitulate the data to which it was fit.
+In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to correctly classify the data to which it was fit.
 Posterior predictive checks are a type of sensitivity analysis because we are checking the quality of model's fit to the data.
@@
 
-Cross-validation is a procedure for estimating a models expected out-of-sample error, which is a similar goal to metrics such as AIC or WAIC \citep{ESL,Gelman2013d}.
-Measures like AIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}.
-Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, then out-of-sample predictions will be very biased and inaccurate \citep{ESL}.
-Similarly, a high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).
-Cross-validation is preferable to simple metrics such as AIC because instead of a single value, we have an entire posterior distribution of estimates.
+Cross-validation is a procedure for estimating a models expected out-of-sample error.
+Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}.
+Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, out-of-sample predictions will be biased and inaccurate \citep{ESL}.
+A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).
+Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces, an entire posterior distribution of estimates.
 
@@
 See our code repository at https://github.com/psmits/trident for full code details.
-This entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}.
-Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.
+The entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}.
+All of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.
 
@@
 
-The focus of this study is on understanding how well our models forecast future extinction events by comparing our in-sample and out-of-sample forecast estimates.
-A presentation of the posterior estimates for the regression coefficient estimates from our VP model (Table \ref{tab:model_def}) is available our Supplemental Materials (Section \ref{sec:supp_res}).
+The primary focus of this study is on understanding how well our models forecast future extinction events by comparing our in-sample and out-of-sample forecast estimates.
+A presentation of the posterior estimates for the regression coefficient estimates from our VP model (Table \ref{tab:model_def}) is available our Supplemental Materials (Section S2).
 
@@
 
-However, the actual difference in forecast AUC result between the VP model and the other three models is extremely small (0.01 AUC unit), and all the in-sample AUC estimates from our models are concentrated between an AUC value of 0.775 and 0.795 (Fig.
+However, the actual difference in forecast AUC result between the VP model and the other three models is extremely small (0.01 AUC unit), and all of the in-sample AUC estimates from our models are concentrated between an AUC value of 0.775 and 0.795 (Fig.
 \ref{fig:auc_hist}).
@@
 \ref{fig:fold_auc}).
-One of the most striking results is that the models that include either the historical covariates (CP model) or allows covariate effects to vary over time (V model) produce less biased out-of-sample forecasts than the model which includes both historical covariates and allows covariate effects to vary over time (VP model) (Fig.
+These results confirm that past extinction patterns can provide valuable information about which extant species are most threatened with extinction in the near geological future.
+%One of the most striking results is that the models that include either the historical covariates (CP model) or allows covariate effects to vary over time (V model) produce less biased out-of-sample forecasts than the model which includes both historical covariates and allows covariate effects to vary over time (VP model) (Fig.
 \ref{fig:auc_compare}).
@@
 
-We find that three of our four models are practically identical in their ability to make in-sample and out-of-sample forecasts.
+Three of the four models we evaluated are practically identical in their ability to make in-sample and out-of-sample forecasts.
 Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig.
 \ref{fig:auc_hist}).
-While the model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time, the out-of-sample forecasts from this mode demonstrate this model yields biased forecasts, which indicate that this model is overfit to our data \ref{fig:fold_auc}).
-Including historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to a model that allows covariate effects to vary over time but does not include historical covariates.
+The model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time.
+However, the out-of-sample forecasts from this model are biased forecasts, indicating that it is overfit to our data \ref{fig:fold_auc}).
+The model that includes historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to the model that allows covariate effects to vary over time but does not include historical covariates.
 
-It is interesting to note that both historical covariates like change in geographic range and allowing covariate effects to vary over time can be thought of as different ways of encoding the past: explicitly or implicitly.
+We note that both including historical covariates such as change in geographic range and allowing covariate effects to vary over time are different ways of encoding information from the past.
 Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time.
@@
 
-It is imporatant to remember that our model with historical covariates assumes that extinction selection based on these covarites is constant over time.
-If human impacts substatially alter present and future extinction risk dynamics, the assumption the effects of covariates have not changed over time may be limit or bias our ability to predict extinction in truly novel environmental regimes.
+An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predicatable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
+Our model with historical covariates assumes that extinction selectivity with respect to these covariates is constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes.
 Thus, it might be preferable to model the variation in extinction risk and selectivity over time instead of relying solely on measures from a taxon's past such as change in geographic range over time.
-Thus, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.
+For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.
 
-The relative quality and consistency between three of our models' in-sample and out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
+The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
 Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity.
@@
 
-In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of taxonomic, geographic, and historical predictors.
-We find that our VP model chich includes the historical covariates such as change in geographic range between observations and past global temperature information while also allowing covariate effects to vary over time is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models.
+In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors.
+Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models.
+The remaining three models yield nearly equivalent out-of-sample forecasts, suggesting that including historical information via either explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions.
+The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data.
 
-We find that three of our models yield functional equivalent out-of-sample forecasts: our model with historical covariates but assume covariate effects are constant over time (CP model), our model which allows for covariate effects to vary over time but does not include the historical covarites (V model), and our model without historical covariates and does not allow covariate effects to vary over time (C model).
-This result implies that explicit modeling of historical covariate effects on extinction yields similar forecasts to modeling how covariate effects have changed over time.
-We interpret this to mean that paleobiological information, such as variation in covariate effects on extinction risk over time, can improve our ability to forecast future extinctions, however including both historical coviarates and allowing covariate effects to vary over time yields an overfit model which produces biased forecasts.
-Importantly, while our model with no historical covariates and no variation on effects over time yielded similar out-of-sample forecasts to our model with historical covariates but constant effect and our model with varying effects but no historical covariates, the inclusion of this additional information does not harm these models' forecast ability.
-The results of this simple exercise suggest that conservation decisions would be bolstered by including fossil data.
+\section{Acknowledgements}
+
+We thank Jonathan Payne, Thomas Ezard, Erin Saup, and Peter Roopnarine for useful discussions.
+Wolfgang Kiessling, Matthew Clapham, and an anonymous reviewer provided thoughful reviews that strengthened the manuscript consideraly.
+This work was funded by a David and Lucile Packard Fellowship to SF.
 
@@
 \end{refsection}
-\clearpage
+%\clearpage
 
-\beginsupplement
-\begin{refsection}
+\end{document}
 
+%\beginsupplement
+%\begin{refsection}
+%
 %\section{Supplemental Figures}
-
-\setcounter{page}{1}
-\section{Supplement to Materials and Methods}
-
-\subsection{Data Specifications} \label{sec:data_desc}
-
-\subsubsection{Binning fossil occurrences}
-
-The estimated age of each occurrence is based on the core-specific age-model that observation is from and can be overly precise.
+%
+%\setcounter{page}{1}
+%\section{Supplement to Materials and Methods}
+%
+%\subsection{Data Specifications} \label{sec:data_desc}
+%
+%\subsubsection{Binning fossil occurrences}
+%
+%The estimated age of each occurrence is based on the core-specific age-model that observation is from and can be overly precise.
 To alleviate this overprecision, we coarsened our temporal information in an effort to limit the effects of between-core heterogeneity in age.
@@
 In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores \citep{Mekik2018,Broecker1999,Mekik2014,Peng1984}.
-
-\subsubsection{Covariate transformation and standardization}
-
-Prior to analysis, geographic range was then log-plus-one transformed and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges.
+%
+%
+%\subsubsection{Covariate transformation and standardization}
+%
+%Prior to analysis, geographic range was then log-plus-one transformed and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges.
 This standardization means that a regression coefficient associated with each covariate describes the change in extinction probability per change in standard deviation of that covariate, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}.
 Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.
-
-Temperature was also transformed and standardized the in the same manner as geographic range.
+%
+%Temperature was also transformed and standardized the in the same manner as geographic range.
 The change in temperature between an observation and its previous observation was measured from the standardized temperature values and was not standardized separately.
-
-\subsection{Model Specifications} \label{sec:model_desc}
-
-In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and covariate information.
+%
+%
+%\subsection{Model Specifications} \label{sec:model_desc}
+%
+%In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and covariate information.
 The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals \citep{Tutz2016}.
@@
 The discrete time hazard function is defined as
-\begin{equation}
- \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k.
- \label{eq:hazard}
-\end{equation}
-
-The hazard function (Eq.
+%\begin{equation}
+% \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k.
+% \label{eq:hazard}
+%\end{equation}
+%
+%The hazard function (Eq.
 \ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\) \citep{Tutz2016}.
@@
 In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes \citep{ARM}.
-
-Our covariates matrix \(X\) is a \(N \times D\) matrix where \(N\) is the total number of observations and \(D\) is the total number of covariates.
+%
+%Our covariates matrix \(X\) is a \(N \times D\) matrix where \(N\) is the total number of observations and \(D\) is the total number of covariates.
 The first column of \(X\) is entirely 1's as it corresponds to the intercept term in the regression model.
@@
 mean temperature during interval \(t - 1\).) As with change to geographic range, the lag of temperature is based on the transformed and standardized temperature estimates.
-
-The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla.
+%
+%The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla.
 The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each phylum \(p\).
@@
 \(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.
-
-The effect of species age on the log-odds of species extinction is modeled as a non-nested random intercept \(A\) \citep{Tutz2016}.
+%
+%The effect of species age on the log-odds of species extinction is modeled as a non-nested random intercept \(A\) \citep{Tutz2016}.
 This term describes how the log-odds of extinction varies along a species duration, and how this effect can differ between the phyla.
@@
 The covariation between the elements of vector \(\delta\) are modeled as a multivariate normal distribution with a mean vector of all 0s and covariance matrix \(\Sigma_{\delta}\).
-
-To complete the generative model, we need to assign final priors to the ``top-level'' parameters.
+%
+%To complete the generative model, we need to assign final priors to the ``top-level'' parameters.
 In general, we favored weakly informative priors which help regularize our estimates.
@@
 standard deviations), we used half-Cauchy distributed priors with heavy tails but the majority of probability density near 0.
-
-Our top-level intercept was given a more diffuse prior than our regression coefficients, which reflects our greater degree of uncertainty about its value.
+%
+%Our top-level intercept was given a more diffuse prior than our regression coefficients, which reflects our greater degree of uncertainty about its value.
 Our top-level regression coefficient for the effect of geographic range was given an informative prior reflecting the overwhelming amount of evidence that species with a larger than average geographic range have a lower risk of extinction than species with an average or less than average geographic range.
@@
 larger than average geographic range decreases extinction risk).
-
-Instead of assigning a prior distribution for each of the covariance matrices in the model, we instead decomposed the covariance matrices (e.g.
+%
+%Instead of assigning a prior distribution for each of the covariance matrices in the model, we instead decomposed the covariance matrices (e.g.
 \(\Sigma_{B}\)) which allows us to assign independent priors for the scale and correlation aspects of covariance.
@@
 This approach to assigning priors to a covariance matrix is recommended by the Stan Manual \citep{StanManual}.
-
-In total, our model can be expressed as: 
-\begin{equation}
- \begin{aligned}
-  t_{i} &\sim \text{Bernoulli}(\Theta) \\
-  \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{l[i], p[i]}) \\
-  B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
-  \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
-  % double check this\dots is A MVN dist?
-  A_{l, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
-  \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
-  \mu_{d} &\sim 
-  \begin{cases}
-   N(-2, 5) & \text{if } d = \text{intercept} \\
-   N(-1, 1) & \text{if } d = \text{geo. range} \\
-   N(0, 1) & \text{else } \\
-  \end{cases} \\
-  \delta &\sim N(0, 1) \\
-  \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
-  \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
-  \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
-  \tau_{B} &\sim C^{+}(1) \\
-  \tau_{\alpha} &\sim C^{+}(1) \\
-  \tau_{A} &\sim C^{+}(1) \\
-  \Omega_{B} &\sim LKJ(1) \\
-  \Omega_{\alpha} &\sim LKJ(1) \\
-  \Omega_{A} &\sim LKJ(1) \\
- \end{aligned}
- \label{eq:model}
-\end{equation}
-with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.
-
-\subsection{Model Parameter Estimation} \label{sec:model_est}
-
-We implemented our model (Eq.
+%
+%In total, our model can be expressed as: 
+%\begin{equation}
+% \begin{aligned}
+%  t_{i} &\sim \text{Bernoulli}(\Theta) \\
+%  \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{l[i], p[i]}) \\
+%  B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
+%  \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
+%  % double check this\dots is A MVN dist?
+%  A_{l, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
+%  \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
+%  \mu_{d} &\sim 
+%  \begin{cases}
+%   N(-2, 5) & \text{if } d = \text{intercept} \\
+%   N(-1, 1) & \text{if } d = \text{geo. range} \\
+%   N(0, 1) & \text{else } \\
+%  \end{cases} \\
+%  \delta &\sim N(0, 1) \\
+%  \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
+%  \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
+%  \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
+%  \tau_{B} &\sim C^{+}(1) \\
+%  \tau_{\alpha} &\sim C^{+}(1) \\
+%  \tau_{A} &\sim C^{+}(1) \\
+%  \Omega_{B} &\sim LKJ(1) \\
+%  \Omega_{\alpha} &\sim LKJ(1) \\
+%  \Omega_{A} &\sim LKJ(1) \\
+% \end{aligned}
+% \label{eq:model}
+%\end{equation}
+%with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.
+%
+%
+%\subsection{Model Parameter Estimation} \label{sec:model_est}
+%
+%We implemented our model (Eq.
 \ref{eq:model} using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}.
@@
 In order to prevent divergent transitions adapt delta was increased to 0.999999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.
-
-To implement our VP model in \begin{texttt}rstanarm\end{texttt}, where ``data'' is a data.frame object of all necessary data (response, covariates), is coded as:
-\begin{verbatim}
-form <- event ~ 
-        range + range_diff1 + range_diff2 + range_diff3 + 
-        temp + temp_lag + 
-        (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
-        (1 | age/phylum), 
-stan_glmer(formula = form,
-      data = data, 
-      family = 'binomial',
-      prior = normal(c(-1, 0, 0, 0, 0, 0), rep(1, 6), autoscale = FALSE), 
-      prior_intercept = normal(-2, 5, autoscale = FALSE), 
-      prior_aux = cauchy(0, 1, autoscale = FALSE), 
-      chains = 4,
-      thin = 4,
-      adapt_delta = 0.999999)
-\end{verbatim}
-
-Similarly, our VP model can be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
-\begin{verbatim}
-priors <- c(set_prior('normal(-2, 5)', class= 'Intercept'),
-      set_prior('normal(0, 1)', class = 'b'),
-      set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
-      set_prior('cauchy(0, 1)', class = 'sd'),
-      set_prior('lkj(1)', class = 'cor'))
-form <- bf(event ~ 
-           range + range_diff1 + range_diff2 + range_diff3 + 
-           temp + temp_lag +
-           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
-           (1 | age/phylum))
-brmfit <- brm(formula = form,
-       data = data, 
-       family = bernoulli(), 
-       prior = priors,
-       chains = 4, 
-       thin = 4,
-       control = list(adapt_delta = 0.999999)
-\end{verbatim}
-
-Posterior convergence was determined using the general and HMC-specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)).
+%
+%To implement our VP model in \begin{texttt}rstanarm\end{texttt}, where ``data'' is a data.frame object of all necessary data (response, covariates), is coded as:
+%\begin{verbatim}
+%form <- event ~ 
+%        range + range_diff1 + range_diff2 + range_diff3 + 
+%        temp + temp_lag + 
+%        (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
+%        (1 | age/phylum), 
+%stan_glmer(formula = form,
+%      data = data, 
+%      family = 'binomial',
+%      prior = normal(c(-1, 0, 0, 0, 0, 0), rep(1, 6), autoscale = FALSE), 
+%      prior_intercept = normal(-2, 5, autoscale = FALSE), 
+%      prior_aux = cauchy(0, 1, autoscale = FALSE), 
+%      chains = 4,
+%      thin = 4,
+%      adapt_delta = 0.999999)
+%\end{verbatim}
+%
+%Similarly, our VP model can be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
+%\begin{verbatim}
+%priors <- c(set_prior('normal(-2, 5)', class= 'Intercept'),
+%      set_prior('normal(0, 1)', class = 'b'),
+%      set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
+%      set_prior('cauchy(0, 1)', class = 'sd'),
+%      set_prior('lkj(1)', class = 'cor'))
+%form <- bf(event ~ 
+%           range + range_diff1 + range_diff2 + range_diff3 + 
+%           temp + temp_lag +
+%           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
+%           (1 | age/phylum))
+%brmfit <- brm(formula = form,
+%       data = data, 
+%       family = bernoulli(), 
+%       prior = priors,
+%       chains = 4, 
+%       thin = 4,
+%       control = list(adapt_delta = 0.999999)
+%\end{verbatim}
+%
+%Posterior convergence was determined using the general and HMC-specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)).
 For further explanation of these diagnostic criteria, see the Stan Manual \citep{StanManual}.
-
-\section{Supplement to Results} \label{sec:supp_res}
-
-Here we present the parameter estimates for our from our VP model (Table \ref{tab:model_def}).
+%
+%\section{Supplement to Results} \label{sec:supp_res}
+%
+%Here we present the parameter estimates for our from our VP model (Table \ref{tab:model_def}).
 We choose to present the estimates from this model because it is our most inclusive model and its parameter estimates are indicative of parameter estimates from other models with a subset of those covariates included in this model.
-
-First, we present the group-level estimates of our regression coefficients and intercept (Fig.
+%
+%First, we present the group-level estimates of our regression coefficients and intercept (Fig.
 \ref{fig:param_est}).
@@
 In addition to the average effect of our covariates, we also included the estimate of the group-level intercept, which is the average log-odds of extinction for an average observation over time and across taxonomic groups.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/effect_est}
-  \caption{Group-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
+%
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/effect_est}
+%  \caption{Group-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
 The posterior distribution of our parameter estimates are presented as densities.
@@
 Estimates are on the log-odds scale.}
-  \label{fig:param_est}
-\end{figure}
-
-Second, we present the population-level estimates for our regression coefficients along with the population-level intercept estimates (Fig.
+%  \label{fig:param_est}
+%\end{figure}
+%
+%Second, we present the population-level estimates for our regression coefficients along with the population-level intercept estimates (Fig.
 \ref{fig:param_est_time_group}.
@@
 In this case, the intercept estimate describes the log-odds of extinction for an average observation at that point in time and taxonomic group.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/eff_time_group}
-  \caption{Population-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
+%
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/eff_time_group}
+%  \caption{Population-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
 Posterior estimates are presented as a time series.
@@
 Estimates are on the log-odds scale.}
-  \label{fig:param_est_time_group}
-\end{figure}
-
-\printbibliography[title={Supplementary References}]
-\end{refsection}
-
-\end{document}
+%  \label{fig:param_est_time_group}
+%\end{figure}
+%
+%
+%
+%\printbibliography[title={Supplementary References}]
+%\end{refsection}

### 2019-07-08 doc/napc2019_talk.tex
process for resubmit

@@
     \begin{LARGE}
-      Extinction is \alert{hard} to predict, but is \alert{important} to conservation decisions.
+      Extinction is \alert{hard} to predict, \\but is \alert{important} to conservation decisions.
     \end{LARGE}
@@
 
+  \begin{Large}
     \begin{itemize}[<+->]
@@
     \end{itemize}
+  \end{Large}
 
@@
 
+  \begin{Large}
     \begin{itemize}%[<+->]
@@
         \begin{itemize}
+            \begin{large}
             \item Change in geographic range between current observation and previous observation.
@@
             \item Age in millions of years at time of observation.
+            \end{large}
         \end{itemize}
-    %\item \alert{Compare} models using WAIC/LOOIC.
       \item Explore model adequacy using posterior predictive distribution.
@@
     \begin{column}{0.6\textwidth}
+      \begin{large}
+
       The area represents the probability of correct ranking of a random ``extant''-``extinct'' pair.
@@
       \]
+
+      \end{large}
 
@@
 
+  \begin{large}
     \begin{itemize}
@@
         \end{itemize}
-    \item \alert{But not that much\dots}
+      \item \alert{But extinction is still hard to predict}
         \begin{itemize}
@@
     \end{itemize}
+  \end{large}
 

### 2019-07-08 doc/supplement.tex
process for resubmit

--- 2019-06-19 doc/paper.tex
@@
 \begin{document}
-
-\begin{refsection}
-
-\maketitle
-
-\linenumbers{}
-\modulolinenumbers[3]
-
-\begin{abstract}
-  A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions.
-Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past.
-To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
-We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
-Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time.
-We find that models which include either historical covariates or account for variation in covariate effects over time yield equivalent forecasts, but a model including both is overfit and yields biased forecasts.
-An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
-
-\end{abstract}
-
-{\bf Keywords:} conservation, palaeobiology, extinction, forecasting
-
-\section{Introduction}
-
-The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the near future.
-Predicting which species will go extinct is difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented.
-This has led to the suggestion that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}.
-The fossil record preserves information about the full histories, include ultimate extinction, of thousands of lineages, and this information can help to augment the shorter-term higher-resolution data used to make risk assessments of extant taxa.
-
-Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
-This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how capable we are at predicting future extinction events.
-
-We ask how accurate risk predictions based on fossil data are when estimating future extinctions.
-Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions.
-However, we can address the question by taking a given point in the geological past, developing a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g.
-``future'', from the point of view of the model) extinction/survival events.
-
-Extinction intensity (average rate) and selectivity (difference in risk among taxa) vary through time, and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}.
-This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
-Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.
-
-Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
-The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}.
-In the former case, knowledge of the specific past trajectory of a species -- whether its range has expanded or contracted over a given time span -- might help to improve assessments of its current risk.
-In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past.
-Discriminating among these alternative models is thus very important for determining how best to incorporate fossil data in present risk assessments.
-
-For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores).
-These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}.
-This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomy synonymization strategy for delaing wiht 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible.
-Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.
-
-\section{Materials and Methods}
-
-\subsection{Data Specifications}
-
-We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
-This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic foraminifera, and radiolarians.
-Occurrences were filtered to include only those species who's first occurrence is no earlier than 63 Mya (millions of years ago).
-This filtering criterion excludes taxa that survived the K/Pg extinction or arose during this recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).
-
-All fossil occurrences were assigned to 1 My (million year) bins based on the estimated age of the fossil occurrence as listed in the Neptune Database.
-After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred.
-Geographic range was calculated as the minimum spanning tree distance between all observations of that taxon during that temporal bin; this distance was measured in kilometers.
-Minimum spanning tree distance was caluclated using the \texttt{GeoRange} package for R \citep{GeoRange}.
-
-We also included how a taxon's geographic range has changed since its last three observation times.
-We measured this change in geographic range by calculating the difference in geographic range between an observation and that taxon's three previous occurrences.
-If there are not enough previous observations to calculate, then that value is recorded as a 0.
-These differences were calculated after minimum spanning tree distance was transformed and standardized (see Section \ref{sec:data_desc}).
-
-Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}.
-We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
-Estimating temperature over long periods of time from Mg/Ca ratios also suffers from complications because Mg/Ca based temperature estimates requires benthic foram and seawater Mg/Ca isotope ratio information.
-Because seawater Mg/Ca ratio has changed over time, the method to estimate temperature used \citet{Cramer2011} attempts to account for this unknown in order to obtain accurate, albeit uncertain, temperature estimates.
-Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My interval between 0 to 63 Mya.
-The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.
-
-We also included the global temperature from the previous time that taxon was observed.
-If there are not enough previous observations to calculate, then that value is recorded as a 0.
-This lag was calculated after global temperature was transformed and standardized (see Section \ref{sec:data_desc}).
-
-Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature.
-The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit.
-Ideally, we would have detailed ocean surface water temperature estimates for the entire globe for the entire Cenozoic.
-Unfortunately, that type of data does not exist.
-So, we interpret our temperature estimates as reflecting the global climate state that an organisms experiences, and not as a descriptor of that taxon's local environmental ecology.
-
-See Section \ref{sec:data_desc} for a further explanation on how observations were temporally binned, and how our covariates were standardized and transformed prior to analysis.
-
-%\begin{figure}[ht]
-% \centering
-% \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
-% \caption{Comparison of initial temperature estimates from \citet{Cramer2011} (goldenrod) versus the binned values used in this analysis (blue).
-The initial values are for every 0.1 My while our bins are defined for every 1 My.}
-% \label{fig:temp_curve}
-%\end{figure}
-
-\subsection{Model Specifications}
-
-We used a discrete-time survival modelling framework to estimate how well we can predict extinction risk at one million year time scales.
-At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}.
-We considered four different models involving different permutations of covariate effects (fixed or time-varying) and historical covariates: covariate effects constant over time and no historical covariates included (Model C), covariate effects allowed to vary over time but no historical covariates included (Model V), covariate effects constant over time and historical covariates included (Model CP), and covariate effects allowed to vary over time and historical covariates are included (Model VP).
-The C and P models attempt to predict based only on present state, whereas the CP and VP models allow for the possibility of non-Markovian behaviour by including change in state from the previous time increment.
-
-We always included species age at time of observation (i.e.
-observed prior duration) as a varying-intercept term.
-This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.
-The effect of species age is allowed to vary by taxonomic group.
-
-Similarly, we included time of obseration as an additional varying-intercept term to account for changes average global extinction risk over time that are not related to the covariates included in this model.
-This varying-intercept is further allowed to vary by taxonomic group.
-This varying-intercept term allows us to tease apart the differences in extinction risk associated with time of observation versus age since first observation.
-An important note is that for our V and VP models, the covariation between this varying-intercept and the varying-slopes of our covariates is explicitly modeled (see Section \ref{sec:model_desc}).
-
-See Table \ref{tab:model_def} for further explanation of how the four models we considered differ from each other.
-A complete description of the statistical model used in this analysis is available in Section \ref{sec:model_desc}.
-Additionally, the full description of how these models were implemented and coded, including choice of priors, is available in Section \ref{sec:model_est}.
-
-\begin{table}[ht]
- \caption{Models and their definitions}
- \begin{threeparttable}
-  {
-   \def\arraystretch{1.5}
-   \begin{tabular}{ l p{3cm} l l }
-    Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
-    \hline
-    C & Constant effects, no historical cov.
-& \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ time\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}}) +\\(1 $|$ age\tnote{g}\phantom{\textsuperscript{g}}/phylum)} \\
-    V & Varying effects, no historical cov.
-& \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ time/phylum) + (1 $|$ age/phylum)} \\
-    CP & Constant effects, historical cov.
-& \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ + range\_diff1\tnote{g}\phantom{\textsuperscript{h}} + range\_diff2\tnote{h}\phantom{\textsuperscript{h}} + range\_diff3\tnote{h}\phantom{\textsuperscript{h}} + \\temp + temp\_lag\tnote{i}\phantom{\textsuperscript{i}} +\\(1 $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
-    VP & Varying effects, historical cov.
-& \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff1 + range\_diff2 + range\_diff3 + \\temp + temp\_lag +\\(1 + range + range\_diff1 + range\_diff2 + range\_diff3 +\\temp + temp\_lag $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
-    \hline
-   \end{tabular}
-  }
-  \begin{tablenotes}
-  \item[a] See Equation \ref{eq:model} for full statistical model definition.
-  \item[b] Species observation where 1 if time of last observation, otherwise 0.
-  \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
-  \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
-  \item[e] Time of observation.
-  \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcareous nannoplankton).
-  \item[g] Age at observation.
-  \item[h] Change in geographic range since last observation (number indicates how lags).
-  \item[i] Temperature at previous observation.
-  \end{tablenotes}
- \end{threeparttable}
- \label{tab:model_def}
-\end{table}
-
-\subsection{In-sample and out-of-sample forecasting}
-
-We are interested in our models' performance in two contexts: in-sample performance, and out-of-sample predictive performance (i.e. forecasting).
-
-In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to recapitulate the data to which it was fit.
-Posterior predictive checks are a type of sensitivity analysis because we are checking the quality of model's fit to the data.
-If our models have poor in-sample forecasting performance, than our models are not adequate descriptors of the data and will most likely make poor out-of-sample predictions.
-In-sample forecasting measures, however, are not the same as understanding our models' ability to forecast future extinctions or if our models are overfit to our data and produce biased out-of-sample estimates \citep{ESL}.
-
-We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data).
-To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation.
-For time-series data, the folds (data partitions) are approximately equal segments of time.
-Each fold represents a sequence of time points.
-With 63 time points, each of the five folds represents approximately 13 million-year time increments.
-It is important to bear in mind, however, that each time increment includes many (100s-1000s) individual observations.
-
-\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016,ESL}.
-Prior to cross-valdiation, the data is divided into \textit{k} nearly even segments or folds -- for a time series, this means the data is divided into \textit{k} continuous sequences.
-Next, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e.
-the future).
-Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold.
-Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold.
-Next, the model is fit to the first four folds combined and then is used to forecast the fifth fold.
-This process continues until $k - 1$ folds are included in the fitting the model and the final fold is predicted from this model.
-When combined, the results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.
-In 5-fold cross validaition, the data is divided into five folds the cross-validation procedure yields predictions for four of the folds.
-
-Cross-validation is a procedure for estimating a models expected out-of-sample error, which is a similar goal to metrics such as AIC or WAIC \citep{ESL,Gelman2013d}.
-Measures like AIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}.
-Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, then out-of-sample predictions will be very biased and inaccurate \citep{ESL}.
-Similarly, a high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).
-Cross-validation is preferable to simple metrics such as AIC because instead of a single value, we have an entire posterior distribution of estimates.
-
-The relative adequacy of the four model variants was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}.
-This measure is commonly used to measure the performance of classification models as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives.
-AUC ranges between 0.5 and 1, with 0.5 indicating no difference in classification from random and 1 indicating perfect classification.
-AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of a randomly selected extinct-extant species pair \citep{Fawcett2006,Mason2002}.
-AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we consider values between between 0.7 and 0.8 as ``fair,'' and values between 0.6 and 0.7 as ``poor.''
-
-%The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time.
-These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.
-
-See our code repository at https://github.com/psmits/trident for full code details.
-This entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}.
-Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.
-
-\section{Results}
-
-The focus of this study is on understanding how well our models forecast future extinction events by comparing our in-sample and out-of-sample forecast estimates.
-A presentation of the posterior estimates for the regression coefficient estimates from our VP model (Table \ref{tab:model_def}) is available our Supplemental Materials (Section \ref{sec:supp_res}).
-
-% ROC model comparison
-\begin{figure}[ht]
- \begin{subfigure}[ht]{0.45\textwidth}
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
-  \caption{In-sample}
-  \label{fig:auc_hist}
- \end{subfigure}
- \begin{subfigure}[ht]{0.45\textwidth}
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
-  \caption{Out-of-sample}
-  \label{fig:fold_auc}
- \end{subfigure}
- \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation.
-The area under the receiver operating charactered curve (AUC) was calculated for each model.
-these estimates are calculated from the models posterior predictive distribution (\ref{fig:auc_hist}) or from predictions made to new data (\ref{fig:fold_auc}), respectively.
-Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in our dataset.
-Models with higher AUC values indicate better performance over models with lower AUC values.
-AUC is bounded between 0.5 and 1.
-See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
- \label{fig:auc_compare}
-\end{figure}
-
-\subsection{In-sample forecasting adequacy}
-
-The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis.
-Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that the parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, while there is substantial overlap in the posterior distributions of the forecasts from the other three models (Fig.
-\ref{fig:auc_hist}).
-
-However, the actual difference in forecast AUC result between the VP model and the other three models is extremely small (0.01 AUC unit), and all the in-sample AUC estimates from our models are concentrated between an AUC value of 0.775 and 0.795 (Fig.
-\ref{fig:auc_hist}).
-This result indicates that the practical difference in performanc between these models might be so small that there is no practical benefit that the VP model over the other three.
-Ultimately, determining which of these models produces the best forecasts of future extinctions requires comparing these in-sample results to our out-of-sample results (see below).
-
-The in-sample forecasts from our four models over time are broadly similar between taxonomic groups (Fig.
-\ref{fig:auc_taxon_time}).
-Our in-sample forecasts for Diatoms are the weakest of the taxonomic groups as all four models have an equal number of intervals with no predictive power (AUC approximately 0.5).
-In contrast, our best in-sample forecast results are for Radiolarians where for any of our models there is at most 1 interval with almost no predictive power.
-Ultimately, our in-sample forecasts over time by taxonimc group are broadly consistent between our four models.
-
-\begin{figure}[ht]
- \centering
- \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
- \caption{Understanding model adequacy over time and taxonomic group by comparing in-sample forecasting performance measured by AUC for each of the four models.
-These estimates reflect each model's fit to the various taxonomic groups over time.
-The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend.
-In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.
-The grey intervals mark the geologic ages of the Cenozoic.
-See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
- \label{fig:auc_taxon_time}
-\end{figure}
-
-\subsection{Out-of-sample forecasting performance}
-
-Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}.
-Our out-of-sample forecast AUC estimates demonstrate a broader range of results, with AUC estimates ranging between approximately 0.7 and 0.85 (Fig.
-\ref{fig:auc_hist}, \ref{fig:fold_auc}).
-While our VP model was the model with greatest in-sample forecasting performance (Fig.
-\ref{fig:auc_hist}), this model has the greatest decrease in out-of-sample forecasting performance (Fig.
-\ref{fig:fold_auc}).
-In comparison, the other three models demontrstrate a much smaller difference between in-sample forecast and out-of-sample forecast AUC values which indicates that our VP model is overfit to our data, and that one of the simpler model would be preferable for predicting future extinctions.
-This result means that our model which includes the historical covariates (e.g.
-change in geographic range) and allowing these effects to vary over time produces biased extinciton forecasts.
-Interestingly, including either historical covariates like change in geographic range or allowing the effects of geographic range and other covariate effects to vary over time have approximately equal effect on forecasting future extinction events (Fig.
-\ref{fig:auc_compare}).
-
-In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig.
-\ref{fig:auc_taxon_time}).
-This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts.
-The major exception to this pattern are our estimates for the diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random.
-The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
-\begin{figure}[ht]
- \centering
- \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
- \caption{Comparison of our models ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
-The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds.
-This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals.
-The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.
-The grey intervals mark the geologic ages of the Cenozoic.
-See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
- \label{fig:fold_auc_taxon_time}
-\end{figure}
-
-We compared the difference in our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig.
-\ref{fig:oos_ins_diff}).
-A difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts.
-A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast.
-Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig.
-\ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).
-The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig.
-\ref{fig:oos_ins_diff}).
-
-\begin{figure}[ht]
- \centering
- \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
- \caption{Comparing our models' ability to forecast out-of-sample extinction compared to our models' in-sample forecasts.
-Congruence between these measures indicates that our models are not necessarily overfit to the data.
-This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}.
-A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast.
-See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
- \label{fig:oos_ins_diff}
-\end{figure}
-
-\section{Discussion}
-
-We find that all of our models are expected to correctly forecast which species of randomly selected extinct-extant pair is more likely to go extinct between 70\% to 80\% of the time (Fig.
-\ref{fig:fold_auc}).
-One of the most striking results is that the models that include either the historical covariates (CP model) or allows covariate effects to vary over time (V model) produce less biased out-of-sample forecasts than the model which includes both historical covariates and allows covariate effects to vary over time (VP model) (Fig.
-\ref{fig:auc_compare}).
-However, we might expect a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect.
-An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
-
-We find that three of our four models are practically identical in their ability to make in-sample and out-of-sample forecasts.
-Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig.
-\ref{fig:auc_hist}).
-While the model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time, the out-of-sample forecasts from this mode demonstrate this model yields biased forecasts, which indicate that this model is overfit to our data \ref{fig:fold_auc}).
-Including historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to a model that allows covariate effects to vary over time but does not include historical covariates.
-
-It is interesting to note that both historical covariates like change in geographic range and allowing covariate effects to vary over time can be thought of as different ways of encoding the past: explicitly or implicitly.
-Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time.
-Allowing covariate effects to vary over time, on the other hand, does not explicitly encode ``the past'' into our model but instead models how covariate effects vary over time which allows the past to be implicitly included in our model.
-By modeling this variation, forecasts made for future extinction events can allow for a wide range of potential effects of our covariates on predicting extinction.
-Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig.
-\ref{fig:fold_auc}).
-Our results supplement those of \citet{Kiessling2016} which examined using differences in geographic range over time to predict extinction risk.
-
-It is imporatant to remember that our model with historical covariates assumes that extinction selection based on these covarites is constant over time.
-If human impacts substatially alter present and future extinction risk dynamics, the assumption the effects of covariates have not changed over time may be limit or bias our ability to predict extinction in truly novel environmental regimes.
-Thus, it might be preferable to model the variation in extinction risk and selectivity over time instead of relying solely on measures from a taxon's past such as change in geographic range over time.
-Thus, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.
-
-The relative quality and consistency between three of our models' in-sample and out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
-Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity.
-The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study.
-Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}.
-However, comparable information does not exist for all foram species, nor does this type of data exist for the other three taxonomic groups studied here.
-Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.
-
-In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of taxonomic, geographic, and historical predictors.
-We find that our VP model chich includes the historical covariates such as change in geographic range between observations and past global temperature information while also allowing covariate effects to vary over time is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models.
-
-We find that three of our models yield functional equivalent out-of-sample forecasts: our model with historical covariates but assume covariate effects are constant over time (CP model), our model which allows for covariate effects to vary over time but does not include the historical covarites (V model), and our model without historical covariates and does not allow covariate effects to vary over time (C model).
-This result implies that explicit modeling of historical covariate effects on extinction yields similar forecasts to modeling how covariate effects have changed over time.
-We interpret this to mean that paleobiological information, such as variation in covariate effects on extinction risk over time, can improve our ability to forecast future extinctions, however including both historical coviarates and allowing covariate effects to vary over time yields an overfit model which produces biased forecasts.
-Importantly, while our model with no historical covariates and no variation on effects over time yielded similar out-of-sample forecasts to our model with historical covariates but constant effect and our model with varying effects but no historical covariates, the inclusion of this additional information does not harm these models' forecast ability.
-The results of this simple exercise suggest that conservation decisions would be bolstered by including fossil data.
-
-\printbibliography
-\end{refsection}
-\clearpage
-
 \beginsupplement
@@
 
-Here we present the parameter estimates for our from our VP model (Table \ref{tab:model_def}).
+Here we present the parameter estimates for our from our VP model (Table 1).
 We choose to present the estimates from this model because it is our most inclusive model and its parameter estimates are indicative of parameter estimates from other models with a subset of those covariates included in this model.
@@
   \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/effect_est}
-  \caption{Group-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
+  \caption{Group-level parameter posterior estimates from our VP model (Table 1).
 The posterior distribution of our parameter estimates are presented as densities.
@@
   \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/eff_time_group}
-  \caption{Population-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
+  \caption{Population-level parameter posterior estimates from our VP model (Table 1).
 Posterior estimates are presented as a time series.

### 2019-07-09 doc/article.tex
seth discussion comments

@@
 Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past.
-To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
+To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (Foramanifera, Radiolaria, diatoms, and calcareous nanoplankton).
 We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
@@
 
-Here we take this approach, using as a model system the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores).
+Here we take this approach, using as a model system the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, diatoms, and calcareous nannoplankton including coccolithophores).
 This record has several key strengths for our purposes: planktonic microorganisms are widespread and abundant in pelagic habitats, have high preservation potential, and because of their utility for biostratigraphic, paleoclimatic, and oceanographic study they have been the focus of an extensive international coring and study effort \citep{Lazarus1994,SpencerCervato1999}.
@@
 We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
-This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic foraminifera, and radiolarians.
+This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic Foraminifera, and radiolarians.
 Occurrences were filtered to include only those species with first occurrences no earlier than 63 Mya (millions of years ago).
@@
   \item[e] Time of observation.
-  \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcareous nannoplankton).
+  \item[f] Taxonomic group of species (i.e. Foraminifera, diatoms, Radiolaria, calcareous nannoplankton).
   \item[g] Age at observation.
@@
 
-We are interested in our models' performance in two distinct contexts: in-sample performance, and out-of-sample predictive performance (i.e. forecasting).
+We are interested in model performance (i.e. forecasting) in two distinct contexts: in-sample performance, and out-of-sample predictive performance.
 
@@
 
-Cross-validation is a procedure for estimating a models expected out-of-sample error.
+Cross-validation is a procedure for estimating a model's expected out-of-sample error.
 Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}.
@@
 these estimates are calculated from the models posterior predictive distribution (\ref{fig:auc_hist}) or from predictions made to new data (\ref{fig:fold_auc}), respectively.
-Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in our dataset.
+Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in the dataset.
 Models with higher AUC values indicate better performance over models with lower AUC values.
@@
 \ref{fig:auc_hist}).
-This result indicates that the practical difference in performanc between these models might be so small that there is no practical benefit that the VP model over the other three.
+This result indicates that the practical difference in performance between these models might be so small that there is no practical benefit that the VP model over the other three.
 Ultimately, determining which of these models produces the best forecasts of future extinctions requires comparing these in-sample results to our out-of-sample results (see below).
 
-The in-sample forecasts from our four models over time are broadly similar between taxonomic groups (Fig.
+In-sample forecasts from the four models over time are broadly similar between taxonomic groups (Fig.
 \ref{fig:auc_taxon_time}).
-Our in-sample forecasts for Diatoms are the weakest of the taxonomic groups as all four models have an equal number of intervals with no predictive power (AUC approximately 0.5).
-In contrast, our best in-sample forecast results are for Radiolarians where for any of our models there is at most 1 interval with almost no predictive power.
-Ultimately, our in-sample forecasts over time by taxonimc group are broadly consistent between our four models.
+In-sample forecasts for diatoms are the weakest of the four taxonomic groups as all four models have several intervals with no predictive power (AUC not significantly greater than 0.5).
+The best in-sample forecast results are for radiolarians, for which all models have at most 1 interval with little predictive power.
+In-sample forecasts over time by taxonomic group are broadly consistent between the four models.
 
@@
 
-Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}.
-Our out-of-sample forecast AUC estimates demonstrate a broader range of results, with AUC estimates ranging between approximately 0.7 and 0.85 (Fig.
+We estimated expected out-of-sample forecasting performance using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}.
+Out-of-sample forecast AUC estimates exhibit a broader range than in-sample estimates, with AUC ranging between approximately 0.7 and 0.85 (Fig.
 \ref{fig:auc_hist}, \ref{fig:fold_auc}).
-While our VP model was the model with greatest in-sample forecasting performance (Fig.
-\ref{fig:auc_hist}), this model has the greatest decrease in out-of-sample forecasting performance (Fig.
+While the VP model has the best in-sample forecasting performance (Fig.
+\ref{fig:auc_hist}) this model performs poorly at out-of-sample forecasting compared to the other models (Fig.
 \ref{fig:fold_auc}).
-In comparison, the other three models demontrstrate a much smaller difference between in-sample forecast and out-of-sample forecast AUC values which indicates that our VP model is overfit to our data, and that one of the simpler model would be preferable for predicting future extinctions.
-This result means that our model which includes the historical covariates (e.g.
-change in geographic range) and allowing these effects to vary over time produces biased extinciton forecasts.
-Interestingly, including either historical covariates like change in geographic range or allowing the effects of geographic range and other covariate effects to vary over time have approximately equal effect on forecasting future extinction events (Fig.
+The poor out-of-sample performance suggests that this most complex model is overfit and that one of the simpler models would be preferable for predicting future extinctions.
+Thus the models that include both historical covariates (e.g.
+change in geographic range) and time-varying effects produce biased extinction forecasts.
+Interestingly, models that include either historical covariates but assume constant effects (CP) or do not include historical covariates but include time-varying effects (V) perform similarly when forecasting future extinction events (Fig.
 \ref{fig:auc_compare}).
 
-In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig.
+As noted above there were some time intervals in which in-sample forecasts were no better than random (Fig.
 \ref{fig:auc_taxon_time}).
-This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts.
-The major exception to this pattern are our estimates for the diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random.
+Such intervals are generally much rarer for out-of-sample forecasts.
+The major exception to this pattern are diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random.
 The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
@@
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
- \caption{Comparison of our models ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
+ \caption{Comparison of the four models' ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
 The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds.
@@
 
-We compared the difference in our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig.
-\ref{fig:oos_ins_diff}).
-A difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts.
-A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast.
-Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig.
+We compared the difference in AUC estimates from the out-of-sample forecasts to the AUC estimates from in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig.
+\ref{fig:oos_ins_diff}); a difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts.
+A positive difference indicates that out-of-sample forecasts actually outperform in-sample forecasts, whereas negative difference indicates poorer out-of-sample performance than in-sample forecast.
+Divergences between out-of-sample and in-sample forecasts are rare and tend not to cluster in time, consistent with the broad visual congruence between the in-sample and out-of-sample performance (Fig.
 \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).
@@
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
- \caption{Comparing our models' ability to forecast out-of-sample extinction compared to our models' in-sample forecasts.
-Congruence between these measures indicates that our models are not necessarily overfit to the data.
+ \caption{Comparing out-of-sample and in-sample forecasts.
+Congruence between in-sample and out-of-sample indicates that a model is not necessarily overfit to the data.
 This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}.
-A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast.
+A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that the out-of-sample forecasts are actually higher performing than the in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast.
 See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
@@
 The model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time.
-However, the out-of-sample forecasts from this model are biased forecasts, indicating that it is overfit to our data \ref{fig:fold_auc}).
+However, the out-of-sample forecasts from this model are biased, indicating that it is overfit to the data \ref{fig:fold_auc}).
 The model that includes historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to the model that allows covariate effects to vary over time but does not include historical covariates.
@@
 \ref{fig:fold_auc}).
-Our results supplement those of \citet{Kiessling2016} which examined using differences in geographic range over time to predict extinction risk.
-
-An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predicatable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
-Our model with historical covariates assumes that extinction selectivity with respect to these covariates is constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes.
-Thus, it might be preferable to model the variation in extinction risk and selectivity over time instead of relying solely on measures from a taxon's past such as change in geographic range over time.
-For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.
+Our results complement those of \citet{Kiessling2016} who found that including historical range trajectories improved the performance of extinction risk forecasts.
 
@@
 However, comparable information does not exist for all foram species, nor does this type of data exist for the other three taxonomic groups studied here.
-Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.
+Future analyses including this type of information and focused more narrowly on the Foraminifera may be informative.
 
-In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors.
+An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predicatable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
+The CP model assumes that the effects of the historical covariates are constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes.
+Thus, it might be preferable to use a model similar to our V model which allows extinction risk and selectivity to vary over time.
+For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.
+
+On a related note, it is notable that there are no obvious consistent changes in average model performance during episodes of climatic environmental change such as the mid-Eocene to early Oligocene (Fig.
+\ref{fig:fold_auc}), although this interval is characterized by elevated extinction among several planktonic groups \citep{Prothero1994,Wade2008,Kamikuri2012}.
+This suggests that relationships between the simple covariates included in our models and extinction risk were relatively stable though this interval.
+However, it is likely that more complex models accounting for other aspects of ecology and biogeography might exhibit reduced forecasting performance during this and other climate change episodes.
+
+To summarize, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors.
 Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models.

### 2019-07-24 doc/response_review.tex
repsonse reviews. second round started.

@@
 
-Below are our responses to the editor and reviewer comments that we recieved on our original manuscrpt.
+Below are our responses to the editor and reviewer comments that we received on our original manuscript.
 We would like to thank the editor and the three reviewers for their thoughtful and insightful comments which have greatly improved the manuscript.
@@
 
-Thank you for this suggestion.
-We have update our analysis to include change in geographic range from the last three observations.
-This change adds two more covariates to all models that include our ``historical covaraites.'' This addition to our analysis did change our results, and we now consider our most complex model that includes historical covariates and variation in covariate effect over time to be overfit to our data.
-Our models that include either historical covariates or variation in effects over time, however, are found to have functionaly equivalent out-of-sample performance.
+Thank you for this suggestion; see above re. the change we made in the timespace of historical covariates.
 
@@
 We have updated our analysis using this new range metric.
-The change in range metric did not change our interpretation of the results.
+The change in range metric did not substantively change our interpretation of the results.
 
@@
 
-``In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to correctly classify the data to which it was fit.
+Line 173: ``In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to correctly classify the data to which it was fit.
 Posterior predictive checks are a type of sensitivity analysis because we are checking the quality of model's fit to the data.
@@
 
-``The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis.
+Line 232: ``The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis.
 Comparison between the posterior distributions \dots''
@@
 
-WAITING FOR SETH
+Thank you for your suggestion to expand upon the implciations of the human impact on extinction risk and how that might affect how well we can predict future extinction events.
+We have added a section to our discussion in an attempt to address this.
+
+From the updated manuscript:
+
+Line 318: ``An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
+The CP model assumes that the effects of the historical covariates are constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes.
+Thus, it might be preferable to use a model similar to our V model which allows extinction risk and selectivity to vary over time.
+For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.''
+
+On a related note, it is notable that there are no obvious consistent changes in average model performance during episodes of climatic environmental change such as the mid-Eocene to early Oligocene (Fig.
+1b), although this interval is characterized by elevated extinction among several planktonic groups \citep{Prothero1994,Wade2008,Kamikuri2012}.
+This suggests that relationships between the simple covariates included in our models and extinction risk were relatively stable though this interval.
+However, it is likely that more complex models accounting for other aspects of ecology and biogeography might exhibit reduced forecasting performance during this and other climate change episodes.
 
@@
 
-``Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature.
+Line 134: ``Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature.
 The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit.

### 2019-07-24 doc/article.tex
repsonse reviews. second round started.

@@
 We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
-Estimating temperature over long periods of time from Mg/Ca ratios also suffers from complications because Mg/Ca based temperature estimates requires benthic foram and seawater Mg/Ca isotope ratio information.
+Estimating temperature over long periods of time from Mg/Ca ratios also suffers from complications because Mg/Ca based temperature estimates requires benthic Foraminifera and seawater Mg/Ca isotope ratio information.
 Because seawater Mg/Ca ratio has changed over time, the method to estimate temperature used \citet{Cramer2011} attempts to account for this unknown in order to obtain accurate, albeit uncertain, temperature estimates.
@@
 
-Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature.
+Mg/Ca based temperature estimates are measured from benthic Foraminifera, and are an estimate of deep water ocean temperature.
 The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit.
@@
 A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).
-Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces, an entire posterior distribution of estimates.
+Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces an entire posterior distribution of estimates.
 
@@
 
+To reiterate, the primary focus of this study is on understanding how well our models forecast future extinction events by comparing our in-sample and out-of-sample forecast estimates.
+A presentation of the posterior estimates for the regression coefficient estimates from our VP model (Table \ref{tab:model_def}) is available our Supplemental Materials (Section S2).
+
 See our code repository at https://github.com/psmits/trident for full code details.
@@
 
-The primary focus of this study is on understanding how well our models forecast future extinction events by comparing our in-sample and out-of-sample forecast estimates.
-A presentation of the posterior estimates for the regression coefficient estimates from our VP model (Table \ref{tab:model_def}) is available our Supplemental Materials (Section S2).
-
 % ROC model comparison
@@
  \begin{subfigure}[ht]{0.45\textwidth}
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
+   %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{auc_hist_full.png}
   \caption{In-sample}
@@
  \begin{subfigure}[ht]{0.45\textwidth}
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
+   %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{fold_auc_full.png}
   \caption{Out-of-sample}
@@
 
-The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis.
 Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that the parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, while there is substantial overlap in the posterior distributions of the forecasts from the other three models (Fig.
@@
 
-In-sample forecasts from the four models over time are broadly similar between taxonomic groups (Fig.
+In-sample forecasts from the four models over time are broadly similar among taxonomic groups (Fig.
 \ref{fig:auc_taxon_time}).
@@
 The best in-sample forecast results are for radiolarians, for which all models have at most 1 interval with little predictive power.
-In-sample forecasts over time by taxonomic group are broadly consistent between the four models.
+The pattern of high and low in-sample forecast performance is broadly similar among the four models.
 
@@
  \centering
- \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
+ %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{auc_taxon_time_full.png}
  \caption{Understanding model adequacy over time and taxonomic group by comparing in-sample forecasting performance measured by AUC for each of the four models.
@@
 
-We estimated expected out-of-sample forecasting performance using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}.
-Out-of-sample forecast AUC estimates exhibit a broader range than in-sample estimates, with AUC ranging between approximately 0.7 and 0.85 (Fig.
+Out-of-sample forecast AUC estimates, based on five-fold cross validation \citep{Arlot2010,Bergmeir2016}, exhibit a broader range than in-sample estimates, with AUC ranging between approximately 0.7 and 0.85 (Fig.
 \ref{fig:auc_hist}, \ref{fig:fold_auc}).
@@
 
-As noted above there were some time intervals in which in-sample forecasts were no better than random (Fig.
+As noted above, there were some time intervals in which in-sample forecasts were no better than random (Fig.
 \ref{fig:auc_taxon_time}).
@@
  \centering
- \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
+ %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{fold_auc_taxon_time_full.png}
  \caption{Comparison of the four models' ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
@@
 \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).
-The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig.
+One of the few multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model between 35 Mya and approximately 28 Mya (Fig.
 \ref{fig:oos_ins_diff}).
+There exist similar periods of worse out-of-sample forecasting performance for other combinations of taxonomic group and predictive model, but none are as long as that for Radiolaria.
 
@@
  \centering
- \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
+ %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{auc_diff.png}
  \caption{Comparing out-of-sample and in-sample forecasts.
@@
 The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study.
-Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}.
-However, comparable information does not exist for all foram species, nor does this type of data exist for the other three taxonomic groups studied here.
+Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many Foraminifera species \citep{Ezard2011}.
+However, comparable information does not exist for all Foraminifera species, nor does this type of data exist for the other three taxonomic groups studied here.
 Future analyses including this type of information and focused more narrowly on the Foraminifera may be informative.
 
-An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predicatable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
+An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
 The CP model assumes that the effects of the historical covariates are constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes.

### 2019-07-24 doc/letter_editor.tex
repsonse reviews. second round started.

@@
 
-Please find enclosed our manuscript entitled ``How predictable is extinction?
+Please find enclosed our revised manuscript entitled ``How predictable is extinction?
 Forecasting species survival at million-year timescales'' by myself (Peter D.
 Smits) and Seth Finnegan.
+We would like to thank the editor and the three reviewers for their thoughtful comments which greatly improved our manuscript.
+
 In this study we asked a relatively simple yet fundamental question for the emerging field of conservation paleobiology: how well do models conditioned on past observations predict future extinction events?
-To answer this question we analyzed the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) We examined how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
+To answer this question we analyzed the well-sampled fossil record of Cenozoic planktonic microfossil taxa.
+We examined how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
+Our models were found to have a 70-80\% probability of correctly forecasting the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction risk have varied only modestly through time.
 
-We used cross-validation to estimate the expected forecast performance of our model for future observations.
-Our models were found to have a 70-80\% probability of correctly forecasting the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction risk have varied only modestly through time.
-Including specific information about individual species trajectories resulted in only extremely marginal improvement in fit, suggesting that, at least at the million-year timescale, extinction risk is best described by dominantly Markovian (e.g.
-memoryless) models.
-
-The relative quality and consistency of our models’ out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
+The relative consistency of our models’ out-of-sample forecasting performance is encouraging given that these estimates are based on limited biological and environmental information.
 The results of this simple exercise bolster the case that fossil data can meaningfully inform present and future conservation decisions.
@@
 
-  \encl{Article, supplementary text}.
+  \encl{Article, supplementary text, response to reviewers}.
 

### 2019-07-24 doc/model_def.tex
repsonse reviews. second round started.

--- 2019-07-08 doc/article.tex
@@
 
-\newcommand{\beginsupplement}{
- \setcounter{section}{0}
- \renewcommand{\thesection}{S\arabic{section}}
- \setcounter{table}{0}
- \renewcommand{\thetable}{S\arabic{table}}
- \setcounter{figure}{0}
- \renewcommand{\thefigure}{S\arabic{figure}}
- \setcounter{equation}{0}
- \renewcommand{\theequation}{S\arabic{equation}}
-}
-
-% https://3d.bk.tudelft.nl/hledoux/blog/fiddling-biblatex/
-%-- formatting hell for biblatex
-%-- remove "In:"
-\renewbibmacro{in:}{}
-%-- no "quotes" around titles of chapters/article titles
-\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]
-{title}{#1}
-%-- no punctuation after volume
-\DeclareFieldFormat[article]{volume}{ {#1} } 
-%-- puts number/issue between brackets
-\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]{number}{\mkbibparens{#1}} 
-%-- and then for articles directly the pages w/o any "pages" or "pp." 
-\DeclareFieldFormat[article]{pages}{#1}
-%-- for some types replace "pages" by "p."
-\DeclareFieldFormat[inproceedings, incollection, inbook]{pages}{p. #1}
-
-% https://tex.stackexchange.com/questions/81569/biblatex-parentheses-around-the-volume-number-of-an-article
-\renewbibmacro*{volume+number+eid}{%
- \printfield{volume}%
- \printfield{number}%
- \printfield{eid}}
-\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
-
-% bibliographic information
-%\bibliographystyle{abbrvnat}
-%\bibliography{citations}
-\addbibresource{citations.bib}
-
-\title{How predictable is extinction? Forecasting species survival at million-year timescales}
-\author{
- Smits, Peter\\
- %Department of Integrative Biology, University of California Berkeley\\
- \texttt{psmits@berkeley.edu} 
- \and
- Finnegan, Seth\\
- %Department of Integrative Biology, University of California Berkeley\\
- \texttt{sethf@berkeley.edu}
-}
-\date{}
-
 \begin{document}
-
-\begin{refsection}
-
-\maketitle
-
-\linenumbers{}
-\modulolinenumbers[3]
-
-\begin{abstract}
-  A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions.
-Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past.
-To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton).
-We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.
-Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time.
-We find that models which include either historical covariates or account for variation in covariate effects over time yield equivalent forecasts, but a model including both is overfit and yields biased forecasts.
-An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
-
-\end{abstract}
-
-{\bf Keywords:} conservation, palaeobiology, extinction, forecasting
-
-\section{Introduction}
-
-The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the near future.
-Predicting which species will go extinct is difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented.
-This has led to the suggestion that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}.
-The fossil record preserves information about the full histories, include ultimate extinction, of thousands of lineages, and this information can help to augment the shorter-term higher-resolution data used to make risk assessments of extant taxa.
-
-Extinction intensity (average rate) and selectivity (difference in risk among taxa) have varied greatly through time, and the relative risk of extinction exhibited by different taxonomic and ecological groups can provide insights into the drivers of both background and mass extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019,Wang2008a,Knoll2007}.
-Many studies have examined the effects of various potential predictors on extinction risk through time \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Harnik2012a,Ezard2011,Foote2006} or refined methods for identifying and measuring these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.
-These studies have produced a growing body of knowledge regarding which factors have been general determinates of extinction risk in the geological past.
-
-A related question that has received much less attention is how successful we might expect to be when using this knowledge to attempt to predict future extinction events.
-
-Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions.
-However, we can take a given point in the geological past, develop a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on subsequent (e.g.
-``future'', from the point of view of the model) extinction/survival events.
-Putting aside the very important question of how human activities will alter the determinants of future extinction risk, such an approach provides a framework for evaluating the expected accuracy of future risk assessments based on past extinction events.
-
-%This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future?
-Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.
-
-Here we take this approach, using as a model system the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores).
-This record has several key strengths for our purposes: planktonic microorganisms are widespread and abundant in pelagic habitats, have high preservation potential, and because of their utility for biostratigraphic, paleoclimatic, and oceanographic study they have been the focus of an extensive international coring and study effort \citep{Lazarus1994,SpencerCervato1999}.
-A compliation of these data is readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}.
-This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomic synonymization framework for dealing with 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible.
-Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.
-
-The overall question of how well models based on past extinction patterns perform at forecasting future extinctions depends in part on model complexity.
-Simple models requiring only a few parameters are in general preferable because more complex models run a greater risk of being overfit to the observations on which they are trained.
-In addition, many traits that might influence extinction risk among extant species are difficult to assign confidently to extinct species.
-For these reasons we elect to focus on “baseline” models which include only a few parameters that have been shown to be important and/or consistent determinants of extinction risk in the marine fossil record.
-Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species’ geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}.
-In addition to geographic range, we also considered global climate state and change in climate state since previous observation in order to evaluate the influence of climate or climate change trajectory on extinction risk.
-Finally, we included species age, both because previous studies of planktonic taxa have found it to be a determinant of extinction risk and because its inclusion in our models is critical to their nature as survival models (see Model Specifications below).
-We reiterate that our primary objective is to evaluate the predictive performance of simple models that include only a few general parameters; more complex models including other likely determinants of extinction risk such as skeletal mineralogy, trophic ecology, and thermal tolerance range might well perform better.
-
-There are a number of ways in which past extinction patterns might be used to model present risk.
-The simplest case assumes that relationships between predictors (hereafter covariates) and extinction risk have been constant through time.
-A more complex but more realistic case allows relationships between covariates and extinction risk to vary through time, consistent with evidence for temporal variation in extinction selectivity regime.
-Finally, an important consideration is the degree to which species’ geographic range trajectories exhibit deterministic versus Markovian behavior \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}.
-In the former case, knowledge of the specific past trajectory of a species – whether its range has expanded or contracted from some point in the past to the present – might help to improve assessments of its current risk.
-In the latter case only the current geographic range of the species would convey useful information about current and future risk (although that assessment would still be based on the relative extinction risk of species that had similar ranges at different points in the geological past).
-
-Below, we evaluate four models along a spectrum from simplest (fixed covariate effects, Markovian range dynamics) to most complex (varying covariate effects, deterministic range dynamics).
-We ask (1) how well they perform at classifying species as extinctions or survivors in the data they were fit to, and (2) how well they perform at classifying species as extinctions or survivors in “future” data that were not used in fitting the models.
-
-\section{Materials and Methods}
-
-\subsection{Data Specifications}
-
-We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}.
-This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic foraminifera, and radiolarians.
-Occurrences were filtered to include only those species with first occurrences no earlier than 63 Mya (millions of years ago).
-This filtering criterion excludes taxa that survived the K/Pg extinction or arose during this recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).
-
-All fossil occurrences were assigned to 1 My (million year) bins based on the estimated age of the fossil occurrence as listed in the Neptune Database.
-After binning, each species' geographic range was calculated for each of the 1 My bins in which it occurred.
-Geographic range was calculated as the minimum spanning tree distance between all observations of that taxon during that temporal bin; this distance was measured in kilometers.
-Minimum spanning tree distance was calculated using the \texttt{GeoRange} package for R \citep{GeoRange}.
-
-We also included how a taxon's geographic range has changed since its last three observation times.
-We measured this change in geographic range by calculating the difference in geographic range between an observation and that taxon's three previous occurrences.
-Change between the most recent and the three previous occurrences was calculated individually for each of those lags.
-If there are not enough previous observations to calculate, then that value is recorded as a 0.
-These differences were calculated after minimum spanning tree distance was transformed and standardized (see Supplement Section S1.1.2)
-
-Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}.
-We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic.
-Estimating temperature over long periods of time from Mg/Ca ratios also suffers from complications because Mg/Ca based temperature estimates requires benthic foram and seawater Mg/Ca isotope ratio information.
-Because seawater Mg/Ca ratio has changed over time, the method to estimate temperature used \citet{Cramer2011} attempts to account for this unknown in order to obtain accurate, albeit uncertain, temperature estimates.
-Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My interval between 0 to 63 Mya.
-The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.
-
-We also included the global temperature from the previous time that taxon was observed.
-If there are not enough previous observations to calculate, then that value is recorded as a 0.
-This lag was calculated after global temperature was transformed and standardized (see Supplement Section S1.1.2).
-
-Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature.
-The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit.
-Ideally, we would have detailed ocean surface water temperature estimates for the entire globe for the entire Cenozoic.
-Unfortunately, that type of data does not exist.
-So, we interpret our temperature estimates as reflecting the global climate state that an organisms experiences, and not as a descriptor of that taxon's local environmental ecology.
-
-See Supplemental Section S1.1 for a further explanation on how observations were temporally binned, and how our covariates were standardized and transformed prior to analysis.
-
-%\begin{figure}[ht]
-% \centering
-% \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
-% \caption{Comparison of initial temperature estimates from \citet{Cramer2011} (goldenrod) versus the binned values used in this analysis (blue).
-The initial values are for every 0.1 My while our bins are defined for every 1 My.}
-% \label{fig:temp_curve}
-%\end{figure}
-
-\subsection{Model Specifications}
-
-We used a discrete-time survival modelling framework to estimate how well we can predict extinction risk at one million year time scales.
-At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}.
-We considered four different models involving different permutations of covariate effects (fixed or time-varying) and historical covariates: covariate effects constant over time and no historical covariates included (Model C), covariate effects allowed to vary over time but no historical covariates included (Model V), covariate effects constant over time and historical covariates included (Model CP), and covariate effects allowed to vary over time and historical covariates are included (Model VP).
-The C and P models attempt to predict based only on present state, whereas the CP and VP models allow for the possibility of non-Markovian behaviour by including change in state from the previous time increment.
-
-We always included species age at time of observation (i.e.
-observed prior duration) as a varying-intercept term.
-This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.
-The effect of species age is allowed to vary by taxonomic group.
-
-Similarly, we included time of obseration as an additional varying-intercept term to account for changes average global extinction risk over time that are not related to the covariates included in this model.
-This varying-intercept is further allowed to vary by taxonomic group.
-This varying-intercept term allows us to tease apart the differences in extinction risk associated with time of observation versus age since first observation.
-An important note is that for our V and VP models, the covariation between this varying-intercept and the varying-slopes of our covariates is explicitly modeled (see Supplement Section S1.2).
-
-See Table \ref{tab:model_def} for further explanation of how the four models we considered differ from each other.
-A complete description of the statistical model used in this analysis is available in Supplement Section S1.2.
-Additionally, the full description of how these models were implemented and coded, including choice of priors, is available in Supplement Section S1.2.
 
@@
 
-\subsection{In-sample and out-of-sample forecasting}
-
-We are interested in our models' performance in two distinct contexts: in-sample performance, and out-of-sample predictive performance (i.e. forecasting).
-
-In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to correctly classify the data to which it was fit.
-Posterior predictive checks are a type of sensitivity analysis because we are checking the quality of model's fit to the data.
-If our models have poor in-sample forecasting performance, than our models are not adequate descriptors of the data and will most likely make poor out-of-sample predictions.
-In-sample forecasting measures, however, are not the same as understanding our models' ability to forecast future extinctions or if our models are overfit to our data and produce biased out-of-sample estimates \citep{ESL}.
-
-We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data).
-To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation.
-For time-series data, the folds (data partitions) are approximately equal segments of time.
-Each fold represents a sequence of time points.
-With 63 time points, each of the five folds represents approximately 13 million-year time increments.
-It is important to bear in mind, however, that each time increment includes many (100s-1000s) individual observations.
-
-\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016,ESL}.
-Prior to cross-valdiation, the data is divided into \textit{k} nearly even segments or folds -- for a time series, this means the data is divided into \textit{k} continuous sequences.
-Next, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e.
-the future).
-Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold.
-Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold.
-Next, the model is fit to the first four folds combined and then is used to forecast the fifth fold.
-This process continues until $k - 1$ folds are included in the fitting the model and the final fold is predicted from this model.
-When combined, the results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.
-In 5-fold cross validaition, the data is divided into five folds the cross-validation procedure yields predictions for four of the folds.
-
-Cross-validation is a procedure for estimating a models expected out-of-sample error.
-Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}.
-Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, out-of-sample predictions will be biased and inaccurate \citep{ESL}.
-A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).
-Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces, an entire posterior distribution of estimates.
-
-The relative adequacy of the four model variants was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}.
-This measure is commonly used to measure the performance of classification models as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives.
-AUC ranges between 0.5 and 1, with 0.5 indicating no difference in classification from random and 1 indicating perfect classification.
-AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of a randomly selected extinct-extant species pair \citep{Fawcett2006,Mason2002}.
-AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we consider values between between 0.7 and 0.8 as ``fair,'' and values between 0.6 and 0.7 as ``poor.''
-
-%The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time.
-These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.
-
-See our code repository at https://github.com/psmits/trident for full code details.
-The entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}.
-All of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.
-
-\section{Results}
-
-The primary focus of this study is on understanding how well our models forecast future extinction events by comparing our in-sample and out-of-sample forecast estimates.
-A presentation of the posterior estimates for the regression coefficient estimates from our VP model (Table \ref{tab:model_def}) is available our Supplemental Materials (Section S2).
-
-% ROC model comparison
-\begin{figure}[ht]
- \begin{subfigure}[ht]{0.45\textwidth}
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
-  \caption{In-sample}
-  \label{fig:auc_hist}
- \end{subfigure}
- \begin{subfigure}[ht]{0.45\textwidth}
-  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
-  \caption{Out-of-sample}
-  \label{fig:fold_auc}
- \end{subfigure}
- \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation.
-The area under the receiver operating charactered curve (AUC) was calculated for each model.
-these estimates are calculated from the models posterior predictive distribution (\ref{fig:auc_hist}) or from predictions made to new data (\ref{fig:fold_auc}), respectively.
-Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in our dataset.
-Models with higher AUC values indicate better performance over models with lower AUC values.
-AUC is bounded between 0.5 and 1.
-See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
- \label{fig:auc_compare}
-\end{figure}
-
-\subsection{In-sample forecasting adequacy}
-
-The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis.
-Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that the parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, while there is substantial overlap in the posterior distributions of the forecasts from the other three models (Fig.
-\ref{fig:auc_hist}).
-
-However, the actual difference in forecast AUC result between the VP model and the other three models is extremely small (0.01 AUC unit), and all of the in-sample AUC estimates from our models are concentrated between an AUC value of 0.775 and 0.795 (Fig.
-\ref{fig:auc_hist}).
-This result indicates that the practical difference in performanc between these models might be so small that there is no practical benefit that the VP model over the other three.
-Ultimately, determining which of these models produces the best forecasts of future extinctions requires comparing these in-sample results to our out-of-sample results (see below).
-
-The in-sample forecasts from our four models over time are broadly similar between taxonomic groups (Fig.
-\ref{fig:auc_taxon_time}).
-Our in-sample forecasts for Diatoms are the weakest of the taxonomic groups as all four models have an equal number of intervals with no predictive power (AUC approximately 0.5).
-In contrast, our best in-sample forecast results are for Radiolarians where for any of our models there is at most 1 interval with almost no predictive power.
-Ultimately, our in-sample forecasts over time by taxonimc group are broadly consistent between our four models.
-
-\begin{figure}[ht]
- \centering
- \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
- \caption{Understanding model adequacy over time and taxonomic group by comparing in-sample forecasting performance measured by AUC for each of the four models.
-These estimates reflect each model's fit to the various taxonomic groups over time.
-The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend.
-In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.
-The grey intervals mark the geologic ages of the Cenozoic.
-See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
- \label{fig:auc_taxon_time}
-\end{figure}
-
-\subsection{Out-of-sample forecasting performance}
-
-Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}.
-Our out-of-sample forecast AUC estimates demonstrate a broader range of results, with AUC estimates ranging between approximately 0.7 and 0.85 (Fig.
-\ref{fig:auc_hist}, \ref{fig:fold_auc}).
-While our VP model was the model with greatest in-sample forecasting performance (Fig.
-\ref{fig:auc_hist}), this model has the greatest decrease in out-of-sample forecasting performance (Fig.
-\ref{fig:fold_auc}).
-In comparison, the other three models demontrstrate a much smaller difference between in-sample forecast and out-of-sample forecast AUC values which indicates that our VP model is overfit to our data, and that one of the simpler model would be preferable for predicting future extinctions.
-This result means that our model which includes the historical covariates (e.g.
-change in geographic range) and allowing these effects to vary over time produces biased extinciton forecasts.
-Interestingly, including either historical covariates like change in geographic range or allowing the effects of geographic range and other covariate effects to vary over time have approximately equal effect on forecasting future extinction events (Fig.
-\ref{fig:auc_compare}).
-
-In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig.
-\ref{fig:auc_taxon_time}).
-This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts.
-The major exception to this pattern are our estimates for the diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random.
-The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
-\begin{figure}[ht]
- \centering
- \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
- \caption{Comparison of our models ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models.
-The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds.
-This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals.
-The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.
-The grey intervals mark the geologic ages of the Cenozoic.
-See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
- \label{fig:fold_auc_taxon_time}
-\end{figure}
-
-We compared the difference in our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig.
-\ref{fig:oos_ins_diff}).
-A difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts.
-A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast.
-Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig.
-\ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).
-The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig.
-\ref{fig:oos_ins_diff}).
-
-\begin{figure}[ht]
- \centering
- \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
- \caption{Comparing our models' ability to forecast out-of-sample extinction compared to our models' in-sample forecasts.
-Congruence between these measures indicates that our models are not necessarily overfit to the data.
-This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}.
-A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast.
-See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
- \label{fig:oos_ins_diff}
-\end{figure}
-
-\section{Discussion}
-
-We find that all of our models are expected to correctly forecast which species of randomly selected extinct-extant pair is more likely to go extinct between 70\% to 80\% of the time (Fig.
-\ref{fig:fold_auc}).
-These results confirm that past extinction patterns can provide valuable information about which extant species are most threatened with extinction in the near geological future.
-%One of the most striking results is that the models that include either the historical covariates (CP model) or allows covariate effects to vary over time (V model) produce less biased out-of-sample forecasts than the model which includes both historical covariates and allows covariate effects to vary over time (VP model) (Fig.
-\ref{fig:auc_compare}).
-However, we might expect a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect.
-An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
-
-Three of the four models we evaluated are practically identical in their ability to make in-sample and out-of-sample forecasts.
-Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig.
-\ref{fig:auc_hist}).
-The model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time.
-However, the out-of-sample forecasts from this model are biased forecasts, indicating that it is overfit to our data \ref{fig:fold_auc}).
-The model that includes historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to the model that allows covariate effects to vary over time but does not include historical covariates.
-
-We note that both including historical covariates such as change in geographic range and allowing covariate effects to vary over time are different ways of encoding information from the past.
-Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time.
-Allowing covariate effects to vary over time, on the other hand, does not explicitly encode ``the past'' into our model but instead models how covariate effects vary over time which allows the past to be implicitly included in our model.
-By modeling this variation, forecasts made for future extinction events can allow for a wide range of potential effects of our covariates on predicting extinction.
-Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig.
-\ref{fig:fold_auc}).
-Our results supplement those of \citet{Kiessling2016} which examined using differences in geographic range over time to predict extinction risk.
-
-An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predicatable than it has been in the past \citep{Harnik2012a,Finnegan2015}.
-Our model with historical covariates assumes that extinction selectivity with respect to these covariates is constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes.
-Thus, it might be preferable to model the variation in extinction risk and selectivity over time instead of relying solely on measures from a taxon's past such as change in geographic range over time.
-For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.
-
-The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
-Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity.
-The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study.
-Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}.
-However, comparable information does not exist for all foram species, nor does this type of data exist for the other three taxonomic groups studied here.
-Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.
-
-In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors.
-Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models.
-The remaining three models yield nearly equivalent out-of-sample forecasts, suggesting that including historical information via either explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions.
-The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data.
-
-\section{Acknowledgements}
-
-We thank Jonathan Payne, Thomas Ezard, Erin Saup, and Peter Roopnarine for useful discussions.
-Wolfgang Kiessling, Matthew Clapham, and an anonymous reviewer provided thoughful reviews that strengthened the manuscript consideraly.
-This work was funded by a David and Lucile Packard Fellowship to SF.
-
-\printbibliography
-\end{refsection}
-%\clearpage
-
 \end{document}
-
-%\beginsupplement
-%\begin{refsection}
-%
-%\section{Supplemental Figures}
-%
-%\setcounter{page}{1}
-%\section{Supplement to Materials and Methods}
-%
-%\subsection{Data Specifications} \label{sec:data_desc}
-%
-%\subsubsection{Binning fossil occurrences}
-%
-%The estimated age of each occurrence is based on the core-specific age-model that observation is from and can be overly precise.
-To alleviate this overprecision, we coarsened our temporal information in an effort to limit the effects of between-core heterogeneity in age.
-The occurrence histories of each species was then summarized as a series of binary codes indicating the presence or last occurrence of that species.
-For every occurrence of a species, except the last, that species existence and survival is recorded as a 0.
-The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is recorded as 1.
-This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is an overconfident statement of preservation and may be shortening the actual durations of the studied species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}.
-However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}.
-In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores \citep{Mekik2018,Broecker1999,Mekik2014,Peng1984}.
-%
-%
-%\subsubsection{Covariate transformation and standardization}
-%
-%Prior to analysis, geographic range was then log-plus-one transformed and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges.
-This standardization means that a regression coefficient associated with each covariate describes the change in extinction probability per change in standard deviation of that covariate, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}.
-Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.
-%
-%Temperature was also transformed and standardized the in the same manner as geographic range.
-The change in temperature between an observation and its previous observation was measured from the standardized temperature values and was not standardized separately.
-%
-%
-%\subsection{Model Specifications} \label{sec:model_desc}
-%
-%In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and covariate information.
-The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals \citep{Tutz2016}.
-For the discrete time intervals \(T = 1, \cdots, k\), extinction is defined as \(T = t\).
-The discrete time hazard function is defined as
-%\begin{equation}
-% \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k.
-% \label{eq:hazard}
-%\end{equation}
-%
-%The hazard function (Eq.
-\ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\) \citep{Tutz2016}.
-\(h(\Theta)\) is then modeled as with any regression.
-In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes \citep{ARM}.
-%
-%Our covariates matrix \(X\) is a \(N \times D\) matrix where \(N\) is the total number of observations and \(D\) is the total number of covariates.
-The first column of \(X\) is entirely 1's as it corresponds to the intercept term in the regression model.
-The next two columns of \(X\) are two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\).
-Change in geographic range was calculated from the transformed and standardized geographic range values; this means that change in geographic range is in units of changes in standard deviations.
-The final two columns are two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e.
-mean temperature during interval \(t - 1\).) As with change to geographic range, the lag of temperature is based on the transformed and standardized temperature estimates.
-%
-%The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla.
-The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each phylum \(p\).
-These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept.
-\(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.
-%
-%The effect of species age on the log-odds of species extinction is modeled as a non-nested random intercept \(A\) \citep{Tutz2016}.
-This term describes how the log-odds of extinction varies along a species duration, and how this effect can differ between the phyla.
-\(A\) is a \(l\) by \(p\) matrix, where \(l\) is the age at observation of a species and \(p\) is its phylum.
-\(A\) is modeled as following a multivariate normal distribution with phylum means being the vector \(\delta\) and covariance matrix \(\Sigma_{A}\).
-The covariation between the elements of vector \(\delta\) are modeled as a multivariate normal distribution with a mean vector of all 0s and covariance matrix \(\Sigma_{\delta}\).
-%
-%To complete the generative model, we need to assign final priors to the ``top-level'' parameters.
-In general, we favored weakly informative priors which help regularize our estimates.
-In the case of a regression coefficient, this means a Normal distribution with mean 0 and a standard deviation of 3.
-For our scale parameters (e.g.
-standard deviations), we used half-Cauchy distributed priors with heavy tails but the majority of probability density near 0.
-%
-%Our top-level intercept was given a more diffuse prior than our regression coefficients, which reflects our greater degree of uncertainty about its value.
-Our top-level regression coefficient for the effect of geographic range was given an informative prior reflecting the overwhelming amount of evidence that species with a larger than average geographic range have a lower risk of extinction than species with an average or less than average geographic range.
-In the context of this analysis, this means that we are again using a weakly informative prior but instead of centering the density around -1 (i.e.
-larger than average geographic range decreases extinction risk).
-%
-%Instead of assigning a prior distribution for each of the covariance matrices in the model, we instead decomposed the covariance matrices (e.g.
-\(\Sigma_{B}\)) which allows us to assign independent priors for the scale and correlation aspects of covariance.
-The scale parameters were assigned half-Cauchy priors as described above in the context of all other scale parameters.
-The correlation matrices were assigned LKJ priors each with shape parameter set to 1.
-This choice of shape parameter produces a uniform distribution over possible correlation matrices.
-These priors are also slightly more interpretable than other common prior distributions for covariance matrices such as the inverse-Wishart distribution.
-This approach to assigning priors to a covariance matrix is recommended by the Stan Manual \citep{StanManual}.
-%
-%In total, our model can be expressed as: 
-%\begin{equation}
-% \begin{aligned}
-%  t_{i} &\sim \text{Bernoulli}(\Theta) \\
-%  \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{l[i], p[i]}) \\
-%  B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
-%  \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
-%  % double check this\dots is A MVN dist?
-%  A_{l, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
-%  \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
-%  \mu_{d} &\sim 
-%  \begin{cases}
-%   N(-2, 5) & \text{if } d = \text{intercept} \\
-%   N(-1, 1) & \text{if } d = \text{geo. range} \\
-%   N(0, 1) & \text{else } \\
-%  \end{cases} \\
-%  \delta &\sim N(0, 1) \\
-%  \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
-%  \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
-%  \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
-%  \tau_{B} &\sim C^{+}(1) \\
-%  \tau_{\alpha} &\sim C^{+}(1) \\
-%  \tau_{A} &\sim C^{+}(1) \\
-%  \Omega_{B} &\sim LKJ(1) \\
-%  \Omega_{\alpha} &\sim LKJ(1) \\
-%  \Omega_{A} &\sim LKJ(1) \\
-% \end{aligned}
-% \label{eq:model}
-%\end{equation}
-%with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.
-%
-%
-%\subsection{Model Parameter Estimation} \label{sec:model_est}
-%
-%We implemented our model (Eq.
-\ref{eq:model} using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}.
-This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R.
-Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling.
-In order to prevent divergent transitions adapt delta was increased to 0.999999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.
-%
-%To implement our VP model in \begin{texttt}rstanarm\end{texttt}, where ``data'' is a data.frame object of all necessary data (response, covariates), is coded as:
-%\begin{verbatim}
-%form <- event ~ 
-%        range + range_diff1 + range_diff2 + range_diff3 + 
-%        temp + temp_lag + 
-%        (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
-%        (1 | age/phylum), 
-%stan_glmer(formula = form,
-%      data = data, 
-%      family = 'binomial',
-%      prior = normal(c(-1, 0, 0, 0, 0, 0), rep(1, 6), autoscale = FALSE), 
-%      prior_intercept = normal(-2, 5, autoscale = FALSE), 
-%      prior_aux = cauchy(0, 1, autoscale = FALSE), 
-%      chains = 4,
-%      thin = 4,
-%      adapt_delta = 0.999999)
-%\end{verbatim}
-%
-%Similarly, our VP model can be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
-%\begin{verbatim}
-%priors <- c(set_prior('normal(-2, 5)', class= 'Intercept'),
-%      set_prior('normal(0, 1)', class = 'b'),
-%      set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
-%      set_prior('cauchy(0, 1)', class = 'sd'),
-%      set_prior('lkj(1)', class = 'cor'))
-%form <- bf(event ~ 
-%           range + range_diff1 + range_diff2 + range_diff3 + 
-%           temp + temp_lag +
-%           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
-%           (1 | age/phylum))
-%brmfit <- brm(formula = form,
-%       data = data, 
-%       family = bernoulli(), 
-%       prior = priors,
-%       chains = 4, 
-%       thin = 4,
-%       control = list(adapt_delta = 0.999999)
-%\end{verbatim}
-%
-%Posterior convergence was determined using the general and HMC-specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)).
-For further explanation of these diagnostic criteria, see the Stan Manual \citep{StanManual}.
-%
-%\section{Supplement to Results} \label{sec:supp_res}
-%
-%Here we present the parameter estimates for our from our VP model (Table \ref{tab:model_def}).
-We choose to present the estimates from this model because it is our most inclusive model and its parameter estimates are indicative of parameter estimates from other models with a subset of those covariates included in this model.
-%
-%First, we present the group-level estimates of our regression coefficients and intercept (Fig.
-\ref{fig:param_est}).
-Group-level effects are the average effects of that covariate over time and across taxonomic groups.
-These estimates make up the vector $\mu$ described above.
-In addition to the average effect of our covariates, we also included the estimate of the group-level intercept, which is the average log-odds of extinction for an average observation over time and across taxonomic groups.
-%
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/effect_est}
-%  \caption{Group-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
-The posterior distribution of our parameter estimates are presented as densities.
-Below each of these densities is marked the median estimate along with 50\% and 80\% credible intervals.
-Estimates are on the log-odds scale.}
-%  \label{fig:param_est}
-%\end{figure}
-%
-%Second, we present the population-level estimates for our regression coefficients along with the population-level intercept estimates (Fig.
-\ref{fig:param_est_time_group}.
-Our VP model allows for these our regression coefficients to vary over time and between taxonomic groups.
-Population-level estimates are those for a specific time interval and taxonomic group.
-In this case, the intercept estimate describes the log-odds of extinction for an average observation at that point in time and taxonomic group.
-%
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/eff_time_group}
-%  \caption{Population-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}).
-Posterior estimates are presented as a time series.
-The black line represents the median estimate of that parameter.
-In addition, the 50\%, 80\%, and 95\% credible intervals are indicated.
-Estimates are on the log-odds scale.}
-%  \label{fig:param_est_time_group}
-%\end{figure}
-%
-%
-%
-%\printbibliography[title={Supplementary References}]
-%\end{refsection}

### 2019-07-24 R/07_predict_quality.r

@@ -93,12 +93,13 @@ diff_gg <-
   scale_fill_brewer() +
   facet_grid(fossil_group ~ model) +
   scale_x_reverse(name = 'CI', breaks = brks) +
+  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
   theme(legend.position = 'bottom') +
   labs(x = 'Time (Mya)', y = 'Difference between out-of-sample and in-sample AUC') +
   NULL
 
 ggsave(filename = here::here('results', 'figure', 'auc_diff.png'),
        plot = diff_gg,
-       width = 11, height = 8.5)
+       width = 11, height = 11)
 
 

### 2019-07-24 doc/article.tex
waiting for seth comments

@@
 In the latter case only the current geographic range of the species would convey useful information about current and future risk (although that assessment would still be based on the relative extinction risk of species that had similar ranges at different points in the geological past).
+In all cases, we use information about the past to predict the future, the question is wheter and how much historical information (e.g.
+the histories of species over time) improves our ability to forecast future extinction events.
 
@@
 \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).
-One of the few multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model between 35 Mya and approximately 28 Mya (Fig.
+An example multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model between 35 Mya and approximately 28 Mya (Fig.
 \ref{fig:oos_ins_diff}).
-There exist similar periods of worse out-of-sample forecasting performance for other combinations of taxonomic group and predictive model, but none are as long as that for Radiolaria.
+There exist similar periods of worse out-of-sample forecasting performance for other combinations of taxonomic group and predictive model, for example the CP model has worse out-of-sample forecasting than in-sample for the last 5 million years of the Cenozoic.
+In general, however, most out-of-sample and in-sample forecasts are almost identical.
 
@@
 \ref{fig:fold_auc}).
-These results confirm that past extinction patterns can provide valuable information about which extant species are most threatened with extinction in the near geological future.
+These results confirm that past extinction patterns can provide valuable information about which extant species are most threatened with extinction in the near geological future, and that some historical information does not degrade our ability to forecast future extinction risk.
+To reiterate, all of our models are fit to past extinction events from the Cenozoic and covariates like geographic range or global climate state are associated with these extinctions; our models are conditioned on the past.
+Some of our models, however, encode historical information such as how covariate effects have varied over time or the change in a species geographic range over time (Table \ref{tab:model_def}).
 %One of the most striking results is that the models that include either the historical covariates (CP model) or allows covariate effects to vary over time (V model) produce less biased out-of-sample forecasts than the model which includes both historical covariates and allows covariate effects to vary over time (VP model) (Fig.
@@
 \ref{fig:auc_hist}).
-The model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time.
+Our VP model had the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time.
 However, the out-of-sample forecasts from this model are biased, indicating that it is overfit to the data \ref{fig:fold_auc}).
-The model that includes historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to the model that allows covariate effects to vary over time but does not include historical covariates.
+The CP model that includes historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to the V model that allows covariate effects to vary over time but does not include historical covariates.
 
-We note that both including historical covariates such as change in geographic range and allowing covariate effects to vary over time are different ways of encoding information from the past.
-Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time.
-Allowing covariate effects to vary over time, on the other hand, does not explicitly encode ``the past'' into our model but instead models how covariate effects vary over time which allows the past to be implicitly included in our model.
-By modeling this variation, forecasts made for future extinction events can allow for a wide range of potential effects of our covariates on predicting extinction.
+While all of our models are conditioned on past extinction data from the Cenozoic planktonic microfossil record, we used multiple ways to encode historical information which describes the species over time such as change in geographic range or previous climate state or by allowing covariate effects to vary over time.
+Models that include the historical covariates (e.g.
+change in geographic range) but do not allow covariate effects to vary over time (i.e.
+CP model) encode the past explicitly but assumes that covariates effects are constant over time.
+Allowing covariate effects to vary over time, as with our V model, does not explicitly encode the history of individual species into our model but instead models the history of how covariate effects vary over time, thus implicitly encoding historical information about the species in our models.
+By modeling the variation in covariate effects over time, forecasts made for future extinction events are conditioned on a wider range of potential covariate effects which can improve our models flexibility when forecasting extinction in the novel environmental conditions we might expect in the future.
 Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig.
@@
 Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models.
-The remaining three models yield nearly equivalent out-of-sample forecasts, suggesting that including historical information via either explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions.
-The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data.
+The remaining three models yield nearly equivalent out-of-sample forecasts.
+The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data, and that historical information.
+Additionally, including historical information via explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions.
 

### 2019-11-05 doc/article.tex
pretty head

@@
 
-While all of our models are conditioned on past extinction data from the Cenozoic planktonic microfossil record, we used multiple ways to encode historical information which describes the species over time such as change in geographic range or previous climate state or by allowing covariate effects to vary over time.
+While all of our models are conditioned on past extinction data from the Cenozoic planktonic microfossil record, we used multiple approaches to encode historical information.
 Models that include the historical covariates (e.g.
@@
 CP model) encode the past explicitly but assumes that covariates effects are constant over time.
-Allowing covariate effects to vary over time, as with our V model, does not explicitly encode the history of individual species into our model but instead models the history of how covariate effects vary over time, thus implicitly encoding historical information about the species in our models.
+Allowing covariate effects to vary over time, as with our V model, does not explicitly encode the history of individual species into our model but instead models the history of how covariate effects have varied over time, thus implicitly encoding historical information about the species in our models.
 By modeling the variation in covariate effects over time, forecasts made for future extinction events are conditioned on a wider range of potential covariate effects which can improve our models flexibility when forecasting extinction in the novel environmental conditions we might expect in the future.
@@
 \ref{fig:fold_auc}).
-Our results complement those of \citet{Kiessling2016} who found that including historical range trajectories improved the performance of extinction risk forecasts.
+Our results contrast somewhat with those of \citet{Kiessling2016}, who found that including historical range trajectories significantly and substantially improved the performance of extinction risk forecasts.
+Whether this reflects differences in analytical methodology or study system (planktonic microfossil taxa versus marine macroinvertebrates) is not clear and is worth further investigation.
 
-The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
+The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models we considered is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa.
 Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity.
@@
 The remaining three models yield nearly equivalent out-of-sample forecasts.
-The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data, and that historical information.
+The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data.
 Additionally, including historical information via explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions.
@@
 
-We thank Jonathan Payne, Thomas Ezard, Erin Saup, and Peter Roopnarine for useful discussions.
-Wolfgang Kiessling, Matthew Clapham, and an anonymous reviewer provided thoughful reviews that strengthened the manuscript consideraly.
+We thank Jonathan Payne, Thomas Ezard, Erin Saupe, and Peter Roopnarine for useful discussions.
+Wolfgang Kiessling, Matthew Clapham, and an anonymous reviewer provided thoughtful reviews that strengthened the manuscript considerably.
 This work was funded by a David and Lucile Packard Fellowship to SF.
